---
title: "turbidity"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE)
```

# Purpose

This script is meant to explore patterns in turbidity in an attempt to forecast
turbidity at a 1-3 day time horizon at the FC intake.

```{r}
library(tidyverse)
library(cdssr)
library(corrr)
library(gghighlight)

turb <- read_csv("data/collated/FC_intake/fc_intake_munroe_turbidity_quantile_record_2010-2025.csv")

summary(turb)

# read in flow data at canyon mouth
flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = "2010-01-01",
                       end_date = "2025-12-31",
                       timescale = "day") %>%
  select(datetime, canyon_mouth_cfs = value) %>% 
  mutate(date = date(datetime), 
         year = year(date),
         doy = yday(date)) 

turb_q <- full_join(turb, flow_data)
```
Look at the patterns by DOY

```{r}
turb %>% 
  pivot_longer(cols = c(p05:p95), names_to = "quantile", values_to = "turb") %>% 
  ggplot(., aes(x = doy, y = turb)) +
  geom_point() +
  facet_grid(quantile ~ ., scales = "free_y")
```

And look at p95 year-by-year for seasonal patterns

```{r}
map(min(year(turb$date)):max(year(turb$date)),
    ~ {
      turb %>% 
        filter(year(date) == .x) %>% 
        ggplot(., aes(x = doy, y = p95)) +
        geom_point() +
        labs(title = paste0("Intake - ", .x))
    }) 
```

And the correlation between p95 and others:

```{r}
turb %>% 
  select(c(starts_with("p"), -pct_diff)) %>% 
  correlate() %>% 
  stretch() %>% 
  filter(x == "p95" & !is.na(r))
```

Not really a strong correlation across any of the other quantiles.

# Flow + Turbidity

Plot the two together

```{r}
turb_q %>% 
  ggplot(., aes(x = doy)) +
  geom_point(aes(y = p95)) +
  geom_line(aes(y = canyon_mouth_cfs/20)) +
  coord_cartesian(ylim = c(0, 300)) + #limit to keep things reasonable
  facet_wrap(year ~ .)
```

Let's plot turbidity against flow, faceted by month, as I suspect there are some antecedent
conditions to consider that this might start to tease out

```{r}
turb_q %>% 
  pivot_longer(cols = c(p05:p95), names_to = "quantile", values_to = "turb") %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = canyon_mouth_cfs, y = turb)) +
  geom_point() +
  facet_grid(quantile ~ month, scales = "free_y")
```


# Changes in Q

Instead of looking at today's Q, let's look at the change in Q over the previous
24h.

```{r}
turb_apr_sept_lagQ <- turb_q %>% 
  arrange(date) %>% 
  mutate(deltaQ_24h = canyon_mouth_cfs - lag(canyon_mouth_cfs),
         deltaQ_48h = canyon_mouth_cfs - lag(canyon_mouth_cfs, n = 2),
         deltaQ_72h = canyon_mouth_cfs - lag(canyon_mouth_cfs, n = 3))

turb_apr_sept_lagQ %>% 
  ggplot(., aes(x = doy)) +
  geom_line(aes(y = deltaQ_24h/5)) + # standardize a little to for viewing
  geom_point(aes(y = p95-200), color = "purple") +
  coord_cartesian(ylim = c(-200, 100)) + #limit to keep things reasonable
  facet_wrap(year ~ .)

turb_apr_sept_lagQ %>% 
  ggplot(., aes(x = doy)) +
  geom_line(aes(y = deltaQ_48h/5)) + # standardize a little to for viewing
  geom_point(aes(y = p95-200), color = "purple") +
  coord_cartesian(ylim = c(-200, 100)) + #limit to keep things reasonable
  facet_wrap(year ~ .)

turb_apr_sept_lagQ %>% 
  ggplot(., aes(x = doy)) +
  geom_line(aes(y = deltaQ_72h/5)) + # standardize a little to for viewing
  geom_point(aes(y = p95-200), color = "purple") +
  coord_cartesian(ylim = c(-200, 100)) + #limit to keep things reasonable
  facet_wrap(year ~ .)
```

Well, this looks useful... what does it look like as a regression?

```{r}
turb_apr_sept_lagQ %>% 
  pivot_longer(cols = c(p05:p95), names_to = "quantile", values_to = "turb") %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_24h, y = turb)) +
  geom_point() +
  facet_grid(quantile ~ month, scales = "free_y")

turb_apr_sept_lagQ %>% 
  pivot_longer(cols = c(p05:p95), names_to = "quantile", values_to = "turb") %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_48h, y = turb)) +
  geom_point() +
  facet_grid(quantile ~ month, scales = "free_y")

turb_apr_sept_lagQ %>% 
  pivot_longer(cols = c(p05:p95), names_to = "quantile", values_to = "turb") %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_72h, y = turb)) +
  geom_point() +
  facet_grid(quantile ~ month, scales = "free_y")

```

Look a little closer at p95:

```{r}
turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_24h, y = p95)) +
  geom_point() 

turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_48h, y = p95)) +
  geom_point() 

turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_72h, y = p95)) +
  geom_point() 

```

Not so great. Can we see anything if we facet by month?

```{r}
turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_24h, y = p95)) +
  geom_point()  +
  coord_cartesian(xlim = c(-1000,1000),
                  ylim = c(0, 500)) +
  facet_wrap(month ~ .)

turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_48h, y = p95)) +
  geom_point() +
  coord_cartesian(xlim = c(-1000,1000),
                  ylim = c(0, 500)) +
  facet_wrap(month ~ .)

turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_72h, y = p95)) +
  geom_point()+
  coord_cartesian(xlim = c(-1000,1000),
                  ylim = c(0, 500)) +
  facet_wrap(month ~ .)


```

Mayyyyybe at the 72h lag.

