---
title: "turbidity"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, error = FALSE, warning = FALSE)
```

# Purpose

This script is meant to explore patterns in turbidity in an attempt to forecast
turbidity at a 1-3 day time horizon at the FC intake. We know that from a conceptual
standpoint, turbidity is coming from autochthounous and allochthonous sources and
those sources and responses differ over the year. Turbidity can spike during snowmelt
events (primarily on the rising limb of Q), it can spike from reservoir releases,
and it can spike from convective storms. Over fire-affected landscapes, turbidity
can be at a higher magnitude and for a longer period of time, but is dependent on
complex factors that probably can't be incorporated in a simple model.

```{r}
library(tidyverse)
library(cdssr)
library(corrr)
library(gghighlight)
library(RcppRoll)

turb <- read_csv("data/collated/FC_intake/fc_intake_munroe_turbidity_quantile_record_2010-2025.csv")

summary(turb)

# read in flow data at canyon mouth
flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = "2010-01-01",
                       end_date = "2025-12-31",
                       timescale = "day") %>%
  select(datetime, canyon_mouth_cfs = value) %>% 
  mutate(date = date(datetime), 
         year = year(date),
         doy = yday(date)) 

turb_q <- full_join(turb, flow_data)
```

# Data Exploration

Look at the patterns by DOY

```{r}
turb %>% 
  pivot_longer(cols = c(p05:p95), names_to = "quantile", values_to = "turb") %>% 
  ggplot(., aes(x = doy, y = turb, color = location)) +
  geom_point() +
  facet_grid(quantile ~ ., scales = "free_y")
```

And look at p95 year-by-year for seasonal patterns

```{r}
map(min(year(turb$date)):max(year(turb$date)),
    ~ {
      turb %>% 
        filter(year(date) == .x) %>% 
        ggplot(., aes(x = doy, y = p95, color = location)) +
        geom_point() +
        labs(title = paste0("Intake - ", .x))
    }) 
```

And the correlation between p95 and others:

```{r}
turb %>% 
  filter(location == "intake_turb_NTU") %>% 
  select(c(starts_with("p"), -pct_diff)) %>% 
  correlate() %>% 
  stretch() %>% 
  filter(x == "p95" & !is.na(r))

turb %>% 
  filter(location == "munroe_turb_NTU") %>% 
  select(c(starts_with("p"), -pct_diff)) %>% 
  correlate() %>% 
  stretch() %>% 
  filter(x == "p95" & !is.na(r))
```

Not really a strong correlation at the intake - fairly strong at p75/p95 at 
munroe though!

# Flow + Turbidity

Plot the two together

```{r}
turb_q %>% 
  ggplot(., aes(x = doy)) +
  geom_point(aes(y = p95)) +
  geom_line(aes(y = canyon_mouth_cfs/20)) +
  coord_cartesian(ylim = c(0, 300)) + #limit to keep things reasonable
  facet_wrap(year ~ location)
```

Let's plot turbidity against flow, faceted by month, as I suspect there are some antecedent
conditions to consider that this might start to tease out

```{r}
turb_q %>% 
  pivot_longer(cols = c(p05:p95), names_to = "quantile", values_to = "turb") %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = canyon_mouth_cfs, y = turb, color = location)) +
  geom_point() +
  facet_grid(quantile ~ month, scales = "free_y")
```


# Changes in Q

Instead of looking at today's Q, let's look at the change in Q over the previous
24h.

```{r}
turb_apr_sept_lagQ <- turb_q %>% 
  arrange(date) %>% 
  mutate(deltaQ_24h = canyon_mouth_cfs - lag(canyon_mouth_cfs),
         deltaQ_48h = canyon_mouth_cfs - lag(canyon_mouth_cfs, n = 2),
         deltaQ_72h = canyon_mouth_cfs - lag(canyon_mouth_cfs, n = 3),
         .by = location)

turb_apr_sept_lagQ %>% 
  ggplot(., aes(x = doy, color = location)) +
  geom_line(aes(y = deltaQ_24h/5)) + # standardize a little to for viewing
  geom_point(aes(y = p95-200), color = "purple") +
  coord_cartesian(ylim = c(-200, 100)) + #limit to keep things reasonable
  facet_wrap(year ~ location)

turb_apr_sept_lagQ %>% 
  ggplot(., aes(x = doy, color = location)) +
  geom_line(aes(y = deltaQ_48h/5)) + # standardize a little to for viewing
  geom_point(aes(y = p95-200), color = "purple") +
  coord_cartesian(ylim = c(-200, 100)) + #limit to keep things reasonable
  facet_wrap(year ~ location)

turb_apr_sept_lagQ %>% 
  ggplot(., aes(x = doy)) +
  geom_line(aes(y = deltaQ_72h/5)) + # standardize a little to for viewing
  geom_point(aes(y = p95-200), color = "purple") +
  coord_cartesian(ylim = c(-200, 100)) + #limit to keep things reasonable
  facet_wrap(year ~ location)
```

Well, this looks useful... what does it look like as a regression?

```{r}
turb_apr_sept_lagQ %>% 
  pivot_longer(cols = c(p05:p95), names_to = "quantile", values_to = "turb") %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_24h, y = turb, color =location)) +
  geom_point() +
  facet_grid(quantile ~ month, scales = "free_y")

turb_apr_sept_lagQ %>% 
  pivot_longer(cols = c(p05:p95), names_to = "quantile", values_to = "turb") %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_48h, y = turb, color = location)) +
  geom_point() +
  facet_grid(quantile ~ month, scales = "free_y")

turb_apr_sept_lagQ %>% 
  pivot_longer(cols = c(p05:p95), names_to = "quantile", values_to = "turb") %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_72h, y = turb, color =location)) +
  geom_point() +
  facet_grid(quantile ~ month, scales = "free_y")

```

Look a little closer at p95:

```{r}
turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_24h, y = p95, color = location)) +
  geom_point() 

turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_48h, y = p95, color = location)) +
  geom_point() 

turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_72h, y = p95, color = location)) +
  geom_point() 

```

Not so great. Can we see anything if we facet by month?

```{r}
turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_24h, y = p95, color = location)) +
  geom_point()  +
  coord_cartesian(xlim = c(-1000,1000),
                  ylim = c(0, 500)) +
  facet_wrap(month ~ location)

turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_48h, y = p95, color = location)) +
  geom_point() +
  coord_cartesian(xlim = c(-1000,1000),
                  ylim = c(0, 500)) +
  facet_wrap(month ~ location)

turb_apr_sept_lagQ %>% 
  mutate(month = month(date)) %>% 
  filter(between(month, 4, 9)) %>% 
  ggplot(., aes(x = deltaQ_72h, y = p95, color = location)) +
  geom_point()+
  coord_cartesian(xlim = c(-1000,1000),
                  ylim = c(0, 500)) +
  facet_wrap(month ~ location)

```

Mayyyyybe at the 48/72h lag for May and June.

# Precipitation Events

This one is kind of tricky, we're going to grab data from a single precip gauge,
but we know that precip is variable across space (and the impacts of storms are 
too)

```{r}
precip_data <- get_climate_ts(site_id = "US1COLR0815",
                              param = "Precip",
                              start_date = "2010-01-01",
                              end_date = "2025-12-31",
                              timescale = "day") %>%
  select(datetime, redfeather_precip = value) %>% 
  mutate(date = date(datetime), 
         year = year(date),
         doy = yday(date)) 

summary(precip_data)

turb_q_precip <- full_join(turb_q, precip_data)
```

Plot the two together

```{r}
turb_q_precip %>% 
  ggplot(., aes(x = doy, color = location)) +
  geom_point(aes(y = p95)) +
  geom_line(aes(y = redfeather_precip*100)) +
  coord_cartesian(ylim = c(0, 300)) + #limit to keep things reasonable
  facet_wrap(year ~ location)
```

And with a rolling 2-day sum of precipitation

```{r}
turb_q_precip <- turb_q_precip %>%
  arrange(date) %>%
  mutate(precip_2day = roll_sum(replace_na(redfeather_precip, 0), n = 2, fill = NA, align = "right"),
         .by = location)

ggplot(turb_q_precip, aes(x = doy, color = location)) +
  geom_point(aes(y = p95)) +
  geom_line(aes(y = precip_2day * 100)) +
  coord_cartesian(ylim = c(0, 300)) +
  facet_wrap(year ~ location)

ggplot(turb_q_precip, aes(x = p95, y = precip_2day * 100)) +
  geom_point() +
  facet_wrap(year ~ location)

```

Maybe some responses, but nothing linear at near-real time. This is unsurprising, 
but we press on. 

# Thoughts

If we had a lot of other basins, we might be able to do some spatial models to 
incorporate the land heterogeneity/physical connections, but we don't, so we 
probably need a model that would pick up on patterns that aren't super "visible".
This is a pretty good problem for a Long-Short Term Memory (LSTM) model - we have 
year-round data (mostly) and there's a lot of seasonal/antecedent conditions 
that are important that wouldn't be captured in most other model types. 
