---
title: "Baseline TOC + Q"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

# Purpsoe

The purpose of this markdown is to add some complexity to the DOY / distance
up canyon in `modeling/toc/toc_forecast/baseline_TOC.Rmd` by adding canyon mouth
data, which we have forecast data for and can easily turn into an operational 
model that refelects interannual variability.

## Data

```{r}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
library(tidyverse)
library(here)
library(mgcv)
library(Metrics)
library(cdssr)

source(here("src/collation/load_grab_sample_data.R"))

metrics <- function(df, set_name) {
  data.frame(
    dataset = set_name,
    rmse = sqrt(mean((df$toc_doc - df$pred)^2)),
    r2   = cor(df$toc_doc, df$pred)^2
  )
}

# grab chemistry data
chem <- load_grab_sample_data()
# and location info
locations <- read_csv(here("data/raw/spatial/metadata/water_sampling_sites.csv"))

chem <- left_join(chem, locations)

mainstem <- chem %>% 
  filter(Campaign %in% c("Mainstem - Upper", "South Fork") & 
           location_type %in% c("Mainstem", "Stream", "Inflow")) %>% 
  mutate(DOY = yday(Date),
         year = year(Date), 
         toc_doc = if_else(!is.na(TOC), TOC, DOC),
         # to play with categorical variables, they failed.
         identity = if_else(Campaign == "South Fork", 1, 0),
         fire_signal = if_else(year %in% c(2021, 2022), 1, 0)) %>% 
  # limit time, remove fire signal, remove flier
  filter(!is.na(toc_doc), between(DOY, 100, 320), fire_signal == 0, toc_doc < 9) %>% 
  rename(date = Date)

# grab canyon flow
flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(mainstem$Date, na.rm = T) - days(7),
                       end_date = max(mainstem$Date, na.rm = T) + days(7),
                       timescale = "day") %>%
  select(date = datetime, 
         canyon_mouth_cfs = value) %>% 
  filter(date >= min(mainstem$date) - days(7))

ggplot(flow_data, aes(x = date, y = canyon_mouth_cfs)) +
  geom_point() 

mainstem_plus <- left_join(mainstem, flow_data) %>%
  filter(!is.na(canyon_mouth_cfs))

TOC_GAM_more_q <- gam(toc_doc ~ te(DOY, distance_upstream_km, canyon_mouth_cfs), 
                      data = mainstem_plus)
gam.check(TOC_GAM_more_q)

plot(TOC_GAM_more_q, scheme = 2, too.far = 0.1, shade = TRUE)

vis.gam(TOC_GAM_more_q, view = c("DOY", "distance_upstream_km"),
        plot.type = "persp", theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(TOC_GAM_more_q, view = c("distance_upstream_km", "canyon_mouth_cfs"),
        plot.type = "persp", theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(TOC_GAM_more_q, view = c("DOY", "canyon_mouth_cfs"),
        plot.type = "persp", theta = 30, phi = 30,
        ticktype = "detailed")
```

Okay, there seems to bee some good convergence here. Let's do the same method for
split as the baseline GAM for DOY and distance up stream:

- Test: 2025
- Fold 1: 2015 (both campaigns together)
- Folds 2/3: 2023 Mainstem, 2023 South Fork
- Folds 4/5: 2024 Mainstem, 2024 South Fork

GAM diagnostics are interpreted using the gam.check() function. We want to make
sure that:

1) the model converged (second line of the output), the convergence value is very small,
and ideally model rank is last (26/26 for instance)
2) We are confident we are not finding a local minima in the convergence ("Hessian was positive difinitive")
3) that the k is ~ 1 and the p value > 0.05
4) we want the lowest k (default is 25 or c(5, 5)) while not maxing out the wiggles
which is determined by k' (max number of wiggles) and edf (actual wiggles used)

Some notes on interpretation and tuning:

- If the p-value < 0.05, you need to add more complexity to the model (by increasing k)
- The lower the k means it's less likely to be over fit. 
- As long as the p value > 0.05 and k ~ 1, it's unlikely to be underfit.
- A higher k risks being overfit, and not generalizable. 
- If your edf is very close to k' (like, within 1 unit), you probably need to increase k

My rule of thumb for making sure you don't overfit: 
if the edf value of your model is less than the k' of a lower k, you might be overfit:

edf = 20, k' = 48 (implying the k = c(7,7), or 7x7-1), you'd want to try k = c(6,6),
which would have a k' of 35 (6x6-1), and ideally the edf would still be low with
a smaller parameterization space.

## CV Model 1

```{r}
valid_1 <- mainstem_plus %>% filter(year == 2015)
train_1 <- anti_join(mainstem_plus, valid_1) %>% 
  filter(year != 2025)

# fit GAM
fit_1 <- gam(toc_doc ~ te(DOY, distance_upstream_km, canyon_mouth_cfs), data = train_1, method = "GCV.Cp")

# check GAM diagnostics
gam.check(fit_1)  

# 3D perspective
vis.gam(fit_1,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(fit_1,
        view = c("DOY", "canyon_mouth_cfs"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(fit_1,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# generate predictions
valid_1$pred <- predict(fit_1, newdata = valid_1)

metrics(valid_1, "validation")
```

## CV Model 2

```{r}
valid_2 <- mainstem_plus %>% filter(year == 2023 & Campaign == "Mainstem - Upper")
train_2 <- anti_join(mainstem_plus, valid_2) %>% 
  filter(year != 2025)

# fit GAM
fit_2 <- gam(toc_doc ~ te(DOY, distance_upstream_km, canyon_mouth_cfs), data = train_2, method = "GCV.Cp")

# check GAM diagnostics
gam.check(fit_2)  

# 3D perspective
vis.gam(fit_2,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(fit_2,
        view = c("DOY", "canyon_mouth_cfs"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(fit_2,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# generate predictions
valid_2$pred <- predict(fit_2, newdata = valid_2)

metrics(valid_2, "validation")
```

## CV Model 3

```{r}
valid_3 <- mainstem_plus %>% filter(year == 2023 & Campaign == "South Fork")
train_3 <- anti_join(mainstem_plus, valid_3) %>% 
  filter(year != 2025)

# fit GAM
fit_3 <- gam(toc_doc ~ te(DOY, distance_upstream_km, canyon_mouth_cfs), data = train_3, method = "GCV.Cp")

# check GAM diagnostics
gam.check(fit_3)  

# 3D perspective
vis.gam(fit_3,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(fit_3,
        view = c("DOY", "canyon_mouth_cfs"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(fit_3,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# generate predictions
valid_3$pred <- predict(fit_3, newdata = valid_3)

metrics(valid_3, "validation")
```

## CV Model 4

```{r}
valid_4 <- mainstem_plus %>% filter(year == 2024 & Campaign == "Mainstem - Upper")
train_4 <- anti_join(mainstem_plus, valid_4) %>% 
  filter(year != 2025)

# fit GAM
fit_4 <- gam(toc_doc ~ te(DOY, distance_upstream_km, canyon_mouth_cfs), data = train_4, method = "GCV.Cp")

# check GAM diagnostics
gam.check(fit_4)  

# 3D perspective
vis.gam(fit_4,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(fit_4,
        view = c("DOY", "canyon_mouth_cfs"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(fit_4,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# generate predictions
valid_4$pred <- predict(fit_4, newdata = valid_4)

metrics(valid_4, "validation")
```

## CV Model 5

```{r}
valid_5 <- mainstem_plus %>% filter(year == 2024 & Campaign == "South Fork")
train_5 <- anti_join(mainstem_plus, valid_5) %>% 
  filter(year != 2025)

# fit GAM
fit_5 <- gam(toc_doc ~ te(DOY, distance_upstream_km, canyon_mouth_cfs), data = train_5, method = "GCV.Cp")

# check GAM diagnostics
gam.check(fit_5)  

# 3D perspective
vis.gam(fit_5,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(fit_5,
        view = c("DOY", "canyon_mouth_cfs"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

vis.gam(fit_5,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# generate predictions
valid_5$pred <- predict(fit_5, newdata = valid_5)

metrics(valid_5, "validation")
```

## Apply to test set

```{r}
test_2025 <- mainstem_plus %>% filter(year == 2025)

# apply ensemble
test_2025 <- test_2025 %>%
  mutate(
    pred1 = predict(fit_1,newdata = test_2025),
    pred2 = predict(fit_2, newdata = test_2025),
    pred3 = predict(fit_3, newdata = test_2025),
    pred4 = predict(fit_4, newdata = test_2025),
    pred5 = predict(fit_5, newdata = test_2025),
    pred_avg = (pred1 + pred2 + pred3 + pred4 + pred5) / 5
  )

# calculate performance metrics
mae_val  <- mae(test_2025$toc_doc, test_2025$pred_avg)
rmse_val <- rmse(test_2025$toc_doc, test_2025$pred_avg)
bias_val <- bias(test_2025$toc_doc, test_2025$pred_avg)

# r2 from linear regression
r2_val <- summary(lm(test_2025$toc_doc ~test_2025$pred_avg))$r.squared

# format annotation text for plot
metrics_text <- paste0(
  "MAE = ", round(mae_val, 2),
  "\nRMSE = ", round(rmse_val, 2),
  "\nBias = ", round(bias_val, 2),
  "\nR² = ", round(r2_val, 2)
)


# pred - obs scatter plot!
ggplot(test_2025, aes(y = pred_avg, x = toc_doc)) +
  geom_point(color = "blue", alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (2025)",
       subtitle = "TOC ~ DOY + distance upstream + canyon mouth flow") +
  annotate("text", x = 1, y = 10, 
           hjust = 0, vjust = 1, 
           label = metrics_text, size = 4) +
  theme_bw(base_size = 14) +
  coord_cartesian(xlim = c(1, 10), ylim = c(1, 10))

# for kicks, look at performance by site
ggplot(test_2025, aes(y = pred_avg, x = toc_doc)) +
  geom_point(color = "blue", alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (2025)",
       subtitle = "TOC ~ DOY + distance upstream + canyon mouth flow") +
  theme_bw(base_size = 14) +
  coord_cartesian(xlim = c(1, 10), ylim = c(1, 10)) +
  facet_wrap(site_name ~ .)

```

This looks pretty good, but it honestly performs about the same as the simple
model of DOY + distance upstream. 

# Do a quick sensitivity check

We're going to raise the canyon flow by 25% and see if TOC increases here. This
will give us a better idea of whether or not adding flow actually might be better
in an operational scenario.

```{r}
test_2025_sens <- test_2025 %>% 
  mutate(canyon_mouth_cfs = canyon_mouth_cfs*1.25)

# apply ensemble for test set
test_2025_sens <- test_2025_sens %>%
  mutate(
    pred1 = predict(fit_1, newdata = test_2025_sens),
    pred2 = predict(fit_2, newdata = test_2025_sens),
    pred3 = predict(fit_3, newdata = test_2025_sens),
    pred4 = predict(fit_4, newdata = test_2025_sens),
    pred5 = predict(fit_5, newdata = test_2025_sens),
    pred_avg = (pred1 + pred2 + pred3 + pred4 + pred5) / 5
  )

# bias should change
bias_val_sens <- bias(test_2025_sens$toc_doc, test_2025_sens$pred_avg)

# pred - obs scatter plot!
ggplot(test_2025_sens, aes(y = pred_avg, x = toc_doc)) +
  geom_point(color = "blue", alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (2025)",
       subtitle = "TOC ~ DOY + distance upstream + 1.25 * canyon mouth flow") +
  theme_bw(base_size = 14) +
  coord_cartesian(xlim = c(1, 10), ylim = c(1, 10))

test_2025_sens <- test_2025_sens %>% 
  select(date, site_name, toc_doc, pred_sens = pred_avg)

sensitivity <- test_2025 %>% 
  select(date, site_name, toc_doc, pred_avg) %>% 
  left_join(., test_2025_sens) %>% 
  unique()

ggplot(sensitivity, aes(x = pred_avg, y = pred_sens)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  theme_bw() +
  labs(x = "GAM predicted",
       y = "GAM predicted 25% increase in flow")


```

Okay, there's some sensitivity here, let's try this model on 2015 data and see
what it looks like - 2015 had 2x max flow than 2025, and we want to know if we 
under predict or overpredict at all that year. Keep in mind that these data have
already been "seen" by the GAM, so it's kind of cheating, but since we're using
a 5-fold CV, this should give us a bit of an idea about performance in a scenario
with really high Q.

```{r}
test_2015 <- mainstem_plus %>% 
  filter(year(date) == 2015)

# apply ensemble for test set
test_2015 <- test_2015 %>%
  mutate(
    pred1 = predict(fit_1, newdata = test_2015),
    pred2 = predict(fit_2, newdata = test_2015),
    pred3 = predict(fit_3, newdata = test_2015),
    pred4 = predict(fit_4, newdata = test_2015),
    pred5 = predict(fit_5, newdata = test_2015),
    pred_avg = (pred1 + pred2 + pred3 + pred4 + pred5) / 5
  )

# bias should change
bias_val_2015 <- bias(test_2015$toc_doc, test_2015$pred_avg)

# pred - obs scatter plot!
ggplot(test_2015, aes(y = pred_avg, x = toc_doc)) +
  geom_point(color = "blue", alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (2025)",
       subtitle = "TOC ~ DOY + distance upstream + 1.25 * canyon mouth flow") +
  theme_bw(base_size = 14) +
  coord_cartesian(xlim = c(1, 10), ylim = c(1, 10))

```

Well, it's right on. Interesting that even though Q was 2x, the max TOC was
very similar to 2025. There were also many fewer samples in 2015, and it doesn't
look like there were samples during the peak runoff, so interpret with a grain
of salt.

# Visualize the decision space for interpretability

GAMs are pretty interpretable when they have fewer interacting variables. In this
case, we can just create a synthetic space/time space and apply the model across
it to visualize what happens up the canyon over time.

```{r}
# Make a prediction grid across DOY and distance
newdata <- expand.grid(
  DOY = seq(min(mainstem_plus$DOY), 
            max(mainstem_plus$DOY), 
            5),
  distance_upstream_km = seq(min(mainstem_plus$distance_upstream_km),
                             max(mainstem_plus$distance_upstream_km), 10),
  canyon_mouth_cfs = seq(min(mainstem_plus$canyon_mouth_cfs),
                         max(mainstem_plus$canyon_mouth_cfs),
                         100)
)

# Get predictions from your GAMs for the grid
newdata$fit1 <- predict(fit_1, newdata = newdata)
newdata$fit2 <- predict(fit_2, newdata = newdata)
newdata$fit3 <- predict(fit_3, newdata = newdata)
newdata$fit4 <- predict(fit_4, newdata = newdata)
newdata$fit5 <- predict(fit_5, newdata = newdata)

# create ensemble prediction
newdata <- newdata %>% 
  mutate(ave_fit = (fit1 + fit2 + fit3 + fit4 + fit5) / 5)

# heatmap + contour plot
ggplot(newdata, aes(x = DOY, y = distance_upstream_km, fill = ave_fit)) +
  geom_tile() +
  geom_contour(aes(z = ave_fit), color = "white", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    fill = "Partial TOC",
    x = "Day of Year",
    y = "Distance upstream from Canyon Mouth (km)"
  ) +
  theme_minimal(base_size = 14) 

ggplot(newdata, aes(x = distance_upstream_km, y = canyon_mouth_cfs, fill = ave_fit)) +
  geom_tile() +
  geom_contour(aes(z = ave_fit), color = "white", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    fill = "Partial TOC",
    y = "Q at the canyon mouth",
    x = "Distance upstream from Canyon Mouth (km)"
  ) +
  theme_minimal(base_size = 14) 

ggplot(newdata, aes(x = DOY, y = canyon_mouth_cfs, fill = ave_fit)) +
  geom_tile() +
  geom_contour(aes(z = ave_fit), color = "white", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    fill = "Partial TOC",
    x = "Day of Year",
    y = "Q at the canyon mouth"
  ) +
  theme_minimal(base_size = 14) 

```


Save models

```{r}

write_rds(fit_1, here("data/derived/models/TOC_baseline/working/TOC_GAM_addQ_fit1_v2026-01-23.rds"))
write_rds(fit_2, here("data/derived/models/TOC_baseline/working/TOC_GAM_addQ_fit2_v2026-01-23.rds"))
write_rds(fit_3, here("data/derived/models/TOC_baseline/working/TOC_GAM_addQ_fit3_v2026-01-23.rds"))
write_rds(fit_4, here("data/derived/models/TOC_baseline/working/TOC_GAM_addQ_fit4_v2026-01-23.rds"))
write_rds(fit_5, here("data/derived/models/TOC_baseline/working/TOC_GAM_addQ_fit5_v2026-01-23.rds"))

```
