---
title: "FC Intake TOC Exploration"
author: "ROSSyndicate"
date: "2026-01-24"
output: html_document
---

# Purpose

This markdown is meant to explore seasonal maxima of TOC data at the FC intake 
and the relationship between TOC and the SWE from Snotel - this is a baseline
for a seasonal forecast.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(here)
library(cdssr)
library(snotelr)
library(Metrics)
```

# Load in data

```{r, message = FALSE}
# intake TOC
fc_intake <- read_csv(here("data/collated/FC_intake/fc_intake_TOC_record_2010-2025.csv"))

# read in flow data at canyon mouth
flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = "2010-01-01",
                       end_date = "2025-12-31",
                       timescale = "day") %>%
  select(datetime, canyon_mouth_cfs = value) %>% 
  mutate(date = date(datetime), 
         year = year(date),
         doy = yday(date))


# grab snotel data in HUC8
snotel_sites <- snotel_info() %>%
  filter(grepl("10190007", description))

# Download daily SNOTEL data
snotel_data <- snotel_download(site_id = snotel_sites$site_id,
                               internal = TRUE,
                               metric = TRUE) %>% 
  mutate(date = ymd(date),
         doy = yday(date),
         year = year(date))

```

# Summarize data to annual stats

Here, we'll grab the max TOC, max flow, and maximum SWE per year (and per site
for SWE). This summarizes the data pretty succinctly for a simple descriptive 
model of the data on hand. We mostly care about magnitude and timing.

```{r, message = FALSE}
fc_annual_max <- fc_intake %>% 
  arrange(-mean_TOC) %>% 
  slice(1, .by = year) %>% 
  arrange(year) %>% 
  rename(toc_max_date = date,
         toc_max_doy = doy) 

q_max <- flow_data %>% 
  select(-datetime) %>% 
  filter(between(doy, 100, 250)) %>% 
  arrange(-canyon_mouth_cfs) %>% 
  slice(1, .by = year) %>% 
  arrange(year) %>% 
  rename(q_max_date = date,
         q_max_doy = doy) 

swe_max <- snotel_data %>% 
  select(doy, year, site_id, snow_water_equivalent:temperature_mean) %>% 
  arrange(-snow_water_equivalent) %>% 
  slice(1, .by = c(year, site_id)) %>% 
  arrange(year) %>% 
  rename(snotel_doy = doy) %>% 
  mutate(snotel_doy = if_else(snotel_doy > 200, snotel_doy - 365, snotel_doy))

max_q_toc <- full_join(fc_annual_max, q_max) %>% 
  left_join(., swe_max)

```

## Snotel correlations

Let's just look at the snotel data across stations and see where our best 
correlations arefor TOC magnitude and timing.

```{r, message = FALSE}
ggplot(max_q_toc, aes(x = toc_max_doy, y = snotel_doy)) +
  facet_wrap(site_id ~ .) +
  geom_smooth(method = "lm", se = FALSE, color = "grey80", lty = 2) +
  geom_point() +
  labs(x = "TOC maxima DOY",
       y = "day of SWE maximum") +
  theme_bw()

ggplot(max_q_toc, aes(x = toc_max_doy, y = snow_water_equivalent)) +
  facet_wrap(site_id ~ ., scales = "free_y") +
  geom_smooth(method = "lm", se = FALSE, color = "grey80", lty = 2) +
  geom_point() +
  labs(x = "TOC maxima DOY",
       y = "SWE maximum") +
  theme_bw()

ggplot(max_q_toc, aes(x = mean_TOC, y = snow_water_equivalent)) +
  facet_wrap(site_id ~ ., scales = "free_y") +
  geom_smooth(method = "lm", se = FALSE, color = "grey80", lty = 2) +
  geom_point() +
  labs(x = "TOC maxima (mg/L)",
       y = "SWE maximum") +
  theme_bw()

ggplot(max_q_toc, aes(x = canyon_mouth_cfs, y = snow_water_equivalent)) +
  facet_wrap(site_id ~ ., scales = "free_y") +
  geom_smooth(method = "lm", se = FALSE, color = "grey80", lty = 2) +
  geom_point() +
  labs(x = "Q maximum @ canyon mouth (cfs)",
       y = "SWE maximum") +
  theme_bw()

ggplot(max_q_toc, aes(x = q_max_doy, y = snotel_doy)) +
  facet_wrap(site_id ~ ., scales = "free_y") +
  geom_smooth(method = "lm", se = FALSE, color = "grey80", lty = 2) +
  geom_point() +
  labs(x = "day of Q maximum @ canyon mouth",
       y = "day of SWE maximum") +
  theme_bw()

```

Interestingly enough, the SNOTEL station best correlated with max TOC and peak 
canyon Q is Deadman (North Fork). For the sake of this exploration, we are
only going to use the Snotel data from that station. 

```{r, message = FALSE}
max_q_toc <- max_q_toc %>% 
  filter(site_id == 438)
```

### Snotel data reduction

```{r, message = FALSE}
snotel_data %>% 
  filter(year >= 2010, site_id == 438) %>% 
  select(snow_water_equivalent:precipitation) %>% 
  plot()
```

Given the high correlation between snow depth and swe, and min and mean temp, 
we will drop snow depth and mean temp.

```{r, message = FALSE}
max_q_toc <- max_q_toc %>% 
  select(-c(snow_depth, temperature_mean))
```

## Flow/toc correlations

Let's explore some correlations in this dataset between the variables.

```{r, message = FALSE}
ggplot(max_q_toc, aes(x = q_max_doy, y = toc_max_doy, 
                      color = mean_TOC, size = canyon_mouth_cfs)) +
  coord_cartesian(xlim = c(120, 200), ylim = c(120, 200)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0,
              slope = 1,
              lty = 2,
              color = "grey80") +
  labs(x = "day of max Q @ canyon mouth",
       y = "day of max TOC @ intake",
       color = "max TOC (mg/L)",
       size = "max Q @ canyon mouth (cfs)") +
  scale_color_viridis_c() +
  theme_bw()

ggplot(max_q_toc, aes(y = mean_TOC, x = canyon_mouth_cfs)) +
  geom_smooth(se = FALSE, method = "lm", color = "orange", lty = 2) +
  geom_point(aes(size = snow_water_equivalent), alpha = 0.5) +
  theme_bw() +
  labs(x = "max Q @ canyon mouth (cfs)",
       y = "max TOC @ intake (mg/L)",
       size = "max SWE @ deadman") +
  theme(legend.position = "bottom")
```

Great, there are some kind of linear relationships here. 

# Describing magnitude of TOC

Let's just use the magnitude/timing from the Q/SWE data here. I've explored other
combinations, and this one (linear model of maximum Q
at the canyon mouth) seems to be the best performing. 

For reference, I tride a multiple linear regression...

```{r, message = FALSE}
lm_TOC <- lm(mean_TOC ~ canyon_mouth_cfs + snow_water_equivalent,
             data = max_q_toc)

summary(lm_TOC)
```

But the slopes aren't significant.

```{r, message = FALSE}
lm_TOC <- lm(mean_TOC ~ canyon_mouth_cfs,
             data = max_q_toc)

summary(lm_TOC)
```

This is better for significance, so we'll apply it/visualize.

```{r, message = FALSE}
max_q_toc <- max_q_toc %>% 
  mutate(pred_toc = predict(lm_TOC))

ggplot(max_q_toc, aes(x = mean_TOC, 
                      y = pred_toc)) +
  geom_point() +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(6, 13), ylim = c(6, 13)) +
  labs(x = "maximum TOC in runoff period (mg/L)",
       y = "predicted max TOC in runoff period (mg/L)")

RMSE = rmse(max_q_toc$mean_TOC, max_q_toc$pred_toc)
MAE = mae(max_q_toc$mean_TOC, max_q_toc$pred_toc)
BIAS = bias(max_q_toc$mean_TOC, max_q_toc$pred_toc)

```

There is a caveat here, though:


```{r, message = FALSE}
ggplot(max_q_toc, aes(x = toc_max_doy, y = q_max_doy)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  theme_bw() +
  coord_cartesian(xlim = c(90, 200), ylim = c(90, 200)) +
  labs(x = "day of max TOC",
       y = "day of max Q")
```

There are instances where the max Q occurs AFTER max TOC.

# Describing timing of max TOC

We'll have the same caveat of peak Q timing for these data.

```{r, message = FALSE}
lm_TOC_doy <- lm(toc_max_doy ~ q_max_doy +
                   snotel_doy,
                 data = max_q_toc)

summary(lm_TOC_doy)
```

Same issue here with significance.

```{r, message = FALSE}
lm_TOC_swe <- lm(toc_max_doy ~ snotel_doy,
                 data = max_q_toc)

summary(lm_TOC_swe)
```

But this time we loose some correlation. We'll apply both, with the 
acknowledgement that there isn't statistical significance for adding max Q doy.

```{r, message = FALSE}
max_q_toc <- max_q_toc %>% 
  mutate(pred_toc_doy = predict(lm_TOC_doy),
         pred_toc_swe = predict(lm_TOC_swe))

max_q_toc %>% 
  pivot_longer(cols = c(pred_toc_doy, pred_toc_swe),
               names_to = "model",
               values_to = "doy") %>% 
  mutate(model = factor(model, 
                        levels = c("pred_toc_doy", "pred_toc_swe"), 
                        labels = c("swe + q", "swe"))) %>% 
  ggplot(., aes(x = toc_max_doy, y = doy, color = model)) +
  geom_point() +
  theme_bw() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  labs(x = "day of maximum TOC",
       y = "predicted day of maximum TOC",
       color = "covariates") +
  coord_cartesian(xlim = c(120, 165), ylim = c(120, 165)) 

max_q_toc %>% 
  summarize(RMSE_doy = rmse(max_q_toc$toc_max_doy, max_q_toc$pred_toc_doy),
            MAE_doy = mae(max_q_toc$toc_max_doy, max_q_toc$pred_toc_doy),
            BIAS_doy = bias(max_q_toc$toc_max_doy, max_q_toc$pred_toc_doy),
            RMSE_swe = rmse(max_q_toc$toc_max_doy, max_q_toc$pred_toc_swe),
            MAE_swe = mae(max_q_toc$toc_max_doy, max_q_toc$pred_toc_swe),
            BIAS_swe = bias(max_q_toc$toc_max_doy, max_q_toc$pred_toc_swe))

```

And just using SWE date:

```{r, message = FALSE}
ggplot(max_q_toc, aes(x = toc_max_doy, y = pred_toc_swe)) +
  #geom_smooth(method = "lm", se = FALSE, color = "grey80") +
  geom_point() +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  labs(x = "day of maximum TOC",
       y = "predicted day of maximum TOC") +
  coord_cartesian(xlim = c(120, 165), ylim = c(120, 165)) 
```

Since we're using max SWE here, let's look at timing for that like we did 
with Q.

```{r, message = FALSE}
ggplot(max_q_toc, aes(x = toc_max_doy, y = snotel_doy)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  theme_bw() +
  coord_cartesian(xlim = c(90, 200), ylim = c(90, 200)) +
  labs(x = "day of max TOC",
       y = "day of max SWE") 

```

Great. No caveat here.

Probably should try some ridge regression here to see if it allows for the use
of additional variables without overfitting and reducing significance of variables
with colinearity.

