---
title: "Alkalinity Baseline Forecast Exploration"
author: "Danny Duncan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(mgcv)
library(scoringRules)
library(tidyverse)
library(lubridate)
library(kableExtra)

source("src/setup_libraries.R")
```

# Background

Disinfectant byproducts (DBPs) are a concern for water utilities due to their negative health impacts, namely DBPs associated with Chlorine. Alkalinity indirectly affects the efficiency of Chlorine and the production of DBPs from treatment. Per the EPA's Disinfectant Byproduct Rule, water treatment facility operators must follow a TOC removal matrix to reduce the production of DBP and DBP precursors.

```{r toc-removal-table, include=FALSE}
tibble(
  `Source Water TOC (mg/L)` = c(">2.0 to 4.0", ">4.0 to 8.0", ">8.0"),
  `0-60` = c("35.0%", "45.0%", "50.0%"),
  `>60-120` = c("25.0%", "35.0%", "40.0%"),
  `>120` = c("15.0%", "25.0%", "30.0%")
) %>%
  kable(align = "c", caption = "EPA Required Percent Removal of TOC (40 CFR Part 141, Subpart H)") %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"), full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Source Water Alkalinity, mg/L, as CaCO3" = 3)) %>%
  column_spec(1, bold = TRUE)
```

# Data Load

```{r, message = FALSE}
# Historical chemical data from Greeley intake (2001 - 2025)
collate_path <- here("data/collated/chem")
file_path <- list.files(collate_path,
  pattern = "^greeley_intake_chem_data_.*?\\.parquet$",
  full.names = TRUE
) %>% tail(1)

plant_data <- read_parquet(file_path)
```

```{r flow data, warning=FALSE}
# Read in flow data
cm_flow_data <- get_sw_ts(
     abbrev = "CLAFTCCO",
     start_date = min(plant_data$date, na.rm = T),
     end_date = max(plant_data$date, na.rm = T),
     timescale = "day"
 ) %>%
  rename(cm_cfs = value) %>%
  mutate(cm_cfs_rescale = rescale(cm_cfs, to = c(0,100))) %>%
  select(datetime, cm_cfs, cm_cfs_rescale)

nf_flow_data <- get_sw_ts(
  abbrev = "CLANSECO",
  start_date = min(plant_data$date, na.rm = T),
  end_date = max(plant_data$date, na.rm = T),
  timescale = "day"
 ) %>%
  rename(nf_cfs = value) %>%
  mutate(nf_cfs_rescale = rescale(nf_cfs, to = c(0,100))) %>%
  select(datetime, nf_cfs, nf_cfs_rescale)

flow_data <- cm_flow_data %>%
  left_join(nf_flow_data, by = c("datetime")) %>%
  mutate(flow_ratio = nf_cfs / cm_cfs) %>%
  rename(date = datetime)
```

```{r combine data, warning=FALSE}
# Combine flow and plant data
all_data <- plant_data %>%
  left_join(flow_data, by = "date") %>% 
  mutate(date = as.Date(date),
         doy = yday(date),
         year = year(date)) %>% 
  filter(!is.na(alkalinity), !is.na(flow_ratio), !is.na(doy)) %>%
  arrange(date)
```

# GAM Exploration

From our initial exploration, the proportion of North Fork flow to Canyon Mouth flow (flow_ratio) is the best predictor when paired with day-of-year (doy). Unlike the Canyon Mouth, no flow forecast is available for the North Fork. Since this is the case, we'll create a forecast which calculates "flow_ratio" from the Canyon Mouth forecast for n+7 while using the n+0 observation from the North Fork.

## Data Prep
```{r data prep}
# Data Prep
clean_data <- all_data %>%
  mutate(
    date = as.Date(date),
    year = year(date),
    doy = as.numeric(doy)
  ) %>%
  filter(!is.na(alkalinity), !is.na(nf_cfs), !is.na(cm_cfs), alkalinity > 0) %>%
  arrange(date)

# Train/Test Data Split
train_data <- clean_data %>% filter(year <= 2022)
```

## Model Building
```{r gam-model}
# Build GAM with CM/NF flow ratio & Day-of-Year
# Cyclic spline utilized to capture seasonality in DOY
# Gamma distribution with log link for right-skewed Alkalinity data
m1 <- gam(
  alkalinity ~ s(flow_ratio) + s(doy, bs = "cc"),
  data = train_data,
  family = Gamma(link = "log"),
  method = "REML"
)

gam.check(m1)
summary(m1)
```
## Test Data Evaluation (Observed Flow Ratio)

```{r gam obs test}
# Create test data (2023-2025)
test_data <- clean_data %>% filter(year >= 2023)

# Predict using observed flow_ratio and doy
test_preds <- predict(m1, newdata = test_data, type = "response", se.fit = TRUE)

test_data <- test_data %>%
  mutate(
    alk_pred_obs = test_preds$fit,
    se_obs = test_preds$se.fit
  )
```

## Forecast Simulation

```{r gam-forecast-test}
# Create 7-day forecast simulation dataset for 2023-2024
# For every "today", we create a 7-day forecast window.
test_dates <- clean_data %>% 
  filter(year >= 2023) %>% 
  pull(date) %>% 
  unique()

# Initialize the forecast horizon (1 to 7 days for every date)
sim_skeleton <- expand_grid(
  origin_date = test_dates,
  lead_time = 1:7
) %>%
  mutate(target_date = origin_date + lead_time)

daily_sim_df <- sim_skeleton %>%
  # Get the NF flow from the ORIGIN date (Persisted)
  left_join(clean_data %>% select(date, nf_cfs), by = c("origin_date" = "date")) %>%
  rename(nf_flow_obs = nf_cfs) %>%
  # Get the CM flow and Alkalinity from the TARGET date (Forecasted/Actual)
  left_join(clean_data %>% select(date, cm_cfs, alkalinity, doy), by = c("target_date" = "date")) %>%
  rename(cm_flow_fcst = cm_cfs, alkalinity_actual = alkalinity) %>%
  # Calculate predictors for the model
  mutate(
    flow_ratio_sim = nf_flow_obs / cm_flow_fcst,
    # Ensure we drop any rows where we don't have target data
    doy = yday(target_date)
  ) %>%
  filter(!is.na(alkalinity_actual), !is.na(flow_ratio_sim))

# Generate predictions for all simulated forecasts; pass simulated flow ratio and DOY to GAM
alk_preds <- predict(m1, 
                     newdata = data.frame(
                       flow_ratio = daily_sim_df$flow_ratio_sim,
                       doy = daily_sim_df$doy
                     ), 
                     type = "response", 
                     se.fit = TRUE)

daily_sim_df <- daily_sim_df %>%
  mutate(
    alk_pred = alk_preds$fit,
    se = alk_preds$se.fit,
    upper_ci = alk_pred + (1.96 * se),
    lower_ci = alk_pred - (1.96 * se),
    error = alk_pred - alkalinity_actual
  )

# Analyze Performance by Lead Time
# Reveals how much the 'Persisted NF Flow' assumption hurts accuracy over time
performance_summary <- daily_sim_df %>%
  group_by(lead_time) %>%
  summarise(
    RMSE = sqrt(mean(error^2)),
    MAE = mean(abs(error)),
    Bias = mean(error),
    n_simulations = n()
  )

print(performance_summary)
```

```{r gam forecast plot}
# Plot day 1 & 7 forecast against observed Alkalinity
plot_data <- daily_sim_df %>%
  filter(lead_time %in% c(1, 7)) %>% 
  mutate(
    year = year(target_date),
    lead_label = paste0("Day ", lead_time, " Forecast")
  )

ggplot(plot_data) +
  # Observed Alkalinity
  geom_line(aes(x = target_date, y = alkalinity_actual, color = "Actual"), 
            linewidth = 0.8, alpha = 0.5) +
  # Predicted Alkalinity for Day 1 and Day 7 "forecasts"
  geom_line(aes(x = target_date, y = alk_pred, color = lead_label), 
            linewidth = 0.8) +
  facet_wrap(~year, scales = "free_x", ncol = 1) +
  scale_color_manual(values = c("Actual" = "black", 
                                "Day 1 Forecast" = "dodgerblue", 
                                "Day 7 Forecast" = "firebrick")) +
  labs(
    title = "Alkalinity Forecast vs Actual (2023-2025)",
    subtitle = "Comparing the 1-Day Lead (using yesterday's NF) vs 7-Day Lead (using 7-day old NF)",
    x = "Date",
    y = "Alkalinity (mg/L)",
    color = "Legend"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


```{r obs vs. preds scatterplot, fig.dim=c(10, 6)}
# --- X-Y Scatterplot of Measured vs Predicted Alkalinity ---
plot_df <- daily_sim_df %>%
  mutate(horizon_label = factor(paste0("Day ", lead_time, " Forecast"), 
                                levels = paste0("Day ", 1:7, " Forecast")))

ggplot(plot_df, aes(x = alkalinity_actual, y = alk_pred)) +
  geom_point(alpha = 0.4, color = "dodgerblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~horizon_label, ncol = 4) +
  labs(
    title = "Measured vs Predicted Alkalinity",
    subtitle = "Comparing performance across 1-7 day forecast horizons",
    x = "Measured Alkalinity (mg/L)",
    y = "Predicted Alkalinity (mg/L)"
  ) +
  theme_bw() +
  coord_equal()
```

```{r crps calculation}
phi <- summary(m1)$dispersion 

# Calculate CRPS for each forecast point in the daily_sim_df
crps_results <- daily_sim_df %>%
  mutate(
    # The Gamma distribution in R uses shape and scale
    # Mean (mu) = shape * scale
    # Variance = shape * scale^2 = mu^2 * phi
    # Therefore:
    shape_param = 1 / phi,
    scale_param = alk_pred * phi
  ) %>%
  mutate(
    # crps_gamma(actual_observation, shape, scale)
    crps = crps_gamma(alkalinity_actual, shape = shape_param, scale = scale_param)
  )

# Calculate CRPS for Observed Flow (Test Data)
test_crps <- test_data %>%
  mutate(
    shape_param = 1 / phi,
    scale_param = alk_pred_obs * phi,
    crps = crps_gamma(alkalinity, shape = shape_param, scale = scale_param),
    lead_time = 0 # Representing observed flow as lead time 0
  )
```

```{r crps plots, fig.dim=c(12, 6)}
# Plot 1: Observed Flow (Test Data)
ggplot(test_crps, aes(x = "Observed Flow", y = crps)) +
  geom_violin(fill = "lightblue", alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  theme_bw() +
  labs(y = "Continuous Ranked Probability Score (mg/L)",
       x = "Model Condition",
       title = "CRPS: Perfect Foresight",
       subtitle = "Using observed flow_ratio")

# Plot 2: Forecast Simulation (1-7 Days)
ggplot(crps_results, aes(x = as.factor(lead_time), y = crps)) +
  geom_violin(fill = "lightgray", alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  theme_bw() +
  labs(y = "Continuous Ranked Probability Score (mg/L)",
       x = "Forecast Horizon (Days)",
       title = "CRPS: Forecast Horizons",
       subtitle = "1 to 7 Day Horizon")
```

# Extreme Forecast Sensitivity Test
```{r}
# Create a 7-day horizon for every single day in the test period
rolling_sens_df <- expand_grid(
  origin_date = test_dates,
  lead_time = 1:7
) %>%
  mutate(target_date = origin_date + lead_time) %>%
  # Join Origin NF Flow (Persisted)
  left_join(clean_data %>% select(date, nf_cfs), by = c("origin_date" = "date")) %>%
  rename(nf_flow_obs = nf_cfs) %>%
  # Join Target CM Flow (Acting as Forecast) and Actual Alkalinity
  left_join(clean_data %>% select(date, cm_cfs, alkalinity, doy), by = c("target_date" = "date")) %>%
  rename(cm_flow_fcst = cm_cfs, alkalinity_actual = alkalinity) %>%
  filter(!is.na(alkalinity_actual), !is.na(nf_flow_obs)) %>%
  # Sensitivity modification
  mutate(
    # Scenario A: Control (Baseline Forecast)
    flow_ratio_control = nf_flow_obs / cm_flow_fcst,
    # Scenario B: Over-forecast Sensitivity (+25% CM Flow)
    cm_flow_sens_high = cm_flow_fcst * 1.25,
    flow_ratio_sens_high = nf_flow_obs / cm_flow_sens_high,
    # Scenario C: Under-forecast Sensitivity (-25% CM Flow)
    cm_flow_sens_low = cm_flow_fcst * 0.75,
    flow_ratio_sens_low = nf_flow_obs / cm_flow_sens_low
  )

# Generate predictions for control & sensitivity
# Predict Control (Baseline Forecast)
rolling_sens_df$alk_pred_control <- predict(m1, 
                                            newdata = data.frame(
                                              flow_ratio = rolling_sens_df$flow_ratio_control,
                                              doy = rolling_sens_df$doy
                                            ), type = "response")

# Predict Over-forecast (+25%)
rolling_sens_df$alk_pred_sens_high <- predict(m1, 
                                          newdata = data.frame(
                                            flow_ratio = rolling_sens_df$flow_ratio_sens_high,
                                            doy = rolling_sens_df$doy
                                          ), type = "response")

# Predict Under-forecast (-25%)
rolling_sens_df$alk_pred_sens_low <- predict(m1, 
                                          newdata = data.frame(
                                            flow_ratio = rolling_sens_df$flow_ratio_sens_low,
                                            doy = rolling_sens_df$doy
                                          ), type = "response")

# Calculate the Delta (Impact of forecast errors on alkalinity)
rolling_sens_df <- rolling_sens_df %>%
  mutate(
    diff_mgL_high = alk_pred_sens_high - alk_pred_control,
    diff_mgL_low = alk_pred_sens_low - alk_pred_control
  )

# Performance and Sensitivity Analysis
sens_summary <- rolling_sens_df %>%
  group_by(lead_time) %>%
  summarise(
    avg_control_pred = mean(alk_pred_control),
    avg_sens_high_pred = mean(alk_pred_sens_high),
    avg_sens_low_pred = mean(alk_pred_sens_low),
    mean_diff_high = mean(diff_mgL_high),
    pct_impact_high = (mean_diff_high / avg_control_pred) * 100,
    mean_diff_low = mean(diff_mgL_low),
    pct_impact_low = (mean_diff_low / avg_control_pred) * 100
  )

print(sens_summary)

# Visualizing yearly summaries
# Compare the average forecast of scenarios vs Actual across the test years
yearly_summary <- rolling_sens_df %>%
  group_by(target_date) %>%
  summarise(actual = first(alkalinity_actual),
            pred_control = mean(alk_pred_control),
            pred_sens_high = mean(alk_pred_sens_high),
            pred_sens_low = mean(alk_pred_sens_low)) %>%
  mutate(year = year(target_date))

ggplot(yearly_summary, aes(x = target_date)) +
  geom_line(aes(y = actual, color = "Actual"), alpha = 0.4) +
  geom_line(aes(y = pred_control, color = "Baseline Forecast"), linewidth = 0.8) +
  geom_line(aes(y = pred_sens_high, color = "Over-forecast (+25%)"), linewidth = 0.8, linetype = "dashed") +
  geom_line(aes(y = pred_sens_low, color = "Under-forecast (-25%)"), linewidth = 0.8, linetype = "dotted") +
  facet_wrap(~year, scales = "free_x", ncol = 1) +
  scale_color_manual(values = c("Actual" = "gray50", 
                                "Baseline Forecast" = "blue", 
                                "Over-forecast (+25%)" = "red",
                                "Under-forecast (-25%)" = "darkorange")) +
  labs(title = "Forecast Sensitivity: 2023-2025",
       subtitle = "Testing the impact of +/- 25% errors in CM Flow forecasts",
       y = "Alkalinity (mg/L)",
       color = "Scenario") +
  theme_minimal()
```
