---
title: "greeley_baseline_alkalinity"
author: "Danny Duncan"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%")

library(tidyverse)
library(Metrics)
library(mgcv)
library(corrplot)
library(forecast)
library(here)

source("src/setup_libraries.R")
source("src/collation/load_grab_sample_data.R")
source("src/collation/load_sensor_data.R")
```

# Overview

# Data Preparation

## Load Datasets

### Greeley Chemical Plant Intake (2001 - 2025)

```{r load greeley data}
# Load path to collated data folder.
collate_path <- here("data/collated/chem")

# FInd greeley parquet file path in collated data folder.
file_path <- list.files(collate_path,
  pattern = "^greeley_intake_chem_data_.*?\\.parquet$",
  full.names = TRUE
)

# Read in the parquet file.
plant_data <- read_parquet(file_path) %>%
  filter(!is.na(alkalinity) | !is.na(conductivity) | !is.na(TDS)) %>%
  mutate(
    year = year(date),
    doy = yday(date)
  )

plant_data_long <- plant_data %>%
  pivot_longer(
    cols = -c(date, year, doy, time),
    names_to = "variable",
    values_to = "value",
    values_drop_na = T
  )

alk_data_long <- plant_data %>%
  pivot_longer(
    cols = -c(date, year, doy, time, alkalinity),
    names_to = "variable",
    values_to = "value",
    values_drop_na = T
  )

```

### Grab Sample

```{r load grab sample data}
grab_data <- load_grab_sample_data() %>%
  filter(!is.na(Alkalinity),
         #filter to sites where we have sensor data
         site_code %in% c("pbr", "pnf", # close to PMAN/ FC Intake
                          "pbd", # At Greeley Intake
                          "chd", "sfm", 
                          "nfg"))%>%
  select(Date, site_code, site_name, Alkalinity, SC, TSS, TOC, pH)%>%
  mutate(year = year(Date), 
         doy = yday(Date))%>%
  mutate(ws = if_else(site_code == "nfg", "NF", "CLP"))

```

### Flow Data
```{r load flow data}
# Load in flow data for the canyon mouth (CLAFTCCO)
canyon_flow_data <- get_sw_ts(
     abbrev = "CLAFTCCO",
     start_date = min(plant_data$date, na.rm = T),
     end_date = max(plant_data$date, na.rm = T),
     timescale = "day"
 )%>%
  rename(canyon_mouth_cfs = value)%>%
  mutate(canyon_mouth_cfs_rescale = rescale(canyon_mouth_cfs, to = c(0,100)))%>%
  select(datetime, canyon_mouth_cfs, canyon_mouth_cfs_rescale)

# Load in flow data for the North Fork (CLANSECO)
nf_flow_data <- get_sw_ts(
     abbrev = "CLANSECO",
     start_date = min(plant_data$date, na.rm = T),
     end_date = max(plant_data$date, na.rm = T),
     timescale = "day"
 )%>%
  rename(nf_cfs = value)%>%
  mutate(nf_cfs_rescale = rescale(nf_cfs, to = c(0,100)))%>%
  select(datetime, nf_cfs, nf_cfs_rescale)

# Calculate flow ratio of North Fork to canyon mouth
flow_data <- canyon_flow_data %>%
  left_join(nf_flow_data, by = c("datetime"))%>%
  mutate(nf_prop = nf_cfs / canyon_mouth_cfs)%>%
    mutate(year = factor(year(datetime)), 
    doy = yday(datetime))%>%
  filter(doy >= 60 & doy <= 325)
```

### Sensor Data
```{r load sensor data}
sensor_data <- load_sensor_data() %>%
  filter(site %in% c("pbd", "sfm", "pman", "pbr", "chd") & parameter == "Specific Conductivity") %>%
  mutate(date = as_date(DT_round),
         site = if_else(site == "pman", "pnf", site))%>%
  summarise(sensor_sc = mean(mean_cleaned, na.rm = T),
            n_measure = sum(!is.na(mean_cleaned)),
            .by = c( "date", "site"))%>%
  filter(n_measure >= 48) #need > 1/2 of a day's data to use

```

# Data Exploration

## Greeley Plant Chemical Data

### Alkalinity over time

```{r alkalinity over time}
# Plot alkalinity over time.
ggplot(plant_data, aes(x = doy, y = alkalinity)) +
  geom_line(na.rm = T) +
  facet_wrap(~year) +
  labs(
    x = "Day of Year",
    y = "Alkalinity (mg/L as CaCO3)",
    caption = "Figure 1: Alkalinity measurements at the Greeley, CO chemical plant intake between 2001 - 2025."
  ) +
  theme(plot.caption = element_text(hjust = 0.5))

# Plot alkalinity, conductivity, and total dissolved solids over time.
plant_data_long %>%
  filter(variable %in% c("alkalinity", "conductivity", "TDS")) %>%
  ggplot() +
  geom_line(aes(x = doy, y = value, colour = variable), na.rm = T) +
  facet_wrap(~year) +
  labs(
    x = "Day of Year",
    y = "Value",
    caption = "Figure 2: Alkalinity, Conductivity, and Total Dissolved Solids\n measurements at the Greeley, CO chemical plant intake between 2001 - 2025."
  ) +
  theme(plot.caption = element_text(hjust = 0.5))

# Plot seasonality of alkalinity, conductivity, and total dissolved solids.
plant_data_long %>%
  filter(variable %in% c("alkalinity", "conductivity", "TDS")) %>%
  ggplot(aes(x = doy, y = value, colour = year)) +
  geom_point(na.rm = T) +
  geom_smooth(method = "loess", se = FALSE, na.rm = T, color = "red") +
  facet_wrap(~variable)

```

### Alkalinity Relationship

```{r alkalinity relationship}
# Plot alkalinity vs. other chemical parameters.
plant_data %>%
  pivot_longer(
    cols = -c(date, year, doy, time, alkalinity),
    names_to = "variable",
    values_to = "value",
    values_drop_na = T
  ) %>%
  ggplot(aes(x = alkalinity, y = value)) +
  geom_point() +
  facet_wrap(~variable, scales = "free_y") +
  labs(
    x = "Alkalinity (mg/L as CaCO3)",
    y = "Value",
    caption = "Figure 3: Relationship of Alkalinity versus all other recorded water chemistry parameters at the Greeley, CO chemical plant intake (2001 - 2025"
  ) +
  theme(plot.caption = element_text(hjust = 0.5))
```

## Flow Data

### North Fork Conductivity
```{r north fork conductivity}
# Sitewise alkalinity distribution by
ggplot(grab_data, aes(x = Alkalinity, fill = site_code))+
  geom_histogram()+
  geom_vline(xintercept = 60, linetype = "dashed", color = "red")+
  facet_wrap(~ws, scale = "free_x")+
  labs(title = "Alkalinity Distribution by Site",
       x = "Alkalinity (mg/L as CaCO3)", 
       y = "Count", 
       fill = "Site")

alk_data_nf_flows <- plant_data %>%
  left_join(flow_data %>%
              select(datetime, 
                     canyon_mouth_cfs,
                     nf_cfs,
                     nf_prop
                     ),
              by = c("date" = "datetime")) %>%
  filter(nf_prop >= 0
         & !is.na(nf_prop)
         & !is.na(alkalinity) 
         & !is.na(conductivity)
         & !is.na(TDS))

ggplot(alk_data_nf_flows, aes(x = nf_prop,
                              y = alkalinity, 
                              color = factor(year))) +
  geom_point(size = 3) +
  labs(title = "North Fork ~ Cache la Poudre Flow Ratio vs Alkalinity at Greeley Plant Intake",
       x = "nf_cfs / canyon_mouth_cfs ",
       y = "Alkalinity (mg/L as CaCO3)", 
       color = "Year")
```

# Linear Regression

## Alkalinity, Conductivity, and Total Dissolved Solids
```{r}
# Linear model of alkalinity vs. conductivity.
alk_cond_lm <- lm(alkalinity ~ conductivity, data = alk_data_nf_flows)
summary(alk_cond_lm)

# Linear model of alkalinity vs. total dissolved solids.
alk_tds_lm <- lm(alkalinity ~ TDS, data = alk_data_nf_flows)
summary(alk_tds_lm)

# Linear model of alkalinity vs. conductivity and total dissolved solids.
alk_cond_tds_lm <- lm(alkalinity ~ conductivity + TDS, data = alk_data_nf_flows)
summary(alk_cond_tds_lm)

# Linear model of conductivity + flow ratio vs. conductivity
alk_flow_lm <- lm(alkalinity ~ nf_prop + conductivity + TDS, data = alk_data_nf_flows)
summary(alk_flow_lm)

# Linear model of alkalinity vs. conductivity, total dissolved solids, and day of year.
# Day of year accounts for seasonal variation in alkalinity.
alk_flow_doy_lm <- lm(alkalinity ~ nf_prop + conductivity + TDS + doy, data = alk_data_nf_flows)
summary(alk_flow_doy_lm)

alk_preds <- alk_data_nf_flows %>%
  mutate(
    alk_cond_pred = predict(alk_cond_lm, newdata = .),
    alk_cond_residuals = alkalinity - alk_cond_pred,
    alk_tds_pred = predict(alk_tds_lm, newdata = .),
    alk_tds_residuals = alkalinity - alk_tds_pred,
    alk_cond_tds_pred = predict(alk_cond_tds_lm, newdata = .),
    alk_cond_tds_residuals = alkalinity - alk_cond_tds_pred,
    alk_flow_pred = predict(alk_flow_lm, newdata = .),
    alk_flow_residuals = alkalinity - alk_flow_pred,
    alk_flow_doy_pred = predict(alk_flow_doy_lm, newdata = .,),
    alk_flow_doy_residuals = alkalinity - alk_flow_doy_pred
  ) %>%
  filter(!is.na(alk_cond_pred) & !is.na(alk_tds_pred) & !is.na(alk_cond_tds_pred) & !is.na(alk_flow_pred) & !is.na(alk_flow_doy_pred))
```

## Predicted vs. Actual Alkalinity

```{r}
alk_model_data <- alk_preds %>%
  select(year, alkalinity, alk_cond_pred, alk_tds_pred, alk_cond_tds_pred, alk_flow_pred, alk_flow_doy_pred) %>%
  pivot_longer(cols = -c(year, alkalinity), names_to = "model", values_to = "predicted_alkalinity") %>%
  mutate(model = case_when(
    model == "alk_cond_pred" ~ "Alkalinity ~ Conductivity",
    model == "alk_tds_pred" ~ "Alkalinity ~ TDS",
    model == "alk_cond_tds_pred" ~ "Alkalinity ~ Conductivity + TDS",
    model == "alk_flow_pred" ~ "Alkalinity ~ Flow Ratio + Conductivity + TDS",
    model == "alk_flow_doy_pred" ~ "Alkalinity ~ Flow Ratio + Conductivity + TDS + Day of Year"
  )) %>%
  filter(!is.na(predicted_alkalinity))

alk_model_data %>%
  group_split(model) %>%
  map(~ ggplot(., aes(
    x = alkalinity,
    y = predicted_alkalinity,
    color = factor(year)
  )) +
    geom_point() +
    geom_abline(
      slope = 1,
      intercept = 0,
      linetype = "dashed"
    ) +
    geom_hline(yintercept = 60, linetype = "dashed", color = "red") +
    geom_vline(xintercept = 60, linetype = "dashed", color = "red") +
    labs(
      x = "Actual Alkalinity (mg/L as CaCO3)",
      y = "Predicted Alkalinity (mg/L as CaCO3)",
      color = "Year",
      title = unique(.$model)
    ))

ggplot(
  alk_model_data,
  aes(
    x = alkalinity,
    y = predicted_alkalinity,
    color = model
  )
) +
  geom_point() +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  labs(
    x = "Actual Alkalinity (mg/L as CaCO3)",
    y = "Predicted Alkalinity (mg/L as CaCO3)",
    color = "Model",
    title = "Predicted vs. Actual Alkalinity for Different Linear Models"
  )
```

## Model Comparison
```{r}
alk_cond_rmse = Metrics::rmse(alk_preds$alkalinity, alk_preds$alk_cond_pred)
alk_cond_bias = bias(alk_preds$alkalinity, alk_preds$alk_cond_pred)
alk_conde_mse = mse(alk_preds$alkalinity, alk_preds$alk_cond_pred)

alk_tds_rmse = Metrics::rmse(alk_preds$alkalinity, alk_preds$alk_tds_pred)
alk_tds_bias = bias(alk_preds$alkalinity, alk_preds$alk_tds_pred)
alk_tds_mse = mse(alk_preds$alkalinity, alk_preds$alk_tds_pred)

alk_cond_tds_rmse = Metrics::rmse(alk_preds$alkalinity, alk_preds$alk_cond_tds_pred)
alk_cond_tds_bias = bias(alk_preds$alkalinity, alk_preds$alk_cond_tds_pred)
alk_cond_tds_mse = mse(alk_preds$alkalinity, alk_preds$alk_cond_tds_pred)

alk_flow_rmse = Metrics::rmse(alk_preds$alkalinity, alk_preds$alk_flow_pred)
alk_flow_bias = bias(alk_preds$alkalinity, alk_preds$alk_flow_pred)
alk_flow_mse = mse(alk_preds$alkalinity, alk_preds$alk_flow_pred)

alk_flow_doy_rmse = Metrics::rmse(alk_preds$alkalinity, alk_preds$alk_flow_doy_pred)
alk_flow_doy_bias = bias(alk_preds$alkalinity, alk_preds$alk_flow_doy_pred)
alk_flow_doy_mse = mse(alk_preds$alkalinity, alk_preds$alk_flow_doy_pred)

tibble(
  Model = c("Conudctivity", "Total Dissolved Solids", "Conductivity + Total Dissolved Solids", "Conductivity + Total Dissolved Solids + Flow Ratio", "Conductivity + Total Dissolved Solids + Flow Ratio + Day of Year"),
  RMSE = c(alk_cond_rmse, alk_tds_rmse, alk_cond_tds_rmse, alk_flow_rmse, alk_flow_doy_rmse),
  Bias = c(alk_cond_bias, alk_tds_bias, alk_cond_tds_bias, alk_flow_bias, alk_flow_doy_bias),
  MSE = c(alk_conde_mse, alk_tds_mse, alk_cond_tds_mse, alk_flow_mse, alk_flow_doy_mse)
)%>%
  knitr::kable()
```

## Sensor Data Comparison
```{r}
sensor_plant_comp <- sensor_data %>%
  filter(site == "pbd") %>%
  left_join(plant_data, by = c("date" = "date"))

cor_plot <- sensor_plant_comp %>%
  select(sensor_sc, conductivity, alkalinity, TDS) %>%
  drop_na() %>%
  cor()

ggplot(sensor_plant_comp, aes(x = sensor_sc, y = conductivity))+
  geom_point() +
  coord_cartesian(xlim = c(30, 150), ylim = c(30,150))+
  labs(title = "Sensor vs Greeley Plant Intake Conductivity Measurements",
       x = "Sonde Conductivity (µS/cm)",
       y = "Plant Conductivity (µS/cm)")

ggplot(sensor_plant_comp, aes(x = sensor_sc, y = alkalinity)) +
  geom_point() +
  geom_smooth(method = "gam", se = FALSE, color = "red") +
  labs(title = "Sensor Conductivity vs. Greeley Plant Intake Alkalinity",
       x = "Sensor Conductivity (µS/cm)",
       y = "Plant Alkalinity (mg/L as CaCO3)",)
```

## Sensor Model
```{r}
sensor_sc_lm <- lm(alkalinity ~ sensor_sc, data = sensor_plant_comp)
summary(sensor_sc_lm)

sensor_sc_gam <- lm(model = "GAM", alkalinity ~ sensor_sc, data = sensor_plant_comp)
summary(sensor_sc_gam)

sensor_preds <- sensor_plant_comp %>%
  mutate(gam_pred = predict(sensor_sc_gam, newdata = .), 
         gam_res = alkalinity - gam_pred,
         lm_pred = predict(sensor_sc_lm, newdata = .),
         lm_res = alkalinity - lm_pred) %>%
  filter(!is.na(gam_pred) & !is.na(lm_pred)) %>% 
  select(date, alkalinity, lm_pred, gam_pred) %>% 
  pivot_longer(cols = -c(date, alkalinity), names_to = "model", values_to = "alk_pred")

ggplot(sensor_preds, aes(x = date)) +
  geom_line(aes(y = alk_pred, color = model)) +
  geom_line(aes(y = alkalinity, color = "Intake Measurement")) +
  labs(title = "Predicted Alkalinity at Greeley Plant Intake from Sensor Data",
       x = "Date",
       y = "Alkalinity (mg/L as CaCO3)", 
       color = "Legend")
```

# Model Training

```{r}
# split data into training and testing sets.
train_set <- plant_data %>% select(alkalinity, conductivity, TDS, date, year, doy) %>%
  subset(., year < 2020)

test_set <- plant_data %>% select(alkalinity, conductivity, TDS, date, year, doy) %>%
  subset(., year >= 2020)

ggplot() +
  geom_line(data = train_set, aes(x = date, y = alkalinity, color = "Training"), size = 1) +
  geom_line(data = test_set, aes(x = date, y = alkalinity, color = "Testing"), size = 1) +
  labs(
    title = "Alkalinity - Training and Testing Sets",
    x = "Date",
    y = "Alkalinity (mg/L as CaCO3)"
  ) +
  scale_color_manual(values = c("Training" = "#12355B", "Testing" = "#D72638"), name = "Dataset")

```
