---
title: "greeley_baseline_alkalinity"
author: "Danny Duncan"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%")

library(tidyverse)
library(jsonlite)
library(forecast)
library(flowfabricr)
library(Metrics)
library(arrow)
library(here)
library(httr)
library(mgcv)

source("src/setup_libraries.R")
```

# Overview

# Data Preparation

## Load Datasets

### Greeley Chemical Plant Intake (2001 - 2025)

```{r load greeley data}
# Load path to collated data folder.
collate_path <- here("data/collated/chem")

# FInd greeley parquet file path in collated data folder.
file_path <- list.files(collate_path,
  pattern = "^greeley_intake_chem_data_.*?\\.parquet$",
  full.names = TRUE
)

# Read in the parquet file.
plant_data <- read_parquet(file_path) %>%
  filter(!is.na(alkalinity) | !is.na(conductivity) | !is.na(TDS)) %>%
  mutate(
    year = year(date),
    doy = yday(date)
  )

```

### Flow Data
```{r load flow data}
# Load in flow data for the canyon mouth (CLAFTCCO)
canyon_flow_data <- get_sw_ts(
  abbrev = "CLAFTCCO",
  start_date = min(plant_data$date, na.rm = TRUE),
  end_date = max(plant_data$date, na.rm = TRUE),
  timescale = "day"
) %>%
  rename(ftdc2_flow = value) %>%
  mutate(date = as.Date(datetime), ftdc2_flow = as.numeric(ftdc2_flow)) %>%
  select(date, ftdc2_flow)

# Load in flow data for the North Fork (CLANSECO)
nf_flow_data <- get_sw_ts(
  abbrev = "CLANSECO",
  start_date = min(plant_data$date, na.rm = TRUE),
  end_date = max(plant_data$date, na.rm = TRUE),
  timescale = "day"
) %>%
  rename(nsec2_flow = value) %>%
  mutate(date = as.Date(datetime), nsec2_flow = as.numeric(nsec2_flow)) %>%
  select(date, nsec2_flow)

# Calculate flow ratio of North Fork to canyon mouth
flow_data <- canyon_flow_data %>%
  left_join(nf_flow_data, by = "date") %>%
  mutate(flow_ratio = ifelse(ftdc2_flow > 0, nsec2_flow / ftdc2_flow, NA_real_)) %>%
  filter(is.finite(flow_ratio), is.finite(nsec2_flow), is.finite(ftdc2_flow))
```

### Join Plant & Flow Data
```{r}
hist_data <- plant_data %>%
  left_join(flow_data, by = "date") %>%
  select(date, alkalinity, conductivity, nsec2_flow, ftdc2_flow, flow_ratio) %>%
  filter(!is.na(alkalinity), is.finite(flow_ratio), is.finite(nsec2_flow), is.finite(ftdc2_flow))
```

# Model Exploration

## GAM
### Training Data Preparation
```{r}
gam_train <- hist_data %>%
  mutate(date = as.Date(date)) %>%
  filter(date < as.Date("2020-01-01")) %>%
  arrange(date) %>%
  # day-of-year for seasonality
  mutate(doy = yday(date),
         total_flow = pmax(1e-6, nsec2_flow + ftdc2_flow),
         # log transformed flow magnitude predictors
         log_total_flow = log1p(total_flow),
         log_nsec2 = log1p(pmax(0, nsec2_flow)),
         log_ftdc2 = log1p(pmax(0, ftdc2_flow)),
         # lagged flow-ratio predictors
         lag1  = lag(flow_ratio, 1),
         lag2  = lag(flow_ratio, 2),
         lag3  = lag(flow_ratio, 3),
         lag7  = lag(flow_ratio, 7),
         lag14 = lag(flow_ratio, 14),
         lag30 = lag(flow_ratio, 30),
         y_log = log1p(alkalinity)
    ) %>%
  filter(if_all(c("alkalinity","y_log","flow_ratio","lag1","lag2","lag3","lag7","lag14","lag30","log_total_flow","log_nsec2","log_ftdc2","doy"), 
      is.finite))

```

### Generalized Additive Model (GAM) Exploration
#### GAM Model Fitting 1
```{r}
gam_model <- gam(y_log ~ 
                   s(flow_ratio, k = 10) +
                   s(lag1, k = 10) + s(lag2, k = 10) + s(lag3, k = 10) +
                   s(lag7, k = 10) + s(lag14, k = 10) + s(lag30, k = 10) +
                   s(log_total_flow, k = 10) +
                   s(log_nsec2, k = 10) +
                   s(log_ftdc2, k = 10) +
                   s(doy, bs = "cc", k = 20),
                 method = "REML",
                 data = gam_train
                 )

gam.check(gam_model, 0)
```

Model is a good start, but some major K flags for flow_ratio, lag1, lag2, lag3, lag30, and log_nsec2. Minor K flags for lag7 and log_total_flow. Let's try increasing the K values for those predictors to see if we can capture more complex relationships without over fitting.

#### GAM Model Fitting 2
```{r}
gam_model2 <- gam(
  y_log ~
    s(flow_ratio, k = 20) +
    s(lag1, k = 15) + s(lag2, k = 15) + s(lag3, k = 20) +
    s(lag7, k = 20) + s(lag14, k = 15) + s(lag30, k = 20) +
    s(log_total_flow, k = 20) +
    s(log_nsec2, k = 20) +
    s(log_ftdc2, k = 12) +
    s(doy, bs = "cc", k = 20),
  method = "REML",
  data = gam_train
)
gam.check(gam_model2, 0)
```

#### GAM Model Fitting 3
```{r}
gam_model3 <- gam(
  y_log ~
    s(flow_ratio, k = 30) +
    s(lag1, k = 15) + s(lag2, k = 20) + s(lag3, k = 25) +
    s(lag7, k = 30) + s(lag14, k = 20) + s(lag30, k = 30) +
    s(log_total_flow, k = 25) +
    s(log_nsec2, k = 30) +
    s(log_ftdc2, k = 20) +
    s(doy, bs = "cc", k = 20),
  method = "REML",
  data = gam_train
)
gam.check(gam_model3, 0)
summary(gam_model3)
```

#### GAM Model Fitting 4 (simplified)

Many of the lagged flow ratio predictors are showing similar patterns and concurvity issues. Let's only keep lag1, lag7, and lag30 to see if we can reduce concurvity.
```{r}
gam_simple <- gam(
  y_log ~
    s(flow_ratio, k = 25) +
    s(log_total_flow, k = 20) +
    s(lag1, k = 12) +
    s(lag7, k = 20) +
    s(lag30, k = 20) +
    s(doy, bs = "cc", k = 20),
  method = "REML",
  data = gam_train
  )

gam.check(gam_simple, 0)
summary(gam_simple)
```

#### GAM Model Fitting 5 (no lagged predictors)
```{r}
gam_no_lag <- gam(
  y_log ~
    s(flow_ratio, k = 25) +
    s(log_total_flow, k = 20) +
    s(doy, bs = "cc", k = 20),
  data = gam_train,
  method = "REML"
  )

gam.check(gam_no_lag, 0)
```

### GAM Predictions
```{r}
gam_test <- hist_data %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= as.Date("2020-01-01")) %>%
  arrange(date) %>%
  mutate(
    doy = yday(date),
    total_flow = pmax(1e-6, nsec2_flow + ftdc2_flow),
    log_total_flow = log1p(total_flow),
    log_nsec2 = log1p(pmax(0, nsec2_flow)),
    log_ftdc2 = log1p(pmax(0, ftdc2_flow)),
    lag1  = lag(flow_ratio, 1),
    lag2  = lag(flow_ratio, 2),
    lag3  = lag(flow_ratio, 3),
    lag7  = lag(flow_ratio, 7),
    lag14 = lag(flow_ratio, 14),
    lag30 = lag(flow_ratio, 30),
    y_log = log1p(alkalinity)
  ) %>% 
    filter(if_all(c("alkalinity","y_log","flow_ratio","lag1","lag2","lag3",
                    "lag7","lag14","lag30","log_total_flow","log_nsec2","log_ftdc2","doy"), 
      is.finite))

gam_test <- gam_test %>%
  mutate(
    gam_mod_pred = exp(predict(gam_model, newdata = gam_test)) - 1,
    gam_mod2_pred = exp(predict(gam_model2, newdata = gam_test)) - 1,
    gam_mod3_pred = exp(predict(gam_model3, newdata = gam_test)) - 1,
    gam_simple_pred = exp(predict(gam_simple, newdata = gam_test)) - 1,
    gam_no_lag_pred = exp(predict(gam_no_lag, newdata = gam_test)) - 1,
    gam_mod_resid = alkalinity - gam_mod_pred,
    gam_mod2_resid = alkalinity - gam_mod2_pred,
    gam_mod3_resid = alkalinity - gam_mod3_pred,
    gam_simple_resid = alkalinity - gam_simple_pred,
    gam_no_lag_resid = alkalinity - gam_no_lag_pred
  )

gam_preds_plot <- gam_test %>%
  select(
    date, alkalinity,
    gam_mod_pred, gam_mod2_pred, gam_mod3_pred,
    gam_simple_pred, gam_no_lag_pred
  ) %>%
  pivot_longer(
    cols = ends_with("_pred"),
    names_to = "model",
    values_to = "predicted"
  ) %>%
  mutate(
    model = recode(
      model,
      gam_mod_pred = "gam_model",
      gam_mod2_pred = "gam_model2",
      gam_mod3_pred = "gam_model3",
      gam_simple_pred = "gam_simple",
      gam_no_lag_pred = "gam_no_lag"
    )
  )
```

```{r}
# Observed vs Predicted scatterplot
ggplot(gam_preds_plot, aes(x = alkalinity, y = predicted)) +
  geom_point(alpha = 0.45, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  coord_equal() +
  facet_wrap(~ model, scales = "fixed") +
  labs(
    title = "Observed vs Predicted Alkalinity by GAM Model",
    x = "Observed Alkalinity",
    y = "Predicted Alkalinity"
  ) +
  theme_minimal()

```

```{r}
# Time-series overlay of observed and predicted
plot_ts <- gam_preds_plot %>%
  transmute(date, model, observed = alkalinity, predicted) %>%
  pivot_longer(
    cols = c(observed, predicted),
    names_to = "series",
    values_to = "value"
  )

ggplot(plot_ts, aes(x = date, y = value, color = series)) +
  geom_line(linewidth = 0.6, alpha = 0.9) +
  facet_wrap(~ model, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c(observed = "black", predicted = "red")) +
  labs(
    title = "Test Set: Observed vs Predicted Alkalinity",
    x = "Date",
    y = "Alkalinity",
    color = ""
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```

```{r}
# Residuals vs Fitted plot
ggplot(gam_test, aes(x = gam_no_lag_pred, y = gam_no_lag_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residuals vs Fitted Values for GAM Model (No Lag)",
    x = "Fitted Values (Predicted Alkalinity)",
    y = "Residuals (Observed - Predicted)"
  ) +
  theme_minimal()

ggplot(gam_test, aes(x = gam_simple_pred, y = gam_simple_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residuals vs Fitted Values for GAM Model (Simple)",
    x = "Fitted Values (Predicted Alkalinity)",
    y = "Residuals (Observed - Predicted)"
  ) +
  theme_minimal()

ggplot(gam_test, aes(x = gam_mod_pred, y = gam_mod_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Residuals vs Fitted Values for GAM Model 1",
    x = "Fitted Values (Predicted Alkalinity)",
    y = "Residuals (Observed - Predicted)"
  ) +
  theme_minimal()
```

```{r, include=FALSE}
gam_metrics <- gam_test %>%
  pivot_longer(
    cols = ends_with("_pred"),
    names_to = "pred_col",
    values_to = "pred"
  ) %>%
  mutate(
    model = str_remove(pred_col, "_pred$")
  ) %>%
  # Compute metrics
  group_by(model) %>%
  summarise(
    MAE  = mean(abs(alkalinity - pred), na.rm = TRUE),
    RMSE = sqrt(mean((alkalinity - pred)^2, na.rm = TRUE)),
    Bias = mean(pred - alkalinity, na.rm = TRUE),
    R2   = cor(alkalinity, pred, use = "complete.obs")^2,
    MAPE = mean(abs((alkalinity - pred) / pmax(alkalinity, 1e-6)), na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  # Ranking
  mutate(
    absBias      = abs(Bias),
    rank_MAE     = rank(MAE, ties.method = "min"),
    rank_RMSE    = rank(RMSE, ties.method = "min"),
    rank_MAPE    = rank(MAPE, ties.method = "min"),
    rank_absBias = rank(absBias, ties.method = "min"),
    rank_R2      = rank(-R2, ties.method = "min"),
    rank_total   = rank_MAE + rank_RMSE + rank_MAPE + rank_absBias + rank_R2
  ) %>%
  arrange(rank_total, RMSE) %>%
  select(model, MAE, RMSE, MAPE, Bias, R2,
         rank_MAE, rank_RMSE, rank_MAPE, rank_absBias, rank_R2, rank_total)

# Rounded print
gam_metrics %>%
  mutate(
    across(c(MAE, RMSE, MAPE, Bias), ~ round(.x, 3)),
    R2 = round(R2, 4)
  ) %>%
  print(n = Inf)
```

## Sensitivity Testing
```{r}
sens_test <- hist_data %>%
  mutate(date = as.Date(date)) %>%
  filter(date >= as.Date("2020-01-01")) %>%
  arrange(date) %>%
  mutate(
    doy = yday(date),
    ftdc2_flow = ftdc2_flow * 1.25,  # increase canyon mouth flow by 25%
    total_flow = pmax(1e-6, nsec2_flow + ftdc2_flow),
    log_total_flow = log1p(total_flow),
    log_nsec2 = log1p(pmax(0, nsec2_flow)),
    log_ftdc2 = log1p(pmax(0, ftdc2_flow)),
    lag1  = lag(flow_ratio, 1),
    lag2  = lag(flow_ratio, 2),
    lag3  = lag(flow_ratio, 3),
    lag7  = lag(flow_ratio, 7),
    lag14 = lag(flow_ratio, 14),
    lag30 = lag(flow_ratio, 30),
    y_log = log1p(alkalinity)
  ) %>% 
    filter(if_all(c("alkalinity","y_log","flow_ratio","lag1","lag2","lag3",
                    "lag7","lag14","lag30","log_total_flow","log_nsec2","log_ftdc2","doy"), 
      is.finite))

sens_preds <- sens_test %>%
  mutate(
    gam_mod_pred = exp(predict(gam_model, newdata = sens_test)) - 1,
    gam_mod2_pred = exp(predict(gam_model2, newdata = sens_test)) - 1,
    gam_mod3_pred = exp(predict(gam_model3, newdata = sens_test)) - 1,
    gam_simple_pred = exp(predict(gam_simple, newdata = sens_test)) - 1,
    gam_no_lag_pred = exp(predict(gam_no_lag, newdata = sens_test)) - 1,
    gam_mod_resid = alkalinity - gam_mod_pred,
    gam_mod2_resid = alkalinity - gam_mod2_pred,
    gam_mod3_resid = alkalinity - gam_mod3_pred,
    gam_simple_resid = alkalinity - gam_simple_pred,
    gam_no_lag_resid = alkalinity - gam_no_lag_pred
  )

gam_sens_plot <- sens_preds %>%
  select(
    date, alkalinity,
    gam_mod_pred, gam_mod2_pred, gam_mod3_pred,
    gam_simple_pred, gam_no_lag_pred
  ) %>%
  pivot_longer(
    cols = ends_with("_pred"),
    names_to = "model",
    values_to = "predicted"
  ) %>%
  mutate(
    model = recode(
      model,
      gam_mod_pred = "gam_model",
      gam_mod2_pred = "gam_model2",
      gam_mod3_pred = "gam_model3",
      gam_simple_pred = "gam_simple",
      gam_no_lag_pred = "gam_no_lag"
    )
  )
  
```

### Time-series overlay of observed and predicted for sensitivity test
```{r}
# Time-series overlay of observed and predicted
plot_sens <- gam_sens_plot %>%
  transmute(date, model, observed = alkalinity, predicted) %>%
  pivot_longer(
    cols = c(observed, predicted),
    names_to = "series",
    values_to = "value"
  )

ggplot(plot_sens, aes(x = date, y = value, color = series)) +
  geom_line(linewidth = 0.6, alpha = 0.9) +
  facet_wrap(~ model, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c(observed = "black", predicted = "red")) +
  labs(
    title = "Sensitivity Test:: Observed vs Predicted Alkalinity",
    x = "Date",
    y = "Alkalinity",
    color = ""
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r, include=false}
# Add MAPE + multi-metric ranking
sens_metrics <- sens_preds %>%
  pivot_longer(
    cols = ends_with("_pred"),
    names_to = "pred_col",
    values_to = "pred"
  ) %>%
  mutate(
    model = str_remove(pred_col, "_pred$")
  ) %>%
  # Compute metrics
  group_by(model) %>%
  summarise(
    MAE  = mean(abs(alkalinity - pred), na.rm = TRUE),
    RMSE = sqrt(mean((alkalinity - pred)^2, na.rm = TRUE)),
    Bias = mean(pred - alkalinity, na.rm = TRUE),
    R2   = cor(alkalinity, pred, use = "complete.obs")^2,
    MAPE = mean(abs((alkalinity - pred) / pmax(alkalinity, 1e-6)), na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  # Ranking
  mutate(
    absBias      = abs(Bias),
    rank_MAE     = rank(MAE, ties.method = "min"),
    rank_RMSE    = rank(RMSE, ties.method = "min"),
    rank_MAPE    = rank(MAPE, ties.method = "min"),
    rank_absBias = rank(absBias, ties.method = "min"),
    rank_R2      = rank(-R2, ties.method = "min"),
    rank_total   = rank_MAE + rank_RMSE + rank_MAPE + rank_absBias + rank_R2
  ) %>%
  arrange(rank_total, RMSE) %>%
  select(model, MAE, RMSE, MAPE, Bias, R2,
         rank_MAE, rank_RMSE, rank_MAPE, rank_absBias, rank_R2, rank_total)

# Rounded print
sens_metrics %>%
  mutate(
    across(c(MAE, RMSE, MAPE, Bias), ~ round(.x, 3)),
    R2 = round(R2, 4)
  ) %>%
  print(n = Inf)
```

```{r}
# Residuals vs Fitted plot
ggplot(sens_preds, aes(x = gam_no_lag_pred, y = gam_no_lag_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Sensitivity Test: Residuals vs Fitted Values for GAM Model (No Lag)",
    x = "Fitted Values (Predicted Alkalinity)",
    y = "Residuals (Observed - Predicted)"
  ) +
  theme_minimal()

ggplot(sens_preds, aes(x = gam_simple_pred, y = gam_simple_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Sensitivity Test: Residuals vs Fitted Values for GAM Model (Simple)",
    x = "Fitted Values (Predicted Alkalinity)",
    y = "Residuals (Observed - Predicted)"
  ) +
  theme_minimal()

ggplot(sens_preds, aes(x = gam_mod_pred, y = gam_mod_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Sensitivity Test: Residuals vs Fitted Values for GAM Model 1",
    x = "Fitted Values (Predicted Alkalinity)",
    y = "Residuals (Observed - Predicted)"
  ) +
  theme_minimal()
```


