---
title: "Real Time Chl-a Modeling Cache la Poudre"
author: "Sam Struthers - CSU ROSS"
date: "`r Sys.Date()`"
output: html_document
---
# Goal of this Script

We will import our collected sensor datasets, calculate any lagged parameters we want to use as features in our model, and then create a matchup dataset between grab sample chl-a data and sensor data. We will then bin the data into chl-a groups for classification modeling since there isn't much variability for us to model and users may only need to see if chl-a is high or low rather than some particular value since it is used as a general indicator of productivity and potential for algae blooms.
Similar to TOC, we will be testing on 2025 data collected at: CHD, PBR, SFM, PMAN and PBD. 
We will exclude data from springcreek, boxcreek and riverbluffs since those sites have very different hydrology and land use characteristics that may not be well represented by the other sites. We may also consider removing specific conductivity from the model since our lower sites have much higher SC characteristics that may translate to good performance in the Upper CLP

# Setup

```{r}
source("src/setup_libraries.R")
library(glue)
library(shapviz)
source("src/setup_ross_theme.R")
version = "20250112"
# Load functions
#function to load sensor data
source("src/collation/load_sensor_data.R")
#function to load grab sample data
source("src/collation/load_grab_sample_data.R")
#function to apply FDOM temperature correction
source("src/sensor_qaqc/apply_fdom_temp_corr.R")
#function to calculate lagged parameters on sensor data
source( "src/modeling/calc_lagged_parameter.R")
#function to create matchup dataset
source("src/modeling/create_grab_sensor_matchup.R")
#function to normalize training data (sensor) to entire sensor range of values
source("src/modeling/apply_training_scale.R")
# function to help select features for categorical models
source("src/modeling/xgboost_feature_selection_cat.R")
# function to tune xgboost hyperparameters
source("src/modeling/xgboost_hyperparameter_tuning_cat.R")
#function to plot performance of each fold
source("src/plotting/plot_tvt_fold_acc.R")
#function to plot performance of each fold
source("src/plotting/plot_train_test_ensemble_acc.R")
#function to plot model timeseries
source("src/plotting/plot_cat_model_timeseries.R")
# Load CDWR credentials to get flow data (without hitting data limits)
cdwr_creds <- read_yaml(here("creds", "CDWRCreds.yml"))$api_key


#set warnings to F
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.dim = c(12,10))
```

# Load Sensor and Grab Sample Data

## Sensor

```{r}

sensor_data <- load_sensor_data()%>% 
  mutate(DT_hourly_round = round_date(DT_round, "hour")) %>%
  distinct(site, DT_hourly_round, parameter, .keep_all = T) 

#In 2023 & 2024, we colocated some sensors from Virridy with our sites. The Virridy sondes did not have pH or DO so we will "borrow"
#those parameters from our sondes for those time periods to fill in gaps. 
virridy_subs <- sensor_data%>%
  filter(c(between(DT_hourly_round, ymd_hms("2023-09-01 00:00:00", tz = "MST"), ymd_hms("2024-05-31 23:59:59", tz = "MST")) & str_detect(site, "riverbend|cottonwood")) | c(between(DT_hourly_round, ymd_hms("2023-09-01 00:00:00", tz = "MST"), ymd_hms("2024-12-31 23:59:59", tz = "MST")) & str_detect(site, "archery")))%>%
  filter(parameter %in% c("pH", "DO"))%>%
  mutate(site = paste0(site, "_virridy"))

sensor_data <- sensor_data%>%
  bind_rows(virridy_subs)%>%
  pivot_wider(id_cols = c("site","DT_hourly_round"), names_from = "parameter", values_from = "hourly_median")%>%
  #remove rows where everything other than site and DT_round are NA
  filter(!if_all(-c(site, DT_hourly_round), is.na))%>%
  #Remove a few sites that won't have all the parameters for modeling (pH, DO)
  filter(site %nin% c("penn", "lbea", "boxcreek", "springcreek", "mtncampus", "riverbluffs"))


```

## Water Chem

```{r}
water_chem <- load_grab_sample_data()%>%
  filter(!is.na(ChlA) & site_code %in% unique(sensor_data$site) & !is.na(DT_sample))%>%
  #Filter for after sensors were deployed (2023-09-01)
  filter(DT_sample >= ymd_hms("2023-09-01 00:00:00", tz = "MST"))
  
#Create duplicates for PMAN & PBR so we can make match ups with sensor data from ROSS and FC (_fc) sensors
pbr_pman <- water_chem%>%
  filter(site_code %in% c("pbr", "pman")) %>%
  mutate(site_code = paste0(site_code, "_fc"))

water_chem <- water_chem%>%
  #remove pbr and pman samples prior to 2025-01-01 since ROSS sensors were not deployed yet
  filter(!(site_code %in% c("pbr", "pman") & DT_sample <= ymd_hms("2025-01-01 12:00:00")))%>%
  bind_rows(pbr_pman)%>%
  select(site = site_code, DT_sample, ChlA, collector)

```

# Calculate lagged variables on sensor data

Things that might be helpful:
- Recent (24-48 hour) temperatures
- Recent SC variation, i.e. change in depth or water source
- Diel or greater (48 hour) DO variation for productivity proxy
- Minimum recent DO values

```{r}

sensor_lagged <- sensor_data %>%
  mutate(year = year(DT_hourly_round)) %>%
  split(.$year) %>%
  map(., function(df){
    df %>%
      #applying FDOM temperature correcting/standardization
      apply_fdom_temp_corr( fdom_col = "FDOM Fluorescence",
                            temp_col = "Temperature",
                            fdom_corr_col = "FDOMc",
                            Tr = 25,
                            rho = -0.015)%>%
      # recent means (water temp)
      calc_lagged_parameter(site_col = "site", parameter_col = "Temperature",
                            dt_col = "DT_hourly_round", lag_fun = "mean",
                            time_period = "48 hour", new_value_col = "temp_mean_48hr")%>%
      calc_lagged_parameter(site_col = "site", parameter_col = "Temperature",
                            dt_col = "DT_hourly_round", lag_fun = "mean",
                            time_period = "24 hour", new_value_col = "temp_mean_24hr")%>%
      # recent maxes (temp)
      calc_lagged_parameter(site_col = "site", parameter_col = "Temperature",
                            dt_col = "DT_hourly_round", lag_fun = "max",
                            time_period = "24 hour", new_value_col = "temp_max_24hr")%>%
      # recent variance/sd (sc)
      calc_lagged_parameter(site_col = "site", parameter_col = "Specific Conductivity",
                            dt_col = "DT_hourly_round", lag_fun = "var",
                            time_period = "24 hour", new_value_col = "sc_var_24hr")%>%
      mutate(sc_sd_24hr = sqrt(sc_var_24hr))%>%
      #12 hour median turbidity (proxy for recent storm event)
      calc_lagged_parameter(site_col = "site", parameter_col = "Turbidity",
                            dt_col = "DT_hourly_round", lag_fun = "median",
                            time_period = "24 hour", new_value_col = "turb_median_12hr")%>%
      # recent DO data
      calc_lagged_parameter(site_col = "site", parameter_col = "DO",
                            dt_col = "DT_hourly_round", lag_fun = "var",
                            time_period = "48 hour", new_value_col = "do_var_48hr")%>%
      calc_lagged_parameter(site_col = "site", parameter_col = "DO",
                            dt_col = "DT_hourly_round", lag_fun = "min",
                            time_period = "48 hour", new_value_col = "do_min_24hr")
      
  })%>%
  bind_rows()

```

## Add in Flow Data

For now, we will see if the canyon mouth is a good enough "setting" for the CLP or if we need to find Q for each site to try to get Q/Temp relationships better represented.

```{r}
#load flow data using CDSSR

min_date <- min(sensor_lagged$DT_hourly_round, na.rm = T)
max_date <- max(sensor_lagged$DT_hourly_round, na.rm = T)

#TODO:Match up sites with the nearby flow gage
# Note: The sites tibble below is a placeholder. I may be able to figure out a way to get flow data from Contrail or Larimer County but that will need work, 
# may not cover all sites and may be less accurate than CDSSR data.
# For now, I will use the CDSSR data for CLAFTCCO and use the canyon mouth flow data for the CLAFTCCO site as a "general" setting
# sites <- tibble(abbrev = c("MICDCPCO", "JOEBELCO",  "JWCCHACO", "CLASRKCO",  "CLAFTCCO", "CLAFORCO", "CLABOXCO", "CLARIVCO"), 
#                 site = c("joei", "cbri","chd","sfm", "pbd", "udall","elc","riverbluffs"))

sites <- tibble(abbrev = c( "CLAFTCCO"), 
                site = c("pbd"))

all_flow_data <- map(sites$abbrev, function(abbrev){
  get_telemetry_ts(abbrev = abbrev, 
                 parameter = "DISCHRG", 
                 start_date = as.Date(min_date)- days(1),
                 end_date = as.Date(max_date)+ days(1),
                 timescale = "hour",api_key = cdwr_creds
                )%>%
  select(abbrev, DT_hourly_round = meas_date, flow_cfs = meas_value)%>%
  mutate(DT_hourly_round = with_tz(ymd_hms(DT_hourly_round, tz = "America/Denver"), tzone = "MST"))
}) %>%
  bind_rows()

canyon_mouth_flow_daily <- all_flow_data%>%
  filter(abbrev == "CLAFTCCO")%>%
  group_by(date = as_date(DT_hourly_round))%>%
  summarise(canyon_mouth_daily_flow_cfs = mean(flow_cfs, na.rm = T))%>%
  ungroup()


sensor_final <- sensor_lagged%>%
  mutate(date = as_date(DT_hourly_round))%>%
  left_join(canyon_mouth_flow_daily, by = "date")%>%
  select(-date, -year)%>%
  rename(DT_round = DT_hourly_round)

```

# Create Matchup

## Prep and select features 
```{r}

id_cols <- c("site", "DT_round", "DT_hourly_round")
label <- "ChlA"

features <- c("Specific Conductivity","Temperature" , "Turbidity", "canyon_mouth_daily_flow_cfs", "DO", "pH","temp_mean_48hr" , "temp_mean_24hr",
              "temp_max_24hr" , "sc_sd_24hr" , "turb_median_12hr", "do_var_48hr" , "do_min_24hr")


```

## Run matchup function 

We will use the `create_grab_sensor_matchup()` function to create the matchup dataset between grab sample chl-a data and sensor data. We will use a 3 hour acceptable time window for matching. For a "match" to be made, a row within 3 hours of the sample must have  all the parameters in `features` must have a non-NA value.

```{r}

if(file.exists(here("data", "derived", "matchup_datasets", paste0("parsed_chla_sensor_", version, ".parquet")))){
  matchup_results <- read_parquet(here("data", "derived", "matchup_datasets", paste0("parsed_chla_sensor_", version, ".parquet")))
} else {
matchup_results <- create_grab_sensor_matchup(
  water_chem_df = water_chem,
  sensor_df = sensor_final,
  features = features,
  acceptable_time_window_hours = 3
)
}

```

## Check Matches

Expected samples to miss:
- 2025 Data which has not been verified and may be overflagged
- 2024/2023 Data where data has been removed
- Virridy sites where we may not have all parameters
- Instances where lagged parameters cannot be calculated


```{r}

matchup_results%>%
  filter(site %in% unique(sensor_final$site))%>%
  mutate(year = year(sample_datetime))%>%
  summarise(n_missing = sum(is.na(`sensor_datetime`)),
            n_total = n(),
            perc_missing = n_missing/n_total*100, .by = year)

missing_samples <- matchup_results%>%
  mutate(year = year(sample_datetime))%>%
  group_by(year, site) %>%
  summarise(n_total = n(),
            n_missing = sum(is.na(sensor_datetime)),
            .groups = "drop")%>%
  arrange(desc(n_missing))%>%
  filter(n_missing > 0)

cat("Number of samples with no match up:\n", sum(missing_samples$n_missing), " of ", nrow(matchup_results) )
print(missing_samples, n = Inf)

### Save Matchup file

#write_parquet(matchup_results,here("data", "derived", "matchup_datasets", paste0("parsed_chla_sensor_", version, ".parquet")))

```

# Prepare Dataset for Modeling

## Clean up Matchup Dataset

We will also remove any NAs in sensor_datetime (ie no matchup occured) and set the categories for chl-a

```{r}

# Read in all matches data
all_matches <- matchup_results %>%
  #remove NAs
  filter(!is.na(sensor_datetime))%>%
  mutate(month = month(sensor_datetime),
         #create categorical chlA
         ChlA_cat  = cut(ChlA, breaks = c(0, 2, 4, 6, 20), labels = c('low', 'medium', 'high',"extreme")),
         ChlA_cat = as.numeric(as.factor(ChlA_cat))-1
  )

```

## Define ID Columns, Target, and Features

```{r}
# ID columns to keep to use later on 
id_cols <- c("Date", "site", "DT_round", "sensor_datetime", "sample_datetime", "data_avail")
# What we want to predict
target = "ChlA_cat"
# Features selected in previous steps
features <- c("Specific Conductivity","Temperature" , "Turbidity", "canyon_mouth_daily_flow_cfs", "DO", "pH","temp_mean_48hr" , "temp_mean_24hr",
              "temp_max_24hr" , "sc_sd_24hr" , "turb_median_12hr", "do_var_48hr" , "do_min_24hr")
```

## Create Scaling Parameters & Normalize Data

Normalization is applied based on min-max scaling from the complete sensor dataset using the `apply_training_scale()` function.

```{r}
# Apply scaling to dataset
all_matches_norm <- all_matches %>%
  # Only keep columns that would be in the complete sensor dataset
  select(any_of(id_cols), any_of(names(sensor_final)), any_of(target))%>%
  # Apply normalization 
  apply_training_scale(
    new_data = .,
    features = features,
    full_dataset = sensor_final
  ) %>%
  na.omit()

```

## Create Holdout Dataset

For this model version, we will create a holdout testing set from specific sites (chd, sfm, pbr, pman, pbd) in 2025, and use all other data for training/validation. We will also trim out any known bad data or sites we don't want to include in modeling.

```{r}
# Trim out any known bad data or sites we don't want to include in modeling
all_matches_trimmed <- all_matches_norm%>%
  filter(!(site == "cbri" & between(sensor_datetime, ymd("2024-10-08"), ymd("2024-10-10"))))%>% #exclude a cbri sample taken during release
  filter(!(site == "pman" & between(sensor_datetime, ymd("2025-05-28"), ymd("2025-06-01"))))%>% #exclude a pman sample with bad sensor data
# These are not applicable sites for modeling Upper CLP streams and will just introduce error
  filter( site %nin% c("springcreek", "boxcreek", "riverbluffs"))%>%
  filter(!(site %in% c("pbr_fc", "pman_fc") & year(sensor_datetime) == 2025)) #remove pbr/pman fc samples from 2025 since we would be double counting now that ross sensors are deployed at that site

# Create testing set from 2025 data at specific sites
testing <- all_matches_trimmed %>%
  filter((year(sensor_datetime) == 2025 & site %in% c("chd", "sfm", "pbr", "pman", "pbd", "pbr_fc", "pman_fc")))

# Create training/validation set
train_val <- all_matches_trimmed %>%
  anti_join(testing)%>%
  select(site, any_of(features), ChlA_cat, sensor_datetime)%>%
   #combine _virridy site with it's ross pair since they are the same site
  mutate(site = case_when(site == "archery_virridy" ~ "archery",
                          site == "cottonwood_virridy" ~ "cottonwood",
                          site == "riverbend_virridy" ~ "riverbend",
                          TRUE ~ site))

```

## Set up model folds

Since we have relatively little data, we are going to use three CV folds. Create 3 separate folds and 3 models, with one fold used for validation on each model. Generally we are looking for: - Each fold to have \~30-35% of the total data in validation - Each validation set to have a similar ChlA distribution as its training set (median, range) - Each should have a mix of Upper Mainstem (SFM, PFAL, PBD), Upper Trib (CBRI, CHD, LBEA, PENN) & Lower Mainstem (salyer, udall, riverbend, elc, archery, riverbluffs)

```{r}
val_sets <- list(
  
  val_set_1 <- c( "chd" , "udall","cbri", "pman_fc", "cottonwood"),
  
  val_set_2 <- c( "pbd","joei","archery", "pbr_fc" ),
  
  val_set_3 <- c("pfal","sfm","elc","salyer","riverbend")
)
# Function to create plots for a given fold
imap(val_sets,function(val_set_name, fold_id){
  
  val_set <- train_val %>% filter(site %in% val_set_name)
  train_set <- train_val %>% filter(site %nin% val_set_name)
  # Add set labels and combine
  combined <- bind_rows(
    train_set %>% mutate(set = "Train"),
    val_set %>% mutate(set = "Validation")
  )
  # Compute counts and proportions by ChlA_cat and set
  heat_data <- combined %>%
    group_by(set, ChlA_cat) %>%
    summarise(n = n(), .groups = "drop_last") %>%
    mutate(prop = n / sum(n))%>%
    mutate(
      ChlA_cat_range = case_when(
        ChlA_cat == 0 ~ "0-2",
        ChlA_cat == 1 ~ "2-4",
        ChlA_cat == 2 ~ "4-6",
        ChlA_cat == 3 ~ "6+"
      ))
  # Create heat map
  ggplot(heat_data, aes(x = factor(ChlA_cat_range), y = set, fill = prop)) +
    geom_tile(color = "white") +
    geom_text(aes(label = paste0(round(prop * 100, 1), "%")), color = "black", size = 4) +
    scale_fill_gradient(low = "#F5F5F5", high = "#256BF5") +
    labs(
      title = paste0("Fold ", fold_id, " ChlA Category Distribution"),
      subtitle = paste0("Validation sites: ", paste(unique(val_set$site), collapse = ", "), "\nVal: ", nrow(val_set), " samples, Train: ", nrow(train_set), " samples"),
      x = "ChlA (ug/L) Category",
      y = "",
      fill = "Proportion"
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid = element_blank()
    )
})%>%
  # plot all folds
  ggarrange(plotlist = .,  ncol = 1,common.legend = T, legend = "right")


# Create a lookup table mapping each id to its fold
fold_lookup <- imap_dfr(val_sets, ~
                          tibble(site = .x, fold_id = as.integer(gsub("\\D", "", .y))))

# Assign folds automatically using join
train_val <- train_val %>%
  left_join(fold_lookup, by = "site")

```

## Visualize ChlA distribution in training/validation vs testing sets

Currently, our 2025 data had only a few ~ 1% samples in the "extreme" category. This is not ideal, but we will proceed with this split for now given limited data. Our sites with consistently high chl-a, riverbluffs, spring creek or boxcreek, are excluded from this modeling because they contain little comparison value with our target sites (Upper CLP). 

```{r}
heat_data <- train_val %>%
  mutate(group = "Train/Val") %>%
  bind_rows(
    testing %>%
      mutate(group = "Testing")
  ) %>%
  group_by(group, ChlA_cat) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(prop = n / sum(n))%>%
  mutate(
    ChlA_cat_range = case_when(
      ChlA_cat == 0 ~ "0-2",
      ChlA_cat == 1 ~ "2-4",
      ChlA_cat == 2 ~ "4-6",
      ChlA_cat == 3 ~ "6+"
    ))

# Plot as heat map
ggplot(heat_data, aes(x = factor(ChlA_cat_range), y = group, fill = prop)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(round(prop * 100, 1), "%")), color = "black", size = 3.5) +
  scale_fill_gradient(low = "#F5F5F5", high = "#256BF5") +
  labs(
    title = paste0(
      "Training/Validation Set: ", nrow(train_val), " samples              ",
      "Testing Set: ", nrow(testing), " samples"
    ),
    x = "ChlA (ug/L) Category",
    y = "Model Group",
    fill = "Proportion"
  ) +
  ROSS_theme +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank()
  )

```

# Model Training

For our model training we will first do some feature tuning to see which features are most useful for predicting ChlA We will then do hyperparameter tuning on the selected features to optimize model performance.

## Feature Tuning

### View correlations between features and target

Target: (ChlA and by extension ChlA_cat)


```{r}

# Combine training/validation and testing data
modeling_data <- train_val %>%
  mutate(group = "Train/Val") %>%
  bind_rows(
    testing %>%
      mutate(group = "Testing")
  ) %>%
  distinct()%>%
  #adding in the raw ChlA values for correlation plotting
  left_join(all_matches%>%select(ChlA, site, sensor_datetime)%>%
              mutate(site = case_when(site == "archery_virridy" ~ "archery",
                          site == "cottonwood_virridy" ~ "cottonwood",
                          site == "riverbend_virridy" ~ "riverbend",
                          TRUE ~ site)),
             by = c("site", "sensor_datetime"))%>%
  mutate(
    watershed = ifelse(
      site %in% c("joei", "cbri", "chd", "pfal", "pbr", "sfm", "pman", "pbd", "pbr_fc", "pman_fc"),
      "UpperPoudre", 
      "LowerPoudre"
    )
  )
map(features, function(feat) {

  cor <- round(cor(modeling_data[[feat]], modeling_data$ChlA, use = "complete.obs"), 2)
  
  ggplot(modeling_data, aes(x = .data[[feat]], y = ChlA, color = watershed)) +
    geom_point() +
    annotate(
      "text",
      x = mean(range(modeling_data[[feat]], na.rm = TRUE)), # middle of x range
      y = max(modeling_data$ChlA, na.rm = TRUE) * 0.95,      # 95% of y max (near top)
      label = paste0("Corr: ", round(cor(modeling_data[[feat]], modeling_data$ChlA, use = "complete.obs"), 2)),
      hjust = 0.5,
      vjust = 1,
      size = 4
    ) +
    labs(x = feat, y = "ChlA (ug/L)", color = "Watershed") +
    #facet_wrap(~watershed, scales = "free") +
    ROSS_theme
})%>%
  wrap_plots(., guides = "collect")

```

### Pruning Features with high correlations

We want to avoid using too many highly correlated features in the model, as this can lead to overfitting and instability. We will look at the correlation matrix and remove some of the more redundant features.

```{r}
# get a correlation matrix across all features
corr_matrix <- modeling_data %>% 
  select(all_of(features)) %>% 
  cor(use = "pairwise.complete.obs")

# we just want half the matrix, so recode those in the lower triangle to NA
corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter for those with correlation > 0.9
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.9) %>%
  arrange(desc(abs(correlation)))

# look at the list and make some educated decisions about features
high_corr_pairs

fewer_features <- features[features %nin% c( "temp_mean_24hr", "Temperature")]

# do the correlation matrix again to make sure that is okay
corr_matrix <- modeling_data %>% 
  select(all_of(fewer_features)) %>% 
  cor(use = "pairwise.complete.obs")

corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.85) %>%
  arrange(desc(abs(correlation)))

high_corr_pairs

```

`high_corr_pairs` has features we should be cautious of using together. We will try to avoid using more than one of these in the same model.

#### Run across folds testing different features to see which improve performance

Here I run the function (`xgboost_feature_selection_cat()`) which just runs a basic xgboost model with the specified features and default hyperparameters. It returns feature importance, shap values, and best iteration info for each fold.
Then I am going to look at feature importance and shap values to see which features are most useful, then trim features back to reduce our correlative and overfitting issues.

```{r}
 
#xgboost hyper parameters

 default_hyper_params <- list(
  objective        = "multi:softprob",
  eval_metric      = "mlogloss",
  eta              = 0.05,
  gamma            = 0.6,
  alpha            = 0,
  lambda           = 1,
  max_depth        = 4,
  subsample        = 0.5,
  colsample_bytree = 0.5,
  min_child_weight  = 2,
  num_class        = length(unique(train_val$ChlA_cat)) 
)

# Weighting function for ChlA categories to emphasize high ChlA values
high_weight <- function(df) {
  case_when(
    df$ChlA_cat == 3 ~ 4, # highest ChlA category = highest weight
    df$ChlA_cat == 2 ~ 2, 
    .default = 1          # lowest categories (including 1 or NA) = base weight
  )
}

#set aside sensor data that are generally useful
general_features <- c("Specific Conductivity", "Temperature","Turbidity","canyon_mouth_daily_flow_cfs","DO", "pH")

feature_results <- map(
  #trying all the combinations of features we think might be useful (commented out were tested but found to not improve base model)
  list(
    gen_feat = general_features, 
    fewer_features = c(fewer_features),
    fewer_features_no_sc = setdiff(fewer_features, c("Specific Conductivity", "sc_sd_24hr")), 
    fewest_features = setdiff(fewer_features, c("Specific Conductivity", "sc_sd_24hr", "temp_max_24hr"))
    
  ) , ~xgboost_feature_selection_cat(
    train_val_df = train_val,
    features_to_test = .x,
    target_col = "ChlA_cat",
    default_hyper_params = default_hyper_params
  ))
```

#### Feature Set Results

##### mlogloss

We compare across folds to see if there is big differences in folds (ie unbalanced data) and also compare each feature set against "base" model of general features (sensors). Looking for features that make meaningful improvements in mlogloss across all folds. If our mlogloss improves but that is related to only a single fold, we may want to reevaluate the feature set as it is likely overfitting to a given site(s). Keep in mind that this is weighted mlogloss so it may be higher than unweighted mlogloss you see in other places.

```{r}
# Highlight differences between folds in mlogloss
mlogloss_table <- map_dfr(feature_results, ~{
  # Compute feature set description
  unique_features <- setdiff(.x$features, general_features)
  final_features <- c("general_features", unique_features)
  feature_desc <- paste(final_features, collapse = ", ")
  
  # Take best_msg tibble and rename eval_mlogloss by fold
  .x$best_msg %>%
    select(fold_id, eval_mlogloss) %>%
    mutate(fold_col = paste0("mlogloss_fold_", fold_id)) %>%
    select(-fold_id) %>%
    mutate(eval_mlogloss = round(eval_mlogloss, 2)) %>%
    pivot_wider(names_from = fold_col, values_from = eval_mlogloss) %>%
    mutate(
      features = feature_desc,
      mean_mlogloss = round(mean(c_across(starts_with("mlogloss_fold_")), na.rm = TRUE), 2)
    )%>%
    select(features, mean_mlogloss, mlogloss_fold_1, mlogloss_fold_2, mlogloss_fold_3)
  
}, .id = "feature_set")%>%
  
  #check if new feature is better than baseline
  mutate(
    baseline_mlogloss = mean_mlogloss[feature_set == "gen_feat"],
    improves_mlogloss = if_else(mean_mlogloss < baseline_mlogloss, round(mean_mlogloss - baseline_mlogloss, 2), 0)
  ) %>%
  select(-baseline_mlogloss, -features)%>%
  arrange(improves_mlogloss)%>%
  
  print(mlogloss_table, n = Inf, width = Inf)
```

##### Feature Importance

This comes directly from the xgboost package and is based on Gain, Cover, and Frequency of splits. We summarize across folds to see which features are most important across all folds.

```{r}
#table with all importances
importance_tables <- map(feature_results, ~{
  imp <- .x$model_importance
  
  # Identify fold columns dynamically
  gain_cols <- grep("^Gain_fold_", names(imp), value = TRUE)
  cover_cols <- grep("^Cover_fold_", names(imp), value = TRUE)
  freq_cols <- grep("^Frequency_fold_", names(imp), value = TRUE)
  
  imp %>%
    mutate(
      mean_gain = rowMeans(across(all_of(gain_cols)), na.rm = TRUE),
      mean_cover = rowMeans(across(all_of(cover_cols)), na.rm = TRUE),
      mean_frequency = rowMeans(across(all_of(freq_cols)), na.rm = TRUE)
    ) %>%
    mutate(new_feat = ifelse(Feature %in% general_features, "", "*")) %>%
    select(Feature, new_feat, mean_gain, mean_cover, mean_frequency) %>%
    
    arrange(desc(mean_gain))
}, .id = "feature_set")

importance_tables
#Get features with less than 0.02 gain on average across folds in fewer_features to consider removing
cat("Features with mean gain < 0.02 in 'fewer_features':\n")
importance_tables$fewest_features%>%
  filter(mean_gain < 0.02)

```

##### Shapley Values

```{r}
# #table with all shap values
shap_summary <- map_dfr(feature_results, ~{
  shap <- .x$shap_values

  val_shap_cols <- grep("^val_mean_abs_shap_fold_", names(shap), value = TRUE)
  train_shap_cols <- grep("^train_mean_abs_shap_fold_", names(shap), value = TRUE)

  shap %>%
    mutate(
      mean_val_shap = round(rowMeans(across(all_of(val_shap_cols)), na.rm = TRUE), 3),
      mean_train_shap = round(rowMeans(across(all_of(train_shap_cols)), na.rm = TRUE), 3)
    ) %>%
    mutate(val_train_diff = mean_val_shap - mean_train_shap) %>%
    select(Feature, mean_val_shap, mean_train_shap, val_train_diff)
}, .id = "feature_set")

shap_long <- shap_summary %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "type",
    values_to = "mean_shap"
  ) %>%
  mutate(
    type = recode(type,
                  mean_val_shap = "Validation",
                  mean_train_shap = "Training"),
    feature_set = factor(feature_set, levels = unique(feature_set))
  ) %>%
  mutate(Feature = factor(Feature, levels = c( rev(feature_results$fewer_features$features),"Temperature")))

ggplot(shap_long, aes(x = feature_set, y = Feature, fill = mean_shap)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_viridis_c(
    option = "magma",
    limits = c(0, max(shap_long$mean_shap, na.rm = TRUE)),
    name = "Mean |SHAP|",
    direction = -1  # reverses so dark = low, bright = high (optional)
  )+
  facet_wrap(~type) +
  labs(
    x = "Feature Set",
    y = "Feature",
    title = "Mean SHAP Values by Feature and Feature Set"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )

```

### TV performance of best feature set

After trimming down a couple of features (minimal improvement/low shap values), we should take a quick look at model performance in TV space with our base model (default hyperparameters)

```{r fig.dim=c(10,8)}

best_models <- feature_results$fewest_features$model
ft_tv_plot <- map(1:3, ~plot_tvt_fold_acc(fold_model = best_models[[.x]],
                                         fold_num = .x,
                                         train_df = train_val %>% filter(fold_id != .x),
                                         val_df = train_val %>% filter(fold_id == .x),
                                         title = "Default Hyperparameters ",
                                         target_col = "ChlA_cat", 
                                         units = "ug/L",class_levels = c("0-2","2-4","4-6","6+")))%>%
   unlist()%>%
  wrap_plots(., ncol = 2, guides = "collect") &
  theme(legend.position = "right")

 ft_tv_plot

```

Fold one and three appear to be performing the worst on the validation set (lower kappa & accuracy). Across the board, there is just too many samples in class 0 (0-2) so the model is over predicting that class. We will try hyperparameter & weight tuning next to see if we can improve performance.

## Hyperparameter Tuning

Now that we have a feature set, we can do some hyperparameter tuning to see if we can improve model performance. This will take a while to run, so we will save the results and load them in later.

### Run hyperparameter tuning

```{r}

features <- best_models[[1]]$feature_names

# Reduce tune grid for reruns (ie. General best params)
tune_grid = expand.grid(
  objective = "multi:softprob",
  eval_metric = "mlogloss",
  nrounds = 10000,
  max_depth = c(2, 3, 4),
  eta = c( 0.01, 0.05, 0.1),
  gamma = c(0.8),
  lambda = c(1, 10),
  alpha = c(0, 1),
  colsample_bytree = c(0.8),
  min_child_weight = c(2, 4),
  subsample = c(0.5, 0.8),
  num_class = length(unique(train_val$ChlA_cat))
)

high_weight <- function(df) {
  case_when(
    df$ChlA_cat == 3 ~ 4, # highest ChlA category = highest weight
    df$ChlA_cat == 2 ~ 2, 
    .default = 1          # lowest categories (including 1 or NA) = base weight
  )
}
set.seed(123)
# 
hyperp_results <- xgboost_hyperparameter_tuning_cat(
  data = train_val %>% select(ChlA_cat, id = site, fold_id, any_of(features)),
  weights = high_weight,
  tune_grid = tune_grid,
  target_col = "ChlA_cat",
  site_col = "id",
  units = "mg/L",
  class_levels = c("0-2", "2-4", "4-8", "6+"))

```

### Choosing best grids

### Fold 1

```{r}

hyperp_results[[1]]$perf_dist
hyperp_results[[1]]$eval_plot
hyperp_results[[1]]$tv_plot

best_fold_1_grid <- 38
```

### Fold 2

```{r}
hyperp_results[[2]]$perf_dist
hyperp_results[[2]]$eval_plot
hyperp_results[[2]]$tv_plot

best_fold_2_grid <- 60

```

### Fold 3

```{r}
hyperp_results[[3]]$perf_dist
hyperp_results[[3]]$eval_plot
hyperp_results[[3]]$tv_plot

best_fold_3_grid <- 65

```

## Extract Best Models

```{r}

full_model <- map(1:3, function(i) {

  best_grid <- paste0("grid_", get(glue("best_fold_{i}_grid")))
  model_info <- hyperp_results[[i]]$grid_results[[best_grid]]
  list(
    model = model_info$model,
    perf = model_info$perf,
    params = model_info$params,
    training_data = train_val %>% filter(fold_id != i),
    validation_data = train_val %>% filter(fold_id == i)
  )
})
# Read in model to look at results (Comment out code from last few chunks)
#full_model <- readRDS(here("data", "derived", "models","chla_realtime", "in_dev", paste0("chla_categorical_xgboost_models_",version,".rds")))

```

### Checking Results

#### Feature Importance

```{r}
importance_list <- lapply(seq_along(full_model), function(i) {
  fold_model <- full_model[[i]]$model
  features <- fold_model$feature_names
  importance_matrix <- xgb.importance(feature_names = features, model = fold_model)
  importance_df <- as.data.frame(importance_matrix)
  importance_df$Fold <- paste("Fold", i)
  return(importance_df)
})

print(importance_list)
```

#### Shapley Importance

```{r}

# # Function to plot SHAP values for a given fold
plot_shap <- function(fold_set, train_data, val_data, fold_model) {
  #get features from model
  features <- fold_model$feature_names
  # Predictions on validation/training set
  explainer <- function(object, newdata) {
    preds <- predict(object, as.matrix(newdata))
    
    # Convert to matrix: rows = samples, cols = classes
    preds_mat <- matrix(preds, ncol = fold_model$params$num_class, byrow = TRUE)
    
    # Get the maximum probability for each sample
    max_probs <- apply(preds_mat, 1, max)
    
    return(max_probs)
  }
  
  shap_values_train <- fastshap::explain(
    object = fold_model ,
    X = as.data.frame(train_data[, features]),
    pred_wrapper = explainer,
    nsim = 10,
    adjust = TRUE,
    shap_only = FALSE
  )
  sv <- shapviz(shap_values_train)
  train_plot <- sv_importance(sv, kind = "beeswarm")
  
  shap_values_val <- fastshap::explain(
    object = fold_model,
    X = as.data.frame(val_data[, features]),
    pred_wrapper = explainer,
    nsim = 10,
    adjust = TRUE,
    shap_only = FALSE
  )
  
  sv <- shapviz(shap_values_val)
  val_plot <- sv_importance(sv, kind = "beeswarm")
  
  train_plot + val_plot 
}
# 
map(1:3, ~ plot_shap(
  fold_set = .x,
  fold_model = full_model[[.x]]$model,
  train_data = full_model[[.x]]$training_data, 
  val_data = full_model[[.x]]$validation_data
))%>%
  wrap_plots(ncol = 1, guides = "collect")+
  plot_annotation(
    title = paste("SHAP Values"),
    subtitle = "Left: Training Set | Right: Validation Set"
  )

```

### Plots of TV Performance

Now let's look at our performance in TV space.

```{r}

hyperparam_tv_plots <- map(1:3, ~plot_tvt_fold_acc(train_df = full_model[[.x]]$training_data,
                                                  val_df = full_model[[.x]]$validation_data,
                                                  fold_model = full_model[[.x]]$model, # a list where each element is a model for that fold  
                                                  fold_num = .x,
                                                  title = "",class_levels = c("0-2","2-4","4-6","6+"),
                                                  target_col = "ChlA_cat", units = "ug/L")
 )%>%
  unlist()

hp_tv_plot<- wrap_plots(hyperparam_tv_plots, ncol = 2, guides = "collect")&
  theme(legend.position = "bottom")
#ggsave(here("docs/uclp_dss/figs/realtime_chla_modeling/performance/chla_cat_model_fold_validation_performance.png"), width = 14, height = 10)

hp_tv_plot

```

#### Against base hyperparameters

We should hope that all our tuning efforts have improved the model!!

```{r fig.dim = c(14,14)}
# Compare best hyperparameter model to default hyperparameter model
final_plot <- wrap_plots(
  ft_tv_plot,
  hp_tv_plot,
  ncol = 2,
  guides = "collect"
) &
  theme(legend.position = "bottom")
final_plot <- final_plot + 
  plot_annotation(
    title = "Model Performance Comparison: Default (left) vs Tuned Hyperparameters (right)"
  )

final_plot

```

Looking at the plots, we can see that the tuned hyperparameter model has improved performance (kappa) in all folds compared to the default hyperparameter model. The validation accuracy has decreased in all folds but we have increased our accuracy in classes other than 0-2 (likely due to weighting). In all cases, we don't seem to be predicting well at higher classes (few samples) and generally don't have good model prediction. We saw the highest improvement in fold three where kappa increased from 0.23 to 0.4 but still had an overall accuracy of only 70% & a F1 score of 0.6 on validation


## Save as "in_dev" and remove unneeded objects

If we want to test additional ideas for bias correction or look at model performance on timeseries data, we can just run this script until `model_training` and then load in the "in_dev" model file to avoid having to rerun all the tuning again.

```{r}
#commented out so we don't overwrite existing model files
#saveRDS(full_model, file = here("data", "derived", "models","chla_realtime", "in_dev", paste0("chla_categorical_xgboost_models_",version,".rds")))
#if we want to reload model
#full_model <- readRDS(here("data", "derived", "models","chla_realtime", "in_dev", paste0("chla_categorical_xgboost_models_",version,".rds")))

rm(list = setdiff(ls(), c("train_val", "testing", "full_model","version","id_cols",
                          "complete_sensor_data","sensor_final", 
                          #keeping functions around
                          "apply_training_scale","plot_tvt_fold_acc","plot_train_test_ensemble_acc",
                          "load_grab_sample_data","load_sensor_data", "plot_cat_model_timeseries")))
```
# Evaluate on Test Set

We should only do this once we feel good about our model and are ready to see how it performs on the unseen test set. This should only be run after big changes to the model or if we get new data (ie The model should be in a "final/working" version)

## Visualize Train Val Test

### Individual folds

```{r}
plots <- map(1:3, ~{
  p <- plot_tvt_fold_acc(train_df = full_model[[.x]]$training_data, 
                    val_df = full_model[[.x]]$validation_data, 
                    test_df = testing, 
                    fold_model = full_model[[.x]]$model, 
                    fold_num = .x,
                    target_col = "ChlA_cat",
                    units = "ug/L",
                    class_levels = c("0-2","2-4","4-6","6+"))$test_plot
  
  # ggsave(glue(here("docs", "uclp_dss", "figs", "realtime_chla_modeling", "performance", "ross_only_model_fold_{.x}.png")),
  #      plot = p , width = 10, height = 6, dpi = 500)
  p
})
#Print plots
plots

```

### All folds together

```{r, fig.dim= c(10,8)}
# Single plot with all folds
wrap_plots(plots, ncol = 1, guides = "collect") &
  theme(legend.position = "bottom")
# ggsave(here("docs","uclp_dss", "figs", "realtime_chla_modeling","performance",paste0("ross_only_cat_model_all_folds_test", version, ".png")),
#        width = 14, height = 8, dpi = 500)
```

### Ensemble predictions

```{r}
# Plot ensemble
ensemble <- plot_train_test_ensemble_acc(
  train_df = train_val,
  test_df = testing,
  models = map(full_model, "model"),
  target_col = "ChlA_cat",
  units = "ug/L",voting_method = "soft",class_levels = c("0-2", "2-4", "4-6", "6+")
)

ensemble_plot <- wrap_plots(ensemble, ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")

ensemble_plot

ggsave(here("docs","uclp_dss", "figs", "realtime_chla_modeling","performance",paste0("ross_only_cat_model_ensemble", version, ".png")),
       plot = ensemble_plot, width = 12, height = 10, dpi = 500)
```

Overall, we have quite poor performance on the test set (Accuracy: 67%, F1: 0.6, Kappa: 0.26). I do not feel particularly confident in this model predicting any high chl-a values in the upper clp. We may need to remove lower CLP samples from training so we can only represent conditions in the UCLP better. Additionally, we may need to get more high chl-a samples in the training set to better represent those conditions.

Our model ensemble do preform decently well on the training set (Accuracy: 81%, F1: 0.78, Kappa: 0.63) but still struggles to predict higher chl-a categories and distinguishing between 0-2 & 2-4 mg/L

# Timeseries Space Performance

## Prep sensor timeseries & water chem

```{r}
#loading in all sensor data and applying training scaling
sensor_ts <- apply_training_scale(new_data = sensor_final, 
                                  full_dataset = sensor_final, 
                                  features = full_model[[1]]$model$feature_names)%>%
  select(site, DT_round, any_of(full_model[[1]]$model$feature_names))

water_chem <- load_grab_sample_data()%>%
  mutate(ChlA_cat = cut(ChlA, breaks = c(0, 2, 4, 6, 30), labels = c("0-2", "2-4", "4-6", "6+")), 
         ChlA_cat = as.numeric(factor(ChlA_cat))-1)
```

## Example 1: CHD 2025

```{r}

chd_plot <- plot_cat_model_timeseries(model_input_df = sensor_ts, 
                                      water_chem_df = water_chem%>%filter(collector == "ROSS"), 
                                      summary_interval = "24 hour",
                                      site_sel = "chd", 
                                      site_title = "Chambers Lake outflow",
                                      model = map(full_model, "model"),
                                       voting_method = "soft",
                                      target_col = "ChlA_cat",
                                      units = "ug/L",
                                      start_DT = "2025-05-01 12:00",
                                      end_DT = "2025-09-05 00:00",
                                      cat_colors = c("0-2" = "#a6cee3", "2-4" = "#b2df8a", "4-6" = "orange", "6+" = "red2"),
                                      subtitle_arg = "No Site Flow")
# ggsave(here("docs", "uclp_dss", "figs", "realtime_chla_modeling", "timeseries", paste0("chambers_outflow_20250501_20250705_cat_chla.png")),
#        plot = chd_plot,
#        width = 12,
#        height = 8,
#        dpi = 300)
chd_plot



```

## Example 2: PFAL 2025

```{r}

pfal_plot <- plot_cat_model_timeseries(
  model_input_df = sensor_ts, 
  water_chem_df = water_chem, 
  summary_interval = "24 hour",
  site_sel = "pfal", 
  site_title = "Poudre @ Poudre Falls",
  model = map(full_model, "model"),
  voting_method = "soft",
  target_col = "ChlA_cat",
  units = "ug/L",
  start_DT = "2025-04-16 00:00",
  end_DT = "2025-10-10 00:00",
  subtitle_arg = "Baseline Model",
  cat_colors = c("0-2" = "#a6cee3", "2-4" = "#b2df8a", "4-6" = "orange", "6+" = "red2")
)
# ggsave(
#   here("docs", "uclp_dss", "figs", "realtime_chla_modeling", "timeseries",
#        paste0("pfal_20250416_20250710_cat_chla.png")),
#   plot = pfal_plot, width = 12, height = 8, dpi = 300
# )
pfal_plot




```

## Example 3: PBR 2025

```{r}

pbr_plot <- plot_cat_model_timeseries(
  model_input_df = sensor_ts, 
  water_chem_df = water_chem, 
  summary_interval = "24 hour",
  site_sel = "pbr", 
  site_title = "Poudre @ Rustic",
  model = map(full_model, "model"),
  voting_method = "soft",
  target_col = "ChlA_cat",
  units = "ug/L",
  start_DT = "2025-04-16 00:00",
  end_DT = "2025-10-10 00:00",
  subtitle_arg = "Baseline Model",
  cat_colors = c("0-2" = "#a6cee3", "2-4" = "#b2df8a", "4-6" = "orange", "6+" = "red2")
)
# ggsave(
#   here("docs", "uclp_dss", "figs", "realtime_chla_modeling", "timeseries",
#        paste0("pbr_20250416_20251010_cat_chla.png")),
#   plot = pbr_plot, width = 12, height = 8, dpi = 300
# )

pbr_plot




```

## Example 4: PBD 2025

```{r}

pbd_plot <- plot_cat_model_timeseries(
  model_input_df = sensor_ts, 
  water_chem_df = water_chem, 
  summary_interval = "12 hour",
  site_sel = "pbd", 
  site_title = "Poudre @ Canyon Mouth",
  model = map(full_model, "model"),
  voting_method = "soft",
  target_col = "ChlA_cat",
  units = "ug/L",
  start_DT = "2025-08-16 00:00",
  end_DT = "2025-11-20 00:00",
  subtitle_arg = "Baseline Model",
  cat_colors = c("0-2" = "#a6cee3", "2-4" = "#b2df8a", "4-6" = "orange", "6+" = "red2")
)
# ggsave(
#   here("docs", "uclp_dss", "figs", "realtime_chla_modeling", "timeseries",
#        paste0("pbd_20250816_20251120_cat_chla.png")),
#   plot = pbd_plot, width = 12, height = 8, dpi = 300
# )

pbd_plot




```
