---
title: "01_data_prep"
author: "Sam Struthers- CSU ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

# Setup & Custom Functions

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(dplyr.summarise.inform = FALSE)
# load libraries
source("src/setup_libraries.R")
# custom function to reduce overflagging
source("src/apply_cleaning_filters.R")
# custom function to apply interpolation for missing data
source("src/apply_interpolation_missing_data.R")
# custom function to apply low pass binomial filter
source("src/apply_low_pass_binomial_filter.R")
# custom function to calculate medians in data
source("src/apply_timestep_median.R")
# custom function to calculate lagged parameters
source("src/calc_lagged_parameter.R")
#custom function to apply fdom temp correction
source("src/apply_fdom_temp_corr.R")
```

The purpose of this script is to pull in all verified sensor data from 2023 and 2024, as well as auto QAQC'ed data from 2025, apply some additional cleaning steps (interpolation, smoothing, hourly median) and then combine into a single long format dataset as well as a wide format dataset with some additional lagged parameters calculated. Finally, we will add in flow data from the Canyon Mouth guage. 

This script also includes the loading/parsing of ROSS and FC sample data into a single file: "data/upper_clp_dss/modeling/ROSS_FC_water_chemistry.parquet"

# Pull in & Clean up sensor data

## Set Data Targets (sites and parameters)

```{r}
site_levels <- c("salyer", "udall", "riverbend", "cottonwood","riverbend_virrridy", "cottonwood_virridy", "elc",
                 "archery", "archery_virridy" ,  "riverbluffs", "joei", "cbri", "chd", "pfal", "sfm", "pbr", "pman", "pbd", "penn", "lbea", "springcreek", "boxcreek")

parameter_levels <- c("Chl-a Fluorescence", "Depth", "DO", "ORP", "pH","Specific Conductivity", "Temperature", "Turbidity", "FDOM Fluorescence") 

all_combinations <- crossing(
  site = site_levels,
  parameter = parameter_levels
)
site_level_string <- paste(site_levels, collapse = "|")
parameter_level_string <- paste(parameter_levels, collapse = "|")
```

## 2023 dataset

2023 data was manually verified after passing auto-QAQC, data is binned into two categories of PASS or FAIL based on the flag. If a flagged point is PASSED, it is of poor quality and should be omitted from the final dataset. If a flagged point is FAILED, it is of good quality and should be retained in the final dataset. If a non-flagged point is PASSED, it is of good quality and should be retained in the final dataset. If a non-flagged point is FAILED, it is of poor quality and should be omitted from the final dataset.

```{r}
#load in 2023 verified data
# Pulling in the manually verified data
verified_names <- list.files(
  here("data", "raw", "sensor", "manual_data_verification", "2023_cycle", 
       "in_progress", "virridy_verification", "verified_directory"),
  full.names = T)

finalized_dataset_2023 <- verified_names %>% 
  map(\(file_path){
    site_parameter_df <- read_rds(file_path) %>%
      fix_site_names()
    return(site_parameter_df)
  }) %>%
  set_names(map_chr(., ~ paste0(unique(.x$site), "-", unique(.x$parameter)))) %>%
  keep_at(imap_lgl(., ~ grepl(site_level_string, .y))) %>%
  keep_at(imap_lgl(., ~ grepl(parameter_level_string, .y))) 
# Pulling in the post verified data
post_verified_names <- list.files(
  here("data", "raw", "sensor",  "manual_data_verification", "2023_cycle",
       "in_progress", "virridy_verification", "post_verified_directory"),
  full.names = T)
post_finalized_dataset_2023 <- post_verified_names %>% 
  map(\(file_path){
    site_parameter_df <- read_rds(file_path) %>%
      fix_site_names()
    return(site_parameter_df)
  }) %>%
  set_names(map_chr(., ~ paste0(unique(.x$site), "-", unique(.x$parameter)))) %>%
  keep_at(imap_lgl(., ~ grepl(site_level_string, .y))) %>%
  keep_at(imap_lgl(., ~ grepl(parameter_level_string, .y))) 
# Combining all 2023 data by preferentially using post verified data
post_names <- intersect(names(post_finalized_dataset_2023), names(finalized_dataset_2023))
non_post_names <- setdiff(names(finalized_dataset_2023), post_names)
dataset_2023 <- c(
  post_finalized_dataset_2023[post_names], 
  finalized_dataset_2023[non_post_names]
)
all_data_2023 <- dataset_2023 %>%
  bind_rows() %>% 
  data.table() %>%
  fix_site_names() %>% 
  mutate(
    clean_mean = case_when(
      is.na(flag) & verification_status == "PASS" ~ mean,
      is.na(flag) & verification_status == "FAIL" ~ NA,
      !is.na(flag) & verification_status == "PASS" ~ NA,
      !is.na(flag) & verification_status == "FAIL" ~ mean
    ),
    clean_flag = case_when(
      is.na(flag) & verification_status == "PASS" ~ NA,
      is.na(flag) & verification_status == "FAIL" ~ "MANUAL FLAG",
      !is.na(flag) & verification_status == "PASS" ~ flag,
      !is.na(flag) & verification_status == "FAIL" ~ NA
    )
  ) %>%
  filter( !is.na(site),
    # Filter based on DT. The 2023 data is in MST.
    DT_round >= as.POSIXct("2023-01-01 00:00:00", tz = "MST") & DT_round <= as.POSIXct("2023-12-31 11:59:59", tz = "MST")
  ) %>%
  #fixing virridy site names
  mutate(site = gsub(" virridy","_virridy", site  ))%>%
  dplyr::select(DT_round, DT_join, site, parameter, mean = clean_mean, flag = clean_flag,last_site_visit)
```

### Cleaning tasks (interpolation, smoothing, hourly median)

```{r}
all_data_2023_final <- apply_interpolation_missing_data(df = all_data_2023, new_value_col = "mean_filled", max_gap = 4, method = "spline")%>%
  apply_low_pass_binomial_filter(value_col = "mean_filled", new_value_col = "smoothed_mean")%>%
  apply_timestep_median(df = ., timestep  = "1 hour", value_col = "smoothed_mean", new_value_col = "hourly_median")

```

### Save 2023 Data & Clean Up

```{r}
write_parquet(all_data_2023_final, here("data", "collated", "sensor", paste0("all_2023_sensor_data_", Sys.Date(),".parquet")))

rm(all_data_2023, dataset_2023, finalized_dataset_2023, post_finalized_dataset_2023, post_verified_names)
gc()
```


## 2024 dataset

This dataset's verification is still in progress and may need to be rerun as additional data is confirmed. This data has three possible verification statuses: PASS, FLAGGED, and OMIT. If a point is marked as OMIT, it should be removed from the final dataset. If a point is marked as PASS, it is of good quality and should be retained in the final dataset. If a point is marked as FLAGGED, it is of medium quality and should be used cautiously/smoothed in the final dataset. Unverified data points are retained if they do not have an auto-flag, but are removed if they do have an auto-flag.

```{r}
#load in 2024 verified data
# Pulling in the manually verified data
verified_names <- list.files(
  here("data","raw","sensor", "manual_data_verification", "2024_cycle", 
       "in_progress", "verified_directory"),
  full.names = T)

verified_dataset_2024 <- verified_names %>% 
  map(\(file_path){
    site_parameter_df <- read_rds(file_path)
    return(site_parameter_df)
  }) %>%
  set_names(map_chr(., ~ paste0(unique(.x$site), "-", unique(.x$parameter)))) %>%
  keep_at(imap_lgl(., ~ grepl(site_level_string, .y))) %>%
  keep_at(imap_lgl(., ~ grepl(parameter_level_string, .y))) 
# Pulling in the pre verified data
pre_verified_names <- list.files(
  here("data", "raw", "sensor", "manual_data_verification", "2024_cycle",
       "in_progress", "all_data_directory"),
  full.names = T)
pre_verified_dataset_2024 <- pre_verified_names %>% 
  map(\(file_path){
    site_parameter_df <- read_rds(file_path)
    return(site_parameter_df)
  }) %>%
  set_names(map_chr(., ~ paste0(unique(.x$site), "-", unique(.x$parameter)))) %>%
  keep_at(imap_lgl(., ~ grepl(site_level_string, .y))) %>%
  keep_at(imap_lgl(., ~ grepl(parameter_level_string, .y))) 


# Combining all 2024 data by preferentially using post verified data
ver_names <- intersect(names(verified_dataset_2024), names(pre_verified_dataset_2024))
non_ver_names <- setdiff(names(pre_verified_dataset_2024), ver_names)
dataset_2024 <- c(
  verified_dataset_2024[ver_names], 
  pre_verified_dataset_2024[non_ver_names]
)
all_data_2024 <- dataset_2024 %>%
  bind_rows() %>% 
  data.table() %>%
  mutate(
    clean_mean = case_when(
      #verified data
      verification_status == "OMIT"  ~ NA,
      verification_status == "PASS" ~ mean,
      verification_status == "FLAGGED" ~ mean, 
      #unverified data
      is.na(verification_status) & is.na(flag) ~ mean, # no auto flag and not verified
      is.na(verification_status) & !is.na(flag) ~ NA #  auto flag and not verified
    ),
    clean_flag = case_when(
      #verified data
      verification_status == "OMIT"  ~ NA,
      verification_status == "PASS" ~  NA,
      verification_status == "FLAGGED" ~ user_flag,
      #unverified data
      is.na(verification_status) & is.na(flag) ~ NA, # no auto flag and not verified
      is.na(verification_status) & !is.na(flag) ~ flag #  auto flag and not verified
    )
  ) %>%
  filter( !is.na(site),
    # Filter based on DT. The 2024 data is in MST.
    DT_round >= as.POSIXct("2024-01-01 00:00:00", tz = "MST") & DT_round <= as.POSIXct("2024-12-31 11:59:59", tz = "MST")
  ) %>%
  mutate(last_site_visit = with_tz(last_site_visit, tzone = "MST"))%>%
  dplyr::select(DT_round, DT_join, site, parameter, mean = clean_mean, flag = clean_flag, last_site_visit)
```


### Cleaning tasks (interpolation, smoothing, hourly median)

```{r}
all_data_2024_final <- all_data_2024%>%
  apply_interpolation_missing_data(df = ., new_value_col = "mean_filled", max_gap = 4, method = "spline")%>%
  apply_low_pass_binomial_filter(value_col = "mean_filled", new_value_col = "smoothed_mean")%>%
  apply_timestep_median(df = ., timestep  = "1 hour", value_col = "smoothed_mean", new_value_col = "hourly_median")

```

### Save 2024 Data & Clean Up

```{r}
write_parquet(all_data_2024_final, here("data", "collated", "sensor", paste0("all_2024_sensor_data_", Sys.Date(),".parquet")))

rm(all_data_2024, dataset_2024, pre_verified_dataset_2024, pre_verified_names, verified_dataset_2024, verified_names)
gc()
```

## 2025 dataset (auto qaqc only)

Loading in pre-processed 2025 data. This is created by the script `data_organization/Process_2025.Rmd` in the `poudre_sonde_network` repository. Please see that script and update as needed to recollate data from logs/HydroVu

```{r}
#load in 2025 unverified but Auto QAQC'ed data
# Pulling in the non-manually verified data

flagged_2025_files <- list.files(
  here("data","raw","sensor",  "manual_data_verification", "2025_cycle", 
       "hydro_vu_pull", "flagged_data"),
  full.names = T)

flagged_2025_dataset <- flagged_2025_files %>% 
  map(\(file_path){
    site_parameter_df <- read_csv(file_path, show_col_types = F)
    return(site_parameter_df)
  }) %>%
  set_names(map_chr(., ~ paste0(unique(.x$site), "-", unique(.x$parameter)))) %>%
  keep_at(imap_lgl(., ~ grepl(site_level_string, .y))) %>%
  keep_at(imap_lgl(., ~ grepl(parameter_level_string, .y))) 


all_data_2025 <- flagged_2025_dataset %>%
  bind_rows() %>% 
  data.table() %>%
  filter( !is.na(site),
    # Filter based on DT. The 2025 data is in UTC
    DT_round >= as.POSIXct("2025-01-01 00:00:00", tz = "UTC") & DT_round <= as.POSIXct("2025-12-31 11:59:59", tz = "UTC")
  ) %>%
  #converting to MST for later use
  mutate(
    DT_round = with_tz(DT_round, tz = "MST"),
    DT_join = as.character(DT_round), 
    last_site_visit = with_tz(last_site_visit, "MST"), 
    mean = if_else(is.na(mal_flag), mean, NA) # remove reported sensor malfunctions
  )%>%
  dplyr::select(DT_round, DT_join, site, parameter, mean , auto_flag,mal_flag,  last_site_visit)
```


### Cleaning tasks (reducing overflagging, interpolation, smoothing, hourly median)
```{r}
all_data_2025_final <- all_data_2025%>%
  apply_cleaning_filters(df = ., new_value_col = "mean_cleaned")%>%
  apply_interpolation_missing_data(df = ., value_col = "mean_cleaned", new_value_col = "mean_filled", max_gap = 4, method = "spline")%>%
  apply_low_pass_binomial_filter(df = ., value_col = "mean_filled", new_value_col = "smoothed_mean")%>%
  apply_timestep_median(df = ., timestep  = "1 hour", value_col = "smoothed_mean", new_value_col = "hourly_median")
  
```

###  Save 2025 Data
```{r}

write_parquet(all_data_2025_final, here("data", "collated", "sensor", paste0("all_2025_sensor_data", Sys.Date(), ".parquet")))

rm(all_data_2025, flagged_2025_dataset, flagged_2025_files)
gc()
```

# Create long sensor dataset

```{r}

all_sensor_data <- bind_rows(all_data_2023_final, all_data_2024_final, all_data_2025_final)%>%
  select(-auto_flag, -mal_flag, -clean_flag, - flag )
  
write_parquet(all_sensor_data, here("data", "collated", "sensor", paste0("all_sensor_data_long", Sys.Date(), ".parquet")))
```

# Clean up space/memory
```{r}

#remove everything from memory
rm(list=setdiff(ls(), c("calc_lagged_parameter","apply_fdom_temp_corr")))
gc()

```


# Create wide sensor dataset

This comes from the script above and should be updated when an API pull, QAQC is updated or as more data is verified.

I think we could have a function (`update_sensor_data()`) that would essentially run all the code above + code in this section and saves it to come consistent file path. We could then have a `load_sensor_data()` function that always pulls in the most recent data.

### Get most recent files

```{r}
#Define folder path
collated_sensor_folder <- here("data", "collated", "sensor")

#Get most recent files
all_data_2023_wide_file <- list.files(path = collated_sensor_folder, pattern = "all_2023_sensor_data_.*\\.parquet$", full.names = TRUE)%>%tail(1)
all_data_2024_wide_file <- list.files(path = collated_sensor_folder, pattern = "all_2024_sensor_data_.*\\.parquet$", full.names = TRUE)%>%tail(1)
all_data_2025_wide_file <- list.files(path = collated_sensor_folder, pattern = "all_2025_sensor_data_.*\\.parquet$", full.names = TRUE)%>%tail(1)
```

### Read in recent files

```{r}
# Read in all datasets, subset to hourly medians and then pivot to wide format
all_data <- list(read_parquet(file = all_data_2023_wide_file), 
                 read_parquet(file = all_data_2024_wide_file), 
                 read_parquet(file = all_data_2025_wide_file))

#combine each year's data
all_data_wide <- all_data %>%
  map(~ .x %>%
        rename(DT_hourly_round = DT_group) %>%
        distinct(site, DT_hourly_round, parameter, hourly_median) %>%
        pivot_wider(
          id_cols = c(site, DT_hourly_round),
          names_from = parameter,
          values_from = hourly_median
        )%>%
      #cleaning up turbidity values (rounding to nearest value, setting min to 1 and max to 1000)
      mutate(Turbidity = if_else(Turbidity <= 0, 0.1, if_else(Turbidity>=1000, 1000, Turbidity)))
      )
```


## Compute additional lagged parameters

```{r}

final_dataset <-all_data_wide%>%
  map(., function(df){
    df %>%
      #applying FDOM temperature correcting/standardization
      apply_fdom_temp_corr( fdom_col = "FDOM Fluorescence",
                            temp_col = "Temperature",
                            fdom_corr_col = "FDOM_corr_25c",
                            Tr = 25,
                            rho = -0.015)%>%
      # recent means (water temp)
      calc_lagged_parameter(site_col = "site", parameter_col = "Temperature",
                            dt_col = "DT_hourly_round", lag_fun = "mean",
                            time_period = "48 hour", new_value_col = "temp_mean_48hr")%>%
      calc_lagged_parameter(site_col = "site", parameter_col = "Temperature",
                            dt_col = "DT_hourly_round", lag_fun = "mean",
                            time_period = "24 hour", new_value_col = "temp_mean_24hr")%>%
      # recent maxes (temp, fdom)
      calc_lagged_parameter(site_col = "site", parameter_col = "FDOM Fluorescence",
                            dt_col = "DT_hourly_round", lag_fun = "max",
                            time_period = "6 hour", new_value_col = "fdom_max_6hr")%>%
      calc_lagged_parameter(site_col = "site", parameter_col = "FDOM_corr_25c",
                            dt_col = "DT_hourly_round", lag_fun = "max",
                            time_period = "6 hour", new_value_col = "fdom_corr_max_6hr")%>%
      calc_lagged_parameter(site_col = "site", parameter_col = "Temperature",
                            dt_col = "DT_hourly_round", lag_fun = "max",
                            time_period = "24 hour", new_value_col = "temp_max_24hr")%>%
      # recent variance/sd (sc)
      calc_lagged_parameter(site_col = "site", parameter_col = "Specific Conductivity",
                            dt_col = "DT_hourly_round", lag_fun = "var",
                            time_period = "24 hour", new_value_col = "sc_var_24hr")%>%
      mutate(sc_sd_24hr = sqrt(sc_var_24hr))%>%
      #12 hour median turbidity (proxy for recent storm event)
      calc_lagged_parameter(site_col = "site", parameter_col = "Turbidity",
                            dt_col = "DT_hourly_round", lag_fun = "median",
                            time_period = "24 hour", new_value_col = "turb_median_12hr")
  })%>%
  bind_rows()


```

## Get Flow data for CLP & join with sensor data

```{r}

min_date <- min(final_dataset$DT_hourly_round, na.rm = T)
max_date <- max(final_dataset$DT_hourly_round, na.rm = T)

canyon_mouth_flow <- get_telemetry_ts(abbrev = "CLAFTCCO", 
                 parameter = "DISCHRG", 
                 start_date = as.Date(min_date)- days(1),
                 end_date = as.Date(max_date)+ days(1),
                 timescale = "hour")%>%
  select(DT_hourly_round = meas_date, hourly_canyon_flow_cfs = meas_value)%>%
  mutate(DT_hourly_round = with_tz(ymd_hms(DT_hourly_round, tz = "America/Denver"), tzone = "MST"))

canyon_mouth_flow_daily <- get_telemetry_ts(abbrev = "CLAFTCCO", 
                 parameter = "DISCHRG", 
                 start_date = as.Date(min_date)- days(1),
                 end_date = as.Date(max_date)+ days(1),
                 timescale = "day")%>%
  select(date = datetime, daily_canyon_flow_cfs = meas_value)%>%
  mutate( date = as.Date(date))

final_dataset <- final_dataset%>%
  mutate(date = as_date(DT_hourly_round))%>%
  left_join(canyon_mouth_flow, by = "DT_hourly_round")%>%
  left_join(canyon_mouth_flow_daily, by = "date")%>%
  select(-date)
  

```

## Save wide dataset

This will be used downstream to determine matchups/modeling

```{r}
write_parquet(final_dataset, here("data", "collated", "sensor", paste0("all_sensor_data_wide_", Sys.Date(),".parquet")))

```

# Pull in & clean up grab sample data

This has been migrated to `src/generate_chem` scripts and should use these scripts to generate updated chemistry datasets. 
(ie water_chem <- `load_clp_chem_dataset(zenodo_data_version, save_folder_dir, fc_data_path)`)


## ROSS

```{r}

#discrete sample data and location metadata from most recent pub

source("src/pull_ROSS_zenodo_data.R")
zenodo_data <- pull_ROSS_zenodo_data(data_version = "v2025.07.01", DOI = "15883685", save_folder_dir = here("data", "raw", "chem", "ross_clp_chem"))

# Grab the water chemistry dataset

ross_water_chem <- zenodo_data[["most_recent_chem"]]%>%
  mutate(site_code = tolower(site_code), 
         #fixing site names based on sonde deployments with ROSS/Virridy sondes
         site_code = case_when(site_code == "archery" & DT_mst <= ymd("2024-11-30") ~ "archery_virridy",
                          site_code == "timberline" & DT_mst <= ymd("2024-11-30") ~ "riverbend_virridy",
                          site_code == "prospect" & DT_mst <= ymd("2024-11-30") ~ "cottonwood_virridy",
                          #updating to new names
                          site_code == "timberline" & DT_mst >= ymd("2024-11-30") ~ "riverbend",
                          site_code == "prospect" & DT_mst >= ymd("2024-11-30") ~ "cottonwood",
                          site_code == "lincoln" ~ "udall", 
                          site_code == "legacy" ~ "salyer",
                          site_code == "boxelder" ~ "elc",
                          site_code == "tamasag" ~ "bellvue",
                          T ~ site_code))%>%
  filter(DT_mst >= as_date("2023-09-01"))%>%
  distinct()%>%
  dplyr::select(site_code, 
         Date, 
         DT_sample = DT_mst,
         TOC, TN, DOC, ChlA, NO3, NH4, PO4, SC,Cl, lab_turb = Turbidity, TSS,DT_mst_char)%>%
  mutate(collector = "ROSS")
```

## Fort Collins UCLP Data

This is given to us by the City of Fort Collins Watershed Team (Diana Schmidt and Jared Heath). 
2025 Dataset is not yet finalized but can likely still be used for our first models


```{r}

FC_chem_2022_2024 <- read_csv(here("data", "raw","chem", "fc_clp_chem", "UCLP_CWQMP_database_2022-24.csv"), show_col_types = F)%>%
  mutate(Date = as.POSIXct(Date, format = "%m/%d/%y"), 
         NO3 = NO3_N/ 0.2259, 
         SC = TDS/ 0.65, #converting TDS to SC
         Cl = as.numeric(Cl), 
         lab_turb = as.numeric(Turbidity),
         TN = as.numeric(TN_calc),
         ChlA = NA_integer_,
         DOC = NA_integer_, 
         NH4 = NA_integer_, 
         TSS = NA_integer_,
         site_code = tolower(ShortDesc))%>%
  #filter to the period of our sensor deployment and sites where we have sensors
  filter(site_code %in% c("chd", 'pbd', "sfm") & Date >= as_date("2023-09-01"))%>%
  #grab columns of interest
  dplyr::select(site_code, Date,
         TOC, TN, DOC, ChlA, NO3, NH4, PO4 = oPhos, SC,Cl, lab_turb, TSS)%>%
  mutate(collector = "FC")

FC_field_data_2022_2024 <- readxl::read_xlsx(path = here("data", "raw","chem", "fc_clp_chem", "UCLP_field_data_2022-24.xlsx"))%>%
  mutate(
    Date = as.Date(`Date (MM/DD/YYYY)`),
    # Combine date and time into one datetime column from MT to MST for consistency 
    dt_mst = with_tz(ymd_hms(paste(
      format(`Date (MM/DD/YYYY)`, "%Y-%m-%d"), 
      format(`Time (HH:MM:SS)`, "%H:%M:%S")
    ), tz = "America/Denver"), tzone = "MST"),
    # Remove numbers from site name
    site_code = tolower(gsub("\\d+", "", `Site Name`)), 
    DT_mst_char = as.character(dt_mst)
  )%>%
  dplyr::select(site_code, Date, DT_sample = dt_mst, DT_mst_char)
  
# Combine the chemistry data with the field data
FC_chem_2022_2024 <- FC_chem_2022_2024 %>%
  left_join(FC_field_data_2022_2024, by = c("site_code", "Date")) 


# Repeat for 2025 data (TOC only!)

FC_chem_2025 <- readxl::read_xlsx(path = here("data", "raw","chem", "fc_clp_chem", "UCLP_CWQMP_TOC_202504-202506.xlsx"))%>%
  filter(grepl(x = `Parameter Name`, pattern = "Total Organic Carbon", ignore.case = TRUE), 
          `Entered Unit` == "mg/L",
         grepl(x = `Location Name`, pattern = "CHD|PBD|SFM|PBR", ignore.case = TRUE))%>%
  dplyr::select(
    Date = `Sampled Date`, 
    site_name = `Location Name`,
    value = `Corrected Result`
  )%>%
  mutate(site_code = case_when(
    grepl("PBR", site_name, ignore.case = TRUE) ~ "pbr",
    grepl("CHD", site_name, ignore.case = TRUE) ~ "chd",
    grepl("PBD", site_name, ignore.case = TRUE) ~ "pbd",
    grepl("SFM", site_name, ignore.case = TRUE) ~ "sfm",
    TRUE ~ NA_character_),
         TOC = as.numeric(value))%>%
  mutate(collector = "FC")%>%
  dplyr::select(site_code, Date, TOC, collector)

FC_field_data_2025 <- readxl::read_xlsx(path = here("data", "raw", "chem", "fc_clp_chem", "UCLP_field_data_04-062025.xlsx"))%>%
  mutate(
    Date = as.Date(`Date (MM/DD/YYYY)`),
    # Combine date and time into one datetime column from MT to MST for consistency 
    dt_mst = with_tz(ymd_hms(paste(
      format(`Date (MM/DD/YYYY)`, "%Y-%m-%d"), 
      format(`Time (HH:MM:SS)`, "%H:%M:%S")
    ), tz = "America/Denver"), tzone = "MST"),
    # Remove numbers from site name
    site_code = tolower(gsub("\\d+", "", `Site Name`)), 
    DT_mst_char = as.character(dt_mst)
  )%>%
  dplyr::select(site_code, Date, DT_sample = dt_mst, DT_mst_char)


FC_chem_2025 <- FC_chem_2025 %>%
  left_join(FC_field_data_2025, by = c("site_code", "Date"))%>%
  na.omit()


FC_chem <- bind_rows(FC_chem_2022_2024, FC_chem_2025)


#write_csv(FC_chem, "data/FC_clp_chem/FC_testing_data.csv")
```

## Combine data
```{r}

water_chem <- bind_rows(ross_water_chem, FC_chem)

#save for later use
write_parquet(water_chem, here("data" "collated", "chem", "ROSS_FC_water_chemistry.parquet")


# ggplot(water_chem%>%
#          filter(site_code %in% c("sfm", "chd", "pbr")), aes(x = Date, y = TOC, color = collector)) +
#   geom_point() +
#   geom_line()+
#   facet_wrap(~site_code, scales = "free_y") +
#   labs(title = "Water Chemistry Data from Ross and FC",
#        x = "Date",
#        y = "TOC (mg/L)") +
#   theme_minimal()

```

