---
title: "Testing TOC preds on Yampa data & transfer learning"
author: "ROSSyndicate"
date: "2025-12-16"
output: html_document
---

# Puprose

The purpose of this markdown is to test a generalized model to predict TOC on
the Yampa data. Here, we build a single xgboost model for the Poudre and use that 
model to estimate the Yampa, then follow up with a sequential training using 
a holdout of yampa data as the test.


# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("src/setup_libraries.R")
source("src/setup_ross_theme.R")

#these are just packages for testing within this script
library(gridExtra)
library(grid)
library(Ckmeans.1d.dp)
library(DiagrammeR)
library(Metrics)
library(glue)
library(SHAPforxgboost)

set.seed(17)
#set version for saving (update when changes are significant)
#version = "20251211"
# function to apply sensor transformations for TOC modeling
source("src/modeling/apply_sensor_transformations_toc.R")
#function to apply training scale
source("src/modeling/apply_training_scale.R")
# function to tune xgboost hyperparameters
source("src/modeling/xgboost_hyperparameter_tuning.R")
#function to plot performance of each fold
source("src/plotting/plot_tv_fold_perf.R")
#function to plot performance of each fold
source("src/plotting/plot_tvt_fold_perf.R")
#function to plot performance of each fold
source("src/plotting/plot_train_test_ensemble_perf.R")
#function to apply bias correction
source("src/modeling/apply_ml_bias_correction.R")
#load grab sample data function
source("src/collation/load_grab_sample_data.R")
#function to plot model timeseries
source("src/plotting/plot_model_timeseries.R")
```

## Read in sensor dataset

Here, we read in the sensor data, apply min-max scaling defined in the 
`apply_sensor_transformations_toc()` function, and store the dataset for use 
later

```{r}
# Read in all data to create scaling parameters
wide_data_file <- list.files(here("data","collated", "sensor"), 
                             pattern = "all_sensor_data_wide_.*\\.parquet", 
                             full.names = TRUE) %>%
  tail(1)

# Use the complete sensor dataset for scaling parameters (assuming sensor has all necessary columns already)
complete_sensor_data <- read_parquet(wide_data_file) %>%
  apply_sensor_transformations_toc(., dt_col = "DT_hourly_round") %>%
  select(-DO, -Depth, -ORP, -pH) #remove unused sensors


```


## Read in matchup dataset

Matchup dataset is created in 02_sensor_grab_matchup, where we match data from 
the sensors with each available lab-anlyzed TOC sample. In this script we will 
use the dataset that also contains the Virridy Yampa Samples. 

```{r}
matchup_dataset <- list.files(here("data","derived", "matchup_datasets"), 
                              pattern = "parsed_toc_sensor_virridy_clp_", 
                              full.names = TRUE) %>%
  tail(1)

# Read in all matches data
all_matches <- read_csv(file = matchup_dataset, show_col_types = F) %>%
  apply_sensor_transformations_toc(dt_col = "sensor_datetime") %>%
  #remove NAs
  filter(!is.na(TOC)) %>%
  mutate(clean_sensor_datetime = parse_date_time(sensor_datetime, 
                                                 orders = c("%Y-%m-%d %H:%M:%S", 
                                                            "%m/%d/%y %H:%M", 
                                                            "%Y-%m-%d"))) %>%
  mutate(sensor_datetime = force_tz(clean_sensor_datetime, tzone = 'MST'), 
         month = month(sensor_datetime), 
         doy = yday(sensor_datetime),
         sin_doy =  sin(2 * pi * doy / 365), 
         basin = if_else(collector == "Virridy", "Yampa", "Poudre"))
```

## Adding in flow for Yampa Samples
Chose YAMSTECO as it has it accumulates the majority of the basin's flow but is not severly impacted by diversions. This will be tested as our "daily_canyon_flow_cfs" feature in the model.
https://dwr.state.co.us/Tools/Stations/YAMSTECO?params=DISCHRG

```{r}
yampa_data <- all_matches %>% filter(basin == "Yampa")

max_yampa_date = as.character(as.Date(max(yampa_data$sensor_datetime)) + days(1))
min_yampa_date = as.character(as.Date(min(yampa_data$sensor_datetime)) - days(1))

# read in flow data for Yampa sites
yampa_data_w_flow <- yampa_data %>%
  mutate(date = as.Date(sensor_datetime)) %>%
  select(-daily_canyon_flow_cfs)%>%
  left_join(
    #use CDSSR to read in flow data for yampa
    get_telemetry_ts(abbrev = "YAMSTECO", 
                     parameter = "DISCHRG",
                     start_date = min_yampa_date,
                     end_date = max_yampa_date,
                     timescale = "day",
                     api_key = NULL) %>%
      select(date = datetime, daily_canyon_flow_cfs = meas_value),
    by = c( "date")
  )

all_matches <- all_matches %>% 
  # remove old yampa data
  filter(basin != "Yampa")%>%
  # add in new yampa data
  bind_rows(yampa_data_w_flow)

```

## Define ID Columns, Target, and Features

```{r}
# ID columns to keep to use later on 
id_cols <- c("Date", "id","site", "DT_hourly_round", "sensor_datetime","lab_datetime",  "collector","basin")
#What we want to predict
target <- "TOC"
# Get feature columns from complete sensor data but remove id cols and target
features <- setdiff(names(complete_sensor_data), c(id_cols, target)) %>% 
  #remove unused sensors
  setdiff(., c("DO", "Depth", "ORP", "pH")) %>% 
  #remove variables not in the yampa dataset
  setdiff(., c("temp_mean_48hr", "temp_mean_24hr", "fdom_max_6hr", 
               "fdom_corr_max_6hr", "temp_max_24hr", "sc_var_24hr",
               "sc_sd_24hr", "turb_median_12hr", "hourly_canyon_flow_cfs", 
               "data_avail", "daily_canyon_flow")) 
features
```

## Create Scaling Parameters & Normalize Data
We do not have access to the full sensor dataset for the Yampa sites, so we 
will use the scaling parameters from the complete Poudre sensor dataset to 
normalize the matchup data. This may yield some out of samples results when we 
apply the model to the Yampa data. 

```{r}
# Apply scaling to dataset
all_matches_norm <- all_matches %>%
  # Only keep columns that would be in the complete sensor dataset
  select(any_of(id_cols), any_of(features), any_of(target))%>%
  # apply normalization 
  apply_training_scale(
    new_data = .,
    features = features,
    full_dataset = complete_sensor_data
  ) %>%
  #remove any rows with NAs
  na.omit()

```

## Create TVT sets

```{r}
# Trim out any known bad data or sites we don't want to include in modeling
all_matches_trimmed <- all_matches_norm %>%
  #exclude a cbri sample taken during release
  filter(!(id == "cbri" & 
             between(sensor_datetime, ymd("2024-10-08"), ymd("2024-10-10")))) %>% 
  #exclude a pman sample with bad sensor data
  filter(!(id == "pman" & 
             between(sensor_datetime, ymd("2025-05-28"), ymd("2025-06-01")))) %>% 
  filter(id %nin% c("springcreek", "boxcreek")) %>%
  # remove the Yampa data from training/validation set
  filter(collector != "Virridy")

yampa <- all_matches_norm %>%
  filter(collector == "Virridy")

# Create testing set from 2025 data at specific sites
testing <- all_matches_trimmed %>%
  filter(id %in% c("chd", "sfm", "pbr", "pman", "pbd"))

# Create training/validation set
train_val <- all_matches_trimmed %>%
  anti_join(testing) %>%
  select(id, any_of(features), TOC, sensor_datetime, collector) %>%
  # combine archery and archery_virridy since they are the same site
  mutate(id = if_else(id == "archery_virridy", "archery", id))

```

# Model Training

## Feature Tuning 

__View correlations between features and target (TOC)__

Most products with FDOM in them are going to have higher correlation with TOC, 
as well as seasonal (sin_doy) and flow related features. We want to be sure 
that we don't put too many of these in the model as calibration error or sensor 
drift could lead to compounding errors.

```{r}
# Combine training/validation and testing data
modeling_data <- train_val %>%
  mutate(group = "Train/Val") %>%
  bind_rows(
    testing %>%
      mutate(group = "Testing")
  ) %>%
  bind_rows(
    yampa %>%
      mutate(group = "Yampa")
  ) 
```


__Features differing in Yampa Set__

We can see that there are some basin differences in the Yampa samples which may 
influence the performance of the model on the Yampa data. Features like doy, 
f_sc, f_x_sc, daily canyon flow, and f_x_turb show some differences in distribution 
compared to the Poudre samples. We do still see FDOM as a useful starting point 
for predicting TOC in the Yampa as well.

```{r}
plots <-set_names(features) %>%
  map( function(feat) {
    
    cor <- round(cor(modeling_data[[feat]], modeling_data$TOC, use = "complete.obs"), 2)
    
    ggplot(modeling_data, aes(x = .data[[feat]], y = TOC, color = collector)) +
      geom_point() +
      annotate(
        "text",
        x = mean(range(modeling_data[[feat]], na.rm = TRUE)), # middle of x range
        y = max(modeling_data$TOC, na.rm = TRUE) * 0.95,      # 95% of y max (near top)
        label = paste0("Corr: ", round(cor(modeling_data[[feat]], modeling_data$TOC, use = "complete.obs"), 2)),
        hjust = 0.5,
        vjust = 1,
        size = 4
      ) +
      labs(x = feat, y = "TOC (mg/L)", color = "Collector") +
      #facet_wrap(~watershed, scales = "free") +
      ROSS_theme
  })


plots%>%
  wrap_plots(., guides = "collect")

wrap_plots(plots$Sensor_Cond, plots$sin_doy, plots$f_sc, 
           plots$FDOMc, plots$daily_canyon_flow_cfs, plots$f_x_turb, ncol = 3, 
           guides = "collect") &
  theme(legend.position = "bottom")

```

__Pruning Features with high correlations__

We want to avoid using too many highly correlated features in the model, as this
can lead to overfitting and instability. We will look at the correlation matrix
and remove some of the more redundant features.

```{r}
# get a correlation matrix across all features
corr_matrix <- modeling_data %>% 
  select(all_of(features)) %>% 
  cor(use = "pairwise.complete.obs")

# we just want half the matrix, so recode those in the lower triangle to NA
corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter for those with correlation > 0.9
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.9) %>%
  arrange(desc(abs(correlation))) %>% 
  # remove redundant vars
  filter(feature1 %nin% c("FDOM", "doy"))

# look at the list and make some educated decisions about features
high_corr_pairs

fewer_features <- features[features %nin% c("FDOM", "doy", 
                                            # remove vars with high correlation 
                                            # with others
                                            "Temp", "f_log",
                                            "Sensor_Turb")]

# do the correlation matrix again to make sure that is okay
corr_matrix <- modeling_data %>% 
  select(all_of(fewer_features)) %>% 
  cor(use = "pairwise.complete.obs")

corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.85) %>%
  arrange(desc(abs(correlation)))

high_corr_pairs

```

`high_corr_pairs` has features we should be cautious of using together. We will 
try to avoid using more than one of these in the same model.


First, we will start with our base model (live sensor data only) and then test 
adding in other features to see if they improve performance. We will use the 
default hyperparameters below for all models to keep things consistent. Once we 
have tested all the features together, we will want to trim it down a bit so we 
are not using too many highly correlated features or overfitting the model due 
to lack of data

```{r}
#xgboost hyper parameters
tune_grid = expand.grid(
  nrounds = 10000,
  max_depth = c(3, 4),
  eta = c(0.01, 0.005),
  gamma = c(0.05, 0.1, 0.2),
  lambda = c(1,10),
  alpha = c(0, 1),
  colsample_bytree = c(0.8, 0.9),
  min_child_weight = c(2, 3),
  subsample = c(0.5,0.8)
)


# Weighting function for TOC
toc_high_weights <-  function(df) {
  case_when(df$TOC < 4 ~ 1, 
            between(df$TOC, 4, 6) ~ 2,
            df$TOC > 6 ~ 5)
}

```

## CLP hypertuning

Let's do some hyper tuning on the CLP data - this step can be skipped if not changing anything, all models have been saved at the end and the "best" model is read in at the top of the next section.

```{r}
data <- train_val %>% 
  select(TOC, id, any_of(fewer_features)) 

target_col = "TOC"
units = "mg/L"
site_col = "id"

# Prepare features
features <- setdiff(names(data), 
                    c(target_col, site_col))

#setup performance dataframe
perf <- data.frame()

train_data <- train_val
val_data   <- testing

w_train <- toc_high_weights(train_data)
w_val   <- toc_high_weights(val_data)

# set up data matrix for xgb
dtrain <- xgb.DMatrix(
  data = as.matrix(train_data[, features]),
  label = train_data[[target_col]],
  weight = w_train # add weights
)

dval <- xgb.DMatrix(
  data = as.matrix(val_data[, features]),
  label = val_data[[target_col]],
  weight = w_val # add weights
)

# Set train vs val for early stopping
watchlist <- list(train = dtrain, eval = dval)


# only re-run if the models don't exist
if (!file.exists(here("data/derived/models/TOC_transfer/all_basic_CLP_for_transfer.rds"))) {
  
  models <- list()
  
  #run through all hyper parameters and save to fold_models with name foldi_gridj
  for (j in 1:nrow(tune_grid)) {
    
    #setup parameters for tune
    params <- list(
      objective = "reg:squarederror",
      eval_metric = "rmse",
      tree_method = "exact",
      eta = tune_grid$eta[j],
      gamma = tune_grid$gamma[j],
      alpha = tune_grid$alpha[j],
      lambda = tune_grid$lambda[j],
      max_depth = tune_grid$max_depth[j],
      subsample = tune_grid$subsample[j],
      colsample_bytree = tune_grid$colsample_bytree[j],
      min_child_weight  = tune_grid$min_child_weight[j]
    )
    
    #train model with hyper parameters
    model_ij <- xgb.train(
      params = params,
      data = dtrain,
      nrounds = tune_grid$nrounds[j],
      watchlist = watchlist,
      #change early stopping rounds based on eta (smaller eta, larger rounds)
      early_stopping_rounds = ifelse(tune_grid$eta[j] >= 0.1, 250,
                                     ifelse(tune_grid$eta[j] >= 0.01, 500,
                                            1000)),
      verbose = 0,
      nthread = 1
    )
    
    models[[j]] <- model_ij
    
    # pred_col <- paste0(target_col, "_guess")
    # # Predictions on validation set
    # val_data[[pred_col]] <- predict(
    #   model_ij, dval,
    #   iterationrange = c(1, model_ij$best_iteration)
    # )
    # rmse_val  <- rmse(val_data[[target_col]], val_data[[pred_col]])
    # mae_val   <- mae( val_data[[target_col]], val_data[[pred_col]])
    # bias_val  <- bias(val_data[[target_col]], val_data[[pred_col]])
    # 
    # # Predictions on training set (to measure overfitting)
    # train_data[[pred_col]] <- predict(
    #   model_ij, dtrain,
    #   iterationrange = c(1, model_ij$best_iteration)
    # )
    # rmse_train  <- rmse(train_data[[target_col]], train_data[[pred_col]])
    # mae_train   <- mae( train_data[[target_col]], train_data[[pred_col]])
    # bias_train  <- bias(train_data[[target_col]], train_data[[pred_col]])
    
    val_rmse <- as.numeric(sub(".*eval-rmse:([0-9.]+).*", "\\1", model_ij$best_msg))
    train_rmse <- as.numeric(sub(".*train-rmse:([0-9.]+).*", "\\1", model_ij$best_msg))
    
    perf <- rbind(perf, 
                  data.frame(
                    grid_id = j,
                    best_iter = model_ij$best_iteration,  # from early stopping
                    rmse_val = val_rmse,
                    rmse_train = train_rmse,
                    diff = abs(val_rmse - train_rmse)
                  ))
    
    if(j/50 == round(j/50)){
      cat(paste0("  Completed ", j, " of ", nrow(tune_grid), " hyperparameter combinations...\n"))
    }
    
  }
  
  write_rds(models, here("data/derived/models/TOC_transfer/all_basic_CLP_for_transfer.rds"))
  write_csv(perf, here("data/derived/models/TOC_transfer/all_basic_CLP_for_transfer_performance.csv"))
  
} else {
  
  models <- read_rds(here("data/derived/models/TOC_transfer/all_basic_CLP_for_transfer.rds"))
  perf <- read_csv(here("data/derived/models/TOC_transfer/all_basic_CLP_for_transfer_performance.csv"))
  
}
```

```{r}
ggplot(perf) +
  geom_histogram(aes(x = rmse_val, fill = "Val")) +
  geom_histogram(aes(x = rmse_train, fill = "Train")) +
  labs(title = paste("Val & Train RMSE by Grid ID"),
       x = "RMSE", y = "density", fill = "Group") +
  theme_minimal()

best_rows  <- perf[order(perf$diff), ][1:50,]
best_rows

```

Since we aren't going to use an explicit testing set here, we'll grab the model
with the least difference between training and validation, assuming that it 
represents the most balanced model in training.

```{r}
if (!file.exists(here("data/derived/models/TOC_transfer/best_basic_CLP_for_transfer.RDS"))) {
  
  best_model <- best_rows$grid_id[1]
  
  xgb_model <- models[[best_model]]
  
  write_rds(xgb_model, here("data/derived/models/TOC_transfer/best_basic_CLP_for_transfer.RDS"))
  
} else {
  
  xgb_model <- read_rds(here("data/derived/models/TOC_transfer/best_basic_CLP_for_transfer.RDS"))
  
}

xgb_model

```

## Naive Predict on Yampa

Using the CLP model, we'll do a blind application of that to the Yampa data just
to get an idea of general performance.

```{r}
# set up data matrix for xgb
xgb_yampa <- xgb.DMatrix(
  data = as.matrix(yampa[, features]),
  label = yampa[[target_col]]
)

yampa$naive_pred <- predict(xgb_model, xgb_yampa)
yampa %>% 
  summarise(rmse = rmse(TOC, naive_pred),
            .by = id)
rmse(yampa$TOC, yampa$naive_pred)

ggplot(yampa, aes(x = TOC, y = naive_pred)) +
  geom_point()+
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red")

```

Since there is just one year of yampa data, and I don't know much about the WS, 
I'm just going to do a random split of the sites for training (4 sites)/validation (2 sites)/test (2 sites)

```{r}
if (!file.exists(here("data/derived/models/TOC_transfer/yampa_simple_TOC_transfer_tvt.csv"))) {
  
  yampa_ids <- unique(yampa$id)
  train_yamp <- sample(yampa_ids, 4)
  val_yamp <- sample(yampa_ids[!yampa_ids %in% train_yamp], 2)
  test_yamp <- yampa_ids[!yampa_ids %in% c(train_yamp, val_yamp)]
  
  yampa_train <- yampa %>% filter(id %in% train_yamp)
  yampa_val <- yampa %>% filter(id %in% val_yamp)
  yampa_test <- yampa %>% filter(id %in% test_yamp)
  
  yampa_tvt <- reduce(list(yampa_train %>% mutate(group = "train"),
                           yampa_val %>% mutate(group = "val"),
                           yampa_test %>% mutate(group = "test")),
                      full_join)
  
  write_csv(yampa_tvt, here("data/derived/models/TOC_transfer/yampa_simple_TOC_transfer_tvt.csv")) 
  
} else {
  
  yampa_tvt <- read_csv(here("data/derived/models/TOC_transfer/yampa_simple_TOC_transfer_tvt.csv"))
  
  yampa_train <- yampa_tvt %>% filter(group == "train") %>% select(-group)
  yampa_val <- yampa_tvt %>% filter(group == "val") %>% select(-group)
  yampa_test <- yampa_tvt %>% filter(group == "test") %>% select(-group)
  
}

yampa_t_w <- toc_high_weights(yampa_train)
yampa_v_w <- toc_high_weights(yampa_val)

xgb_yampa_t <- xgb.DMatrix(
  data = as.matrix(yampa_train[, features]),
  label = yampa_train[[target_col]],
  weight = yampa_t_w
)

xgb_yampa_v <- xgb.DMatrix(
  data = as.matrix(yampa_val[, features]),
  label = yampa_val[[target_col]],
  weight = yampa_v_w
)

yampa_watch <- list(train = xgb_yampa_t, 
                    eval = xgb_yampa_v)
```

## Sequential training

In sequential training there are some things we **don't** want to change (like
max depth, eta, and some of the other regularization parameters), so this tune
grid is pretty small.

```{r}

if (!file.exists(here("data/derived/models/TOC_transfer/best_yampa_transfer_from_CLP.RDS"))) {
  
  models_yampa <- list()
  seq_perf <- data.frame()
  
  seq_params <- list(
    max_depth = xgb_model$params$max_depth,
    eta = xgb_model$params$eta,
    gamma = xgb_model$params$gamma,
    colsample_bytree = xgb_model$params$colsample_bytree,
    min_child_weight = xgb_model$params$min_child_weight,
    subsample = xgb_model$params$subsample
  )
  
  # Only vary REGULARIZATION parameters (lambda and alpha)
  # These can be changed because they don't affect tree structure
  warmstart_tune_grid <- expand.grid(
    lambda = c(0.5, 1, 2, 5), # L2 regularization
    alpha = c(0, 0.5, 1, 2)# L1 regularization
  )
  
  #run through all hyper parameters and save to fold_models with name foldi_gridj
  for (j in 1:nrow(warmstart_tune_grid)) {
    
    # Combine fixed structure with varying regularization
    params <- c(
      list(
        objective = "reg:squarederror",
        eval_metric = "rmse"
      ),
      seq_params,  # Fixed structural params
      list(
        alpha = warmstart_tune_grid$alpha[j],   # Varying
        lambda = warmstart_tune_grid$lambda[j]  # Varying
      )
    )  
    
    model_ij <- xgb.train(
      params = params,
      data = xgb_yampa_t,
      nrounds = 10000,  
      watchlist = yampa_watch,
      early_stopping_rounds = 2000,
      verbose = 0,
      nthread = 1,
      xgb_model = xgb_model  # Warm start from base model
    )
    
    models_yampa[[j]] <- model_ij
    
    val_rmse <- as.numeric(sub(".*eval-rmse:([0-9.]+).*", "\\1", model_ij$best_msg))
    train_rmse <- as.numeric(sub(".*train-rmse:([0-9.]+).*", "\\1", model_ij$best_msg))
    
    seq_perf <- rbind(seq_perf, data.frame(
      grid_id = j,
      lambda = warmstart_tune_grid$lambda[j],
      alpha = warmstart_tune_grid$alpha[j],
      best_iter = model_ij$best_iteration,
      rmse_val = val_rmse,
      rmse_train = train_rmse,
      diff = abs(val_rmse - train_rmse)
    ))
    
  }
  
  ggplot(seq_perf) +
    geom_histogram(aes(x = rmse_val, fill = "Val")) +
    geom_histogram(aes(x = rmse_train, fill = "Train")) +
    labs(title = paste("Val & Train RMSE by Grid ID"),
         x = "RMSE", y = "density", fill = "Group") +
    theme_minimal()
  
  seq_perf <- seq_perf[order(seq_perf$diff), ]
  
  write_rds(models_yampa, here("data/derived/models/TOC_transfer/yampa_transfer_models_from_CLP.rds"))
  write_csv(seq_perf, here("data/derived/models/TOC_transfer/performance_yampa_transfer_from_CLP.csv"))
  
  best_seq_model <- seq_perf$grid_id[1]
  seq_model <- models_yampa[[best_seq_model]]
  
  write_rds(seq_model, here("data/derived/models/TOC_transfer/best_yampa_transfer_from_CLP.RDS"))
  
} else {
  
  seq_model <- read_rds(here("data/derived/models/TOC_transfer/best_yampa_transfer_from_CLP.RDS"))
  
}

```

And now apply to the test set.

```{r}

xgb_yampa_test <- xgb.DMatrix(
  data = as.matrix(yampa_test[, features]),
  label = yampa_test[[target_col]]
)

yampa_test$naive_pred <- predict(xgb_model, xgb_yampa_test)
yampa_test$seq_pred <- predict(seq_model, xgb_yampa_test)
rmse(yampa_test$TOC, yampa_test$naive_pred)
rmse(yampa_test$TOC, yampa_test$seq_pred)

ggplot(yampa_test) +
  geom_point(aes(x = TOC, y = naive_pred, color = "naive")) +
  geom_point(aes(x = TOC, y = seq_pred, color = "sequential")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_bw() +
  scale_color_manual(values = c("sequential" = "black",
                                "naive" = "grey")) +
  labs(x = "measured TOC (mg/L)",
       y = "predicted TOC (mg/L)",
       color = "transfer method") +
  coord_cartesian(xlim = c(3, 9),
                  ylim = c(3, 9)) +
  theme(legend.position = c(0.05, 0.95),      # inside plot (x, y)
        legend.justification = c(0, 1),       # anchor top-left
        legend.background = element_rect(
          fill = "white",
          color = "black"))

```

