---
title: "Categorical TOC Predictions"
author: "Sam Struthers-ROSSyndicate"
date: "2025-11-01"
output: html_document
---

# Purpose

We are testing the development of a categorical TOC model for the UCLP DSS using sensor data and grab samples from the our and Fort Collins Cache la Poudre campaigns. This model will classify TOC into four categories: low (0-2 mg/L), medium (2-4 mg/L), high (4-8 mg/L), and extreme (>8 mg/L). These categories correspond to the TOC and alkalinity matrix that treatment operators much use to assess TOC removal rates. The goal is to apply this to live sensor data to provide decision support/heads up to operators. We are hoping that my using a categorical approach we can increase trust in the system and have higher accuracy than in our continuous predictions.

General workflow:

1.   Load in datasets created in previous scripts (01/02)

2.   Set up training/validation and testing datasets

3.  Create folds for cross-validation

4.   Do feature testing to see which features improve model performance

5.  Tune hyperparameters and weights

6.   Assess model performance on validation

    1.  Repeat steps 3-6 to attempt to improve performance on validation

    2.  Save model when working version is finalized

7.  Apply model to testing set

8.  Apply model to timeseries space & visualize

# Set Up

## Package Load

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, out.width="100%")
source("src/setup_libraries.R")
source("src/setup_ross_theme.R")

#these are just packages for testing within this script
library(gridExtra)
library(grid)
library(Ckmeans.1d.dp)
library(DiagrammeR)
library(Metrics)
library(glue)
library(SHAPforxgboost)
library(fastshap)
library(shapviz)
set.seed(123)
```

## Load Functions

```{r, include=FALSE}
# function to apply sensor transformations for TOC modeling
source("src/modeling/apply_sensor_transformations_toc.R")
#function to apply training scale
source("src/modeling/apply_training_scale.R")
# function to tune xgboost hyperparameters
source("src/modeling/xgboost_hyperparameter_tuning_cat.R")
#function to plot performance of each fold
source("src/plotting/plot_tvt_fold_acc.R")
#function to plot performance of each fold
source("src/plotting/plot_train_test_ensemble_acc.R")
#load grab sample data function
source("src/collation/load_grab_sample_data.R")
#function to plot model timeseries
source("src/plotting/plot_cat_model_timeseries.R")

```

## Set Version

```{r}
#set version for saving (update when changes are significant)
version = "20251105"
```

## Read in sensor dataset

Here, we read in the sensor data, apply min-max scaling defined in the `apply_sensor_transformations_toc()` function, and store the dataset for use later

```{r}
# Read in all data to create scaling parameters
wide_data_file <- list.files(here("data","collated", "sensor"), pattern = "all_sensor_data_.*\\.parquet", full.names = TRUE)%>%
  tail(1)

# Use the complete sensor dataset for scaling parameters (assuming sensor has all necessary columns already)
complete_sensor_data <- read_parquet(wide_data_file) %>%
  apply_sensor_transformations_toc(., dt_col = "DT_hourly_round")%>%
  select(-DO, -Depth, -ORP, -pH) #remove unused sensors


```

## Read in matchup dataset

```{r}

matchup_dataset <- list.files(here("data","derived", "matchup_datasets"), pattern = "parsed_toc_sensor_clp_", full.names = TRUE)%>%
  tail(1)

# Read in all matches data
all_matches <- read_csv(file = matchup_dataset, show_col_types = F) %>%
  apply_sensor_transformations_toc(dt_col = "sensor_datetime")%>%
  #remove NAs
  filter(!is.na(TOC))%>%
  mutate(clean_sensor_datetime = parse_date_time(sensor_datetime, orders = c("%Y-%m-%d %H:%M:%S", "%m/%d/%y %H:%M","%Y-%m-%d" )))%>%
  mutate(sensor_datetime = force_tz(clean_sensor_datetime, tzone = 'MST'), 
         month = month(sensor_datetime),
         #create categorical TOC
         TOC_cat  = cut(TOC, breaks = c(0, 2, 4, 8, 15), labels = c('low', 'medium', 'high',"extreme")),
         TOC_cat = as.numeric(as.factor(TOC_cat))-1
  )

```

## Define ID Columns, Target, and Features

```{r}
# ID columns to keep to use later on 
id_cols <- c("Date", "id", "site", "DT_hourly_round", "sensor_datetime", "lab_datetime", "collector", "data_avail")
# What we want to predict
target = "TOC_cat"
# Get feature columns from complete sensor data but remove id cols and target
features <- setdiff(colnames(complete_sensor_data), id_cols)
features <- features[features %nin% target]
```

## Create Scaling Parameters & Normalize Data

Normalization is applied based on min-max scaling from the complete sensor dataset using the `apply_training_scale()` function.

```{r}
# Apply scaling to dataset
all_matches_norm <- all_matches %>%
  # Only keep columns that would be in the complete sensor dataset
  select(any_of(id_cols), any_of(names(complete_sensor_data)), any_of(target))%>%
  # Apply normalization 
  apply_training_scale(
    new_data = .,
    features = features,
    full_dataset = complete_sensor_data
  ) %>%
  na.omit()

```

## Create Holdout Dataset

For this model version, we will create a holdout testing set from specific sites (chd, sfm, pbr, pman, pbd) in 2025, and use all other data for training/validation. We will also trim out any known bad data or sites we don't want to include in modeling.

```{r}
# Trim out any known bad data or sites we don't want to include in modeling
all_matches_trimmed <- all_matches_norm%>%
  filter(!(id == "cbri" & between(sensor_datetime, ymd("2024-10-08"), ymd("2024-10-10"))))%>% #exclude a cbri sample taken during release
  filter(!(id == "pman" & between(sensor_datetime, ymd("2025-05-28"), ymd("2025-06-01"))))%>% #exclude a pman sample with bad sensor data
# These are not applicable sites for modeling Upper CLP streams and will just introduce error
  filter( id %nin% c("springcreek", "boxcreek"))
# Create testing set from 2025 data at specific sites
testing <- all_matches_trimmed %>%
  filter((year(sensor_datetime) == 2025 & id %in% c("chd", "sfm", "pbr", "pman", "pbd")))
# Create training/validation set
train_val <- all_matches_trimmed %>%
  anti_join(testing)%>%
  select(id, any_of(features), TOC_cat, sensor_datetime, collector)%>%
  mutate(id = if_else(id == "archery_virridy", "archery", id)) #combine archery and archery_virridy since they are the same site

```

## Set up model folds

Since we have relatively little data, we are going to use three CV folds. Create 3 separate folds and 3 models, with one fold used for validation on each model. Generally we are looking for: - Each fold to have \~30-35% of the total data in validation - Each validation set to have a similar TOC distribution as its training set (median, range) - Each should have a mix of Upper Mainstem (SFM, PFAL, PBD), Upper Trib (CBRI, CHD, LBEA, PENN) & Lower Mainstem (salyer, udall, riverbend, elc, archery, riverbluffs)

```{r}
val_sets <- list(
  
  val_set_1 <- c( "sfm","chd","lbea","riverbluffs" ),
  
  val_set_2 <- c( "pbd","joei", "penn", "archery" ),
  
  val_set_3 <- c("pfal","udall","cbri","elc","salyer","riverbend")
)
# Function to create plots for a given fold
imap(val_sets,function(val_set_name, fold_id){
  
  val_set <- train_val %>% filter(id %in% val_set_name)
  train_set <- train_val %>% filter(id %nin% val_set_name)
  # Add set labels and combine
  combined <- bind_rows(
    train_set %>% mutate(set = "Train"),
    val_set %>% mutate(set = "Validation")
  )
  # Compute counts and proportions by TOC_cat and set
  heat_data <- combined %>%
    group_by(set, TOC_cat) %>%
    summarise(n = n(), .groups = "drop_last") %>%
    mutate(prop = n / sum(n))%>%
    mutate(
      TOC_cat_range = case_when(
        TOC_cat == 0 ~ "0-2",
        TOC_cat == 1 ~ "2-4",
        TOC_cat == 2 ~ "4-8",
        TOC_cat == 3 ~ "8+"
      ))
  # Create heat map
  ggplot(heat_data, aes(x = factor(TOC_cat_range), y = set, fill = prop)) +
    geom_tile(color = "white") +
    geom_text(aes(label = paste0(round(prop * 100, 1), "%")), color = "black", size = 4) +
    scale_fill_gradient(low = "#F5F5F5", high = "#256BF5") +
    labs(
      title = paste0("Fold ", fold_id, " TOC Category Distribution"),
      subtitle = paste0("Validation sites: ", paste(unique(val_set$id), collapse = ", ")),
      x = "TOC (mg/L) Category",
      y = "",
      fill = "Proportion"
    ) +
    theme_bw(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      panel.grid = element_blank()
    )
})%>%
  # plot all folds
  ggarrange(plotlist = .,  ncol = 1,common.legend = T, legend = "right")


# Create a lookup table mapping each id to its fold
fold_lookup <- imap_dfr(val_sets, ~
                          tibble(id = .x, fold_id = as.integer(gsub("\\D", "", .y))))

# Assign folds automatically using join
train_val <- train_val %>%
  left_join(fold_lookup, by = "id")

```

## Visualize TOC distribution in training/validation vs testing sets

Currently, our 2025 data (testing) is skewed to higher TOC values (April- July) compared to the training/validation set. This is not ideal, but we will proceed with this split for now given limited data. Future data updates in December should help even this out as TOC decreases in the fall.

```{r}
heat_data <- train_val %>%
  mutate(group = "Train/Val") %>%
  bind_rows(
    testing %>%
      mutate(group = "Testing")
  ) %>%
  group_by(group, TOC_cat) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(prop = n / sum(n))%>%
  mutate(
    TOC_cat_range = case_when(
      TOC_cat == 0 ~ "0-2",
      TOC_cat == 1 ~ "2-4",
      TOC_cat == 2 ~ "4-8",
      TOC_cat == 3 ~ "8+"
    ))

# Plot as heat map
ggplot(heat_data, aes(x = factor(TOC_cat_range), y = group, fill = prop)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(round(prop * 100, 1), "%")), color = "black", size = 3.5) +
  scale_fill_gradient(low = "#F5F5F5", high = "#256BF5") +
  labs(
    title = paste0(
      "Training/Validation Set: ", nrow(train_val), " samples              ",
      "Testing Set: ", nrow(testing), " samples"
    ),
    x = "TOC (mg/L) Category",
    y = "Model Group",
    fill = "Proportion"
  ) +
  ROSS_theme +
  theme(
    legend.position = "bottom",
    panel.grid = element_blank()
  )

```

# Model Training

For our model training we will first do some feature tuning to see which features are most useful for predicting TOC. We will then do hyperparameter tuning on the selected features to optimize model performance.

## Feature Tuning

### View correlations between features and target

Target: (TOC and by extension TOC_cat)

Most products with FDOM in them are going to have higher correlation with TOC, as well as seasonal (sin_doy) and flow related features. We want to be sure that we don't put too many of these in the model as calibration error or sensor drift could lead to compounding errors.

```{r out.width="100%"}

# Combine training/validation and testing data
modeling_data <- train_val %>%
  mutate(group = "Train/Val") %>%
  bind_rows(
    testing %>%
      mutate(group = "Testing")
  ) %>%
  #adding in the raw TOC values for correlation plotting
  left_join(all_matches%>%select(TOC, id, sensor_datetime, collector)%>%
              mutate(id = if_else(id %in% c("archery", "archery_virridy"), "archery", id)),
             by = c("id", "sensor_datetime", "collector"))%>%
  mutate(
    watershed = ifelse(
      id %in% c("joei", "cbri", "chd", "pfal", "pbr", "sfm", "pman", "pbd"),
      "UpperPoudre", 
      "LowerPoudre"
    )
  )
map(features, function(feat) {

  cor <- round(cor(modeling_data[[feat]], modeling_data$TOC, use = "complete.obs"), 2)
  
  ggplot(modeling_data, aes(x = .data[[feat]], y = TOC, shape = collector, color = watershed)) +
    geom_point() +
    annotate(
      "text",
      x = mean(range(modeling_data[[feat]], na.rm = TRUE)), # middle of x range
      y = max(modeling_data$TOC, na.rm = TRUE) * 0.95,      # 95% of y max (near top)
      label = paste0("Corr: ", round(cor(modeling_data[[feat]], modeling_data$TOC, use = "complete.obs"), 2)),
      hjust = 0.5,
      vjust = 1,
      size = 4
    ) +
    labs(x = feat, y = "TOC (mg/L)", shape = "Collector", color = "Watershed") +
    #facet_wrap(~watershed, scales = "free") +
    ROSS_theme
})%>%
  wrap_plots(., guides = "collect")

```

### Pruning Features with high correlations

We want to avoid using too many highly correlated features in the model, as this can lead to overfitting and instability. We will look at the correlation matrix and remove some of the more redundant features.

```{r}
# get a correlation matrix across all features
corr_matrix <- modeling_data %>% 
  select(all_of(features)) %>% 
  cor(use = "pairwise.complete.obs")

# we just want half the matrix, so recode those in the lower triangle to NA
corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter for those with correlation > 0.9
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.9) %>%
  arrange(desc(abs(correlation))) %>% 
  # remove redundant vars
  filter(feature1 %nin% c("FDOM", "doy"))

# look at the list and make some educated decisions about features
high_corr_pairs

fewer_features <- features[features %nin% c("FDOM", "doy", 
                                            # remove vars with high correlation 
                                            # with others
                                            "f_log", "temp_mean_48hr", "Temp",
                                            #Removing some turb features that will lead to over reliance on turb/fDOM relationship
                                            "turb_f", "f_turb", "turb_median_12hr",
                                            #Daily flow is more helpful gage of stream flow since sites are not all within 1 hour of canyon
                                            "hourly_canyon_flow_cfs",
                                            "fdom_max_6hr", "temp_max_24hr")]

# do the correlation matrix again to make sure that is okay
corr_matrix <- modeling_data %>% 
  select(all_of(fewer_features)) %>% 
  cor(use = "pairwise.complete.obs")

corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.85) %>%
  arrange(desc(abs(correlation)))

high_corr_pairs

```

`high_corr_pairs` has features we should be cautious of using together. We will try to avoid using more than one of these in the same model.

### Testing Features

First, we will start with our base model (live sensor data only) and then test adding in other features to see if they improve performance. We will use the default hyperparameters below for all models to keep things consistent. Once we have tested all the features together, we will want to trim it down a bit so we are not using too many highly correlated features or overfitting the model due to lack of data

#### Function to test features & defaults

This just runs a basic xgboost model with the specified features and default hyperparameters. It returns feature importance, shap values, and best iteration info for each fold.

```{r}

 test_features <- function(train_val,
                          features_to_test,
                          target,
                          default_hyper_params,
                          weight_fun = NULL) {
  
  # double check that there are no duplicates in features
  features_to_test <- unique(features_to_test)
  
  #create blank dataframes to store importance/shap values
  importance_df <- tibble(Feature = features_to_test)
  shap_val_df <- tibble(Feature = features_to_test)
  best_msgs <- tibble()
  models <- list()
  
  for(i in unique(train_val$fold_id)){
    
    #train val split
    train_data <- train_val %>%
      filter(fold_id != i) %>%
      select(any_of(c(features_to_test, target, "id", "sensor_datetime", "collector")))
    
    val_data <- suppressMessages(
      train_val %>%
        anti_join(train_data)
     )
    
    #weight data or don't if no function provided
    if(is.null(weight_fun)){
      weight_fun <- function(df) {rep(1, nrow(df))}
    }
    w_train <- weight_fun(train_data)
    w_val   <- weight_fun(val_data)
    
    #prep matrices
    dtrain <- xgb.DMatrix(
      data = as.matrix(train_data[, features_to_test]),
      label = train_data[[target]],
      weight = w_train
    )
    
    dval <- xgb.DMatrix(
      data = as.matrix(val_data[, features_to_test]),
      label = val_data[[target]],
      weight = w_val
    )
    
    watchlist <- list(train = dtrain, eval = dval)
    
    #Train model with defaults
    model <- xgb.train(
      params = default_hyper_params,
      data = dtrain,
      nrounds = 10000,
      watchlist = watchlist,
      early_stopping_rounds = 1000,
      print_every_n = 1000,
      verbose = 0,
      nthread = 1
    )
    #Save model
    models[[i]] <- model
    #Save best msg
    best_msg <- tibble(
      fold_id = i,
      best_iteration = model$best_iteration,
      best_msg = model$best_msg
    ) %>%
      #clean up message
      mutate(
        train_mlogloss = as.numeric(str_extract(best_msg, "(?<=train-mlogloss:)\\d+\\.\\d+")),
        eval_mlogloss  = as.numeric(str_extract(best_msg, "(?<=eval-mlogloss:)\\d+\\.\\d+"))
      )%>%
      select(-best_msg)
    #best message
    best_msgs <- bind_rows(best_msgs, best_msg)%>%
      distinct()
    
    
    #Feature importance
    importance_df <- importance_df%>%
      left_join(xgb.importance(feature_names = features_to_test, model = model)%>%
                  dplyr::rename_with( #fix column names
                    ~ glue::glue("{.x}_fold_{i}"),
                    .cols = c(Gain, Cover, Frequency)
                  ), by = "Feature")
    # # SHAP 
    # function to get predictions
    explainer <- function(object, newdata) {
      preds <- predict(object, as.matrix(newdata))
      
      # Convert to matrix: rows = samples, cols = classes
      preds_mat <- matrix(preds, ncol = default_hyper_params$num_class, byrow = TRUE)
      
      # Get the maximum probability for each sample
      max_probs <- apply(preds_mat, 1, max)
      
      return(max_probs)
    }
  shap_values_train <- fastshap::explain(
      object = model,
      X = as.data.frame(train_data[, features_to_test]),
      pred_wrapper = explainer,
      nsim = 10,
      adjust = TRUE,
      shap_only = FALSE
    )["shapley_values"]%>%
      as.data.frame()%>%
     tibble::rownames_to_column("sample")%>%
       select(-sample)%>%
       pivot_longer(cols = everything(), names_to = "Feature", values_to = "train_shap")%>%
       summarize(train_mean_abs_shap = mean(abs(train_shap)), .by = "Feature")%>%
       mutate(Feature = gsub("shapley_values.", "", Feature))
    
     shap_values_val <- fastshap::explain(
      object = model,
      X = as.data.frame(val_data[, features_to_test]),
      pred_wrapper = explainer,
      nsim = 10,
      adjust = TRUE,
      shap_only = FALSE
    )["shapley_values"]%>%
      as.data.frame()%>%
     tibble::rownames_to_column("sample")%>%
       select(-sample)%>%
       pivot_longer(cols = everything(), names_to = "Feature", values_to = "val_shap")%>%
       summarize(val_mean_abs_shap = mean(abs(val_shap)), .by = "Feature")%>%
       mutate(Feature = gsub("shapley_values.", "", Feature))

    shap_val <- shap_values_val%>%
      left_join(shap_values_train, by = "Feature")%>%
      dplyr::rename_with( #fix column names
        ~ glue::glue("{.x}_fold_{i}"),
        .cols = c(val_mean_abs_shap, train_mean_abs_shap)
      )
    shap_val_df <- shap_val_df%>%
      left_join(shap_val, by = c("Feature"))
  }
  
  #output
  list(
    model = models,
    features = features_to_test, 
    best_msg = best_msgs,
    model_importance = importance_df, 
    shap_values = shap_val_df
  )
  
}

#xgboost hyper parameters

 default_hyper_params <- list(
  objective        = "multi:softprob",
  eval_metric      = "mlogloss",
  eta              = 0.05,
  gamma            = 0.6,
  alpha            = 0,
  lambda           = 1,
  max_depth        = 4,
  subsample        = 0.5,
  colsample_bytree = 0.5,
  min_child_weight  = 2,
  num_class        = length(unique(train_val$TOC_cat)) 
)

#weighting function for TOC categories to emphasize high TOC values
high_weight = function(df) {
  ifelse(df$TOC_cat == 3, 4, # highest TOC category = highest weight
       if_else(df$TOC_cat == 2, 2, 
               1)) #lowest two categories = base weight
}

target = "TOC_cat"
```

#### Run across folds testing different features to see which improve performance

Here I run the function and am going to look at feature importance and shap values to see which features are most useful, then trim features back to reduce our correlative and overfitting issues.

```{r}

#set aside sensor data that are generally useful
general_features <- c("FDOMc","Sensor_Turb","Sensor_Cond", "Chl_a", "Temp")

feature_results <- map(
  #trying all the combinations of features we think might be useful (commented out were tested but found to not improve base model)
  list(
    gen_feat = general_features, # sensor values only
    #fdom_corr_max_6hr = c(general_features, "fdom_corr_max_6hr"), # short term max FDOM, might help with receeding spikes or daily variation
    #f_sc = c(general_features, "fdom_corr_max_6hr", "f_sc"), # adding FDOM/sc ratio, high FDOM + low SC = high TOC (snowmelt)
    #f_x_turb = c(general_features, "fdom_corr_max_6hr", "f_sc", "f_x_turb"), # adding FDOM * turbidity, high FDOM + high turb = high TOC (stormflow/snowmelt), might help with high turb events
    #sin_doy = c(general_features, "fdom_corr_max_6hr", "f_sc", "f_x_turb", "sin_doy"), # seasonal signal but not linked to exact DOY
    #temp_x_f = c(general_features, "fdom_corr_max_6hr", "temp_x_f", "f_sc", "f_x_turb", "sin_doy"), # temp * FDOM, high temp + high FDOM = high TOC (summer/autumn allochthonous inputs)
    #temp_x_turb = c(general_features, "fdom_corr_max_6hr", "temp_x_turb", "temp_x_f", "f_sc", "f_x_turb", "sin_doy"), # temp * turbidity, high temp + high turb = high TOC (summer stormflow?)
    # turb_median_12hr = c(general_features, "fdom_corr_max_6hr", "temp_x_turb", "temp_x_f", "f_sc", "f_x_turb", "sin_doy", "turb_median_12hr"), # receeding storm flow proxy
    #daily_canyon_flow_cfs = c(general_features, "fdom_corr_max_6hr", "temp_x_turb", "temp_x_f", "f_sc", "f_x_turb", "sin_doy", "turb_median_12hr", "daily_canyon_flow_cfs"), # climate (flow) conditions across basin (may want to play with increasing/decreasing slopes?)
    # Tested but did not improve model
    # sc_var_24hr = c(general_features, "sc_var_24hr"), # proxy for recent depth changes (sc = groundwater/surface water contributions)
    #temp_mean_24hr = c(general_features, "temp_mean_24hr"), # recent temp changes
    #temp_mean_48hr = c(general_features, "temp_mean_48hr"), #longer term temp changes
    # f2_temp = c(general_features, "f2_temp"), #FDOM^2 * temp, very high FDOM + low temp = very high TOC 
    #f_x_sc = c(general_features, "f_x_sc"), # FDOM * SC, high FDOM + high SC = low TOC (groundwater influence) probably ended up getting confused with so much variation in SC
    #f_turb = c(general_features, "f_turb"), # FDOM / turbidity, high FDOM + low turb = high TOC (snowmelt in high elev reaches), redundant with f_x_turb
    #turb_f = c(general_features, "turb_f"), # turbidity / FDOM,  redundant with f_x_turb
    fewer_features = c(fewer_features), 
    fewer_features_trimmed = c(fewer_features)%>%setdiff(., c("turb_median_12hr","temp_x_f", "temp_mean_48hr","temp_x_turb", "turb_f", "sc_sd_24hr", "temp_mean_24hr", "temp_f"  )),
        fewest = c(fewer_features)%>%setdiff(., c("turb_median_12hr","temp_x_f", "temp_mean_48hr","temp_x_turb", "turb_f", "sc_sd_24hr", "temp_mean_24hr", "temp_f", "fdom_corr_max_6hr" ))
    
    #Trimming down to most important features from all features (based on importance and shap values) to reduce overfitting
    # trimmed_features = c(general_features, "fdom_corr_max_6hr", "f_sc", "f_x_turb", "sin_doy" , "turb_median_12hr", "daily_canyon_flow_cfs")
  ) , ~test_features(
    train_val = train_val,
    features_to_test = .x,
    target = "TOC_cat",
    default_hyper_params = default_hyper_params
  ))
```

#### Feature Set Results

##### mlogloss

We compare across folds to see if there is big differences in folds (ie unbalanced data) and also compare each feature set against "base" model of general features (sensors). Looking for features that make meaningful improvements in mlogloss across all folds. If our mlogloss improves but that is related to only a single fold, we may want to reevaluate the feature set as it is likely overfitting to a given site(s). Keep in mind that this is weighted mlogloss so it may be higher than unweighted mlogloss you see in other places.

```{r}
# Highlight differences between folds in mlogloss
mlogloss_table <- map_dfr(feature_results, ~{
  # Compute feature set description
  unique_features <- setdiff(.x$features, general_features)
  final_features <- c("general_features", unique_features)
  feature_desc <- paste(final_features, collapse = ", ")
  
  # Take best_msg tibble and rename eval_mlogloss by fold
  .x$best_msg %>%
    select(fold_id, eval_mlogloss) %>%
    mutate(fold_col = paste0("mlogloss_fold_", fold_id)) %>%
    select(-fold_id) %>%
    mutate(eval_mlogloss = round(eval_mlogloss, 2)) %>%
    pivot_wider(names_from = fold_col, values_from = eval_mlogloss) %>%
    mutate(
      features = feature_desc,
      mean_mlogloss = round(mean(c_across(starts_with("mlogloss_fold_")), na.rm = TRUE), 2)
    )%>%
    select(features, mean_mlogloss, mlogloss_fold_1, mlogloss_fold_2, mlogloss_fold_3)
  
}, .id = "feature_set")%>%
  
  #check if new feature is better than baseline
  mutate(
    baseline_mlogloss = mean_mlogloss[feature_set == "gen_feat"],
    improves_mlogloss = if_else(mean_mlogloss < baseline_mlogloss, round(mean_mlogloss - baseline_mlogloss, 2), 0)
  ) %>%
  select(-baseline_mlogloss)%>%
  arrange(improves_mlogloss)%>%
  
  print(mlogloss_table, n = Inf, width = Inf)
```

##### Feature Importance

This comes directly from the xgboost package and is based on Gain, Cover, and Frequency of splits. We summarize across folds to see which features are most important across all folds.

```{r}
#table with all importances
importance_tables <- map(feature_results, ~{
  imp <- .x$model_importance
  
  # Identify fold columns dynamically
  gain_cols <- grep("^Gain_fold_", names(imp), value = TRUE)
  cover_cols <- grep("^Cover_fold_", names(imp), value = TRUE)
  freq_cols <- grep("^Frequency_fold_", names(imp), value = TRUE)
  
  imp %>%
    mutate(
      mean_gain = rowMeans(across(all_of(gain_cols)), na.rm = TRUE),
      mean_cover = rowMeans(across(all_of(cover_cols)), na.rm = TRUE),
      mean_frequency = rowMeans(across(all_of(freq_cols)), na.rm = TRUE)
    ) %>%
    mutate(new_feat = ifelse(Feature %in% general_features, "", "*")) %>%
    select(Feature, new_feat, mean_gain, mean_cover, mean_frequency) %>%
    
    arrange(desc(mean_gain))
}, .id = "feature_set")

importance_tables
#Get features with less than 0.02 gain on average across folds in fewer_features to consider removing
cat("Features with mean gain < 0.02 in 'fewer_features':\n")
importance_tables$fewest%>%
  filter(mean_gain < 0.02)

```

##### Shapley Values

Skipping for now!

```{r}
# #table with all shap values
shap_summary <- map_dfr(feature_results, ~{
  shap <- .x$shap_values

  val_shap_cols <- grep("^val_mean_abs_shap_fold_", names(shap), value = TRUE)
  train_shap_cols <- grep("^train_mean_abs_shap_fold_", names(shap), value = TRUE)

  shap %>%
    mutate(
      mean_val_shap = round(rowMeans(across(all_of(val_shap_cols)), na.rm = TRUE), 3),
      mean_train_shap = round(rowMeans(across(all_of(train_shap_cols)), na.rm = TRUE), 3)
    ) %>%
    mutate(val_train_diff = mean_val_shap - mean_train_shap) %>%
    select(Feature, mean_val_shap, mean_train_shap, val_train_diff)
}, .id = "feature_set")

shap_long <- shap_summary %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "type",
    values_to = "mean_shap"
  ) %>%
  mutate(
    type = recode(type,
                  mean_val_shap = "Validation",
                  mean_train_shap = "Training"),
    feature_set = factor(feature_set, levels = unique(feature_set))
  ) %>%
  mutate(Feature = factor(Feature, levels = c( rev(feature_results$fewer_features$features), "Temp")))

ggplot(shap_long, aes(x = feature_set, y = Feature, fill = mean_shap)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_viridis_c(
    option = "magma",
    limits = c(0, max(shap_long$mean_shap, na.rm = TRUE)),
    name = "Mean |SHAP|",
    direction = -1  # reverses so dark = low, bright = high (optional)
  )+
  facet_wrap(~type) +
  labs(
    x = "Feature Set",
    y = "Feature",
    title = "Mean SHAP Values by Feature and Feature Set"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )

```

### TV performance of best feature set

After trimming down a couple of features (minimal improvement/low shap values), we should take a quick look at model performance in TV space with our base model (default hyperparameters)

```{r out.width="100%"}

best_models <- feature_results$fewest$model
ft_tv_plot <- map(1:3, ~plot_tvt_fold_acc(fold_model = best_models[[.x]],
                                         fold_num = .x,
                                         train_df = train_val %>% filter(fold_id != .x),
                                         val_df = train_val %>% filter(fold_id == .x),
                                         title = "Default Hyperparameters ",
                                         target_col = "TOC_cat"))%>%
   unlist()%>%
  wrap_plots(., ncol = 2, guides = "collect") &
  theme(legend.position = "right")

 ft_tv_plot

```

Fold one and two appear to be performing the worst on the validation set (lower kappa & accuracy). Fold one is primarily made up of classes 2-4, 4-8 and appears to be getting a bit confused on which to predict. Fold two has the JOEI samples so it is not doing great on the low end of TOC there. We will see if hyperparameter tuning can help improve this.

## Hyperparameter Tuning

Now that we have a feature set, we can do some hyperparameter tuning to see if we can improve model performance. This will take a while to run, so we will save the results and load them in later.

### Run hyperparameter tuning

```{r}

features <- best_models[[1]]$feature_names

# Reduce tune grid for reruns (ie. General best params)
tune_grid = expand.grid(
  objective = "multi:softprob",
  eval_metric = "mlogloss",
  nrounds = 10000,
  max_depth = c(2, 3, 4),
  eta = c( 0.01, 0.05, 0.1),
  gamma = c(0.8),
  lambda = c(1, 10),
  alpha = c(0, 1),
  colsample_bytree = c(0.8),
  min_child_weight = c(2, 4),
  subsample = c(0.5, 0.8),
  num_class = length(unique(train_val$TOC_cat))
)

high_weight = function(df) {
  ifelse(df$TOC_cat == 3, 4, # highest TOC category = highest weight
       if_else(df$TOC_cat == 2, 2, 
               1)) #lowest two categories = base weight
}
set.seed(123)

hyperp_results <- xgboost_hyperparameter_tuning_cat(
  data = train_val %>% select(TOC_cat, id, fold_id, any_of(features)), 
  weights = high_weight,
  tune_grid = tune_grid,
  target_col = "TOC_cat",
  site_col = "id",
  units = "mg/L", 
  class_levels = c("0-2", "2-4", "4-8", "8+"))

```

### Choosing best grids

### Fold 1

```{r}

hyperp_results[[1]]$perf_dist
hyperp_results[[1]]$eval_plot
hyperp_results[[1]]$tv_plot

best_fold_1_grid <- 96
```

### Fold 2

```{r}
hyperp_results[[2]]$perf_dist
hyperp_results[[2]]$eval_plot
hyperp_results[[2]]$tv_plot

best_fold_2_grid <- 73

```

### Fold 3

```{r}
hyperp_results[[3]]$perf_dist
hyperp_results[[3]]$eval_plot
hyperp_results[[3]]$tv_plot

best_fold_3_grid <- 1

```

## Extract Best Models

```{r}

full_model <- map(1:3, function(i) {
  
  best_grid <- paste0("grid_", get(glue("best_fold_{i}_grid")))
  model_info <- hyperp_results[[i]]$grid_results[[best_grid]]
  list(
    model = model_info$model,
    perf = model_info$perf,
    params = model_info$params,
    training_data = train_val %>% filter(fold_id != i),
    validation_data = train_val %>% filter(fold_id == i)
  )
})

```

### Checking Results

#### Feature Importance

```{r}
importance_list <- lapply(seq_along(full_model), function(i) {
  fold_model <- full_model[[i]]$model
  features <- fold_model$feature_names
  importance_matrix <- xgb.importance(feature_names = features, model = fold_model)
  importance_df <- as.data.frame(importance_matrix)
  importance_df$Fold <- paste("Fold", i)
  return(importance_df)
})

print(importance_list)
```

#### Shapley Importance

```{r}

# # Function to plot SHAP values for a given fold
plot_shap <- function(fold_set, train_data, val_data, fold_model) {
  #get features from model
  features <- fold_model$feature_names
  # Predictions on validation/training set
  explainer <- function(object, newdata) {
    preds <- predict(object, as.matrix(newdata))
    
    # Convert to matrix: rows = samples, cols = classes
    preds_mat <- matrix(preds, ncol = fold_model$params$num_class, byrow = TRUE)
    
    # Get the maximum probability for each sample
    max_probs <- apply(preds_mat, 1, max)
    
    return(max_probs)
  }
  
  shap_values_train <- fastshap::explain(
    object = fold_model ,
    X = as.data.frame(train_data[, features]),
    pred_wrapper = explainer,
    nsim = 10,
    adjust = TRUE,
    shap_only = FALSE
  )
  sv <- shapviz(shap_values_train)
  train_plot <- sv_importance(sv, kind = "beeswarm")
  
  shap_values_val <- fastshap::explain(
    object = fold_model,
    X = as.data.frame(val_data[, features]),
    pred_wrapper = explainer,
    nsim = 10,
    adjust = TRUE,
    shap_only = FALSE
  )
  
  sv <- shapviz(shap_values_val)
  val_plot <- sv_importance(sv, kind = "beeswarm")
  
  train_plot + val_plot 
}
# 
map(1:3, ~ plot_shap(
  fold_set = .x,
  fold_model = full_model[[.x]]$model,
  train_data = full_model[[.x]]$training_data, 
  val_data = full_model[[.x]]$validation_data
))%>%
  wrap_plots(ncol = 1, guides = "collect")+
  plot_annotation(
    title = paste("SHAP Values"),
    subtitle = "Left: Training Set | Right: Validation Set"
  )

```

### Plots of TV Performance

Now let's look at our performance in TV space.

```{r}

hyperparam_tv_plots <- map(1:3, ~plot_tvt_fold_acc(train_df = full_model[[.x]]$training_data,
                                                  val_df = full_model[[.x]]$validation_data,
                                                  fold_model = full_model[[.x]]$model, # a list where each element is a model for that fold  
                                                  fold_num = .x,
                                                  title = "",
                                                  target_col = "TOC_cat", units = "mg/L")
 )%>%
  unlist()

hp_tv_plot<- wrap_plots(hyperparam_tv_plots, ncol = 2, guides = "collect")&
  theme(legend.position = "bottom")
hp_tv_plot
#ggsave("data/upper_clp_dss/figures/toc_modeling/performance/toc_model_fold_validation_performance.png", width = 14, height = 10)

```

#### Against base hyperparameters

We should hope that all our tuning efforts have improved the model!!

```{r}
# Compare best hyperparameter model to default hyperparameter model
final_plot <- wrap_plots(
  ft_tv_plot,
  hp_tv_plot,
  ncol = 2,
  guides = "collect"
) &
  theme(legend.position = "bottom")
final_plot <- final_plot + 
  plot_annotation(
    title = "Model Performance Comparison: Default (left) vs Tuned Hyperparameters (right)"
  )

final_plot

```

Looking at the plots, we can see that the tuned hyperparameter model has improved performance in all folds compared to the default hyperparameter model. The validation accuracy has increased in all folds but it does look like model fold two is slightly less accurate than other folds (possible issue with distribution). Fold three's training accuracy decreased but we did see improvement in the validation set and the highest category was actually guessed correctly, which is more important! Overall, this is a good sign that our tuning efforts have paid off!

## Save as "in_dev" and remove unneeded objects

If we want to test additional ideas for bias correction or look at model performance on timeseries data, we can just run this script until `model_training` and then load in the "in_dev" model file to avoid having to rerun all the tuning again.

```{r}
#commented out so we don't overwrite existing model files
# saveRDS(full_model, file = here("data", "derived", "models","TOC_realtime", "in_dev", paste0("toc_categorical_xgboost_models_",version,".rds")))
#if we want to reload model
#full_model <- readRDS(here("data", "derived", "models","TOC_realtime", "in_dev", paste0("toc_categorical_xgboost_models_",version,".rds")))

rm(list = setdiff(ls(), c("train_val", "testing", "full_model","version","id_cols",
                          "complete_sensor_data",
                          #keeping functions around
                          "apply_training_scale","plot_tvt_fold_acc","plot_train_test_ensemble_acc",
                          "load_grab_sample_data", "plot_cat_model_timeseries")))
```

# Evaluate on Test Set

We should only do this once we feel good about our model and are ready to see how it performs on the unseen test set. This should only be run after big changes to the model or if we get new data (ie The model should be in a "final/working" version)

## Visualize Train Val Test

### Individual folds

```{r}
plots <- map(1:3, ~{
  p <- plot_tvt_fold_acc(train_df = full_model[[.x]]$training_data, 
                    val_df = full_model[[.x]]$validation_data, 
                    test_df = testing, 
                    fold_model = full_model[[.x]]$model, 
                    fold_num = .x,
                    target_col = "TOC_cat",
                    units = "mg/L")$test_plot
  
  # ggsave(glue(here("docs", "uclp_dss", "figs", "realtime_toc_modeling", "performance", "ross_only_model_fold_{.x}.png")),
  #      plot = p , width = 10, height = 6, dpi = 500)
  p
})
#Print plots
plots

```

### All folds together

```{r}
# Single plot with all folds
wrap_plots(plots, ncol = 1, guides = "collect") &
  theme(legend.position = "bottom")
# ggsave(here("docs","uclp_dss", "figs", "realtime_toc_modeling","performance",paste0("ross_only_cat_model_all_folds_test", version, ".png")),
#        width = 14, height = 8, dpi = 500)
```

### Ensemble predictions

```{r}
# Plot ensemble
ensemble <- plot_train_test_ensemble_acc(
  train_df = train_val,
  test_df = testing,
  models = map(full_model, "model"),
  target_col = "TOC_cat",
  units = "mg/L",voting_method = "soft",class_levels = c("0-2", "2-4", "4-8", "8+")
)

ensemble_plot <- wrap_plots(ensemble, ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")

ensemble_plot

ggsave(here("docs","uclp_dss", "figs", "realtime_toc_modeling","performance",paste0("ross_only_cat_model_ensemble", version, ".png")),
       plot = ensemble_plot, width = 12, height = 10, dpi = 500)
```

# Timeseries Space Performance

## Prep sensor timeseries & water chem

```{r}
#loading in all sensor data and applying training scaling
sensor_ts <- apply_training_scale(new_data = complete_sensor_data, 
                                  full_dataset = complete_sensor_data, 
                                  features = full_model[[1]]$model$feature_names)%>%
  select(site, DT_round = DT_hourly_round, any_of(full_model[[1]]$model$feature_names))

water_chem <- load_grab_sample_data()%>%
  mutate(TOC_cat = cut(TOC, breaks = c(0, 2, 4, 8, 15), labels = c("0-2", "2-4", "4-8", "8+")), 
         TOC_cat = as.numeric(factor(TOC_cat))-1)
```

## Example 1: CHD 2025

```{r}

chd_plot <- plot_cat_model_timeseries(model_input_df = sensor_ts, 
                                      water_chem_df = water_chem%>%filter(collector == "ROSS"), 
                                      summary_interval = "4 hour",
                                      site_sel = "chd", 
                                      site_title = "Chambers Lake outflow",
                                      model = map(full_model, "model"),
                                       voting_method = "soft",
                                      target_col = "TOC_cat",
                                      units = "mg/L",
                                      start_DT = "2025-05-01 12:00",
                                      end_DT = "2025-07-05 00:00",
                                      subtitle_arg = "CLP Samples Only")

chd_plot

# ggsave(here("docs", "uclp_dss", "figs", "realtime_toc_modeling", "timeseries", paste0("chambers_outflow_20250501_20250705_cat_toc.png")),
#        plot = chd_plot,
#        width = 12,
#        height = 8,
#        dpi = 300)

```

## Example 2: PFAL 2025

```{r}

pfal_plot <- plot_cat_model_timeseries(
  model_input_df = sensor_ts, 
  water_chem_df = water_chem %>% filter(collector == "ROSS"), 
  summary_interval = "4 hour",
  site_sel = "pfal", 
  site_title = "Poudre @ Poudre Falls",
  model = map(full_model, "model"),
  voting_method = "soft",
  target_col = "TOC_cat",
  units = "mg/L",
  start_DT = "2025-04-16 00:00",
  end_DT = "2025-07-10 00:00",
  subtitle_arg = "CLP Samples Only"
)

pfal_plot

# ggsave(
#   here("docs", "uclp_dss", "figs", "realtime_toc_modeling", "timeseries",
#        paste0("pfal_20250416_20250710_cat_toc.png")),
#   plot = pfal_plot, width = 12, height = 8, dpi = 300
# )


```

## Example 3: PBR

```{r}

pbr_plot <- plot_cat_model_timeseries(
  model_input_df = sensor_ts, 
  water_chem_df = water_chem, 
  summary_interval = "4 hour",
  site_sel = "pbr", 
  site_title = "Poudre @ Indian Meadows",
  model = map(full_model, "model"),
  voting_method = "soft",
  target_col = "TOC_cat",
  units = "mg/L",
  start_DT = "2025-05-01 00:00",
  end_DT = "2025-07-08 00:00",
  subtitle_arg = "CLP Samples Only"
)

pbr_plot

# ggsave(
#   here("docs", "uclp_dss", "figs", "realtime_toc_modeling", "timeseries",
#        paste0("pbr_20250501_20250708_cat_toc.png")),
#   plot = pbr_plot, width = 12, height = 8, dpi = 300
# )
```

## Example 4: SFM 2025

```{r}
sfm_plot <- plot_cat_model_timeseries(
  model_input_df = sensor_ts, 
  water_chem_df = water_chem, 
  summary_interval = "4 hour",
  site_sel = "sfm", 
  site_title = "South Fork Poudre",
  model = map(full_model, "model"),
  voting_method = "soft",
  target_col = "TOC_cat",
  units = "mg/L",
  start_DT = "2025-04-24 00:00",
  end_DT = "2025-07-08 00:00",
  subtitle_arg = "CLP Samples Only"
)

sfm_plot

# ggsave(
#   here("docs", "uclp_dss", "figs", "realtime_toc_modeling", "timeseries",
#        paste0("sfm_20250424_20250708_cat_toc.png")),
#   plot = sfm_plot, width = 12, height = 8, dpi = 300
# )

```

## Example 5: PMAN 2025

```{r}
pman_plot <- plot_cat_model_timeseries(
  model_input_df = sensor_ts, 
  water_chem_df = water_chem, 
  summary_interval = "4 hour",
  site_sel = "pman", 
  site_title = "Poudre @ Manner's Bridge",
  model = map(full_model, "model"),
  voting_method = "soft",
  target_col = "TOC_cat",
  units = "mg/L",
  start_DT = "2025-04-24 00:00",
  end_DT = "2025-07-18 00:00",
  subtitle_arg = "CLP Samples Only"
)

pman_plot

# ggsave(
#   here("docs", "uclp_dss", "figs", "realtime_toc_modeling", "timeseries",
#        paste0("pman_20250424_20250718_cat_toc.png")),
#   plot = pman_plot, width = 12, height = 8, dpi = 300
# )

```

## Example 6: PBD 2025

```{r}

pbd_plot <- plot_cat_model_timeseries(model_input_df = sensor_ts, 
                                  water_chem_df = water_chem, 
                                  summary_interval = "4 hour",
                                  site_sel = "pbd", 
                                  site_title = "Poudre at Canyon Mouth",
                                  model = map(full_model, "model"),
                                  voting_method = "soft",
                                  target_col = "TOC_cat",
                                  units = "mg/L",
                                  start_DT = "2025-04-15 00:00",
                                  end_DT = "2025-07-05 00:00",
                                  subtitle_arg = "CLP Samples Only")

pbd_plot
# ggsave(here("docs", "uclp_dss", "figs", "realtime_toc_modeling", "timeseries", paste0("pbd_20250415_20250705_cat_toc.png")),
#        plot = pbd_plot,
#        width = 12,
#        height = 8,
#        dpi = 300)

```

## Example 6: Udall (Lincoln Ave) 2025

```{r}
udall_plot <- plot_cat_model_timeseries(
  model_input_df = sensor_ts, 
  water_chem_df = water_chem %>% filter(collector == "ROSS"), 
  summary_interval = "4 hour",
  site_sel = "udall", 
  site_title = "Poudre at Lincoln Ave",
  model = map(full_model, "model"),
  voting_method = "soft",
  target_col = "TOC_cat",
  units = "mg/L",
  start_DT = "2025-04-15 00:00",
  end_DT = "2025-07-05 00:00",
  subtitle_arg = "CLP Samples Only"
)

udall_plot

# ggsave(
#   here("docs", "uclp_dss", "figs", "realtime_toc_modeling", "timeseries",
#        paste0("udall_20250415_20250705_cat_toc.png")),
#   plot = udall_plot, width = 12, height = 8, dpi = 300
# )
```
