---
title: "Testing TOC preds on Yampa data - ROSS Only Simple Model"
author: " Sam Struthers-ROSSyndicate"
date: "2025-08-01"
output: html_document
---

# Set Up

## Package Load

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("src/setup_libraries.R")
source("src/setup_ross_theme.R")

#these are just packages for testing within this script
library(gridExtra)
library(grid)
library(Ckmeans.1d.dp)
library(DiagrammeR)
library(Metrics)
library(glue)
library(SHAPforxgboost)
set.seed(123)
#set version for saving (update when changes are significant)
version = "20251211"
# function to apply sensor transformations for TOC modeling
source("src/modeling/apply_sensor_transformations_toc.R")
#function to apply training scale
source("src/modeling/apply_training_scale.R")
# function to tune xgboost hyperparameters
source("src/modeling/xgboost_hyperparameter_tuning.R")
#function to plot performance of each fold
source("src/plotting/plot_tv_fold_perf.R")
#function to plot performance of each fold
source("src/plotting/plot_tvt_fold_perf.R")
#function to plot performance of each fold
source("src/plotting/plot_train_test_ensemble_perf.R")
#function to apply bias correction
source("src/modeling/apply_ml_bias_correction.R")
#load grab sample data function
source("src/collation/load_grab_sample_data.R")
#function to plot model timeseries
source("src/plotting/plot_model_timeseries.R")
```

## Read in sensor dataset

Here, we read in the sensor data, apply min-max scaling defined in the `apply_sensor_transformations_toc()` function, and store the dataset for use later

```{r}
# Read in all data to create scaling parameters
wide_data_file <- list.files(here("data","collated", "sensor"), pattern = "all_sensor_data_wide_.*\\.parquet", full.names = TRUE)%>%
  tail(1)

# Use the complete sensor dataset for scaling parameters (assuming sensor has all necessary columns already)
complete_sensor_data <- read_parquet(wide_data_file) %>%
  apply_sensor_transformations_toc(., dt_col = "DT_hourly_round")%>%
  select(-DO, -Depth, -ORP, -pH) #remove unused sensors


```


## Read in matchup dataset

Matchup dataset is created in 02_sensor_grab_matchup, where we match data from the sensors with each available lab-anlyzed TOC sample. In this script we will use the dataset that also contains the Virridy Yampa Samples. 

```{r}
matchup_dataset <- list.files(here("data","derived", "matchup_datasets"), pattern = "parsed_toc_sensor_virridy_clp_", full.names = TRUE)%>%
  tail(1)

# Read in all matches data
all_matches <- read_csv(file = matchup_dataset, show_col_types = F) %>%
  apply_sensor_transformations_toc(dt_col = "sensor_datetime")%>%
  #remove NAs
  filter(!is.na(TOC))%>%
  mutate(clean_sensor_datetime = parse_date_time(sensor_datetime, orders = c("%Y-%m-%d %H:%M:%S", "%m/%d/%y %H:%M","%Y-%m-%d" )))%>%
  mutate(sensor_datetime = force_tz(clean_sensor_datetime, tzone = 'MST'), 
         month = lubridate::month(sensor_datetime), 
         doy = yday(sensor_datetime),
         sin_doy =  sin(2 * pi * doy / 365), 
         basin = if_else(collector == "Virridy", "Yampa", "Poudre"))
```

## Adding in flow for Yampa Samples
Chose YAMSTECO as it has it accumulates the majority of the basin's flow but is not severly impacted by diversions. This will be tested as our "daily_canyon_flow_cfs" feature in the model.
https://dwr.state.co.us/Tools/Stations/YAMSTECO?params=DISCHRG

```{r}
yampa_data <- all_matches %>% filter(basin == "Yampa")

max_yampa_date = as.character(as.Date(max(yampa_data$sensor_datetime)) + days(1))
min_yampa_date = as.character(as.Date(min(yampa_data$sensor_datetime)) - days(1))
# read in flow data for Yampa sites
yampa_data_w_flow <- yampa_data %>%
  mutate(date = as.Date(sensor_datetime)) %>%
  select(-daily_canyon_flow_cfs)%>%
  left_join(
    #use CDSSR to read in flow data for yampa
    get_telemetry_ts(abbrev = "YAMSTECO", 
                     parameter = "DISCHRG",
                     start_date = min_yampa_date,
                     end_date = max_yampa_date,
                     timescale = "day",
                     api_key = NULL) %>%
      select(date = datetime, daily_canyon_flow_cfs = meas_value),
    by = c( "date")
  )

all_matches <- all_matches %>% 
  # remove old yampa data
  filter(basin != "Yampa")%>%
  # add in new yampa data
  bind_rows(yampa_data_w_flow)

```

## Define ID Columns, Target, and Features

```{r}
# ID columns to keep to use later on 
id_cols <- c("Date", "id","site", "DT_hourly_round", "sensor_datetime","lab_datetime",  "collector","basin")
#What we want to predict
target <- "TOC"
# Get feature columns from complete sensor data but remove id cols and target
features <- setdiff(names(complete_sensor_data), c(id_cols, target))
#remove unused sensors
features <- setdiff(features, c("DO", "Depth", "ORP", "pH")) 
#remove variables not in the yampa dataset
features <- setdiff(features, c("temp_mean_48hr","temp_mean_24hr","fdom_max_6hr", "fdom_corr_max_6hr","temp_max_24hr" ,"sc_var_24hr","sc_sd_24hr"             ,"turb_median_12hr", "hourly_canyon_flow_cfs", "data_avail")) 
features
```

## Create Scaling Parameters & Normalize Data
We do not have access to the full sensor dataset for the Yampa sites, so we will use the scaling parameters from the complete Poudre sensor dataset to normalize the matchup data. This may yield some out of samples results when we apply the model to the Yampa data. 

```{r}
# Apply scaling to dataset
all_matches_norm <- all_matches %>%
  #Only keep columns that would be in the complete sensor dataset
  select(any_of(id_cols), any_of(features), any_of(target))%>%
  #apply normalization 
  apply_training_scale(
    new_data = .,
    features = features,
    full_dataset = complete_sensor_data
  )%>%
  #remove any columns with NAs
  na.omit()

```

## Create Holdout Dataset

```{r}
# Trim out any known bad data or sites we don't want to include in modeling
all_matches_trimmed <- all_matches_norm%>%
  filter(!(id == "cbri" & between(sensor_datetime, ymd("2024-10-08"), ymd("2024-10-10"))))%>% #exclude a cbri sample taken during release
  filter(!(id == "pman" & between(sensor_datetime, ymd("2025-05-28"), ymd("2025-06-01"))))%>% #exclude a pman sample with bad sensor data
  filter( id %nin% c("springcreek", "boxcreek"))%>%
  #Remove the Yampa data from training/validation set
  filter(basin != "Yampa")

yampa_testing <- all_matches_norm %>%
  filter(basin == "Yampa")

# Create testing set from 2025 data at specific sites
testing <- all_matches_trimmed %>%
  filter((year(sensor_datetime) == 2025 & id %in% c("chd", "sfm", "pbr", "pman", "pbd")))
# Create training/validation set
train_val <- all_matches_trimmed %>%
  anti_join(testing)%>%
  select(id, any_of(features),TOC, sensor_datetime, collector)%>%
  mutate(id = ifelse(id == "archery_virridy", "archery", id)) #combine archery and archery_virridy since they are the same site

```

## Set up model folds
Since we have relatively little data, we are going to use three CV folds. Create 3 separate folds and 3 models, with one fold used for validation on each model. Generally we are looking for:
- Each fold to have ~30-35% of the total data in validation
- Each validation set to have a similar TOC distribution as its training set (median, range)
- Each should have a mix of Upper Mainstem (SFM, PFAL, PBD), Upper Trib (CBRI, CHD, LBEA, PENN) & Lower Mainstem (salyer, udall, riverbend, elc, archery, riverbluffs)

```{r}
val_set_1 <- c( "sfm","chd","lbea","riverbluffs" )
 
val_set_2 <- c( "pbd","joei", "penn", "archery" )
 
val_set_3 <- c("pfal","udall","cbri","elc","salyer","riverbend")

# Function to create plots for a given fold
plot_fold <- function(fold_id, train_val) {
  val_set_name <- paste0("val_set_", fold_id)
  val_set <- train_val %>% filter(id %in% get(val_set_name))
  train_set <- train_val %>% filter(id %nin% get(val_set_name))
  
  prop_val <- round(nrow(val_set)/nrow(train_val), 3) * 100
  train_med <- round(median(train_set$TOC, na.rm = TRUE), 1)
  val_med <- round(median(val_set$TOC, na.rm = TRUE), 1)
  
  # Histogram
  hist <- ggplot(train_set, aes(x = TOC)) +
    geom_histogram(bins = 30, fill = "grey") +
    geom_histogram(data = val_set, aes(x = TOC), bins = 30, fill = "#E70870", alpha = 0.5) +
    labs(
      title = paste0("Fold Set ", fold_id, ": ", nrow(val_set), " samples (", prop_val, "%) in validation set"),
      subtitle = paste0("Validation Sites: ", paste(unique(val_set$id), collapse = ", ")),
      x = "TOC (mg/L)"
    ) +
    theme_minimal()
  
  # Boxplot
  box <- ggplot(
    train_set %>% mutate(set = "Train") %>% 
      bind_rows(val_set %>% mutate(set = "Validation")) %>% 
      mutate(set = factor(set, levels = c("Train", "Validation"))),
    aes(x = set, y = TOC, fill = set)
  ) +
    geom_boxplot() +
    scale_fill_manual(values = c("Train" = "#256BF5", "Validation" = "#E70870")) +
    labs(
      title = paste0("Train Median TOC: ",train_med, "\nVal Median TOC: ", val_med),
      x = "",
      y = "TOC (mg/L)",
      fill = "TV Group",
    ) +
    theme_minimal()
  
  # Combine plots
  ggarrange(hist, box, nrow = 1)
}

# plot all folds
ggarrange(plotlist = map(1:3, ~ plot_fold(.x, train_val)), ncol = 1)


#Update train_val with a fold_id column based on val_sets
train_val <- train_val %>%
  mutate(fold_id = case_when(
    id %in% val_set_1 ~ 1,
    id %in% val_set_2 ~ 2,
    id %in% val_set_3 ~ 3
  ))

```


## Visualize TOC distribution in training/validation vs testing sets

Currently, our 2025 data (testing) is skewed to higher TOC values (April- July) compared to the training/validation set. This is not ideal, but we will proceed with this split for now given limited data. Future data updates in December should help even this out as TOC decreases in the fall. 

```{r}
train_val%>%
  mutate(group = "Train/Val")%>%
  bind_rows(
    testing%>%
      mutate(group = "Testing")
  )%>%
ggplot(., aes(x = TOC, fill = collector)) +
  geom_histogram(bins = 30) +
  labs(
    title = paste0("Training/Validation Set: ", nrow(train_val), " samples              Testing Set: ", nrow(testing), " samples"),
    subtitle = paste0("Training TOC Range: ", round(min(train_val$TOC, na.rm = TRUE),2), " - ", round(max(train_val$TOC, na.rm = TRUE),2) , " mg/L (Median: ", round(median(train_val$TOC), 2), ")\nTesting TOC Range: ", round(min(testing$TOC, na.rm = TRUE),2), " - ", round(max(testing$TOC, na.rm = TRUE),2), " mg/L (Median: ", round(median(testing$TOC), 2),")"),
    x = "TOC (mg/L)", 
    fill = "Sample Collector"
  ) +
  facet_wrap(~group, ncol = 1)+
  ROSS_theme+
  theme(legend.position = "bottom")

```


# Model Training

For our model training we will first do some feature tuning to see which features are most useful for predicting TOC. We will then do hyperparameter tuning on the selected features to optimize model performance.

## Feature Tuning 

### View correlations between features and target (TOC)
Most products with FDOM in them are going to have higher correlation with TOC, as well as seasonal (sin_doy) and flow related features. We want to be sure that we don't put too many of these in the model as calibration error or sensor drift could lead to compounding errors.

```{r}
# Combine training/validation and testing data
  modeling_data <- train_val %>%
    mutate(group = "Train/Val") %>%
    bind_rows(
      testing %>%
        mutate(group = "Testing")
    ) %>%
  bind_rows(
      yampa_testing %>%
        mutate(group = "Yampa")
    ) %>%
  
    mutate(
      watershed = case_when(
        id %in% c("joei", "cbri", "chd", "pfal", "pbr", "sfm", "pman", "pbd") ~ "UpperPoudre",
        group == "Yampa" ~ "Yampa",
        TRUE ~ "LowerPoudre"
      )
    )
plots <-set_names(features) %>%
  map( function(feat) {
  

  cor <- round(cor(modeling_data[[feat]], modeling_data$TOC, use = "complete.obs"), 2)

  ggplot(modeling_data, aes(x = .data[[feat]], y = TOC, shape = collector, color = watershed)) +
    geom_point() +
    annotate(
    "text",
    x = mean(range(modeling_data[[feat]], na.rm = TRUE)), # middle of x range
    y = max(modeling_data$TOC, na.rm = TRUE) * 0.95,      # 95% of y max (near top)
    label = paste0("Corr: ", round(cor(modeling_data[[feat]], modeling_data$TOC, use = "complete.obs"), 2)),
    hjust = 0.5,
    vjust = 1,
    size = 4
  ) +
    labs(x = feat, y = "TOC (mg/L)", shape = "Collector", color = "Watershed") +
    #facet_wrap(~watershed, scales = "free") +
    ROSS_theme
})


plots%>%
  wrap_plots(., guides = "collect")
```

### Features differing in Yampa Set
We can see that there are some basin differences in the Yampa samples which may influence the performance of the model on the Yampa data. Features like doy, f_sc, f_x_sc, daily canyon flow, and f_x_turb show some differences in distribution compared to the Poudre samples. We do still see FDOM as a useful starting point for predicting TOC in the Yampa as well.

```{r}
wrap_plots(plots$Sensor_Cond, plots$sin_doy, plots$f_sc, plots$FDOMc, plots$daily_canyon_flow_cfs, plots$f_x_turb, ncol = 3, guides = "collect") &
  theme(legend.position = "bottom")

```

### Pruning Features with high correlations 

We want to avoid using too many highly correlated features in the model, as this can lead to overfitting and instability. We will look at the correlation matrix and remove some of the more redundant features.

```{r}
# get a correlation matrix across all features
corr_matrix <- modeling_data %>% 
  select(all_of(features)) %>% 
  cor(use = "pairwise.complete.obs")

# we just want half the matrix, so recode those in the lower triangle to NA
corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter for those with correlation > 0.9
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.9) %>%
  arrange(desc(abs(correlation))) %>% 
  # remove redundant vars
  filter(feature1 %nin% c("FDOM", "doy"))

# look at the list and make some educated decisions about features
high_corr_pairs

fewer_features <- features[features %nin% c("FDOM", "doy", 
                                            # remove vars with high correlation 
                                            # with others
                                            "Temp", "f_log",
                                            "Sensor_Turb")]

# do the correlation matrix again to make sure that is okay
corr_matrix <- modeling_data %>% 
  select(all_of(fewer_features)) %>% 
  cor(use = "pairwise.complete.obs")

corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.85) %>%
  arrange(desc(abs(correlation)))

high_corr_pairs

```

`high_corr_pairs` has features we should be cautious of using together. We will try to avoid using more than one of these in the same model.


### Testing Features

First, we will start with our base model (live sensor data only) and then test adding in other features to see if they improve performance. We will use the default hyperparameters below for all models to keep things consistent. Once we have tested all the features together, we will want to trim it down a bit so we are not using too many highly correlated features or overfitting the model due to lack of data

#### Function to test features & defaults
This just runs a basic xgboost model with the specified features and default hyperparameters. It returns feature importance, shap values, and best iteration info for each fold.

```{r}
test_features <- function(train_val,
                          features_to_test,
                          target,
                          default_hyper_params,
                          weight_fun = toc_high_weights) {
  # double check that there are no duplicates in features
  features_to_test <- unique(features_to_test)
  
  #create blank dataframes to store importance/shap values
  importance_df <- tibble(Feature = features_to_test)
  shap_val_df <- tibble(Feature = features_to_test)
  best_msgs <- tibble()
  models <- list()
  
  for(i in unique(train_val$fold_id)){
    
    #train val split
    train_data <- train_val %>%
      filter(fold_id != i) %>%
      select(any_of(c(features_to_test, target, "id", "sensor_datetime", "collector")))
    
    val_data <- suppressMessages(
      train_val %>%
        anti_join(train_data)
    )
    
    #weight data
    w_train <- weight_fun(train_data)
    w_val   <- weight_fun(val_data)
    
    #prep matrices
    dtrain <- xgb.DMatrix(
      data = as.matrix(train_data[, features_to_test]),
      label = train_data[[target]],
      weight = w_train
    )
    
    dval <- xgb.DMatrix(
      data = as.matrix(val_data[, features_to_test]),
      label = val_data[[target]],
      weight = w_val
    )
    
    watchlist <- list(train = dtrain, eval = dval)
    
    #Train model with defaults
    model <- xgb.train(
      params = default_hyper_params,
      data = dtrain,
      nrounds = 10000,
      watchlist = watchlist,
      early_stopping_rounds = 1000,
      print_every_n = 1000,
      verbose = 0,
  nthread = 1
    )
    #Save model
    models[[i]] <- model
    #Save best msg
    best_msg <- tibble(
      fold_id = i,
      best_iteration = model$best_iteration,
      best_msg = model$best_msg
    ) %>%
      #clean up message
      mutate(
        train_rmse = str_extract(best_msg, "(?<=train-rmse:)\\d+\\.\\d+"),
        eval_rmse  = str_extract(best_msg, "(?<=eval-rmse:)\\d+\\.\\d+"),
        train_rmse = as.numeric(train_rmse),
        eval_rmse  = as.numeric(eval_rmse)
      )%>%
      select(-best_msg)
    #best message
    best_msgs <- bind_rows(best_msgs, best_msg)%>%
      distinct()
    
    
    #Feature importance
    importance_df <- importance_df%>%
      left_join(xgb.importance(feature_names = features_to_test, model = model)%>%
                  dplyr::rename_with( #fix column names
                    ~ glue::glue("{.x}_fold_{i}"),
                    .cols = c(Gain, Cover, Frequency)
                  ), by = "Feature")
    
    # SHAP (Training)
    shap_long_train <- shap.prep(xgb_model = model, X_train = as.matrix(train_data[, features_to_test]))
    shap_vals_train <- shap.values(xgb_model = model, X_train = dtrain)$mean_shap_score %>%
      as.data.frame() %>%
      tibble::rownames_to_column("Feature") %>%
      rename(train_mean_abs_shap = ".")
    
    # SHAP (Validation)
    shap_long_val <- shap.prep(xgb_model = model, X_train = as.matrix(val_data[, features_to_test]))
    shap_vals_val <- shap.values(xgb_model = model, X_train = dval)$mean_shap_score %>%
      as.data.frame() %>%
      tibble::rownames_to_column("Feature") %>%
      rename(val_mean_abs_shap = ".")
    
    shap_val <- shap_vals_val%>%
      left_join(shap_vals_train, by = "Feature")%>%
      dplyr::rename_with( #fix column names
        ~ glue::glue("{.x}_fold_{i}"),
        .cols = c(val_mean_abs_shap, train_mean_abs_shap)
      )
    shap_val_df <- shap_val_df%>%
      left_join(shap_val, by = c("Feature"))
    
    # p_train <- shap.plot.summary(shap_long_train)
    # p_val <- shap.plot.summary(shap_long_val)
    # 
    
  }
  
  #output
  list(
    model = models,
    features = features_to_test, 
    best_msg = best_msgs,
    model_importance = importance_df,
    shap_values = shap_val_df
  )
  
}

#xgboost hyper parameters

default_hyper_params <- list(
  objective        = "reg:squarederror",
  eval_metric      = "rmse",
  eta              = 0.05,
  gamma            = 0.6,
  alpha            = 0,
  lambda           = 1,
  max_depth        = 4,
  subsample        = 0.5,
  colsample_bytree = 0.5,
  min_child_weight  = 2
)

# Weighting function for TOC
toc_high_weights <-  function(df) {
  ifelse(df$TOC <= 4,1,                  # flat weight for low TOC
         1 + 3 *log1p(df$TOC - 4))  # grows slowly, no cap needed
}


target = "TOC"
```

#### Run across folds testing different features to see which improve performance

Here I run the function and am going to look at feature importance and shap values to see which features are most useful, then trim features back to reduce our correlative and overfitting issues. 

```{r}
#starting with all features then trimming down to more highly correlated ones
#set aside sensor data that are generally useful
general_features <- c("FDOMc","Sensor_Turb","Sensor_Cond", "Chl_a", "Temp")

feature_results <- map(
  #trying all the combinations of features we think might be useful (commented out were tested but found to not improve base model)
  list(
  gen_feat = general_features, 
 fewer_features = c(fewer_features), 
 #Removing these features based on high correlation and low importance in prior tests and because they are not available in the Yampa set
 fewer_features_yampa = c(fewer_features)%>%setdiff(., c("turb_median_12hr","temp_x_f", "temp_mean_48hr","temp_x_turb", "turb_f", "sc_sd_24hr", "temp_mean_24hr", "temp_f"  ))
) , ~test_features(
  train_val = train_val,
  features_to_test = .x,
  target = "TOC",
  default_hyper_params = default_hyper_params
))
```

#### Feature Set Results

##### Weighted RMSE

We compare across folds to see if there is big differences in folds (ie unbalanced data) and also compare each feature set against "base" model of general features (sensors). Looking for features that make meaningful improvements in RMSE across all folds. If our RMSE improves but that is related to only a single fold, we may want to reevaluate the feature set as it is likely overfitting to a given site(s). 
Keep in mind that this is weighted RMSE so it may be higher than unweighted RMSE you see in other places.

```{r}
# Highlight differences between folds in rmse
map_dfr(feature_results, ~{
  # Compute feature set description
  unique_features <- setdiff(.x$features, general_features)
  final_features <- c("general_features", unique_features)
  feature_desc <- paste(final_features, collapse = ", ")
  
  # Take best_msg tibble and rename eval_rmse by fold
  .x$best_msg %>%
    select(fold_id, eval_rmse) %>%
    mutate(fold_col = paste0("rmse_fold_", fold_id)) %>%
    select(-fold_id) %>%
    mutate(eval_rmse = round(eval_rmse, 2)) %>%
    pivot_wider(names_from = fold_col, values_from = eval_rmse) %>%
    mutate(
      features = feature_desc,
      mean_rmse = round(mean(c_across(starts_with("rmse_fold_")), na.rm = TRUE), 2)
    )%>%
    select(features, mean_rmse, rmse_fold_1, rmse_fold_2, rmse_fold_3)
  
}, .id = "feature_set") %>%
#check if new feature is better than baseline
  mutate(
    baseline_rmse = mean_rmse[feature_set == "gen_feat"],
    improves_rmse = if_else(mean_rmse < baseline_rmse, round(mean_rmse - baseline_rmse, 2), 0)
  ) %>%
  select(-baseline_rmse)%>%
  arrange(improves_rmse)
```

##### Feature Importance

This comes directly from the xgboost package and is based on Gain, Cover, and Frequency of splits. We summarize across folds to see which features are most important across all folds. 

```{r}
#table with all importances
importance_tables <- map(feature_results, ~{
  imp <- .x$model_importance
  
  # Identify fold columns dynamically
  gain_cols <- grep("^Gain_fold_", names(imp), value = TRUE)
  cover_cols <- grep("^Cover_fold_", names(imp), value = TRUE)
  freq_cols <- grep("^Frequency_fold_", names(imp), value = TRUE)
  
  imp %>%
    mutate(
      mean_gain = rowMeans(across(all_of(gain_cols)), na.rm = TRUE),
      mean_cover = rowMeans(across(all_of(cover_cols)), na.rm = TRUE),
      mean_frequency = rowMeans(across(all_of(freq_cols)), na.rm = TRUE)
    ) %>%
    mutate(new_feat = ifelse(Feature %in% general_features, "", "*")) %>%
    select(Feature, new_feat, mean_gain, mean_cover, mean_frequency) %>%
    
    arrange(desc(mean_gain))
}, .id = "feature_set")

#Get features with less than 0.02 gain on average across folds in fewer_features to consider removing
cat("Features with mean gain < 0.02 in 'fewer_features_yampa':\n")
importance_tables$fewer_features_yampa%>%
  filter(mean_gain < 0.02)

```


##### Shapley Values

Let's look at training and validation shap values for features across all folds. This gives us a sense of which features are most important in the model predictions. We summarize across folds to see which features are most important across all folds. In general we would expect features to have similar importance in training and validation, if a feature has high importance in training but low in validation it may be overfitting to the training data and we may want to re-evaluate. 

```{r}
#table with all shap values
shap_summary <- map_dfr(feature_results, ~{
  shap <- .x$shap_values
  
  val_shap_cols <- grep("^val_mean_abs_shap_fold_", names(shap), value = TRUE)
  train_shap_cols <- grep("^train_mean_abs_shap_fold_", names(shap), value = TRUE)
  
  shap %>%
    mutate(
      mean_val_shap = round(rowMeans(across(all_of(val_shap_cols)), na.rm = TRUE), 3),
      mean_train_shap = round(rowMeans(across(all_of(train_shap_cols)), na.rm = TRUE), 3)
    ) %>%
    mutate(val_train_diff = mean_val_shap - mean_train_shap) %>%
    select(Feature, mean_val_shap, mean_train_shap, val_train_diff)
}, .id = "feature_set")

shap_long <- shap_summary %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "type",
    values_to = "mean_shap"
  ) %>%
  mutate(
    type = recode(type,
                  mean_val_shap = "Validation",
                  mean_train_shap = "Training"),
    feature_set = factor(feature_set, levels = unique(feature_set))
  ) %>%
  mutate(Feature = factor(Feature, levels = c( rev(feature_results$fewer_features$features),"Sensor_Turb","Temp")))

ggplot(shap_long, aes(x = feature_set, y = Feature, fill = mean_shap)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_viridis_c(
  option = "magma",
  limits = c(0, 1),
  name = "Mean |SHAP|",
  direction = -1  # reverses so dark = low, bright = high (optional)
)+
  facet_wrap(~type) +
  labs(
    x = "Feature Set",
    y = "Feature",
    title = "Mean SHAP Values by Feature and Feature Set"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )



```

### TV performance of best feature set

After trimming down a couple of features (minimal improvement/low shap values), we should take a quick look at model performance in TV space with our base model (default hyperparameters)

```{r}

best_models <- feature_results$fewer_features_yampa$model

feature_tv_plots <- map(1:3, ~plot_tv_fold_perf(fold_set = .x,
                             train_val_df = train_val,
                             fold_model = best_models[[.x]], # a list where each element is a model for that fold
                             target_col = "TOC", units = "mg/L")
                            )
ft_tv_plot<- wrap_plots(feature_tv_plots, ncol = 1, guides = "collect")&
  theme(legend.position = "bottom")
ft_tv_plot


```


## Hyperparameter Tuning
Now that we have a feature set, we can do some hyperparameter tuning to see if we can improve model performance. This will take a while to run, so we will save the results and load them in later.

### Setup

```{r}

folds <- tibble(
  fold = 1:3,
  val_ids = list(val_set_1, val_set_2, val_set_3)
)
features <- best_models[[1]]$feature_names

# Reduce tune grid for reruns (ie. General best params)
tune_grid = expand.grid(
  nrounds = 10000,
  max_depth = c(2, 3),
  eta = c( 0.01, 0.1),
  gamma = c(0.8),
  lambda = c(1,10),
  alpha = c(0, 1),
  colsample_bytree = c(0.8),
  min_child_weight = c(2, 4),
  subsample = c(0.5,0.8)
)

```

### Run hyperparameter tuning

```{r}
set.seed(123)
hyperp_results <- xgboost_hyperparameter_tuning(
  data = train_val %>% select(TOC, id, any_of(features)), 
  weights = toc_high_weights,
  tune_grid = tune_grid,
  target_col = "TOC",
  site_col = "id",
  fold_ids =  folds, 
  units = "mg/L")

```

### Choosing best grids

### Fold 1

```{r}
hyperp_results[[1]]$perf_dist
hyperp_results[[1]]$eval_plot
hyperp_results[[1]]$tv_plot

best_fold_1_grid <- 11
```

### Fold 2

```{r}
hyperp_results[[2]]$perf_dist
hyperp_results[[2]]$eval_plot
hyperp_results[[2]]$tv_plot

best_fold_2_grid <- 55

```

### Fold 3

```{r}
hyperp_results[[3]]$perf_dist
hyperp_results[[3]]$eval_plot
hyperp_results[[3]]$tv_plot

best_fold_3_grid <- 3

```

## Extract best models for each fold

```{r}

full_model <- map(1:3, function(i) {
  
  best_grid <- paste0("grid_", get(glue("best_fold_{i}_grid")))
  model_info <- hyperp_results[[i]]$grid_results[[best_grid]]
  list(
    model = model_info$model,
    perf = model_info$perf,
    params = model_info$params,
    training_data = train_val %>% filter(fold_id != i),
    validation_data = train_val %>% filter(fold_id == i)
  )
})

```

### Checking Results


#### Feature Importance

```{r}
importance_list <- lapply(seq_along(full_model), function(i) {
  fold_model <- full_model[[i]]$model
  features <- fold_model$feature_names
  importance_matrix <- xgb.importance(feature_names = features, model = fold_model)
  importance_df <- as.data.frame(importance_matrix)
  importance_df$Fold <- paste("Fold", i)
  return(importance_df)
})

print(importance_list)
```

#### Shapley Importance

```{r}

# Function to plot SHAP values for a given fold
plot_shap <- function(fold_set, training_data, fold_model) {
  #get features from model
  features <- fold_model$feature_names
  # Predictions on validation/training set
  dtrain <- as.matrix(training_data[, features])
  
  shap_values <- shap.values(xgb_model = fold_model, X_train = dtrain)
  
  # Long-format SHAP values
  shap_long <- shap.prep(
    shap_contrib = shap_values$shap_score,
    X_train = dtrain
  )
  #return shapley plot
  shap.plot.summary(shap_long)
}

map(1:3, ~ plot_shap(
  fold_set = .x,
  fold_model = full_model[[.x]]$model,
  training_data = full_model[[.x]]$training_data
))%>%
  wrap_plots(., ncol = 1, guides = "collect") &
  theme(legend.position = "bottom")

```

### Plots of TV Performance
Now let's look at our performance in TV space.

```{r}

hyperparam_tv_plots <- map(1:3, ~plot_tv_fold_perf(fold_set = .x,
                             train_val_df = train_val,
                             fold_model = full_model[[.x]]$model, # a list where each element is a model for that fold
                             target_col = "TOC", units = "mg/L")
                            )

hp_tv_plot<- wrap_plots(hyperparam_tv_plots, ncol = 1, guides = "collect")&
  theme(legend.position = "bottom")
hp_tv_plot
#ggsave("data/upper_clp_dss/figures/toc_modeling/performance/toc_model_fold_validation_performance.png", width = 14, height = 10)

```

#### Against base hyperparameters

We should hope that all our tuning efforts have improved the model!! 

```{r}
# Compare best hyperparameter model to default hyperparameter model
final_plot <- wrap_plots(
  ft_tv_plot,
  hp_tv_plot,
  ncol = 2,
  guides = "collect"
) &
  theme(legend.position = "bottom")
final_plot <- final_plot + 
  plot_annotation(
    title = "Model Performance Comparison: Default (left) vs Tuned Hyperparameters (right)"
  )

final_plot

```

Looking at the plots, we can see that the tuned hyperparameter model has improved performance in all folds compared to the default hyperparameter model. The validation RMSE has decreased in all folds but it does look like model fold three is slightly less accurate at the higher ends (possible overtraining?). Overall, this is a good sign that our tuning efforts have paid off!

## Save as "in_dev" and remove unneeded objects

If we want to test additional ideas for bias correction or look at model performance on timeseries data, we can just run this script until `model_training` and then load in the "in_dev" model file to avoid having to rerun all the tuning again.

```{r}
#commented out so we don't overwrite existing model files
saveRDS(full_model, file = here("data", "derived", "models","TOC_realtime", "in_dev", paste0("ross_only_toc_xgboost_models_instantaneous_",version,".rds")))
#if we want to reload model
#full_model <- readRDS(here("data", "derived", "models","TOC_realtime", "in_dev", paste0("ross_only_toc_xgboost_models_instantaneous_",version,".rds")))

rm(list = setdiff(ls(), c("train_val", "testing", "yampa_testing", "full_model","version","id_cols",
                          "complete_sensor_data",
                          #keeping functions around
                          "apply_training_scale","plot_tvt_fold_perf","plot_train_test_ensemble_perf",
                          "apply_ml_bias_correction", "load_grab_sample_data", "plot_model_timeseries" )) )
```


# Evaluate on Test Set

We should only do this once we feel good about our model and are ready to see how it performs on the unseen test set. This should only be run after big changes to the model or if we get new data (ie The model should be in a "final/working" version)

## Visualize Train Val Test

### Individual folds

```{r}
plots<- map(1:3, ~{
    plot_tvt_fold_perf(train_val, testing, fold_set = .x, fold_model = full_model[[.x]]$model, target_col = "TOC", units = "mg/L", subtitle_arg = "CLP Samples only")
})
plots

  # ggsave(glue(here("docs", "uclp_dss", "figs", "toc_modeling", "performance", "ross_only_model_fold_{i}.png")),
  #        plot = p, width = 10, height = 6, dpi = 500)

```

### All folds together

```{r}
# Single plot with all folds
wrap_plots(plots, ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")
# ggsave(here("docs","uclp_dss", "figs", toc_modeling","performance",paste0("ross_only_model_all_folds_TVT", version, ".png"),
#        width = 14, height = 8, dpi = 500)
```

### Ensemble predictions

```{r}
# Plot ensemble
ensemble <- plot_train_test_ensemble_perf(train_val_df = train_val,
                                   test_df = testing,
                                   fold_models = map(full_model, "model"),
                                   target_col = "TOC",
                                   units = "mg/L",
                                   subtitle_arg = "CLP Samples only; No lagged parameters")
plot(ensemble)
# ggsave(here("docs","uclp_dss", "figs", toc_modeling","performance",paste0("ross_only_model_ensemble", version, ".png"),
#        plot = ensemble, width = 12, height = 10, dpi = 500)
```

# Evaluate on Yampa Test Samples

## Fold performance
```{r}
### Individual folds
yampa_plots<- map(1:3, ~{
    plot_tvt_fold_perf(train_val, yampa_testing, fold_set = .x, fold_model = full_model[[.x]]$model, target_col = "TOC", units = "mg/L", subtitle_arg = "CLP Samples only")
})
yampa_plots

  # ggsave(glue(here("docs", "uclp_dss", "figs", "toc_modeling", "performance", "ross_only_model_fold_{i}.png")),
  #        plot = p, width = 10, height = 6, dpi = 500)

```

### All folds together

```{r}
# Single plot with all folds
wrap_plots(yampa_plots, ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")
ggsave(here("docs","uclp_dss", "figs", "realtime_toc_modeling","performance",paste0("ross_only_yampa_model_all_folds_TVT", version, ".png")),
       width = 14, height = 8, dpi = 500)
```

### Ensemble

```{r}
ensemble <- plot_train_test_ensemble_perf(train_val_df = train_val,
                                   test_df = yampa_testing,
                                   fold_models = map(full_model, "model"),
                                   target_col = "TOC",
                                   units = "mg/L",
                                   title_arg = "Yampa Samples Ensemble Predictions (Mean ± Range across folds)",
                                   subtitle_arg = "CLP Samples only in training; No lagged parameters")
plot(ensemble)

ggsave(here("docs","uclp_dss", "figs", "realtime_toc_modeling", "performance", paste0("ross_only_model_yampa_ensemble_", version, ".png")),
       plot = ensemble, width = 12, height = 10, dpi = 500)

```







