---
title: "Testing TOC preds"
author: "Matthew Ross & Sam Struthers-ROSSyndicate"
date: "2025-08-01"
output: html_document
---

# Set Up

## Package Load

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("src/setup_libraries.R")

#these are just packages for testing within this script
library(gridExtra)
library(grid)
library(Ckmeans.1d.dp)
library(DiagrammeR)
library(Metrics)
library(glue)
library(SHAPforxgboost)
set.seed(123)
#set version for saving (update when changes are significant)
version = "20251023"

source("src/apply_sensor_transformations_toc.R")
# function to tune xgboost hyperparameters
source("src/xgboost_hyperparameter_tuning.R")
#function to plot performance of each fold
source("src/plot_tv_fold_perf.R")
#function to plot performance of each fold
source("src/plot_tvt_fold_perf.R")
#function to plot performance of each fold
source("src/plot_tvt_ensemble_perf.R")
```

## Create scaling based on all sensor data

We should functionalize this chunk so that we can easily apply the scaling as we get new sensor data in the future.

```{r}
# Read in all data to create scaling parameters
wide_data_file <- list.files(here("data","collated", "sensor"), pattern = "all_sensor_data_.*\\.parquet", full.names = TRUE)%>%
  tail(1)


all_sensor_data_wide <- read_parquet(wide_data_file)

# Use the complete sensor dataset for scaling parameters (assuming sensor has all necessary columns already)
complete_sensor_data <- all_sensor_data_wide %>%
  apply_sensor_transformations_toc(., dt_col = "DT_hourly_round")


# Save the original min/max values for each numeric column, removing NAs
scaling_params <- complete_sensor_data %>%
  select(where(is.numeric)) %>%
  summarise(across(
    everything(),
    list(min = ~min(.x, na.rm = TRUE), max = ~max(.x, na.rm = TRUE)),
    .names = "{.col}_{.fn}"
  ))


# Save these parameters
#saveRDS(scaling_params, "data/upper_clp_dss/modeling/scaling_parameters_20251007.rds")
# Load saved scaling parameters
#scaling_params <- readRDS("data/upper_clp_dss/modeling/scaling_parameters_20250820.rds")

# Function to apply the same scaling to new data
apply_training_scale <- function(new_data, scaling_params) {
  numeric_cols <- names(select(new_data, where(is.numeric)))
  
  scaled_data <- new_data
  
  for (col in numeric_cols) {
    min_val <- scaling_params[[paste0(col, "_min")]]
    max_val <- scaling_params[[paste0(col, "_max")]]
    
    # Apply same min-max scaling as training data
    scaled_data[[col]] <- (new_data[[col]] - min_val) / (max_val - min_val)
  }
  
  return(scaled_data)
}
```


## Read in matchup dataset

```{r}
matchup_dataset <- list.files(here("data","derived", "matchup_datasets"), pattern = "parsed_toc_sensor_clp_", full.names = TRUE)%>%
  tail(1)

# Read in all matches data
all_matches <- read_csv(file = matchup_dataset, show_col_types = F) %>%
  apply_sensor_transformations_toc(dt_col = "sensor_datetime")%>%
  #remove NAs
  filter(!is.na(TOC))%>%
  mutate(clean_sensor_datetime = parse_date_time(sensor_datetime, orders = c("%Y-%m-%d %H:%M:%S", "%m/%d/%y %H:%M","%Y-%m-%d" )))%>%
  mutate(sensor_datetime = force_tz(clean_sensor_datetime, tzone = 'MST'), 
         month = lubridate::month(sensor_datetime))

id_cols <- c("Date", "id", "sensor_datetime","lab_datetime",  "collector", "data_avail")

# Apply scaling to dataset
all_matches_norm <- all_matches %>%
  #Only keep columns that would be in the complete sensor dataset
  select(any_of(id_cols), any_of(names(complete_sensor_data)))%>%
  #normalize data based on full sensor set from 01_sensor_data_prep.rmd and the chunk above
  apply_training_scale(., scaling_params)%>%
  #add back in TOC (non normalized)
  mutate(TOC = all_matches$TOC)%>%
  na.omit()

```

## Create Holdout Dataset

```{r}

target  = "TOC"
features <- setdiff(colnames(all_matches_norm), c(id_cols, target))

all_matches_trimmed <- all_matches_norm%>%
  filter(!(id == "cbri" & between(sensor_datetime, ymd("2024-10-08"), ymd("2024-10-10"))))%>% #exclude a cbri sample taken during release
  filter(!(id == "pman" & between(sensor_datetime, ymd("2025-05-28"), ymd("2025-06-01"))))%>% #exclude a pman sample with bad sensor data
  filter( id %nin% c("springcreek", "boxcreek"))


testing <- all_matches_trimmed %>%
  filter((year(sensor_datetime) == 2025 & id %in% c("chd", "sfm", "pbr", "pman", "pbd")))


train_val <- all_matches_trimmed %>%
  anti_join(testing)%>%
  select(id, any_of(features),TOC, sensor_datetime, collector)%>%
  mutate(id = ifelse(id == "archery_virridy", "archery", id)) #combine archery sites

```


# Model Training


## Set up model folds
Since we have relatively little data, we are going to use three CV folds. Create 3 separate folds and 3 models, with one fold used for validation on each model. Generally we are looking for:
- Each fold to have ~30-35% of the total data in validation
- Each validation set to have a similar TOC distribution as its training set (median, range)
- Each should have a mix of Upper Mainstem (SFM, PFAL, PBD), Upper Trib (CBRI, CHD, LBEA, PENN) & Lower Mainstem (salyer, udall, riverbend, elc, archery, riverbluffs)

```{r}
val_set_1 <- c( "sfm","chd","lbea","riverbluffs" )
 
val_set_2 <- c( "pbd","joei", "penn", "archery" )
 
val_set_3 <- c("pfal","udall","cbri","elc","salyer","riverbend")

# Function to create plots for a given fold
plot_fold <- function(fold_id, train_val) {
  val_set_name <- paste0("val_set_", fold_id)
  val_set <- train_val %>% filter(id %in% get(val_set_name))
  train_set <- train_val %>% filter(id %nin% get(val_set_name))
  
  prop_val <- round(nrow(val_set)/nrow(train_val), 3) * 100
  train_med <- round(median(train_set$TOC, na.rm = TRUE), 1)
  val_med <- round(median(val_set$TOC, na.rm = TRUE), 1)
  
  # Histogram
  hist <- ggplot(train_set, aes(x = TOC)) +
    geom_histogram(bins = 30, fill = "grey") +
    geom_histogram(data = val_set, aes(x = TOC), bins = 30, fill = "#E70870", alpha = 0.5) +
    labs(
      title = paste0("Fold Set ", fold_id, ": ", nrow(val_set), " samples (", prop_val, "%) in validation set"),
      subtitle = paste0("Validation Sites: ", paste(unique(val_set$id), collapse = ", ")),
      x = "TOC (mg/L)"
    ) +
    theme_minimal()
  
  # Boxplot
  box <- ggplot(
    train_set %>% mutate(set = "Train") %>% 
      bind_rows(val_set %>% mutate(set = "Validation")) %>% 
      mutate(set = factor(set, levels = c("Train", "Validation"))),
    aes(x = set, y = TOC, fill = set)
  ) +
    geom_boxplot() +
    scale_fill_manual(values = c("Train" = "#256BF5", "Validation" = "#E70870")) +
    labs(
      title = paste0("Train Median TOC: ",train_med, "\nVal Median TOC: ", val_med),
      x = "",
      y = "TOC (mg/L)",
      fill = "TV Group",
    ) +
    theme_minimal()
  
  # Combine plots
  ggarrange(hist, box, nrow = 1)
}

# plot all folds
ggarrange(plotlist = map(1:3, ~ plot_fold(.x, train_val)), ncol = 1)


#Update train_val with a fold_id column based on val_sets
train_val <- train_val %>%
  mutate(fold_id = case_when(
    id %in% val_set_1 ~ 1,
    id %in% val_set_2 ~ 2,
    id %in% val_set_3 ~ 3
  ))

```


### Visualize TOC distribution in training/validation vs testing sets

Currently, our 2025 data (testing) is skewed to higher TOC values (April- July) compared to the training/validation set. This is not ideal, but we will proceed with this split for now given limited data. Future data updates in December should help even this out as TOC decreases in the fall. 

```{r}
ggplot(train_val, aes(x = TOC, fill = collector)) +
  geom_histogram(bins = 30) +
  geom_histogram(data = testing, aes(x = TOC), bins = 30, fill = "#E70870", alpha = 0.5) +
  labs(
    title = paste0("Training/Validation Set: ", nrow(train_val), " samples                Testing Set: ", nrow(testing), " samples"),
    subtitle = paste0("Training TOC Range: ", round(min(train_val$TOC, na.rm = TRUE),2), " - ", round(max(train_val$TOC, na.rm = TRUE),2) , " mg/L (Median: ", round(median(train_val$TOC), 2), ")\nTesting TOC Range: ", round(min(testing$TOC, na.rm = TRUE),2), " - ", round(max(testing$TOC, na.rm = TRUE),2), " mg/L (Median: ", round(median(testing$TOC), 2),")"),
    x = "TOC (mg/L)"
  ) +
  theme_minimal()

```

## Feature Tuning 

### View correlations between features and target (TOC)
Most products with FDOM in them are going to have higher correlation with TOC, as well as seasonal (sin_doy) and flow related features. We want to be sure that we don't put too many of these in the model as calibration error or sensor drift could lead to compounding errors.

```{r}

#Looking at features across all data
modeling_data <- all_matches_norm %>%
  mutate(watershed = ifelse(id %in% c("joei", "cbri", "chd", "pfal","pbr",  "sfm", "pman", "pbd"), "UpperPoudre", "LowerPoudre"))%>%
  filter(!id %in% c("springcreek", "boxcreek"))

plot_list <- list()
for(i in features){
  
  plot_list[[i]] <- ggplot(modeling_data, 
                           aes(x = .data[[i]], y = TOC, color = id)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    stat_cor(
      method = "pearson",
      label.x.npc = "left", label.y.npc = 0.9,
      aes(label = paste("r =", ..r..)),
      color = "blue"
    ) +
    labs(x = i, y = "TOC (mg/L)") +
    facet_wrap(~watershed, scales = "free") +
    ROSS_theme
  
  
  if(i %in% c("turb_median_12hr", "turb_mean_12hr")){
    plot_list[[i]] <- plot_list[[i]] + xlim(0,100)
  }
  if(i %in% c("f_temp")){
    plot_list[[i]] <- plot_list[[i]] + xlim(0,2.5)
  }
}

# Combine with patchwork
g <- wrap_plots(plot_list, guides = "collect")+
  plot_annotation(
    caption = "Color is Site, Dashed line is linear regression fit with adjusted R2") 

g

#plot just seasonal vars
wrap_plots(
  plots = list(
    plot_list$sin_doy,
    plot_list$hourly_canyon_flow_cfs,
    plot_list$daily_canyon_flow_cfs
  ), ncol = 1,guides = "collect" ) +
  plot_annotation(
    caption = "Color is Site, Dashed line is linear regression fit with adjusted R²"
  ) &
  theme(legend.position = "right")  

```

### Pruning Features with high correlations 

We want to avoid using too many highly correlated features in the model, as this can lead to overfitting and instability. We will look at the correlation matrix and remove some of the more redundant features.

```{r}
# get a correlation matrix across all features
corr_matrix <- modeling_data %>% 
  select(all_of(features)) %>% 
  cor(use = "pairwise.complete.obs")

# we just want half the matrix, so recode those in the lower triangle to NA
corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter for those with correlation > 0.9
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.9) %>%
  arrange(desc(abs(correlation))) %>% 
  # remove redundant vars
  filter(feature1 %nin% c("FDOM", "doy"))

# look at the list and make some educated decisions about features
high_corr_pairs

fewer_features <- features[features %nin% c("FDOM", "doy", 
                                            # remove vars with high correlation 
                                            # with others
                                            "Temp", "f_log",
                                            "Sensor_Turb",
                                            "hourly_canyon_flow_cfs",
                                            "fdom_max_6hr", "temp_max_24hr")]

# do the correlation matrix again to make sure that is okay
corr_matrix <- modeling_data %>% 
  select(all_of(fewer_features)) %>% 
  cor(use = "pairwise.complete.obs")

corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.85) %>%
  arrange(desc(abs(correlation)))

high_corr_pairs

```

`high_corr_pairs` has features we should be cautious of using together. We will try to avoid using more than one of these in the same model.


### Testing Features

First, we will start with our base model (live sensor data only) and then test adding in other features to see if they improve performance. We will use the default hyperparameters below for all models to keep things consistent. Once we have tested all the features together, we will want to trim it down a bit so we are not using too many highly correlated features or overfitting the model due to lack of data

#### Function to test features & defaults
This just runs a basic xgboost model with the specified features and default hyperparameters. It returns feature importance, shap values, and best iteration info for each fold.

```{r}
test_features <- function(train_val,
                          features_to_test,
                          target,
                          default_hyper_params,
                          weight_fun = toc_high_weights) {
  # double check that there are no duplicates in features
  features_to_test <- unique(features_to_test)
  
  #create blank dataframes to store importance/shap values
  importance_df <- tibble(Feature = features_to_test)
  shap_val_df <- tibble(Feature = features_to_test)
  best_msgs <- tibble()
  models <- list()
  
  for(i in unique(train_val$fold_id)){
    
    #train val split
    train_data <- train_val %>%
      filter(fold_id != i) %>%
      select(any_of(c(features_to_test, target, "id", "sensor_datetime", "collector")))
    
    val_data <- suppressMessages(
      train_val %>%
        anti_join(train_data)
    )
    
    #weight data
    w_train <- weight_fun(train_data)
    w_val   <- weight_fun(val_data)
    
    #prep matrices
    dtrain <- xgb.DMatrix(
      data = as.matrix(train_data[, features_to_test]),
      label = train_data[[target]],
      weight = w_train
    )
    
    dval <- xgb.DMatrix(
      data = as.matrix(val_data[, features_to_test]),
      label = val_data[[target]],
      weight = w_val
    )
    
    watchlist <- list(train = dtrain, eval = dval)
    
    #Train model with defaults
    model <- xgb.train(
      params = default_hyper_params,
      data = dtrain,
      nrounds = 10000,
      watchlist = watchlist,
      early_stopping_rounds = 1000,
      print_every_n = 1000,
      verbose = 0,
  nthread = 1
    )
    #Save model
    models[[i]] <- model
    #Save best msg
    best_msg <- tibble(
      fold_id = i,
      best_iteration = model$best_iteration,
      best_msg = model$best_msg
    ) %>%
      #clean up message
      mutate(
        train_rmse = str_extract(best_msg, "(?<=train-rmse:)\\d+\\.\\d+"),
        eval_rmse  = str_extract(best_msg, "(?<=eval-rmse:)\\d+\\.\\d+"),
        train_rmse = as.numeric(train_rmse),
        eval_rmse  = as.numeric(eval_rmse)
      )%>%
      select(-best_msg)
    #best message
    best_msgs <- bind_rows(best_msgs, best_msg)%>%
      distinct()
    
    
    #Feature importance
    importance_df <- importance_df%>%
      left_join(xgb.importance(feature_names = features_to_test, model = model)%>%
                  dplyr::rename_with( #fix column names
                    ~ glue::glue("{.x}_fold_{i}"),
                    .cols = c(Gain, Cover, Frequency)
                  ), by = "Feature")
    
    # SHAP (Training)
    shap_long_train <- shap.prep(xgb_model = model, X_train = as.matrix(train_data[, features_to_test]))
    shap_vals_train <- shap.values(xgb_model = model, X_train = dtrain)$mean_shap_score %>%
      as.data.frame() %>%
      tibble::rownames_to_column("Feature") %>%
      rename(train_mean_abs_shap = ".")
    
    # SHAP (Validation)
    shap_long_val <- shap.prep(xgb_model = model, X_train = as.matrix(val_data[, features_to_test]))
    shap_vals_val <- shap.values(xgb_model = model, X_train = dval)$mean_shap_score %>%
      as.data.frame() %>%
      tibble::rownames_to_column("Feature") %>%
      rename(val_mean_abs_shap = ".")
    
    shap_val <- shap_vals_val%>%
      left_join(shap_vals_train, by = "Feature")%>%
      dplyr::rename_with( #fix column names
        ~ glue::glue("{.x}_fold_{i}"),
        .cols = c(val_mean_abs_shap, train_mean_abs_shap)
      )
    shap_val_df <- shap_val_df%>%
      left_join(shap_val, by = c("Feature"))
    
    # p_train <- shap.plot.summary(shap_long_train)
    # p_val <- shap.plot.summary(shap_long_val)
    # 
    
  }
  
  #output
  list(
    model = models,
    features = features_to_test, 
    best_msg = best_msgs,
    model_importance = importance_df,
    shap_values = shap_val_df
  )
  
}

#xgboost hyper parameters

default_hyper_params <- list(
  objective        = "reg:squarederror",
  eval_metric      = "rmse",
  eta              = 0.05,
  gamma            = 0.6,
  alpha            = 0,
  lambda           = 1,
  max_depth        = 4,
  subsample        = 0.5,
  colsample_bytree = 0.5,
  min_child_weight  = 2
)

# Weighting function for TOC
toc_high_weights <-  function(df) {
  ifelse(df$TOC <= 4,1,                  # flat weight for low TOC
         1 + 3 *log1p(df$TOC - 4))  # grows slowly, no cap needed
}


target = "TOC"
```

#### Run across folds testing different features to see which improve performance

Here I run the function and am going to look at feature importance and shap values to see which features are most useful, then trim features back to reduce our correlative and overfitting issues. 

```{r}
#starting with all features then trimming down to more highly correlated ones
# general_features <-  setdiff(features, c("FDOM", "fdom_max_6hr", #repeats due to FDOM Correction
#                                          "sin_doy", "doy", "daily_canyon_flow_cfs", #seasonal
#                                          "temp_f","temp_x_f", "f2_temp", #sensor products
#                                          "f_log", "f_x_sc", "f_turb", "turb_f", "temp_x_turb", #sensor products
#                                          "sc_sd_24hr", "temp_max_24hr", "temp_mean_24hr", "temp_mean_48hr", "turb_median_12hr" #lagged variables
# ))

#set aside sensor data that are generally useful
general_features <- c("FDOMc","Sensor_Turb","Sensor_Cond", "Chl_a", "Temp")

feature_results <- map(
  #trying all the combinations of features we think might be useful (commented out were tested but found to not improve base model)
  list(
  gen_feat = general_features, # sensor values only
  #fdom_corr_max_6hr = c(general_features, "fdom_corr_max_6hr"), # short term max FDOM, might help with receeding spikes or daily variation
  #f_sc = c(general_features, "fdom_corr_max_6hr", "f_sc"), # adding FDOM/sc ratio, high FDOM + low SC = high TOC (snowmelt)
  #f_x_turb = c(general_features, "fdom_corr_max_6hr", "f_sc", "f_x_turb"), # adding FDOM * turbidity, high FDOM + high turb = high TOC (stormflow/snowmelt), might help with high turb events
  #sin_doy = c(general_features, "fdom_corr_max_6hr", "f_sc", "f_x_turb", "sin_doy"), # seasonal signal but not linked to exact DOY
  #temp_x_f = c(general_features, "fdom_corr_max_6hr", "temp_x_f", "f_sc", "f_x_turb", "sin_doy"), # temp * FDOM, high temp + high FDOM = high TOC (summer/autumn allochthonous inputs)
  #temp_x_turb = c(general_features, "fdom_corr_max_6hr", "temp_x_turb", "temp_x_f", "f_sc", "f_x_turb", "sin_doy"), # temp * turbidity, high temp + high turb = high TOC (summer stormflow?)
 # turb_median_12hr = c(general_features, "fdom_corr_max_6hr", "temp_x_turb", "temp_x_f", "f_sc", "f_x_turb", "sin_doy", "turb_median_12hr"), # receeding storm flow proxy
  #daily_canyon_flow_cfs = c(general_features, "fdom_corr_max_6hr", "temp_x_turb", "temp_x_f", "f_sc", "f_x_turb", "sin_doy", "turb_median_12hr", "daily_canyon_flow_cfs"), # climate (flow) conditions across basin (may want to play with increasing/decreasing slopes?)
  # Tested but did not improve model
  # sc_var_24hr = c(general_features, "sc_var_24hr"), # proxy for recent depth changes (sc = groundwater/surface water contributions)
  #temp_mean_24hr = c(general_features, "temp_mean_24hr"), # recent temp changes
  #temp_mean_48hr = c(general_features, "temp_mean_48hr"), #longer term temp changes
  # f2_temp = c(general_features, "f2_temp"), #FDOM^2 * temp, very high FDOM + low temp = very high TOC 
  #f_x_sc = c(general_features, "f_x_sc"), # FDOM * SC, high FDOM + high SC = low TOC (groundwater influence) probably ended up getting confused with so much variation in SC
  #f_turb = c(general_features, "f_turb"), # FDOM / turbidity, high FDOM + low turb = high TOC (snowmelt in high elev reaches), redundant with f_x_turb
  #turb_f = c(general_features, "turb_f"), # turbidity / FDOM,  redundant with f_x_turb
 fewer_features = c(fewer_features), 
 fewer_features_trimmed = c(fewer_features)%>%setdiff(., c("turb_median_12hr","temp_x_f", "temp_mean_48hr","temp_x_turb", "turb_f", "sc_sd_24hr", "temp_mean_24hr", "temp_f"  ))
 
  #Trimming down to most important features from all features (based on importance and shap values) to reduce overfitting
 # trimmed_features = c(general_features, "fdom_corr_max_6hr", "f_sc", "f_x_turb", "sin_doy" , "turb_median_12hr", "daily_canyon_flow_cfs")
) , ~test_features(
  train_val = train_val,
  features_to_test = .x,
  target = "TOC",
  default_hyper_params = default_hyper_params
))
```

#### Feature Set Results

##### Weighted RMSE

We compare across folds to see if there is big differences in folds (ie unbalanced data) and also compare each feature set against "base" model of general features (sensors). Looking for features that make meaningful improvements in RMSE across all folds. If our RMSE improves but that is related to only a single fold, we may want to reevaluate the feature set as it is likely overfitting to a given site(s). 
Keep in mind that this is weighted RMSE so it may be higher than unweighted RMSE you see in other places.

```{r}
# Highlight differences between folds in rmse
map_dfr(feature_results, ~{
  # Compute feature set description
  unique_features <- setdiff(.x$features, general_features)
  final_features <- c("general_features", unique_features)
  feature_desc <- paste(final_features, collapse = ", ")
  
  # Take best_msg tibble and rename eval_rmse by fold
  .x$best_msg %>%
    select(fold_id, eval_rmse) %>%
    mutate(fold_col = paste0("rmse_fold_", fold_id)) %>%
    select(-fold_id) %>%
    mutate(eval_rmse = round(eval_rmse, 2)) %>%
    pivot_wider(names_from = fold_col, values_from = eval_rmse) %>%
    mutate(
      features = feature_desc,
      mean_rmse = round(mean(c_across(starts_with("rmse_fold_")), na.rm = TRUE), 2)
    )%>%
    select(features, mean_rmse, rmse_fold_1, rmse_fold_2, rmse_fold_3)
  
}, .id = "feature_set") %>%
#check if new feature is better than baseline
  mutate(
    baseline_rmse = mean_rmse[feature_set == "gen_feat"],
    improves_rmse = if_else(mean_rmse < baseline_rmse, round(mean_rmse - baseline_rmse, 2), 0)
  ) %>%
  select(-baseline_rmse)%>%
  arrange(improves_rmse)
```

##### Feature Importance

This comes directly from the xgboost package and is based on Gain, Cover, and Frequency of splits. We summarize across folds to see which features are most important across all folds. 

```{r}
#table with all importances
importance_tables <- map(feature_results, ~{
  imp <- .x$model_importance
  
  # Identify fold columns dynamically
  gain_cols <- grep("^Gain_fold_", names(imp), value = TRUE)
  cover_cols <- grep("^Cover_fold_", names(imp), value = TRUE)
  freq_cols <- grep("^Frequency_fold_", names(imp), value = TRUE)
  
  imp %>%
    mutate(
      mean_gain = rowMeans(across(all_of(gain_cols)), na.rm = TRUE),
      mean_cover = rowMeans(across(all_of(cover_cols)), na.rm = TRUE),
      mean_frequency = rowMeans(across(all_of(freq_cols)), na.rm = TRUE)
    ) %>%
    mutate(new_feat = ifelse(Feature %in% general_features, "", "*")) %>%
    select(Feature, new_feat, mean_gain, mean_cover, mean_frequency) %>%
    
    arrange(desc(mean_gain))
}, .id = "feature_set")

#Get features with less than 0.02 gain on average across folds in fewer_features to consider removing
cat("Features with mean gain < 0.02 in 'fewer_features':\n")
importance_tables$fewer_features%>%
  filter(mean_gain < 0.02)

```


##### Shapley Values

Let's look at training and validation shap values for features across all folds. This gives us a sense of which features are most important in the model predictions. We summarize across folds to see which features are most important across all folds. In general we would expect features to have similar importance in training and validation, if a feature has high importance in training but low in validation it may be overfitting to the training data and we may want to re-evaluate. 

```{r}
#table with all shap values
shap_summary <- map_dfr(feature_results, ~{
  shap <- .x$shap_values
  
  val_shap_cols <- grep("^val_mean_abs_shap_fold_", names(shap), value = TRUE)
  train_shap_cols <- grep("^train_mean_abs_shap_fold_", names(shap), value = TRUE)
  
  shap %>%
    mutate(
      mean_val_shap = round(rowMeans(across(all_of(val_shap_cols)), na.rm = TRUE), 3),
      mean_train_shap = round(rowMeans(across(all_of(train_shap_cols)), na.rm = TRUE), 3)
    ) %>%
    mutate(val_train_diff = mean_val_shap - mean_train_shap) %>%
    select(Feature, mean_val_shap, mean_train_shap, val_train_diff)
}, .id = "feature_set")

shap_long <- shap_summary %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "type",
    values_to = "mean_shap"
  ) %>%
  mutate(
    type = recode(type,
                  mean_val_shap = "Validation",
                  mean_train_shap = "Training"),
    feature_set = factor(feature_set, levels = unique(feature_set))
  ) %>%
  mutate(Feature = factor(Feature, levels = c( rev(feature_results$fewer_features$features),"Sensor_Turb","Temp")))

ggplot(shap_long, aes(x = feature_set, y = Feature, fill = mean_shap)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_viridis_c(
  option = "magma",
  limits = c(0, 1),
  name = "Mean |SHAP|",
  direction = -1  # reverses so dark = low, bright = high (optional)
)+
  facet_wrap(~type) +
  labs(
    x = "Feature Set",
    y = "Feature",
    title = "Mean SHAP Values by Feature and Feature Set"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )



```

### TV performance of best feature set

After trimming down a couple of features (minimal improvement/low shap values), we should take a quick look at model performance in TV space with our base model (default hyperparameters)

```{r}

best_models <- feature_results$fewer_features_trimmed$model

feature_tv_plots <- map(1:3, ~plot_tv_fold_perf(fold_set = .x,
                             train_val_df = train_val,
                             fold_model = best_models[[.x]], # a list where each element is a model for that fold
                             target_col = "TOC", units = "mg/L")
                            )
ft_tv_plot<- wrap_plots(feature_tv_plots, ncol = 1, guides = "collect")&
  theme(legend.position = "bottom")
ft_tv_plot


```


## Hyperparameter Tuning
Now that we have a feature set, we can do some hyperparameter tuning to see if we can improve model performance. This will take a while to run, so we will save the results and load them in later.

### Setup

```{r}

folds <- tibble(
  fold = 1:3,
  val_ids = list(val_set_1, val_set_2, val_set_3)
)
features <- best_models[[1]]$feature_names

# Reduce tune grid for reruns (ie. General best params)
tune_grid = expand.grid(
  nrounds = 10000,
  max_depth = c(2, 3, 4),
  eta = c( 0.01, 0.1),
  gamma = c(0.8),
  lambda = c(1,10),
  alpha = c(0, 1),
  colsample_bytree = c(0.5,0.8),
  min_child_weight = c(2, 4),
  subsample = c(0.5,0.8)
)

```

### Run hyperparameter tuning

```{r}
set.seed(123)
hyperp_results <- xgboost_hyperparameter_tuning(
  data = train_val %>% select(TOC, id, any_of(features)), 
  weights = toc_high_weights,
  tune_grid = tune_grid,
  target_col = "TOC",
  site_col = "id",
  fold_ids =  folds, 
  units = "mg/L")

```

### Choosing best grids

### Fold 1

```{r}

hyperp_results[[1]]$perf_dist
hyperp_results[[1]]$eval_plot
hyperp_results[[1]]$tv_plot

best_fold_1_grid <- 136
```

### Fold 2

```{r}
hyperp_results[[2]]$perf_dist
hyperp_results[[2]]$eval_plot
hyperp_results[[2]]$tv_plot

best_fold_2_grid <- 165

```

### Fold 3

```{r}
hyperp_results[[3]]$perf_dist
hyperp_results[[3]]$eval_plot
hyperp_results[[3]]$tv_plot

best_fold_3_grid <- 77

```

## Extract best models for each fold

```{r}

full_model <- map(1:3, function(i) {
  
  best_grid <- paste0("grid_", get(glue("best_fold_{i}_grid")))
  model_info <- hyperp_results[[i]]$grid_results[[best_grid]]
  list(
    model = model_info$model,
    perf = model_info$perf,
    params = model_info$params,
    training_data = train_val %>% filter(fold_id != i),
    validation_data = train_val %>% filter(fold_id == i)
  )
})

```

### Checking Results


#### Feature Importance

```{r}
importance_list <- lapply(seq_along(full_model), function(i) {
  fold_model <- full_model[[i]]$model
  features <- fold_model$feature_names
  importance_matrix <- xgb.importance(feature_names = features, model = fold_model)
  importance_df <- as.data.frame(importance_matrix)
  importance_df$Fold <- paste("Fold", i)
  return(importance_df)
})

print(importance_list)
```

#### Shapley Importance

```{r}

# Function to plot SHAP values for a given fold
plot_shap <- function(fold_set, training_data, fold_model) {
  #get features from model
  features <- fold_model$feature_names
  # Predictions on validation/training set
  dtrain <- as.matrix(training_data[, features])
  
  shap_values <- shap.values(xgb_model = fold_model, X_train = dtrain)
  
  # Long-format SHAP values
  shap_long <- shap.prep(
    shap_contrib = shap_values$shap_score,
    X_train = dtrain
  )
  #return shapley plot
  shap.plot.summary(shap_long)
}

map(1:3, ~ plot_shap(
  fold_set = .x,
  fold_model = full_model[[.x]]$model,
  training_data = full_model[[.x]]$training_data
))%>%
  wrap_plots(., ncol = 1, guides = "collect") &
  theme(legend.position = "bottom")

```

### Plots of TV Performance
Now let's look at our performance in TV space.

```{r}

hyperparam_tv_plots <- map(1:3, ~plot_tv_fold_perf(fold_set = .x,
                             train_val_df = train_val,
                             fold_model = full_model[[.x]]$model, # a list where each element is a model for that fold
                             target_col = "TOC", units = "mg/L")
                            )

hp_tv_plot<- wrap_plots(hyperparam_tv_plots, ncol = 1, guides = "collect")&
  theme(legend.position = "bottom")
hp_tv_plot
#ggsave("data/upper_clp_dss/figures/toc_modeling/performance/toc_model_fold_validation_performance.png", width = 14, height = 10)

```

#### Against base hyperparameters

We should hope that all our tuning efforts have improved the model!! 

```{r}
# Compare best hyperparameter model to default hyperparameter model
final_plot <- wrap_plots(
  ft_tv_plot,
  hp_tv_plot,
  ncol = 2,
  guides = "collect"
) &
  theme(legend.position = "bottom")
final_plot <- final_plot + 
  plot_annotation(
    title = "Model Performance Comparison: Default (left) vs Tuned Hyperparameters (right)"
  )

final_plot

```

Looking at the plots, we can see that the tuned hyperparameter model has improved performance in all folds compared to the default hyperparameter model. The validation RMSE has decreased in all folds but it foes look like model fold three is slightly less accurate at the higher ends (possible overtraining?). Overall, this is a good sign that our tuning efforts have paid off!

## Save as "in_dev" and remove unneeded objects

If we want to test additional ideas, we can just run 
```{r}
saveRDS(full_model, file = here("data", "derived", "models","TOC_realtime", "in_dev", paste0("ross_only_toc_xgboost_models_full_",version,".rds")))

rm(list = setdiff(ls(), c("train_val", "testing", "full_model","version","id_cols", "all_sensor_data_wide", 
                          "apply_traning_scale","plot_tvt_fold_perf","plot_tvt_ensemble_perf", "apply_ml_bias_correction" )) )
```


# Bias Correction
Based on the methods outlined in this paper (https://www.sciencedirect.com/science/article/pii/S1364815221000499), we will apply a bias correction to our model predictions to improve accuracy on high and low values (where we do worse). We will first calcate the bias corrections on our validation set and if they seem to work, they will be applied to the predictions from the testing set to hopefully improve performance there as well.

## Function for bias correction methods

```{r}
bias_correct <- function(train_val_df, testing_df, fold_model, fold_num, target_col = "TOC") {

  #Get features
  features <- fold_model$feature_names
  # compute min and max TOC across ALL input data for ML2-res scaling
  toc_min <- min(train_val_df[[target_col]], testing_df[[target_col]])
  toc_max <- max(train_val_df[[target_col]], testing_df[[target_col]])
  
  #Get validation data
  val_data <- train_val_df %>%
    filter(fold_id == fold_num)%>%
    mutate(
      #Make predictions on set
      predicted = predict(
        fold_model,
        newdata = as.matrix(select(., all_of(features)))), 
      group = "Validation"
    )
  val_preds <- val_data$predicted
  val_obs <- val_data[[target_col]]
  
  # Function to perform Empirical Distribution Mapping (EDM) bias correction
  # For each guess in pred_vals: find its percentile in ref_pred_vals; then take that percentile quantile of ref_obs_vals
  #Def: 
  #pred_vals: our model predictions (in this case, from validation or testing set)
  #ref_pred_vals: reference predictions (in this case, again from validation set)
  #ref_obs_vals: reference observations (in this case, observed TOC from validation set)
  ecdf_pred <- ecdf(val_preds) #calculate the empirical cumulative distribution function (ecdf) of the reference predictions
  
  val_corr <- val_data %>%
    #get the percentiles of our predictions based on the reference predictions, then convert to quantiles of observed values
    mutate(edm = quantile(val_obs, probs = ecdf_pred(predicted), type = 8))
  
  #ROE correction
  mroe = cov(val_preds, val_obs) / var(val_preds)
  broe = mean(val_obs) - mroe * mean(val_preds)
  
  val_corr <- val_corr%>%
    mutate(roe = predicted * mroe + broe)
  
  #LTF correction
  
  mltf = cov(val_preds, val_corr$edm) / var(val_preds)
  bltf = mean(val_obs) - mltf * mean(val_preds)
  
  val_corr <- val_corr%>%
    mutate(ltf = predicted * mltf + bltf)
  
  #Z-score correction
  mzz = sqrt(var(val_obs))/sqrt(var(val_preds))
  bzz = mean(val_obs) - mzz * mean(val_preds)
  
  val_corr <- val_corr%>%
    mutate(zz = predicted * mzz + bzz)
  
  #ml2-res correction
  X_res <- val_data %>% 
    mutate(predicted = rescale(val_preds, to = c(0,1), from = c(toc_min,toc_max)))%>%
    select(all_of(features), predicted) %>% 
    as.matrix()
  #residuals
  y_res <- val_obs - val_preds
  
  #Xgboost inputs/settings for ML2 Res model
  dres_train <- xgb.DMatrix(data = X_res, label = y_res)
  default_hyper_params <- list(
    objective        = "reg:squarederror",
    eval_metric      = "rmse",
    eta              = 0.1,
    gamma            = 0.6,
    alpha            = 0,
    lambda           = 1,
    max_depth        = 4,
    subsample        = 0.5,
    colsample_bytree = 0.5,
    min_child_weight  = 2
  )
  #train model on original features + predicted values  and try to predict residuals
  ml2_res_model <- xgb.train(default_hyper_params, data = dres_train, nrounds = 100,verbose = 0)
  #predict residuals on validation set
  val_corr <- val_corr%>%
    mutate(ml2_res_pred = predict(ml2_res_model, 
                                  xgb.DMatrix(as.matrix(val_data %>% 
                                                          mutate(predicted = rescale(val_preds, to = c(0,1), from = c(toc_min,toc_max)))%>%
                                                          select(all_of(features), predicted)))), 
           #add in residual predictions to original predictions
           ml2 = predicted + ml2_res_pred, 
           residuals = !!sym(target_col) - predicted
    )

  # Apply to testing set
  test_corr <- testing_df%>%
    select(id, all_of(features),!!sym(target_col))%>%
    mutate(
      #apply model to testing set
      predicted = predict(
        fold_model,
        newdata = as.matrix(select(., all_of(features)))
      ), 
      #EDM
      edm = quantile(val_obs, probs = ecdf_pred(predicted), type = 8),
      #ROE
      roe = predicted * mroe + broe,
      #LTF
      ltf = predicted * mltf + bltf,
      #Z-score
      zz = predicted * mzz + bzz,
      #ML2-res
      ml2_res_pred = predict(
        ml2_res_model,
        xgb.DMatrix(as.matrix(
          select(., all_of(features)) %>% mutate(predicted = rescale(predicted, to = c(0,1), from = c(toc_min,toc_max))))
        )
      ), 
      ml2 = predicted + ml2_res_pred,
      #calculate residuals
      residuals = !!sym(target_col) - predicted, 
      group = "Test"
    )
  #return as joined dataset
  bind_rows(val_corr, test_corr)
}
```

## Apply function to each fold and visualize results on Validation Set

Results are visualized in the observed vs predicted space. In the paper above, they did not do TVT and so what they refer to as "Training" is actually our validation set here. We will look at each fold separately and see how the bias correction methods perform. We will also compute RMSE, MAE, and Bias metrics for each method and fold to quantify performance. 

```{r}

fold_corrs <- map_dfr(1:3, ~bias_correct(train_val_df = train_val, 
                                         testing_df = testing, 
                                        fold_model = full_model[[.x]]$model,
                                        fold_num = .x))
fold_corrs_val <- fold_corrs%>%
  filter(group == "Validation")%>%
  pivot_longer(cols = c( "edm", "roe", "ltf", "zz", "ml2"),
               names_to = "method",
               values_to = "corrected_pred")%>%
    mutate(fold = paste0("Fold ", fold_id))%>%
    select(id, residuals, predicted, obs = TOC, fold, group, method, corrected_pred)

#order the method factor levels
fold_corrs_val$method <- factor(fold_corrs_val$method, levels = c("edm", "roe", "ltf", "zz", "ml2"))
# Compute rmse, mae and bias metrics for each method and fold

metrics_summary <- fold_corrs_val %>%
  group_by(fold, method) %>%
  summarise(
    rmse_pred = rmse(obs, predicted),
    rmse_corr = rmse(obs, corrected_pred),
    mae = mae(obs, corrected_pred),
    bias = bias(obs, corrected_pred),
    .groups = "drop"
  ) %>%
  mutate( # creat label for plot
    label = paste0("Pre-RMSE: ", round(rmse_pred, 2), "\n",
                   "Post-RMSE: ", round(rmse_corr, 2)), 
    improvement = if_else(rmse_corr <= rmse_pred, "RMSE Improved", "RMSE Not Improved")
  )

ggplot(fold_corrs_val, aes(x = obs)) +
  geom_point(aes( y = corrected_pred, color = "Corrected Prediction"), size = 3) +
  geom_point(aes(y = predicted, color = "Original ML Prediction"), alpha = 0.7, size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    # Add RMSE annotations
  geom_text(
    data = metrics_summary,
    aes(
      x = 2.5,  y = 8,
      label = label, 
      color = improvement
    ),
    inherit.aes = FALSE,
    hjust = 0, vjust = 1,
    size = 4
  ) +
  scale_color_manual(
    values = c("RMSE Improved" = "green2", "RMSE Not Improved" = "red",
               "Corrected Prediction" = "#E70870", "Original ML Prediction" = "#256BF5")
  ) +
  #facetgrid by method and fold
  facet_grid(vars(fold), vars(method)) +
  labs(
    x = "Observed TOC (mg/L)",
    y = "Predicted TOC (mg/L)",
    title = "Comparison of Bias Correction Methods on Validation Set", 
    subtitle = "Blue points are original ML predictions;", 
    color = "Legend", 
  ) +
  theme_bw() +
  theme(legend.position = "bottom") +
  xlim(2, 9.5) +
  ylim(2, 9.5)


```


# Evaluate on Test Set

We should only do this once we feel good about our model and are ready to see how it performs on the unseen test set. This should only be run after big changes to the model or if we get new data (ie The model should be in a "final/working" version)

## Visualize Train Val Test

### Individual folds

```{r}
plots<- map(1:3, ~{
    plot_tvt_fold_perf(train_val, testing, fold_set = .x, fold_model = full_model[[.x]]$model, target_col = "TOC", units = "mg/L", subtitle_arg = "CLP Samples only")
})
plots

  # ggsave(glue(here("docs", "uclp_dss", "figs", "toc_modeling", "performance", "ross_only_model_fold_{i}.png")),
  #        plot = p, width = 10, height = 6, dpi = 500)

```

### All folds together

```{r}
# Single plot with all folds
wrap_plots(plots, ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")
# ggsave(here("docs","uclp_dss", "figs", toc_modeling","performance",paste0("ross_only_model_all_folds_TVT", version, ".png"),
#        width = 14, height = 8, dpi = 500)
```

### Ensemble predictions

```{r}
# Plot ensemble
ensemble <- plot_tvt_ensemble_perf(train_val_df = train_val,
                                   test_df = testing,
                                   fold_models = map(full_model, "model"),
                                   target_col = "TOC",
                                   units = "mg/L",
                                   subtitle_arg = "CLP Samples only")
plot(ensemble)
# ggsave(here("docs","uclp_dss", "figs", toc_modeling","performance",paste0("ross_only_model_ensemble", version, ".png"),
#        plot = ensemble, width = 12, height = 10, dpi = 500)
```

## Visualize performance in timeseries space

COMING SOON! Need to functionalize aspects of 04 script to make this concise enough

# Save final models
We will extract just the model objects from the list to save space.

```{r}

small_models <- lapply(full_model, function(x) {
  m <- x$model   # extract model (xgboost object)
  
  #remove large unnecessary elements 
  if ("trainingData" %in% names(m)) m$trainingData <- NULL
  if ("call" %in% names(m)) m$call <- NULL
  if ("terms" %in% names(m)) m$terms <- NULL
  
  return(m)
})
saveRDS(small_models, file = here("data", "derived", "models","TOC_realtime", "in_dev", paste0("ross_only_toc_xgboost_models_light_",version,".rds")))

```







