---
title: "Real-Time TOC model for 2026"
author: "Sam Struthers, B Steele & Matt Ross-ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

# Set Up

## Package Load

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("src/setup_libraries.R")
source("src/setup_ross_theme.R")

#these are just packages for testing within this script
library(gridExtra)
library(grid)
library(Ckmeans.1d.dp)
library(DiagrammeR)
library(Metrics)
library(glue)
library(SHAPforxgboost)
set.seed(123)
#set version for saving (update when changes are significant)
version = "20261010"

#### Data setup functions ####
#load grab sample data function
source("src/collation/load_grab_sample_data.R")
#load sensor data
source("src/collation/load_sensor_data.R")
#create lagged parameters
source("src/modeling/calc_lagged_parameter.R")
#create matchup dataset
source("src/modeling/create_grab_sensor_matchup.R")
# apply fdom temperature corrections
source("src/sensor_qaqc/apply_fdom_corrections.R")
##### Modeling functions #####
# function to apply sensor transformations for TOC modeling
source("src/modeling/apply_sensor_transformations_toc.R")
#function to apply training scale
source("src/modeling/apply_training_scale.R")
#function to look at xgboost feature tuning
source("src/modeling/xgboost_feature_selection.R")
# function to tune xgboost hyperparameters
source("src/modeling/xgboost_hyperparameter_tuning.R")

#### Performance and plotting functions #####

#function to plot performance of each fold
source("src/plotting/plot_tv_fold_perf.R")
#function to plot performance of each fold
source("src/plotting/plot_tvt_fold_perf.R")
#function to plot performance of each fold
source("src/plotting/plot_train_test_ensemble_perf.R")
#function to plot model timeseries
source("src/plotting/plot_model_timeseries.R")

# Load CDWR credentials to get flow data (without hitting data limits)
cdwr_creds <- read_yaml(here("creds", "CDWRCreds.yml"))$api_key

#set warnings to F
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.dim = c(12,10))
```

## Read in sensor dataset

Here, we read in the sensor data, apply min-max scaling defined in the `apply_sensor_transformations_toc()` function, and store the dataset for use later

```{r}

sensor_data <- load_sensor_data()%>% 
  mutate(DT_hourly_round = floor_date(DT_round, "hour")) %>%
  distinct(site, DT_hourly_round, parameter, .keep_all = T) %>%
  # remove parameters that are not consistently available across sites
  filter(parameter %nin% c("Chl-a ugl","TAL PC Fluorescence","ORP", "pH", "DO"))%>%
  #Remove a FC sites where there is no FDOM sensor
  filter(site %nin% c("pbr_fc", "pman_fc"))

archery_virridy_sub <- sensor_data%>%
  filter(c(between(DT_hourly_round, ymd_hms("2023-09-01 00:00:00", tz = "MST"), ymd_hms("2024-12-31 23:59:59", tz = "MST")) & str_detect(site, "archery_virridy")))%>%
  filter(parameter %in% c("Turbidity", "FDOM Fluorescence", "Chl-a Fluorescence"))%>%
  mutate(site = gsub(pattern = "_virridy", replacement = "", x = site))

riverbend_cottonwood_sub <- sensor_data%>%
  filter(c(between(DT_hourly_round, ymd_hms("2023-09-01 00:00:00", tz = "MST"), ymd_hms("2024-05-31 23:59:59", tz = "MST")) 
           & str_detect(site, "riverbend_virridy|cottonwood_virridy")))%>%
  filter(parameter %in% c( "FDOM Fluorescence", "Chl-a Fluorescence"))%>%
  mutate(site = gsub(pattern = "_virridy", replacement = "", x = site))

sensor_prepped <- sensor_data%>%
  filter(!c(between(DT_hourly_round, ymd_hms("2023-09-01 00:00:00", tz = "MST"), ymd_hms("2024-12-31 23:59:59", tz = "MST")) & str_detect(site, "archery") & parameter == "Chl-a Fluorescence"))%>%
  bind_rows(archery_virridy_sub)%>%
  bind_rows(riverbend_cottonwood_sub)%>%
  filter(site %nin% c("archery_virridy", "riverbend_virridy", "cottonwood_virridy"))%>%
  pivot_wider(id_cols = c("site","DT_hourly_round"), names_from = "parameter", values_from = "hourly_median")%>%
  mutate(Turbidity = if_else(Turbidity == 0, 0.1, Turbidity))%>% #replace 0 turbidity with 0.1 to be able to do turbidity corrections and log transformations
  #remove rows where everything other than site and DT_round are NA
  filter(!if_all(-c(site, DT_hourly_round), is.na))%>%
  #applying FDOM temperature & turbidity correcting/standardization
  apply_fdom_corrections(fdom_col = "FDOM Fluorescence",
                         temp_col = "Temperature",
                         turb_col = "Turbidity",
                         fdom_temp_col = "FDOMc",
                         fdom_turb_col = "FDOM_turb_corr",
                         fdom_final_col = "FDOM_final_corr")

```

## Calculate additional variables on sensor data

Things that might be helpful:

Lagged parameters

- Recent SC variation, i.e. change in depth or water source
- Recent (12 hr) turbidity median, as a proxy for recent storm events that could be driving TOC up  *Maybe impacted by drift affected values*
- Recent (6hr) FDOM median & max, as a proxy for recent changes in OM or storms with OM pulse

Interaction Terms
- FDOM X Turbidity (fdom_x_turb): A proxy for turbidity interference in FDOM signal, rather than correcting for turbidity. 
- FDOM / Turbidity (fdom_turb): Another proxy measurement for FDOM and turbidity interactions
- FDOM X Specific Conductivity (fdom_x_sc): Interaction between SC and FDOM to capture changes in FDOM signal due to changes in water source (ie agricultural vs forested) that are not fully captured FDOM alone
- 

Climate
- Seasonality (sin(doy))
- Daily flow at the watershed mouth



```{r}

sensor_lagged <- sensor_prepped %>%
  mutate(year = year(DT_hourly_round)) %>%
  split(.$year) %>%
  map(., function(df){
    df %>%
      mutate(sin_doy = sin(2*pi*yday(DT_hourly_round)/365.25), 
             fdom_x_turb = FDOMc * Turbidity, 
             fdom_turb = FDOMc/Turbidity, 
             fdom_x_sc = FDOMc * `Specific Conductivity`)%>%
      # recent variance/sd (sc)
      calc_lagged_parameter(site_col = "site", parameter_col = "Specific Conductivity",
                            dt_col = "DT_hourly_round", lag_fun = "var",
                            time_period = "24 hour", new_value_col = "sc_var_24hr")%>%
      mutate(sc_sd_24hr = sqrt(sc_var_24hr))%>%
      #12 hour median turbidity (proxy for recent storm event)
      calc_lagged_parameter(site_col = "site", parameter_col = "Turbidity",
                            dt_col = "DT_hourly_round", lag_fun = "median",
                            time_period = "24 hour", new_value_col = "turb_median_12hr")%>%
      # 6 hour median FDOM
      calc_lagged_parameter(site_col = "site", parameter_col = "FDOMc",
                            dt_col = "DT_hourly_round", lag_fun = "median",
                            time_period = "6 hour", new_value_col = "FDOMc_6hour")%>%
      calc_lagged_parameter(site_col = "site", parameter_col = "Turbidity",
                            dt_col = "DT_hourly_round", lag_fun = "median",
                            time_period = "6 hour", new_value_col = "turb_median_6hour")%>%
      select(-sc_var_24hr)
    
  })%>%
  bind_rows()

```

## Add in Flow Data

Adding in canyon mouth (CLAFTCCO Site) flow as a general setting for the river. Ideally we would use flow data from each site, but there are some significant data gaps at some of the sites and we want to be sure we have a consistent flow dataset across all sites for modeling. We will test adding in flow at the end to see if it improves performance, but it is likely that the sensor data is already capturing much of the flow information (through turbidity, SC, and FDOM) so it may not add much to the model.

```{r}
#load flow data using CDSSR

min_date <- min(sensor_lagged$DT_hourly_round, na.rm = T)
max_date <- max(sensor_lagged$DT_hourly_round, na.rm = T)

sites <- tibble(abbrev = c( "CLAFTCCO"), 
                site = c("pbd"))

all_flow_data <- map(sites$abbrev, function(abbrev){
  get_telemetry_ts(abbrev = abbrev, 
                   parameter = "DISCHRG", 
                   start_date = as.Date(min_date)- days(1),
                   end_date = as.Date(max_date)+ days(1),
                   timescale = "hour",
                   api_key = cdwr_creds
  )%>%
    select(abbrev, DT_hourly_round = meas_date, flow_cfs = meas_value)%>%
    mutate(DT_hourly_round = with_tz(ymd_hms(DT_hourly_round, tz = "America/Denver"), tzone = "MST"))
}) %>%
  bind_rows()

canyon_mouth_flow_daily <- all_flow_data%>%
  filter(abbrev == "CLAFTCCO")%>%
  group_by(date = as_date(DT_hourly_round))%>%
  summarise(canyon_mouth_daily_flow_cfs = mean(flow_cfs, na.rm = T))%>%
  ungroup()


sensor_final <- sensor_lagged%>%
  mutate(date = as_date(DT_hourly_round))%>%
  left_join(canyon_mouth_flow_daily, by = "date")%>%
  select(-date, -year)%>%
  rename(DT_round = DT_hourly_round)

```


## Read in grab sample dataset

```{r}
water_chem <- load_grab_sample_data()%>%
  # Ensure we have a TOC #
  filter(!is.na(TOC),
         #site_code should be in sensor dataset
         site_code %in% unique(sensor_data$site),
         # Ensure an exact DT for matchup
         !is.na(DT_sample),
         #Filter for after sensors were deployed (2023-08-01) and TOC sample collection began
         DT_sample >= ymd("2023-08-01"))%>%
  filter(!c(year(DT_sample) < 2025 & site_code %in% c("pbr", "pman")))%>% #removing pbr/pman samples pre ross sensor deployment
  select(site = site_code, DT_sample, TOC, collector)

```
# Create Matchup

## Prep and select features 
```{r}
#We want to predict TOC
target <- "TOC"
#Sensor Data Columns we don't want to matchup
id_cols <- c("site", "DT_round", "sensor_datetime", "DT_sample", "sample_datetime","Chl-a Fluorescence", "Depth", "collector")
#Get remaining features
features <- setdiff(names(sensor_final), c(id_cols))

```

## Run matchup function 

We will use the `create_grab_sensor_matchup()` function to create the matchup dataset between grab sample chl-a data and sensor data. We will use a 3 hour acceptable time window for matching. For a "match" to be made, a row within 3 hours of the sample must have  all the parameters in `features` must have a non-NA value.

```{r}

if(file.exists(here("data", "derived", "matchup_datasets", paste0("parsed_toc_sensor_", version, ".parquet")))){
  matchup_results <- read_parquet(here("data", "derived", "matchup_datasets", paste0("parsed_toc_sensor_", version, ".parquet")))
} else {
  matchup_results <- create_grab_sensor_matchup(
    water_chem_df = water_chem,
    sensor_df = sensor_final,
    features = features,
    acceptable_time_window_hours = 3
  )
}

```

## Check Matches

```{r}

matchup_results%>%
  filter(site %in% unique(sensor_final$site))%>%
  mutate(year = year(sample_datetime))%>%
  summarise(n_missing = sum(is.na(`sensor_datetime`)),
            n_total = n(),
            perc_missing = n_missing/n_total*100, .by = year)

missing_samples <- matchup_results%>%
  mutate(year = year(sample_datetime))%>%
  group_by(year, site) %>%
  summarise(n_total = n(),
            n_missing = sum(is.na(sensor_datetime)),
            .groups = "drop")%>%
  arrange(desc(n_missing))%>%
  filter(n_missing > 0)

cat("Number of samples with no match up:\n", sum(missing_samples$n_missing), " of ", nrow(matchup_results) )
print(missing_samples, n = Inf)

#Checking specific site/years to see why there are missing matches.
matchup_results%>%
  filter(is.na(sensor_datetime) & site %in% c("pbr", "chd") & year(sample_datetime) == 2025)%>%
  arrange(site, sample_datetime)%>%select(site, sample_datetime, collector)


### Save Matchup file

#write_parquet(matchup_results,here("data", "derived", "matchup_datasets", paste0("parsed_toc_sensor_", version, ".parquet")))

```

## Create Scaling Parameters & Normalize Data

Normalization is applied based on min-max scaling from the complete sensor dataset using the `apply_training_scale()` function.

```{r}
# Apply scaling to dataset
all_matches_norm <- matchup_results %>%
  # Only keep columns that would be in the complete sensor dataset
  select(any_of(id_cols), any_of(names(sensor_final)), any_of(target))%>%
  # Apply normalization 
  apply_training_scale(
    new_data = .,
    features = features,
    full_dataset = sensor_final,
    scaling_param_out_path = paste0("data/derived/model_scaling_parameters/scaling_params_toc_", version)
  ) %>%
  na.omit()

# make distribution of all matches normalized features to check for any weird outliers or issues with scaling
all_matches_norm %>%
  pivot_longer(cols = any_of(features), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  facet_wrap(~feature, scales = "free") +
  labs(title = "Distribution of Normalized Features in Matchup Dataset", x = "Normalized Value", y = "Count") +
  theme_bw()

```

## Create Holdout Dataset

```{r}
# Trim out any known bad data or sites we don't want to include in modeling
all_matches_trimmed <- all_matches_norm%>%
  rename(id = site)%>%
  filter(!( id == "cbri" & between(sensor_datetime, ymd("2024-10-08"), ymd("2024-10-10"))))%>% #exclude a cbri sample taken during release
  filter(!(id == "pman" & between(sensor_datetime, ymd("2025-05-28"), ymd("2025-06-01"))))%>% #exclude a pman sample with bad sensor data
  filter( id %nin% c("springcreek", "boxcreek"))
# Create testing set from 2025 data at specific sites
testing <- all_matches_trimmed %>%
  filter((year(sensor_datetime) == 2025 & id %in% c("chd", "sfm", "pbr", "pman", "pbd")))
# Create training/validation set
train_val <- all_matches_trimmed %>%
  anti_join(testing)%>%
  select(id,sensor_datetime, collector,TOC,  any_of(features))

```

## Set up model folds
Since we have relatively little data, we are going to use three CV folds. Create 3 separate folds and 3 models, with one fold used for validation on each model. Generally we are looking for:
- Each fold to have ~30-35% of the total data in validation
- Each validation set to have a similar TOC distribution as its training set (median, range)
- Each should have a mix of Upper Mainstem (SFM, PFAL, PBD), Upper Trib (CBRI, CHD, LBEA, PENN) & Lower Mainstem (salyer, udall, riverbend, elc, archery, riverbluffs)

```{r}
val_set_1 <- c("sfm","lbea","chd" ,"riverbluffs","salyer", "elc")

val_set_2 <- c( "pbd", "penn", "joei", "archery" )

val_set_3 <- c("pfal","udall","cottonwood", "cbri", "riverbend")

# Function to create plots for a given fold
map(1:3,  function(fold_id) {
  val_set_name <- paste0("val_set_", fold_id)
  val_set <- train_val %>% filter(id %in% get(val_set_name))
  train_set <- train_val %>% filter(id %nin% get(val_set_name))
  
  prop_val <- round(nrow(val_set)/nrow(train_val), 3) * 100
  train_med <- round(median(train_set$TOC, na.rm = TRUE), 1)
  val_med <- round(median(val_set$TOC, na.rm = TRUE), 1)
  
  # Histogram
  hist <- ggplot(train_set, aes(x = TOC)) +
    geom_histogram(bins = 30, fill = "grey") +
    geom_histogram(data = val_set, aes(x = TOC), bins = 30, fill = "#E70870", alpha = 0.5) +
    labs(
      title = paste0("Fold Set ", fold_id, ": ", nrow(val_set), " samples (", prop_val, "%) in validation set"),
      subtitle = paste0("Validation Sites: ", paste(unique(val_set$id), collapse = ", ")),
      x = "TOC (mg/L)"
    ) +
    theme_minimal()
  
  # Boxplot
  box <- ggplot(
    train_set %>% mutate(set = "Train") %>% 
      bind_rows(val_set %>% mutate(set = "Validation")) %>% 
      mutate(set = factor(set, levels = c("Train", "Validation"))),
    aes(x = set, y = TOC, fill = set)
  ) +
    geom_boxplot() +
    scale_fill_manual(values = c("Train" = "#256BF5", "Validation" = "#E70870")) +
    labs(
      title = paste0("Train Median TOC: ",train_med, "\nVal Median TOC: ", val_med),
      x = "",
      y = "TOC (mg/L)",
      fill = "TV Group",
    ) +
    theme_minimal()
  
  # Combine plots
  ggarrange(hist, box, nrow = 1)
})%>%
  # plot all folds
  ggarrange(plotlist = ., ncol = 1)


#Update train_val with a fold_id column based on val_sets
train_val <- train_val %>%
  mutate(fold_id = case_when(
    id %in% val_set_1 ~ 1,
    id %in% val_set_2 ~ 2,
    id %in% val_set_3 ~ 3
  ))

```


## Visualize TOC distribution in training/validation vs testing sets

Currently, our 2025 data (testing) is skewed to higher TOC values (April- July) compared to the training/validation set. This is not ideal, but we will proceed with this split for now given limited data. Future data updates in December should help even this out as TOC decreases in the fall. 

```{r}
train_val%>%
  mutate(group = "Train/Val")%>%
  bind_rows(
    testing%>%
      mutate(group = "Testing")
  )%>%
  ggplot(., aes(x = TOC, fill = collector)) +
  geom_histogram(bins = 30) +
  labs(
    title = paste0("Training/Validation Set: ", nrow(train_val), " samples              Testing Set: ", nrow(testing), " samples"),
    subtitle = paste0("Training TOC Range: ", round(min(train_val$TOC, na.rm = TRUE),2), " - ", round(max(train_val$TOC, na.rm = TRUE),2) , " mg/L (Median: ", round(median(train_val$TOC), 2), ")\nTesting TOC Range: ", round(min(testing$TOC, na.rm = TRUE),2), " - ", round(max(testing$TOC, na.rm = TRUE),2), " mg/L (Median: ", round(median(testing$TOC), 2),")"),
    x = "TOC (mg/L)", 
    fill = "Sample Collector"
  ) +
  facet_wrap(~group, ncol = 1)+
  ROSS_theme+
  theme(legend.position = "bottom")

```


# Model Training

For our model training we will first do some feature tuning to see which features are most useful for predicting TOC. We will then do hyperparameter tuning on the selected features to optimize model performance.

## Feature Tuning 

### View correlations between features and target (TOC)
Most products with FDOM in them are going to have higher correlation with TOC, as well as seasonal (sin_doy) and flow related features. We want to be sure that we don't put too many of these in the model as calibration error or sensor drift could lead to compounding errors.

```{r}

# Combine training/validation and testing data
modeling_data <- train_val %>%
  mutate(group = "Train/Val") %>%
  bind_rows(
    testing %>%
      mutate(group = "Testing")
  ) %>%
  mutate(
    watershed = ifelse(
      id %in% c("joei", "cbri", "chd", "pfal", "pbr", "sfm", "pman", "pbd", "penn", "lbea"),
      "UpperPoudre", 
      "LowerPoudre"
    )
  )

map(features, function(feat) {
  
  
  cor <- round(cor(modeling_data[[feat]], modeling_data$TOC, use = "complete.obs"), 2)
  
  ggplot(modeling_data, aes(x = .data[[feat]], y = TOC, shape = collector, color = watershed)) +
    geom_point() +
    annotate(
      "text",
      x = mean(range(modeling_data[[feat]], na.rm = TRUE)), # middle of x range
      y = max(modeling_data$TOC, na.rm = TRUE) * 0.95,      # 95% of y max (near top)
      label = paste0("Corr: ", round(cor(modeling_data[[feat]], modeling_data$TOC, use = "complete.obs"), 2)),
      hjust = 0.5,
      vjust = 1,
      size = 4
    ) +
    labs(x = feat, y = "TOC (mg/L)", shape = "Collector", color = "Watershed") +
    #facet_wrap(~watershed, scales = "free") +
    ROSS_theme
})%>%
  wrap_plots(., guides = "collect")

```

### Pruning Features with high correlations 

We want to avoid using too many highly correlated features in the model, as this can lead to overfitting and instability. We will look at the correlation matrix and remove some of the more redundant features.

```{r}
# get a correlation matrix across all features
corr_matrix <- modeling_data %>% 
  select(all_of(features)) %>% 
  cor(use = "pairwise.complete.obs")

# we just want half the matrix, so recode those in the lower triangle to NA
corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter for those with correlation > 0.9
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.9) %>%
  arrange(desc(abs(correlation))) %>% 
  # remove redundant vars
  filter(feature1 %nin% c("FDOM", "doy"))

# look at the list and make some educated decisions about features
high_corr_pairs

fewer_features <- features[features %nin% c("FDOM_turb_corr", "FDOM_final_corr", 
                                            # remove vars with high correlation 
                                            # with others
                                            "turb_median_12hr", "FDOMc_6hour",
                                            "FDOM Fluorescence",
                                            "hourly_canyon_flow_cfs",
                                            "fdom_max_6hr", "temp_max_24hr")]

# do the correlation matrix again to make sure that is okay
corr_matrix <- modeling_data %>% 
  select(all_of(fewer_features)) %>% 
  cor(use = "pairwise.complete.obs")

corr_matrix[lower.tri(corr_matrix, diag = TRUE)] <- NA

# Convert to a tidy data frame and filter
high_corr_pairs <- corr_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "feature1") %>%
  pivot_longer(-feature1, names_to = "feature2", values_to = "correlation") %>%
  filter(!is.na(correlation), abs(correlation) > 0.85) %>%
  arrange(desc(abs(correlation)))

high_corr_pairs

```

`high_corr_pairs` has features we should be cautious of using together. We will try to avoid using more than one of these in the same model.

### Testing Features

First, we will start with our base model (live sensor data only) and then test adding in other features to see if they improve performance. We will use the default hyperparameters below for all models to keep things consistent. Once we have tested all the features together, we will want to trim it down a bit so we are not using too many highly correlated features or overfitting the model due to lack of data

#### Run across folds testing different features to see which improve performance

Here I run the `xgboost_feature_selection()` function which runs a basic xgboost model with the specified features and default hyperparameters. It returns feature importance, shap values, and best iteration info for each fold. We will then look at feature importance and shap values to see which features are most useful, then trim features back to reduce our correlative and overfitting issues. 

```{r}

# Weighting function for TOC
toc_high_weights <-  function(df) {
  ifelse(df$TOC <= 4,1,                  # flat weight for low TOC
         1 + 3 *log1p(df$TOC - 4))  # grows slowly, no cap needed
}

#raw sensor features
general_features <- c("FDOMc","Turbidity","Specific Conductivity", "Temperature")

#setup function
feature_results <- map(
  #list of different features we want to test
  list(
    gen_feat = general_features, 
    fewer_features = c(fewer_features), # this is the set of features we are testing that have been pruned based on correlation and domain knowledge, but still includes flow
    # removing FDOM_turb and turb median 6 hour
    fewer_features_doy = c(setdiff(fewer_features, c("fdom_turb", "turb_median_6hour"))), 
    fewest_features = c(setdiff(fewer_features, c("fdom_turb", "turb_median_6hour", "sin_doy")))
    
  ),
  #Setting function parameters
  ~xgboost_feature_selection(
    train_val = train_val,
    features_to_test = .x,
    target = "TOC",
    default_hyper_params =  list(
      objective        = "reg:squarederror",
      eval_metric      = "rmse",
      eta              = 0.05,
      gamma            = 0.6,
      alpha            = 0,
      lambda           = 1,
      max_depth        = 3,
      subsample        = 0.5,
      colsample_bytree = 0.5,
      min_child_weight  = 2 ), 
    weight_fun = toc_high_weights
  ))
```

#### Feature Set Results

##### Weighted RMSE

We compare across folds to see if there is big differences in folds (ie unbalanced data) and also compare each feature set against "base" model of general features (sensors). Looking for features that make meaningful improvements in RMSE across all folds. If our RMSE improves but that is related to only a single fold, we may want to reevaluate the feature set as it is likely overfitting to a given site(s). 
Keep in mind that this is weighted RMSE so it may be higher than unweighted RMSE you see in other places.

```{r}
# Highlight differences between folds in rmse
map_dfr(feature_results, ~{
  # Compute feature set description
  unique_features <- setdiff(.x$features, general_features)
  final_features <- c("general_features", unique_features)
  feature_desc <- paste(final_features, collapse = ", ")
  
  # Take best_msg tibble and rename eval_rmse by fold
  .x$best_msg %>%
    select(fold_id, eval_rmse) %>%
    mutate(fold_col = paste0("rmse_fold_", fold_id)) %>%
    select(-fold_id) %>%
    mutate(eval_rmse = round(eval_rmse, 2)) %>%
    pivot_wider(names_from = fold_col, values_from = eval_rmse) %>%
    mutate(
      features = feature_desc,
      mean_rmse = round(mean(c_across(starts_with("rmse_fold_")), na.rm = TRUE), 2)
    )%>%
    select(features, mean_rmse, contains("rmse_fold_"))
  
}, .id = "feature_set") %>%
  #check if new feature is better than baseline
  mutate(
    baseline_rmse = mean_rmse[feature_set == "gen_feat"],
    improves_rmse = if_else(mean_rmse < baseline_rmse, round(mean_rmse - baseline_rmse, 2), 0)
  ) %>%
  select(-baseline_rmse)%>%
  arrange(improves_rmse)
```

##### Feature Importance

This comes directly from the xgboost package and is based on Gain, Cover, and Frequency of splits. We summarize across folds to see which features are most important across all folds. 

```{r}
#table with all importances
importance_tables <- map(feature_results, ~{
  imp <- .x$model_importance
  
  # Identify fold columns dynamically
  gain_cols <- grep("^Gain_fold_", names(imp), value = TRUE)
  cover_cols <- grep("^Cover_fold_", names(imp), value = TRUE)
  freq_cols <- grep("^Frequency_fold_", names(imp), value = TRUE)
  
  imp %>%
    mutate(
      mean_gain = rowMeans(across(all_of(gain_cols)), na.rm = TRUE),
      mean_cover = rowMeans(across(all_of(cover_cols)), na.rm = TRUE),
      mean_frequency = rowMeans(across(all_of(freq_cols)), na.rm = TRUE)
    ) %>%
    mutate(new_feat = ifelse(Feature %in% general_features, "", "*")) %>%
    select(Feature, new_feat, mean_gain, mean_cover, mean_frequency) %>%
    
    arrange(desc(mean_gain))
}, .id = "feature_set")
importance_tables

#Get features with less than 0.02 gain on average across folds in each model run to consider removing
cat("Features with mean gain < 0.02 in :\n")
iwalk(importance_tables, ~{
  low_gain_features <- .x %>%
    filter(mean_gain < 0.02) %>%
    pull(Feature)
  if(length(low_gain_features) == 0){
    #print the name of the list and say no low gain features
    cat(paste0(.y, ": No low gain features\n"))
  } else {
    cat(.y, paste(low_gain_features, collapse = ", "), "\n")
  }
})

```


##### Shapley Values

Let's look at training and validation shap values for features across all folds. This gives us a sense of which features are most important in the model predictions. We summarize across folds to see which features are most important across all folds. In general we would expect features to have similar importance in training and validation, if a feature has high importance in training but low in validation it may be overfitting to the training data and we may want to re-evaluate. 

```{r}
#table with all shap values
shap_summary <- map_dfr(feature_results, ~{
  shap <- .x$shap_values
  
  val_shap_cols <- grep("^val_mean_abs_shap_fold_", names(shap), value = TRUE)
  train_shap_cols <- grep("^train_mean_abs_shap_fold_", names(shap), value = TRUE)
  
  shap %>%
    mutate(
      mean_val_shap = round(rowMeans(across(all_of(val_shap_cols)), na.rm = TRUE), 3),
      mean_train_shap = round(rowMeans(across(all_of(train_shap_cols)), na.rm = TRUE), 3)
    ) %>%
    mutate(val_train_diff = mean_val_shap - mean_train_shap) %>%
    select(Feature, mean_val_shap, mean_train_shap, val_train_diff)
}, .id = "feature_set")

shap_long <- shap_summary %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "type",
    values_to = "mean_shap"
  ) %>%
  mutate(
    type = recode(type,
                  mean_val_shap = "Validation",
                  mean_train_shap = "Training"),
    feature_set = factor(feature_set, levels = unique(feature_set))
  ) %>%
  mutate(Feature = factor(Feature, levels = c( rev(feature_results$fewer_features$features),"Sensor_Turb","Temp")))

ggplot(shap_long, aes(x = feature_set, y = Feature, fill = mean_shap)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_viridis_c(
    option = "magma",
    limits = c(0, 1),
    name = "Mean |SHAP|",
    direction = -1  # reverses so dark = low, bright = high (optional)
  )+
  facet_wrap(~type) +
  labs(
    x = "Feature Set",
    y = "Feature",
    title = "Mean SHAP Values by Feature and Feature Set"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold")
  )

```

### TV performance of best feature set

After trimming down a couple of features (minimal improvement/low shap values), we should take a quick look at model performance in TV space with our base model (default hyperparameters)

```{r}

best_models <- feature_results$fewest_features$model

feature_tv_plots <- map(1:3, ~plot_tv_fold_perf(fold_set = .x,
                                                train_val_df = train_val,
                                                fold_model = best_models[[.x]], # a list where each element is a model for that fold
                                                target_col = "TOC", units = "mg/L")
)
ft_tv_plot<- wrap_plots(feature_tv_plots, ncol = 1, guides = "collect")&
  theme(legend.position = "bottom")
ft_tv_plot

```


## Hyperparameter Tuning
Now that we have a feature set, we can do some hyperparameter tuning to see if we can improve model performance. This will take a while to run, so we will save the results and load them in later.

### Setup

```{r}

folds <- tibble(
  fold = 1:3,
  val_ids = list(val_set_1, val_set_2, val_set_3)
 )
features <- best_models[[1]]$feature_names

# Reduce tune grid for reruns (ie. General best params)
tune_grid = expand.grid(
  nrounds = 10000,
  max_depth = c(2, 3, 4),
  eta = c( 0.01,0.05, 0.1),
  gamma = c(0.8),
  lambda = c(1,10),
  alpha = c(0, 1),
  colsample_bytree = c(0.5,0.8),
  min_child_weight = c(2, 4),
  subsample = c(0.5,0.8)
)

```

### Run hyperparameter tuning

```{r}
set.seed(123)
hyperp_results <- xgboost_hyperparameter_tuning(
  data = train_val %>% select(TOC, id, any_of(features)), 
  weights = toc_high_weights,
  tune_grid = tune_grid,
  target_col = "TOC",
  site_col = "id",
  fold_ids =  folds, 
  units = "mg/L")

```

### Choosing best grids

### Fold 1

```{r}

hyperp_results[[1]]$perf_dist
hyperp_results[[1]]$eval_plot
hyperp_results[[1]]$tv_plot

best_fold_1_grid <- 145
```

### Fold 2

```{r}
hyperp_results[[2]]$perf_dist
hyperp_results[[2]]$eval_plot
hyperp_results[[2]]$tv_plot

best_fold_2_grid <- 71

```

### Fold 3

```{r}
hyperp_results[[3]]$perf_dist
hyperp_results[[3]]$eval_plot
hyperp_results[[3]]$tv_plot

best_fold_3_grid <- 52

```

## Extract best models for each fold

```{r}

full_model <- map(1:3, function(i) {
  
  best_grid <- paste0("grid_", get(glue("best_fold_{i}_grid")))
  model_info <- hyperp_results[[i]]$grid_results[[best_grid]]
  list(
    model = model_info$model,
    perf = model_info$perf,
    params = model_info$params,
    training_data = train_val %>% filter(fold_id != i),
    validation_data = train_val %>% filter(fold_id == i)
  )
})

```

### Checking Results


#### Feature Importance

```{r}
importance_list <- lapply(seq_along(full_model), function(i) {
  fold_model <- full_model[[i]]$model
  features <- fold_model$feature_names
  importance_matrix <- xgb.importance(feature_names = features, model = fold_model)
  importance_df <- as.data.frame(importance_matrix)
  importance_df$Fold <- paste("Fold", i)
  return(importance_df)
})

print(importance_list)
```

#### Shapley Importance

```{r}

# Function to plot SHAP values for a given fold
plot_shap <- function(fold_set, training_data, fold_model) {
  #get features from model
  features <- fold_model$feature_names
  # Predictions on validation/training set
  dtrain <- as.matrix(training_data[, features])
  
  shap_values <- shap.values(xgb_model = fold_model, X_train = dtrain)
  
  # Long-format SHAP values
  shap_long <- shap.prep(
    shap_contrib = shap_values$shap_score,
    X_train = dtrain
  )
  #return shapley plot
  shap.plot.summary(shap_long)
}

map(1:3, ~ plot_shap(
  fold_set = .x,
  fold_model = full_model[[.x]]$model,
  training_data = full_model[[.x]]$training_data
))%>%
  wrap_plots(., ncol = 1, guides = "collect") &
  theme(legend.position = "bottom")

```

### Plots of TV Performance
Now let's look at our performance in TV space.

```{r}

hyperparam_tv_plots <- map(1:3, ~plot_tv_fold_perf(fold_set = .x,
                                                   train_val_df = train_val,
                                                   fold_model = full_model[[.x]]$model, # a list where each element is a model for that fold
                                                   target_col = "TOC", units = "mg/L")
)

hp_tv_plot<- wrap_plots(hyperparam_tv_plots, ncol = 1, guides = "collect")&
  theme(legend.position = "bottom")
hp_tv_plot
ggsave(here(paste0("docs/uclp_dss/figs/realtime_toc_modeling/performance/toc_model_fold_validation_performance_", version, ".png")), width = 14, height = 10)

```

#### Against base hyperparameters

We should hope that all our tuning efforts have improved the model!! 

```{r}
# Compare best hyperparameter model to default hyperparameter model
final_plot <- wrap_plots(
  ft_tv_plot,
  hp_tv_plot,
  ncol = 2,
  guides = "collect"
) &
  theme(legend.position = "bottom")
final_plot <- final_plot + 
  plot_annotation(
    title = "Model Performance Comparison: Default (left) vs Tuned Hyperparameters (right)"
  )

final_plot

```

Looking at the plots, we can see that the tuned hyperparameter model has slighly improved our results. It appears that fold 2 has the worst performance which is likely due to the distribution of the validation set being the widest and contains the sites (JOEI: lowest TOC, PENN: high TOC, low turb). We will save this model and then move on to testing bias correction methods to see if we can improve performance there. We still see similar performance to previous model versions

## Save as "in_dev" and remove unneeded objects

If we want to test additional ideas for bias correction or look at model performance on timeseries data, we can just run this script until `model_training` and then load in the "in_dev" model file to avoid having to rerun all the tuning again.

```{r}
#commented out so we don't overwrite existing model files
saveRDS(full_model, file = here("data", "derived", "models","TOC_realtime", "in_dev", paste0("ross_only_toc_xgboost_models_full_",version,".rds")))
#if we want to reload model
#full_model <- readRDS(here("data", "derived", "models","TOC_realtime", "in_dev", paste0("ross_only_toc_xgboost_models_full_",version,".rds")))

rm(list = setdiff(ls(), c("train_val", "testing", "full_model","version","id_cols",
                          "sensor_final",
                          #keeping functions around
                          "apply_training_scale","plot_tvt_fold_perf","plot_train_test_ensemble_perf",
                          "load_grab_sample_data", "plot_model_timeseries" )) )
```

# Evaluate on Test Set

We should only do this once we feel good about our model and are ready to see how it performs on the unseen test set. This should only be run after big changes to the model or if we get new data (ie The model should be in a "final/working" version)

## Visualize Train Val Test

### Individual folds

```{r}
plots<- map(1:3, ~{
  plot_tvt_fold_perf(train_val, testing, fold_set = .x, fold_model = full_model[[.x]]$model, target_col = "TOC", units = "mg/L", subtitle_arg = "CLP Samples only")
})
plots

# ggsave(glue(here("docs", "uclp_dss", "figs", "toc_modeling", "performance", "ross_only_model_fold_{i}.png")),
#        plot = p, width = 10, height = 6, dpi = 500)

```

### All folds together

```{r}
# Single plot with all folds
wrap_plots(plots, ncol = 2, guides = "collect") &
  theme(legend.position = "bottom")
ggsave(here("docs","uclp_dss", "figs","realtime_toc_modeling","performance",paste0("ross_only_model_all_folds_TVT", version, ".png")),
       width = 14, height = 8, dpi = 500)
```

### Ensemble predictions

```{r}
# Plot ensemble
ensemble <- plot_train_test_ensemble_perf(train_val_df = train_val,
                                          test_df = testing,
                                          fold_models = map(full_model, "model"),
                                          target_col = "TOC",
                                          units = "mg/L",
                                          subtitle_arg = "CLP Samples only")
plot(ensemble)
ggsave(here("docs","uclp_dss", "figs", "realtime_toc_modeling","performance",paste0("ross_only_model_ensemble", version, ".png")), plot = ensemble, width = 12, height = 10, dpi = 500)
```

# Timeseries Space Performance

## Prep sensor timeseries & water chem

```{r}
#loading in all sensor data and applying training scaling
sensor_ts <- apply_training_scale(new_data = sensor_final, 
                                  full_dataset = sensor_final, 
                                  features = full_model[[1]]$model$feature_names)%>%
  select(site, DT_round, any_of(full_model[[1]]$model$feature_names))

water_chem <- load_grab_sample_data()

time_series_dir <- here("docs", "uclp_dss", "figs", "realtime_toc_modeling", "timeseries")
```

## Example 1: CHD 2025

```{r}

chd_plot <- plot_model_timeseries(model_input_df = sensor_ts, 
                                  water_chem_df = water_chem, 
                                  summary_interval = "4 hour",
                                  site_sel = "chd", 
                                  site_title = "Chambers Lake outflow",
                                  model = map(full_model, "model"),
                                  method = "Ensemble",
                                  target_col = "TOC",
                                  units = "mg/L",
                                  start_DT = "2025-05-01 12:00",
                                  end_DT = "2025-07-05 00:00",
                                  subtitle_arg = "CLP Samples Only", 
                                  plot_model_input = T)


ggarrange(chd_plot$model_plot, chd_plot$model_input_plot,ncol = 2)

ggsave(here(time_series_dir, paste0("chambers_outflow_20250501_20250705_toc", version,".png")),
       plot = chd_plot$model_plot,
       width = 12,
       height = 8,
       dpi = 300)

 ggsave(here(time_series_dir, paste0("chambers_outflow_20250501_20250705_sensor.png")),
        plot = chd_plot$model_input_plot,
       width = 10,
       height = 10,
        dpi = 300)

```


## Example 2: PFAL 2025

```{r}

pfal_plot <- plot_model_timeseries(model_input_df = sensor_ts, 
                                   water_chem_df = water_chem%>%filter(collector == "ROSS"), 
                                   summary_interval = "4 hour",
                                   site_sel = "pfal", 
                                   site_title = "Poudre @ Poudre Falls",
                                   model = map(full_model, "model"),
                                   method = "Ensemble",
                                   target_col = "TOC",
                                   units = "mg/L",
                                   start_DT = "2025-04-16 00:00",
                                   end_DT = "2025-07-10 00:00",
                                   subtitle_arg = "CLP Samples Only", 
                                   plot_model_input = T)

ggarrange(pfal_plot$model_plot, pfal_plot$model_input_plot,ncol = 2)


 ggsave(here(time_series_dir, paste0("pfal_20240415_20240710_toc_", version, ".png")),
        plot = pfal_plot$model_plot,
        width = 12,
        height = 8,
        dpi = 300)


 ggsave(here(time_series_dir, paste0("pfal_20240415_20240710_sensor.png")),
        plot = pfal_plot$model_input_plot,
        width = 10,
        height = 10,
       dpi = 300)

```

## Example 3: PBR

```{r}


pbr_plot <- plot_model_timeseries(model_input_df = sensor_ts, 
                                   water_chem_df = water_chem, 
                                  summary_interval = "4 hour",
                                  site_sel = "pbr", 
                                  site_title = "Poudre @ Indian Meadows",
                                  model = map(full_model, "model"),
                                  method = "Ensemble",
                                  target_col = "TOC",
                                  units = "mg/L",
                                  start_DT = "2025-05-01 00:00",
                                  end_DT = "2025-07-08 00:00",
                                  subtitle_arg = "CLP Samples Only", 
                                  plot_model_input = T)

 ggarrange(pbr_plot$model_plot, pbr_plot$model_input_plot,ncol = 2)


 ggsave(here(time_series_dir, paste0("pbr_20250501_20250708_toc_", version,".png")),
       plot = pbr_plot$model_plot,
       width = 12,
       height = 8,
       dpi = 300)
 
 ggsave(here(time_series_dir, paste0("pbr_20250501_20250708_sensor.png")),
        plot = pbr_plot$model_input_plot,
        width = 10,
        height = 10,
        dpi = 300)
 
```

## Example 4: SFM 2025

```{r}

sfm_plot <- plot_model_timeseries(model_input_df = sensor_ts, 
                                  water_chem_df = water_chem, 
                                  summary_interval = "4 hour",
                                  site_sel = "sfm", 
                                  site_title = "South Fork of the Poudre",
                                  model = map(full_model, "model"),
                                  method = "Ensemble",
                                  target_col = "TOC",
                                  units = "mg/L",
                                  start_DT = "2025-04-15 00:00",
                                  end_DT = "2025-07-01 00:00",
                                  subtitle_arg = "CLP Samples Only", 
                                  plot_model_input = T)

ggarrange(sfm_plot$model_plot, sfm_plot$model_input_plot,ncol = 2)

ggsave(here(time_series_dir, paste0("sfm_20250415_20250701_toc_",version, ".png")),
       plot = sfm_plot$model_plot,
       width = 12,
       height = 8,
       dpi = 300)

ggsave(here(time_series_dir, paste0("sfm_20250415_20250615_sensor.png")),
       plot = sfm_plot$model_input_plot,
       width = 10,
       height = 10,
       dpi = 300)

```



## Example 4: PMAN 2025

```{r}

pman_plot <- plot_model_timeseries(model_input_df = sensor_ts, 
                                   water_chem_df = water_chem%>%filter(collector == "ROSS"), 
                                   summary_interval = "4 hour",
                                   site_sel = "pman", 
                                   site_title = "Poudre at Manners",
                                   model = map(full_model, "model"),
                                   method = "Ensemble",
                                   target_col = "TOC",
                                   units = "mg/L",
                                   start_DT = "2025-04-24 00:00",
                                   end_DT = "2025-07-01 00:00",
                                   subtitle_arg = "CLP Samples Only", 
                                   plot_model_input = T)

ggarrange(pman_plot$model_plot, pman_plot$model_input_plot,ncol = 2)



ggsave(here(time_series_dir, paste0("pman_20250424_20250714_toc_",version,".png")),
       plot = pman_plot$model_plot,
       width = 12,
       height = 8,
       dpi = 300)

ggsave(here(time_series_dir, paste0("pman_20250415_20250705_sensor.png")),
       plot = pman_plot$model_input_plot,
       width = 10,
       height = 10,
       dpi = 300)

```

## Example 5: PBD 2025

```{r}

pbd_plot <- plot_model_timeseries(model_input_df = sensor_ts, 
                                  water_chem_df = water_chem%>%filter(collector == "ROSS"), 
                                  summary_interval = "4 hour",
                                  site_sel = "pbd", 
                                  site_title = "Poudre at Canyon Mouth",
                                  model = map(full_model, "model"),
                                  method = "Ensemble",
                                  target_col = "TOC",
                                  units = "mg/L",
                                  start_DT = "2025-04-15 00:00",
                                  end_DT = "2025-07-05 00:00",
                                  subtitle_arg = "CLP Samples Only", 
                                  plot_model_input = T)

ggarrange(pbd_plot$model_plot, pbd_plot$model_input_plot,ncol = 2)

ggsave(here(time_series_dir, paste0("pbd_20250415_20250705_toc_",version, ".png")),
       plot = pbd_plot$model_plot,
       width = 12,
       height = 8,
       dpi = 300)

ggsave(here(time_series_dir, paste0("pbd_20250520_20250705_sensor.png")),
       plot = pbd_plot$model_input_plot,
       width = 10,
       height = 10,
       dpi = 300)



pbd_plot <- plot_model_timeseries(model_input_df = sensor_ts, 
                                  water_chem_df = water_chem%>%filter(collector == "ROSS"), 
                                  summary_interval = "12 hour",
                                  site_sel = "pbd", 
                                  site_title = "Poudre at Canyon Mouth",
                                  model = map(full_model, "model"),
                                  method = "Ensemble",
                                  target_col = "TOC",
                                  units = "mg/L",
                                  start_DT = "2025-07-05 00:00",
                                  end_DT = "2025-11-05 00:00",
                                  subtitle_arg = "CLP Samples Only", 
                                  plot_model_input = T)

ggarrange(pbd_plot$model_plot, pbd_plot$model_input_plot,ncol = 2)

ggsave(here(time_series_dir, paste0("pbd_20250705_20251105_toc_",version, ".png")),
       plot = pbd_plot$model_plot,
       width = 12,
       height = 8,
       dpi = 300)
```


## Example 6: Udall (Lincoln Ave) 2025

```{r}


udall_plot <- plot_model_timeseries(model_input_df = sensor_ts, 
                                    water_chem_df = water_chem%>%filter(collector == "ROSS"), 
                                    summary_interval = "12 hour",
                                    site_sel = "elc", 
                                    site_title = "Poudre at Lincoln Ave",
                                    model = map(full_model, "model"),
                                    method = "Ensemble",
                                    target_col = "TOC",
                                    units = "mg/L",
                                    start_DT = "2025-04-15 00:00",
                                    end_DT = "2025-07-05 00:00",
                                    subtitle_arg = "CLP Samples Only", 
                                    plot_model_input = T)

ggarrange(udall_plot$model_plot, udall_plot$model_input_plot,ncol = 2)

 ggsave(here(time_series_dir, paste0("udall_20250415_20250705_toc_",version,".png")),
        plot = udall_plot$model_plot,
        width = 12,
        height = 8,
        dpi = 300)

 ggsave(here(time_series_dir, paste0("udall_20250520_20250705_sensor.png")),
        plot = udall_plot$model_input_plot,
        width = 10,
        height = 10,
        dpi = 300)
```



# Save final models

We will extract just the model objects from the list to save space.

```{r}

small_models <- map(full_model, "model")

saveRDS(small_models, file = here("data", "derived", "models","TOC_realtime", "in_dev", paste0("ross_only_toc_xgboost_models_light_",version,".rds")))

```







