---
title: "Apply GAM to intake"
author: "ROSSyndicate"
date: "2026-01-26"
output: html_document
---

# Purpose

This script is meant to apply the distributed TOC GAM model to assess its 
performance at the intake for pre-forecasted Q baseline performance. 

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(here)
library(tidyverse)
library(cdssr)
library(mgcv)
library(Metrics)
```

# Data

Load in the data from the intake and flow data at the canyon mouth.

```{r}
# intake TOC
fc_intake <- read_csv(here("data/collated/FC_intake/fc_intake_TOC_record_2010-2025.csv"))

# read in flow data at canyon mouth
flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = "2010-01-01",
                       end_date = "2025-12-31",
                       timescale = "day") %>%
  select(datetime, canyon_mouth_cfs = value) %>% 
  mutate(date = date(datetime), 
         year = year(date),
         doy = yday(date)) 

intake_flow <- left_join(fc_intake, flow_data) %>% 
  rename(DOY = doy) %>% 
  # remove data outside the window used to build the model
  filter(between(DOY, 107, 320)) %>% 
  # making an estimate of distance as a bit further up from our sampling 
  # location at poudre above north fork
  mutate(distance_upstream_km = 8)

```

And load in the model files. 

```{r}
fit_1 <- read_rds(here("data/derived/models/TOC_baseline/working/TOC_GAM_addQ_fit1_v2026-01-23.rds"))
fit_2 <- read_rds(here("data/derived/models/TOC_baseline/working/TOC_GAM_addQ_fit2_v2026-01-23.rds"))
fit_3 <- read_rds(here("data/derived/models/TOC_baseline/working/TOC_GAM_addQ_fit3_v2026-01-23.rds"))
fit_4 <- read_rds(here("data/derived/models/TOC_baseline/working/TOC_GAM_addQ_fit4_v2026-01-23.rds"))
fit_5 <- read_rds(here("data/derived/models/TOC_baseline/working/TOC_GAM_addQ_fit5_v2026-01-23.rds"))
```

# Model application

And now we'll apply the GAM defined by the data from the watershed (including
both FC data and ROSS data, but not intake data).

```{r}
intake_flow <- intake_flow %>% 
  mutate(pred_toc_1 = predict(fit_1, intake_flow),
         pred_toc_2 = predict(fit_2, intake_flow),
         pred_toc_3 = predict(fit_3, intake_flow),
         pred_toc_4 = predict(fit_4, intake_flow),
         pred_toc_5 = predict(fit_5, intake_flow),
         pred_toc = (pred_toc_1 + pred_toc_2 + pred_toc_3 + pred_toc_4 + pred_toc_5) / 5)
```

And visualize.

```{r}
intake_flow %>% 
  ggplot(., aes(x = mean_TOC, y = pred_toc)) + 
  geom_point() +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  labs(x = "measured TOC @ intake (mg/L)",
       y = "predicted TOC @ intake (mg/L)") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) 

```

And get some performance stats.

```{r}
intake_flow %>% 
  .[complete.cases(.), ] %>% 
  summarize(RMSE = rmse(mean_TOC, pred_toc), 
            MAE = mae(mean_TOC, pred_toc),
            BIAS = bias(mean_TOC, pred_toc))

summary(lm(intake_flow$pred_toc ~ intake_flow$mean_TOC))
```

