---
title: "Baseline TOC + Q"
author: "ROSSyndicate"
date: "2026-02-04"
output: html_document
---

# Purpsoe

The purpose of this markdown is to add some complexity to the DOY / distance
up canyon in `modeling/toc/toc_forecast/baseline_TOC.Rmd` by adding canyon mouth
data, which we have forecast data for and can easily turn into an operational 
model that refelects interannual variability.

## Data

```{r, echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
library(tidyverse)
library(here)
library(mgcv)
library(Metrics)
library(cdssr)

source(here("src/collation/load_grab_sample_data.R"))
```

```{r}
# grab chemistry data
chem <- load_grab_sample_data() %>% 
  # remove intake for now
  filter(site_code != "prw" & !is.na(Campaign)) %>% 
  # check for DOC/TOC data
  filter(!is.na(TOC) | !is.na(DOC)) %>% 
  # only grab columns we need right now
  select(Date, site_name, site_code, collector, TOC, DOC, flag,
         Campaign, location_type, watershed, distance_upstream_km) %>% 
  # drop any duplicates here
  unique()

mainstem <- chem %>% 
  filter(Campaign %in% c("Mainstem - Upper", "South Fork") & 
           location_type %in% c("Mainstem", "Stream", "Inflow") &
           # remove inflows
           !(site_name %in% c("Poudre Above Joe Wright",
                              "Beaver",
                              "Comanche Inflow",
                              "Hourglass Inflow"))) %>% 
  mutate(DOY = yday(Date),
         year = year(Date), 
         toc_doc = if_else(!is.na(TOC), TOC, DOC),
         # to play with categorical variables, they failed.
         identity = if_else(Campaign == "South Fork", 1, 0),
         fire_signal = if_else(year %in% c(2021, 2022), 1, 0)) %>% 
  # limit time, remove fire signal, remove flier
  filter(!is.na(toc_doc), between(DOY, 100, 320), 
         fire_signal == 0) %>% 
  # remove data after the 2013 flood (Sept 9, 2013)
  filter(!(Date >= "2013-09-2013" & year == 2013)) %>% 
  rename(date = Date) 

# let's also make sure we only have one sample per day per site
mainstem <- mainstem %>% 
  summarize(toc_doc = mean(toc_doc),
            .by = c(date, site_name, DOY, year, distance_upstream_km))

ggplot(mainstem, aes(x = DOY, y = toc_doc)) + 
  geom_point() + 
  facet_wrap(site_name ~ .)

# also recode the flier in south fork and manners bridge
mainstem <- mainstem %>% 
  filter(!(site_name == "Poudre South Fork" & DOY > 200 & toc_doc > 9)) %>% 
  filter(!(site_name == "Poudre at Manner's Bridge" & DOY > 200 & toc_doc >5))

ggplot(mainstem, aes(x = DOY, y = toc_doc)) + 
  geom_point() + 
  facet_wrap(site_name ~ .)

# grab canyon flow
flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(mainstem$Date, na.rm = T) - days(7),
                       end_date = max(mainstem$Date, na.rm = T) + days(7),
                       timescale = "day") %>%
  select(date = datetime, 
         canyon_mouth_cfs = value) %>% 
  filter(date >= min(mainstem$date) - days(7))

ggplot(flow_data, aes(x = date, y = canyon_mouth_cfs)) +
  geom_point() +
  theme_bw()

mainstem_plus <- left_join(mainstem, flow_data) %>%
  filter(!is.na(canyon_mouth_cfs)) %>% 
  select(date, year, DOY, distance_upstream_km, canyon_mouth_cfs, site_name, toc_doc) 
```

And now let's walk through some simple GAM architectures here.

```{r}
gam_linear <- gam(toc_doc ~ DOY + distance_upstream_km + canyon_mouth_cfs,
                  bs = c("cc", "tp", "tp"),
                  method = "REML", 
                  family = Gamma(link = "log"),
                  data = mainstem_plus)
gam.check(gam_linear, 0)
summary(gam_linear)

gam_add <- gam(toc_doc ~ s(DOY, bs = "cc") + 
                 s(canyon_mouth_cfs, bs = "tp") +
                 s(distance_upstream_km, bs = "tp", k = 5),
               method = "REML", 
               family = Gamma(link = "log"),
               data = mainstem_plus)
gam.check(gam_add, 0)
# make this a little more aggressive to try to get a gap between k' and
gam_add <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                 s(canyon_mouth_cfs, bs = "tp", k = 15) +
                 s(distance_upstream_km, bs = "tp", k = 5),
               method = "REML", 
               family = Gamma(link = "log"),
               data = mainstem_plus)
gam.check(gam_add, 0)
summary(gam_add)

gam_add_by <- gam(toc_doc ~ s(DOY, k = 15, bs = "cc", by = distance_upstream_km) + 
                    s(canyon_mouth_cfs, k = 15, by = distance_upstream_km, bs = "tp"),
                  method = "REML", 
                  family = Gamma(link = "log"),
                  data = mainstem_plus)
gam.check(gam_add_by, 0)
summary(gam_add_by)


gam_add_int <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                     s(canyon_mouth_cfs, bs = "tp", k = 15) +
                     s(distance_upstream_km, bs = "tp", k = 5) +
                     ti(DOY, canyon_mouth_cfs, distance_upstream_km, 
                        bs = c("cc", "tp", "tp")),
                   data = mainstem_plus,
                   family = Gamma(link = "log"),
                   method = "REML")
gam.check(gam_add_int, 0)
summary(gam_add_int)

gam_add_int_by <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                        s(canyon_mouth_cfs, bs = "tp", k = 15) +
                        ti(DOY, canyon_mouth_cfs, 
                           bs = c("cc", "tp"),
                           by = distance_upstream_km),
                      data = mainstem_plus,
                      family = Gamma(link = "log"),
                      method = "REML")
gam.check(gam_add_int_by, 0)
summary(gam_add_int_by)

gam_tensor <- gam(toc_doc ~ te(DOY, 
                               canyon_mouth_cfs,
                               distance_upstream_km, 
                               bs = c("cc", "tp", "tp")),
                  method = "REML", 
                  family = Gamma(link = "log"),
                  data = mainstem_plus)
gam.check(gam_tensor, 0)
summary(gam_tensor)

AIC(gam_linear, gam_add, gam_add_by, 
    gam_add_int, gam_add_int_by, gam_tensor) %>% 
  arrange(AIC)

```

I think the two to compare here are the additive method and the
additive/interaction with a varying coefficient for the distance upstream. They
have a nice balance of performance and model diagnostics.

Let's do the same method for
split as the baseline GAM for DOY and distance up stream:

- Test: 2010, 2017, 2025
- CV 1: 2008, 2013, 2018
- CV 2: 2009, 2014, 2019
- CV 3: 2011, 2015, 2023
- CV 4: 2012, 2016, 2024

```{r}
test_years <- c(2010, 2017, 2025)
cv_1_years <- c(2008, 2013, 2018)
cv_2_years <- c(2009, 2014, 2019)
cv_3_years <- c(2011, 2015, 2023)
cv_4_years <- c(2012, 2016, 2024)
```

GAM diagnostics are interpreted using the gam.check() function. We want to make
sure that:

1) the model converged (second line of the output), the convergence value is very small,
and ideally model rank is last (26/26 for instance)
2) We are confident we are not finding a local minima in the convergence ("Hessian was positive difinitive")
3) that the k is ~ 1 and the p value > 0.05
4) we want the lowest k (default is 25 or c(5, 5)) while not maxing out the wiggles
which is determined by k' (max number of wiggles) and edf (actual wiggles used)

Some notes on interpretation and tuning:

- If the p-value < 0.05, you need to add more complexity to the model (by increasing k)
- The lower the k means it's less likely to be over fit. 
- As long as the p value > 0.05 and k ~ 1, it's unlikely to be underfit.
- A higher k risks being overfit, and not generalizable. 
- If your edf is very close to k' (like, within 1 unit), you probably need to increase k

My rule of thumb for making sure you don't overfit: 
if the edf value of your model is less than the k' of a lower k, you might be overfit:

edf = 20, k' = 48 (implying the k = c(7,7), or 7x7-1), you'd want to try k = c(6,6),
which would have a k' of 35 (6x6-1), and ideally the edf would still be low with
a smaller parameterization space.

## CV Model 1

```{r}
valid_1 <- mainstem_plus %>% filter(year %in% cv_1_years)
train_1 <- anti_join(mainstem_plus, valid_1) %>% 
  filter(!(year %in% test_years))

gam_add_1 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                   s(canyon_mouth_cfs, bs = "tp", k = 15) +
                   s(distance_upstream_km, bs = "tp", k = 5),
                 data = train_1,
                 family = Gamma(link = "log"),
                 method = "REML")
summary(gam_add_1)

gam_ai_by_1 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                   s(canyon_mouth_cfs, bs = "tp", k = 15) +
                   ti(DOY, canyon_mouth_cfs, 
                      by = distance_upstream_km,
                      bs = c("cc", "tp"), k = c(10, 5)),
                 data = train_1,
                 family = Gamma(link = "log"),
                 method = "REML")
summary(gam_ai_by_1)

AIC(gam_add_1, gam_ai_by_1)

gam.check(gam_add_1)
gam.check(gam_ai_by_1)

# visualize 2d/3d
vis.gam(gam_add_1,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_add_1,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_ai_by_1,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_ai_by_1,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_1$pred_a <- exp(predict(gam_add_1, newdata = valid_1))
valid_1$pred_b <- exp(predict(gam_ai_by_1, newdata = valid_1))
ggplot(valid_1) +
  geom_point(aes(x = toc_doc, y = pred_a), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_b), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() 
ggplot(valid_1) +
  geom_point(aes(x = toc_doc, y = pred_a), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_b), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() +
  facet_grid(year ~ site_name)

# print unadjusted R2, which should be close to deviance explained from gam
summary(gam_add_1)$dev.expl
summary(lm(valid_1$pred_a ~ valid_1$toc_doc))$r.squared
summary(gam_ai_by_1)$dev.expl
summary(lm(valid_1$pred_b ~ valid_1$toc_doc))$r.squared
```

These have a pretty clear shelf at ~ 7mg/L, but we'll keep going. 

## CV Model 2

```{r}
valid_2 <- mainstem_plus %>% filter(year %in% cv_2_years)
train_2 <- anti_join(mainstem_plus, valid_2) %>% 
  filter(!(year %in% test_years))

gam_add_2 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                   s(canyon_mouth_cfs, bs = "tp", k = 15) +
                   s(distance_upstream_km, bs = "tp", k = 5),
                 data = train_2,
                 family = Gamma(link = "log"),
                 method = "REML")
summary(gam_add_2)

gam_ai_by_2 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                   s(canyon_mouth_cfs, bs = "tp", k = 15) +
                   ti(DOY, canyon_mouth_cfs, 
                      by = distance_upstream_km,
                      bs = c("cc", "tp"), k = c(10, 5)),
                 data = train_2,
                 family = Gamma(link = "log"),
                 method = "REML")
summary(gam_ai_by_2)

AIC(gam_add_2, gam_ai_by_2)

gam.check(gam_add_2)
gam.check(gam_ai_by_2)

# visualize 2d/3d
vis.gam(gam_add_2,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_add_2,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_ai_by_2,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_ai_by_2,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_2$pred_a <- exp(predict(gam_add_2, newdata = valid_2))
valid_2$pred_b <- exp(predict(gam_ai_by_2, newdata = valid_2))
ggplot(valid_2) +
  geom_point(aes(x = toc_doc, y = pred_a), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_b), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() 
ggplot(valid_2) +
  geom_point(aes(x = toc_doc, y = pred_a), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_b), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() +
  facet_grid(year ~ site_name)

# print unadjusted R2, which should be close to deviance explained from gam
summary(gam_add_2)$dev.expl
summary(lm(valid_2$pred_a ~ valid_2$toc_doc))$r.squared
summary(gam_ai_by_2)$dev.expl
summary(lm(valid_2$pred_b ~ valid_2$toc_doc))$r.squared
```

2009/2019 aren't great here, either - mostly on the upper end. Granted, performance
is similar to the previous CV, so we'll keep going

## CV Model 3

```{r}
valid_3 <- mainstem_plus %>% filter(year %in% cv_3_years)
train_3 <- anti_join(mainstem_plus, valid_3) %>% 
  filter(!(year %in% test_years))

gam_add_3 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                   s(canyon_mouth_cfs, bs = "tp", k = 15) +
                   s(distance_upstream_km, bs = "tp", k = 5),
                 data = train_3,
                 family = Gamma(link = "log"),
                 method = "REML")
summary(gam_add_3)

gam_ai_by_3 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                   s(canyon_mouth_cfs, bs = "tp", k = 15) +
                   ti(DOY, canyon_mouth_cfs, 
                      by = distance_upstream_km,
                      bs = c("cc", "tp"), k = c(10, 5)),
                 data = train_3,
                 family = Gamma(link = "log"),
                 method = "REML")
summary(gam_ai_by_3)

AIC(gam_add_3, gam_ai_by_3)

gam.check(gam_add_3)
gam.check(gam_ai_by_3)

# visualize 2d/3d
vis.gam(gam_add_3,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_add_3,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_ai_by_3,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_ai_by_3,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_3$pred_a <- exp(predict(gam_add_3, newdata = valid_3))
valid_3$pred_b <- exp(predict(gam_ai_by_3, newdata = valid_3))
ggplot(valid_3) +
  geom_point(aes(x = toc_doc, y = pred_a), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_b), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() 
ggplot(valid_3) +
  geom_point(aes(x = toc_doc, y = pred_a), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_b), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() +
  facet_grid(year ~ site_name)

# print unadjusted R2, which should be close to deviance explained from gam
summary(gam_add_3)$dev.expl
summary(lm(valid_3$pred_a ~ valid_3$toc_doc))$r.squared
summary(gam_ai_by_3)$dev.expl
summary(lm(valid_3$pred_b ~ valid_3$toc_doc))$r.squared
```

This looks pretty good across years

## CV Model 4

```{r}
valid_4 <- mainstem_plus %>% filter(year %in% cv_4_years)
train_4 <- anti_join(mainstem_plus, valid_4) %>% 
  filter(!(year %in% test_years))

gam_add_4 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                   s(canyon_mouth_cfs, bs = "tp", k = 15) +
                   s(distance_upstream_km, bs = "tp", k = 5),
                 data = train_4,
                 family = Gamma(link = "log"),
                 method = "REML")
summary(gam_add_4)

gam_ai_by_4 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 15) + 
                   s(canyon_mouth_cfs, bs = "tp", k = 15) +
                   ti(DOY, canyon_mouth_cfs, 
                      by = distance_upstream_km,
                      bs = c("cc", "tp"), k = c(10, 5)),
                 data = train_4,
                 family = Gamma(link = "log"),
                 method = "REML")
summary(gam_ai_by_4)

AIC(gam_add_4, gam_ai_by_4)

gam.check(gam_add_4)
gam.check(gam_ai_by_4)

# visualize 2d/3d
vis.gam(gam_add_4,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_add_4,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_ai_by_4,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_ai_by_4,
        view = c("canyon_mouth_cfs", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_4$pred_a <- exp(predict(gam_add_4, newdata = valid_4))
valid_4$pred_b <- exp(predict(gam_ai_by_4, newdata = valid_4))
ggplot(valid_4) +
  geom_point(aes(x = toc_doc, y = pred_a), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_b), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() 
ggplot(valid_4) +
  geom_point(aes(x = toc_doc, y = pred_a), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_b), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() +
  facet_grid(year ~ site_name)

# print unadjusted R2, which should be close to deviance explained from gam
summary(gam_add_4)$dev.expl
summary(lm(valid_4$pred_a ~ valid_4$toc_doc))$r.squared
summary(gam_ai_by_4)$dev.expl
summary(lm(valid_4$pred_b ~ valid_4$toc_doc))$r.squared
```

There's some strange nearly vertical predictions here, especially visible in 2024, but the other CV's might balance this.

## Apply to test set

Since performance across each of the CV sets is similar (albeit with wonky-ness 
in some years), we'll calculate for the test set.

```{r}
test <- mainstem_plus %>% filter(year %in% test_years)

# apply ensemble
test <- test %>%
  mutate(
    pred1_a = exp(predict(gam_add_1,newdata = test)),
    pred2_a = exp(predict(gam_add_2, newdata = test)),
    pred3_a = exp(predict(gam_add_3, newdata = test)),
    pred4_a = exp(predict(gam_add_4, newdata = test)),
    pred_avg_a = (pred1_a + pred2_a + pred3_a + pred4_a) / 4,
    pred1_b = exp(predict(gam_ai_by_1, newdata = test)),
    pred2_b = exp(predict(gam_ai_by_2, newdata = test)),
    pred3_b = exp(predict(gam_ai_by_3, newdata = test)),
    pred4_b = exp(predict(gam_ai_by_4, newdata = test)),
    pred_avg_b = (pred1_b + pred2_b + pred3_b + pred4_b) / 4
  )

# format annotation text for plot
metrics_text_a <- paste0(
  "MAE = ", round(mae(test$toc_doc, test$pred_avg_a), 2),
  "\nRMSE = ", round(rmse(test$toc_doc, test$pred_avg_a), 2),
  "\nBias = ", round(bias(test$toc_doc, test$pred_avg_a), 2),
  "\nR² = ", round(summary(lm(test$pred_avg_a ~ test$toc_doc))$r.squared, 2)
)

# format annotation text for plot
metrics_text_b <- paste0(
  "MAE = ", round(mae(test$toc_doc, test$pred_avg_b), 2),
  "\nRMSE = ", round(rmse(test$toc_doc, test$pred_avg_b), 2),
  "\nBias = ", round(bias(test$toc_doc, test$pred_avg_b), 2),
  "\nR² = ", round(summary(lm(test$pred_avg_b ~ test$toc_doc))$r.squared, 2)
)

# pred - obs scatter plot!
ggplot(test, aes(y = pred_avg_a, x = toc_doc)) +
  geom_point(color = "blue", alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (hold out)",
       subtitle = "GAM addtive: TOC ~ DOY + distance upstream + canyon mouth flow") +
  annotate("text", x = 1, y = 12, 
           hjust = 0, vjust = 1, 
           label = metrics_text_a, size = 4) +
  theme_bw(base_size = 14) +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12))

ggplot(test, aes(y = pred_avg_b, x = toc_doc)) +
  geom_point(color = "blue", alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (hold out)",
       subtitle = "GAM addtive/interaction by distance: TOC ~ DOY + canyon mouth flow") +
  annotate("text", x = 1, y = 12, 
           hjust = 0, vjust = 1, 
           label = metrics_text_b, size = 4) +
  theme_bw(base_size = 14) +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12))

```

Okay, this is better than just DOY/distance upstream,
which had the following metrics:

MAE: 1.0
RMSE: 1.35
Bias: 0.04
R2: 0.66

Let's roll with some performance diagnostics with the additive model.

First, a quick plot with each of the members of the ensemble:

```{r}
test %>% 
  select(pred1_a, pred2_a, pred3_a, pred4_a, toc_doc) %>% 
  pivot_longer(cols = c(pred1_a, pred2_a, pred3_a, pred4_a),
               names_to = "cv_model",
               values_to = "pred") %>% 
ggplot(., aes(y = pred, x = toc_doc, shape = cv_model, color = cv_model)) +
  geom_point() +
  scale_color_viridis_d() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM CV Member Predictions vs Observed (hold out)",
       subtitle = "GAM additive: TOC ~ DOY + distance upstream + canyon mouth flow") +
  theme_bw(base_size = 14) +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12))

```

No big outliers here across models, the line around 4 and 3 are from something else!

```{r, fig.dim=c(8, 8)}
# for kicks, look at performance by site
ggplot(test, aes(y = pred_avg_a, x = toc_doc)) +
  geom_point(color = "blue", alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (hold out)",
       subtitle = "TOC ~ DOY + distance upstream + canyon mouth flow") +
  theme_bw(base_size = 14) +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12)) +
  facet_wrap(site_name ~ .)
```

Looks like worst performance is at Poudre Below Rustic and Poudre Below S Fork -
let's see if there are performance differences by year.

```{r}
# and by year
ggplot(test, aes(y = pred_avg_a, x = toc_doc)) +
  geom_point(color = "blue", alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (hold out)",
       subtitle = "TOC ~ DOY + distance upstream + canyon mouth flow") +
  theme_bw(base_size = 14) +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12)) +
  facet_wrap(year ~ .)
```

Ooof. 2017 is pretty bad here. Let's facet by year and site.

```{r, fig.dim=c(12, 8)}
ggplot(test, aes(y = pred_avg_a, x = toc_doc)) +
  geom_point(color = "blue", alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (hold out)",
       subtitle = "TOC ~ DOY + distance upstream + canyon mouth flow") +
  theme_bw(base_size = 14) +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12)) +
  facet_wrap(year ~ site_name)
```

2017 seems to be the problem child here, maybe SS or someone at the city
knows why. It's also possible some of the performance oddities across the data 
might have something to do with snow melt rate, or maybe with total SWE, or 
some other climatalogical thing not included in this model.


# Do a quick sensitivity check

We're going to raise the canyon flow by 25% and see if TOC increases here. This
will give us a better idea of whether or not adding flow actually might be better
in an operational scenario.

```{r}
test_sens <- test %>% 
  mutate(canyon_mouth_cfs = canyon_mouth_cfs*1.25)

# apply ensemble for test set
test_sens <- test_sens %>%
  mutate(
    pred1_25 = exp(predict(gam_add_1, newdata = test_sens)),
    pred2_25 = exp(predict(gam_add_2, newdata = test_sens)),
    pred3_25 = exp(predict(gam_add_3, newdata = test_sens)),
    pred4_25 = exp(predict(gam_add_4, newdata = test_sens)),
    pred_avg_25 = (pred1_25 + pred2_25 + pred3_25 + pred4_25) / 4
  ) %>%
  select(date, DOY, year, site_name, pred_avg_25)

# and join with original file
test_sens <- test %>% 
  left_join(., test_sens)

# bias should change
bias(test_sens$toc_doc, test_sens$pred_avg_a)
bias(test_sens$toc_doc, test_sens$pred_avg_25)

# plot the two pred/obs to see the difference
test_sens %>% 
    select(pred_avg_a, pred_avg_25, toc_doc) %>% 
  pivot_longer(cols = c(pred_avg_a, pred_avg_25),
               names_to = "data",
               values_to = "pred") %>% 
ggplot(., aes(y = pred, x = toc_doc, color = data)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (2025)",
       subtitle = "TOC ~ DOY + distance upstream + 1.25 * canyon mouth flow") +
  theme_bw(base_size = 14) +
  scale_color_viridis_d() +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12))

# and against one another 
ggplot(test_sens, aes(x = pred_avg_a, y = pred_avg_25, color = factor(year))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  theme_bw() +
  labs(x = "GAM predicted",
       y = "GAM predicted 25% increase in flow",
       color = "year") +
  scale_color_viridis_d()

```

Okay, there's some sensitivity here - but there's also some interesting patterns
where an increase in Q actually results in a decrease in TOC, but I can't quite
pin down why... some random combination of values, I guess. 

# Visualize the decision space for interpretability

GAMs are pretty interpretable when they have fewer interacting variables. In this
case, we can just create a synthetic space/time space and apply the model across
it to visualize what happens up the canyon over time.

```{r}
# Make a prediction grid across DOY and distance
newdata <- expand.grid(
  DOY = seq(min(mainstem_plus$DOY), 
            max(mainstem_plus$DOY), 
            5),
  distance_upstream_km = seq(min(mainstem_plus$distance_upstream_km),
                             max(mainstem_plus$distance_upstream_km), 10),
  canyon_mouth_cfs = seq(min(mainstem_plus$canyon_mouth_cfs),
                         max(mainstem_plus$canyon_mouth_cfs),
                         100)
)

# Get predictions from your GAMs for the grid
newdata$fit1 <- exp(predict(gam_add_1, newdata = newdata))
newdata$fit2 <- exp(predict(gam_add_2, newdata = newdata))
newdata$fit3 <- exp(predict(gam_add_3, newdata = newdata))
newdata$fit4 <- exp(predict(gam_add_4, newdata = newdata))

# create ensemble prediction
newdata <- newdata %>% 
  mutate(ave_fit = (fit1 + fit2 + fit3 + fit4) / 4)

# heatmap + contour plot
ggplot(newdata, aes(x = DOY, y = distance_upstream_km, fill = ave_fit)) +
  geom_tile() +
  geom_contour(aes(z = ave_fit), color = "white", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    fill = "Partial TOC",
    x = "Day of Year",
    y = "Distance upstream from Canyon Mouth (km)"
  ) +
  theme_minimal(base_size = 14) 

ggplot(newdata, aes(x = distance_upstream_km, y = canyon_mouth_cfs, fill = ave_fit)) +
  geom_tile() +
  geom_contour(aes(z = ave_fit), color = "white", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    fill = "Partial TOC",
    y = "Q at the canyon mouth",
    x = "Distance upstream from Canyon Mouth (km)"
  ) +
  theme_minimal(base_size = 14) 

ggplot(newdata, aes(x = DOY, y = canyon_mouth_cfs, fill = ave_fit)) +
  geom_tile() +
  geom_contour(aes(z = ave_fit), color = "white", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    fill = "Partial TOC",
    x = "Day of Year",
    y = "Q at the canyon mouth"
  ) +
  theme_minimal(base_size = 14) 

```

This might actually explain that decrease in TOC with an increase in Q!


Save models

```{r}
# write_rds(gam_add_1, here("data/derived/models/TOC_distributed/working/TOC_GAM_Q_add_fit1_v2026-02-04.rds"))
# write_rds(gam_add_2, here("data/derived/models/TOC_distributed/working/TOC_GAM_Q_add_fit2_v2026-02-04.rds"))
# write_rds(gam_add_3, here("data/derived/models/TOC_distributed/working/TOC_GAM_Q_add_fit3_v2026-02-04.rds"))
# write_rds(gam_add_4, here("data/derived/models/TOC_distributed/working/TOC_GAM_Q_add_fit4_v2026-02-04.rds"))

```
