---
title: "Baseline TOC model development in CLP"
date: "2026-01-28"
output: html_document
---

# Purpsoe

The purpose of this markdown is to explore the relationship between TOC and 
snowmelt signals within the CLP, with the intention of creating a baseline 
forecast model to measure further model development against.

```{r, echo = F, error = F, message = F, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
library(tidyverse)
library(here)
library(mgcv)
library(Metrics)
```

```{r}
source(here("src/collation/load_grab_sample_data.R"))
# grab chemistry data
chem <- load_grab_sample_data() %>% 
  # drop intake data
  filter(site_code != "prw") %>% 
  # check for DOC/TOC data
  filter(!is.na(TOC) | !is.na(DOC)) %>% 
  # only grab columns we need right now
  select(Date, site_name, site_code, collector, TOC, DOC, flag,
         Campaign, location_type, watershed, distance_upstream_km) 

# check for flags
unique(chem$flag)
```

No flags. Make sure only unique values, there are some straight-up dupes here.

```{r}
chem <- chem %>% 
  unique()
```

# Exploring DOC/TOC interoperability

Look at relationship between TOC and DOC, does this get us some "more" data that
we can use DOC "as" TOC. 

```{r}
chem %>% 
  filter(!is.na(TOC) & !is.na(DOC)) %>% 
  select(Date, site_name, site_code, collector, TOC, DOC, Campaign, location_type, watershed) %>% 
  unique %>% 
  ggplot(., aes(x = TOC, y = DOC)) + 
  geom_point() + 
  facet_wrap(site_code ~ .)
```

The answer is, kind of yes. Exceptions: basically everything after Canyon Mouth. 
Let's also look at water source type for this:

```{r}
chem %>% 
  filter(!is.na(TOC) & !is.na(DOC)) %>% 
  ggplot(., aes(x = TOC, y = DOC)) + 
  geom_point() + 
  facet_wrap(location_type ~ .)
```

Let's look at mainstem here:

```{r}
chem %>% 
  filter(!is.na(TOC) & !is.na(DOC) & location_type == "Mainstem") %>% 
  ggplot(., aes(x = TOC, y = DOC)) + 
  geom_point() + 
  facet_wrap(Campaign ~ .)
```

Okay, this aligns with the relationship falling apart after the Canyon Mouth.

```{r}
chem %>% 
  filter(!is.na(TOC) & !is.na(DOC) & grepl("Upper", Campaign)) %>% 
  ggplot(., aes(x = TOC, y = DOC)) + geom_point() + 
  facet_wrap(site_code ~ .)
```

Let's also look at the South Fork

```{r}
chem %>% 
  filter(!is.na(TOC) & !is.na(DOC) & Campaign == "South Fork") %>% 
  ggplot(., aes(x = TOC, y = DOC)) + geom_point() + 
  facet_wrap(site_code ~ .)
```

Great, so let's look at DOC across the upper sites and South Fork:

```{r}
chem %>% 
  filter(grepl("Upper", Campaign) | Campaign == "South Fork") %>% 
  mutate(dist_name = paste(distance_upstream_km, site_code, sep = " - ")) %>% 
  ggplot(., aes(x = yday(Date), y = DOC, color = year(Date))) + 
  geom_point() + 
  facet_wrap(dist_name ~.)
```

Okay, these are all right-on - the correlation is something like 0.97, the intercept
is nearly zero and the slope is nearly 1, so we're just going to say that for these
sites, DOC = TOC. 

And removing post-fire 2021, 2022, since we'll assume that DOC is not as dominant 
of a part of TOC during this time, or at the very least might be a different
proportion of TOC during this recovery time. (We only have DOC-TOC matches 
post-fire.)

```{r}
chem %>% 
  filter(!year(Date) %in% c(2021, 2022) & grepl("Upper|South", Campaign)) %>% 
  mutate(dist_name = paste(distance_upstream_km, site_name, sep = " - "),
         ordered_name = factor(site_name, levels = c("Poudre at Bellvue Diversion",
                                                     "Poudre Above North Fork",
                                                     "Poudre at Manner's Bridge",
                                                     "Poudre Below S. Fork",
                                                     "Poudre Below Rustic",
                                                     "Little Beaver",
                                                     "Pennock (Signal Mountain)",
                                                     "Poudre below Poudre Falls",
                                                     "Poudre Above Joe Wright",
                                                     "Sawmill"))) %>%
  ggplot(., aes(x = yday(Date), y = DOC, color = year(Date))) + 
  geom_point() + 
  facet_wrap(site_name + distance_upstream_km + Campaign ~.)
```

The consistency between 2015 and post-fire leads me to believe that there is 
probably consistency in the relationship between DOC and TOC in 2015 data and 
post-fire data, since it doesn't seem significantly larger or smaller than pre-fire.

For now, let's just limit this to Mainstem - Upper, Upper Tributaries, and South 
Fork Campaigns and Mainstem, Stream, Inflow location types

```{r}
chem %>% 
  filter(grepl("Upper|South", Campaign) & 
           location_type %in% c("Mainstem", "Stream", "Inflow") &
           !year(Date) %in% c(2021, 2022)) %>% 
  mutate(dist_name = paste(distance_upstream_km, site_name, sep = " - "),
         ordered_name = factor(site_name, levels = c("Poudre at Bellvue Diversion",
                                                     "Poudre Above North Fork",
                                                     "Poudre at Manner's Bridge",
                                                     "Poudre Below S. Fork", 
                                                     "Poudre Below Rustic",
                                                     "Little Beaver",
                                                     "Poudre South Fork",
                                                     "Pennock (Signal Mountain)",
                                                     "Hourglass Inflow",
                                                     "Beaver",
                                                     "Comanche Inflow",
                                                     "Poudre at Sleeping Elephant",
                                                     "Poudre below Poudre Falls",
                                                     "Poudre Above Joe Wright",
                                                     "Sawmill"))) %>% 
  ggplot(., aes(x = yday(Date), y = DOC, color = year(Date))) + 
  geom_point() + 
  facet_wrap(ordered_name + distance_upstream_km ~.)
```

# GAM time

Great, now let's play with a GAM fit, for now, we won't do a TVT, because I just 
want to see if it works. I have played with including the Upper Tribs, but they
don't add much value here, so leaving out.

```{r}
mainstem <- chem %>% 
  filter(Campaign %in% c("Mainstem - Upper", "South Fork") & 
           location_type %in% c("Mainstem", "Stream", "Inflow") &
           # remove inflows
           !(site_name %in% c("Poudre Above Joe Wright",
                              "Beaver",
                              "Comanche Inflow",
                              "Hourglass Inflow"))) %>% 
  mutate(DOY = yday(Date),
         year = year(Date), 
         toc_doc = if_else(!is.na(TOC), TOC, DOC),
         # to play with categorical variables, they failed.
         identity = if_else(Campaign == "South Fork", 1, 0),
         fire_signal = if_else(year %in% c(2021, 2022), 1, 0)) %>% 
  # limit time, remove fire signal, remove flier
  filter(!is.na(toc_doc), between(DOY, 100, 320), 
         fire_signal == 0) %>% 
  # remove data after the 2013 flood (Sept 9, 2013)
  filter(!(Date >= "2013-09-2013" & year == 2013)) %>% 
  rename(date = Date)

# let's also make sure we only have one sample per day per site
mainstem <- mainstem %>% 
  summarize(toc_doc = mean(toc_doc),
            .by = c(date, site_name, DOY, year, distance_upstream_km))

ggplot(mainstem, aes(x = DOY, y = toc_doc)) + 
  geom_point() + 
  facet_wrap(site_name ~ .) +
  theme_bw()

# also recode the flier in south fork and manners bridge
mainstem <- mainstem %>% 
  filter(!(site_name == "Poudre South Fork" & DOY > 200 & toc_doc > 9)) %>% 
  filter(!(site_name == "Poudre at Manner's Bridge" & DOY > 200 & toc_doc >5))

ggplot(mainstem, aes(x = DOY, y = toc_doc)) + 
  geom_point() + 
  facet_wrap(site_name ~ .) +
  theme_bw()

gam_linear <- gam(toc_doc ~ DOY + distance_upstream_km,
                  bs = c("cc", "tp"),
                  k = c(10, 5),
                  data = mainstem,
                  family = Gamma(link = "log"),
                  method = "REML")
summary(gam_linear, 0)

gam_add <- gam(toc_doc ~ s(DOY, bs = "cc", k = 10) + 
                 s(distance_upstream_km, bs = "tp", k = 5),
               data = mainstem,
               family = Gamma(link = "log"),
               method = "REML")
summary(gam_add, 0)

gam_add_int <- gam(toc_doc ~ s(DOY, bs = "cc", k = 10) + 
                     s(distance_upstream_km, bs = "tp", k = 5) +
                     ti(DOY, distance_upstream_km, bs = c("cc", "tp")),
                   data = mainstem,
                   family = Gamma(link = "log"),
                   method = "REML")
summary(gam_add_int, 0)

gam_tensor <- gam(toc_doc ~ te(DOY, distance_upstream_km, 
                               bs = c("cc", "tp"), k = c(10, 5)),
               data = mainstem,
               family = Gamma(link = "log"),
               method = "REML")
summary(gam_tensor, 0)

AIC(gam_linear, gam_add, gam_add_int, gam_tensor)

gam.check(gam_add_int)
gam.check(gam_tensor)
```

The best models here are the additive gam with an interaction
(gam_add_int) and the tensor gam (gam_tensor). This is a pretty
difficult decision space: AIC and DF are better in the interaction
gam, but the gam.check is better on the tensor. I think we try both,
and see if either performs better in the CV.


```{r}
mainstem %>% 
  summarize(n = n(), 
            .by = year) %>% 
  arrange(year)
```


We'll do some kinda random, but not really random splits here:

- Test: 2010, 2017, 2025
- CV 1: 2008, 2013, 2018
- CV 2: 2009, 2014, 2019
- CV 3: 2011, 2015, 2023
- CV 4: 2012, 2016, 2024

```{r}
test_years <- c(2010, 2017, 2025)
cv_1_years <- c(2008, 2013, 2018)
cv_2_years <- c(2009, 2014, 2019)
cv_3_years <- c(2011, 2015, 2023)
cv_4_years <- c(2012, 2016, 2024)
```

GAM diagnostics are interpreted using the gam.check() function. We want to make
sure that:

1) the model converged (second line of the output), the convergnece value is very small,
and ideally model rank is last (26/26 for instance)
2) We are confident we are not finding a local minima in the convergence ("Hessian was positive difinitive")
3) that the k is ~ 1 and the p value > 0.05
4) we want the lowest k (default is 25 or c(5, 5)) while not maxing out the wiggles
which is determined by k' (max number of wiggles) and edf (actual wiggles used)

Some notes on interpretation and tuning:

- If the p-value < 0.05, you need to add more complexity to the model (by increasing k)
- The lower the k means it's less likely to be over fit. 
- As long as the p value > 0.05 and k ~ 1, it's unlikely to be underfit.
- A higher k risks being overfit, and not generalizable. 
- If your edf is very close to k' (like, within 1 unit), you probably need to increase k

My rule of thumb for making sure you don't overfit: 
if the edf value of your model is less than the k' of a lower k, you might be overfit:

edf = 20, k' = 48 (implying the k = c(7,7), or 7x7-1), you'd want to try k = c(6,6),
which would have a k' of 35 (6x6-1), and ideally the edf would still be low with
a smaller parameterization space.


## CV Model 1 

```{r}
valid_1 <- mainstem %>% filter(year %in% cv_1_years)
train_1 <- anti_join(mainstem, valid_1) %>% 
  filter(!(year %in% test_years))

gam_add_int_1 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 10) + 
                     s(distance_upstream_km, bs = "tp", k = 5) +
                     ti(DOY, distance_upstream_km, bs = c("cc", "tp")),
                   data = train_1,
                   family = Gamma(link = "log"),
                   method = "REML")
summary(gam_add_int_1)

gam_tensor_1 <- gam(toc_doc ~ te(DOY, distance_upstream_km, 
                               bs = c("cc", "tp"), k = c(10, 5)),
               data = train_1,
               family = Gamma(link = "log"),
               method = "REML")
summary(gam_tensor_1)

AIC(gam_add_int_1, gam_tensor_1)

gam.check(gam_add_int_1)
gam.check(gam_tensor_1)

# visualize 2d/3d
vis.gam(gam_add_int_1,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_tensor_1,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_1$pred_i <- exp(predict(gam_add_int_1, newdata = valid_1))
valid_1$pred_t <- exp(predict(gam_tensor_1, newdata = valid_1))
ggplot(valid_1) +
  geom_point(aes(x = toc_doc, y = pred_i), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_t), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() 
ggplot(valid_1) +
  geom_point(aes(x = toc_doc, y = pred_i), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_t), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() +
  facet_grid(year ~ site_name)

# print unadjusted R2, which should be close to deviance explained from gam
summary(gam_add_int_1)$dev.expl
summary(lm(valid_1$pred_i ~ valid_1$toc_doc))$r.squared
summary(gam_tensor_1)$dev.expl
summary(lm(valid_1$pred_t ~ valid_1$toc_doc))$r.squared
```

Similar pattern to the all-data gams for general
performance/diagnostics, except the df are much closer between the
two models. Also 2008 and 2013 aren't performing very
well here, so we're missing some generalization.

## CV Model 2

```{r}
valid_2 <- mainstem %>% filter(year %in% cv_2_years)
train_2 <- anti_join(mainstem, valid_2) %>% 
  filter(!(year %in% test_years))

gam_add_int_2 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 10) + 
                     s(distance_upstream_km, bs = "tp", k = 5) +
                     ti(DOY, distance_upstream_km, bs = c("cc", "tp")),
                   data = train_2,
                   family = Gamma(link = "log"),
                   method = "REML")
summary(gam_add_int_2)

gam_tensor_2 <- gam(toc_doc ~ te(DOY, distance_upstream_km, 
                               bs = c("cc", "tp"), k = c(10, 5)),
               data = train_2,
               family = Gamma(link = "log"),
               method = "REML")
summary(gam_tensor_2)

AIC(gam_add_int_2, gam_tensor_2)

gam.check(gam_add_int_2)
gam.check(gam_tensor_2)

# visualize 2d/3d
vis.gam(gam_add_int_2,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_tensor_2,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_2$pred_i <- exp(predict(gam_add_int_2, newdata = valid_2))
valid_2$pred_t <- exp(predict(gam_tensor_2, newdata = valid_2))
ggplot(valid_2) +
  geom_point(aes(x = toc_doc, y = pred_i), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_t), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() 
ggplot(valid_2) +
  geom_point(aes(x = toc_doc, y = pred_i), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_t), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() +
  facet_grid(year ~ site_name)

# print unadjusted R2, which should be close to deviance explained from gam
summary(gam_add_int_2)$dev.expl
summary(lm(valid_2$pred_i ~ valid_2$toc_doc))$r.squared
summary(gam_tensor_2)$dev.expl
summary(lm(valid_2$pred_t ~ valid_2$toc_doc))$r.squared
```

General under performance in 2014 (arguably in 2019 too)

## CV Model 3

```{r}
valid_3 <- mainstem %>% filter(year %in% cv_3_years)
train_3 <- anti_join(mainstem, valid_3) %>% 
  filter(!(year %in% test_years))

gam_add_int_3 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 10) + 
                     s(distance_upstream_km, bs = "tp", k = 5) +
                     ti(DOY, distance_upstream_km, bs = c("cc", "tp")),
                   data = train_3,
                   family = Gamma(link = "log"),
                   method = "REML")
summary(gam_add_int_3)

gam_tensor_3 <- gam(toc_doc ~ te(DOY, distance_upstream_km, 
                               bs = c("cc", "tp"), k = c(10, 5)),
               data = train_3,
               family = Gamma(link = "log"),
               method = "REML")
summary(gam_tensor_3)

AIC(gam_add_int_3, gam_tensor_3)

gam.check(gam_add_int_3)
gam.check(gam_tensor_3)

# visualize 2d/3d
vis.gam(gam_add_int_3,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_tensor_3,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_3$pred_i <- exp(predict(gam_add_int_3, newdata = valid_3))
valid_3$pred_t <- exp(predict(gam_tensor_3, newdata = valid_3))
ggplot(valid_3) +
  geom_point(aes(x = toc_doc, y = pred_i), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_t), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() 
ggplot(valid_3) +
  geom_point(aes(x = toc_doc, y = pred_i), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_t), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() +
  facet_grid(year ~ site_name)

# print unadjusted R2, which should be close to deviance explained from gam
summary(gam_add_int_3)$dev.expl
summary(lm(valid_3$pred_i ~ valid_3$toc_doc))$r.squared
summary(gam_tensor_3)$dev.expl
summary(lm(valid_3$pred_t ~ valid_3$toc_doc))$r.squared
```

Super not great performance in 2011.

## CV Model 4

```{r}
valid_4 <- mainstem %>% filter(year %in% cv_4_years)
train_4 <- anti_join(mainstem, valid_4) %>% 
  filter(!(year %in% test_years))

gam_add_int_4 <- gam(toc_doc ~ s(DOY, bs = "cc", k = 10) + 
                     s(distance_upstream_km, bs = "tp", k = 5) +
                     ti(DOY, distance_upstream_km, bs = c("cc", "tp")),
                   data = train_4,
                   family = Gamma(link = "log"),
                   method = "REML")
summary(gam_add_int_4)

gam_tensor_4 <- gam(toc_doc ~ te(DOY, distance_upstream_km, 
                               bs = c("cc", "tp"), k = c(10, 5)),
               data = train_4,
               family = Gamma(link = "log"),
               method = "REML")
summary(gam_tensor_4)

AIC(gam_add_int_4, gam_tensor_4)

gam.check(gam_add_int_4)
gam.check(gam_tensor_4)

# visualize 2d/3d
vis.gam(gam_add_int_4,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")
vis.gam(gam_tensor_4,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_4$pred_i <- exp(predict(gam_add_int_4, newdata = valid_4))
valid_4$pred_t <- exp(predict(gam_tensor_4, newdata = valid_4))

ggplot(valid_4) +
  geom_point(aes(x = toc_doc, y = pred_i), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_t), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() 

ggplot(valid_4) +
  geom_point(aes(x = toc_doc, y = pred_i), color = "blue", alpha = 0.8) +
  geom_point(aes(x = toc_doc, y = pred_t), color = "purple", alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") +
  theme_bw() +
  facet_grid(year ~ site_name)

# print unadjusted R2, which should be close to deviance explained from gam
summary(gam_add_int_4)$dev.expl
summary(lm(valid_4$pred_i ~ valid_4$toc_doc))$r.squared
summary(gam_tensor_4)$dev.expl
summary(lm(valid_4$pred_t ~ valid_4$toc_doc))$r.squared
```

Some over prediction in 2024, some under prediction in 2016, some 
wacky data in 2012.

# Apply ensemble to test set

```{r}
test <- mainstem %>% filter(year %in% test_years)

# apply ensemble
test <- test %>%
  mutate(
    pred1_i = exp(predict(gam_add_int_1, newdata = test)),
    pred2_i = exp(predict(gam_add_int_2, newdata = test)),
    pred3_i = exp(predict(gam_add_int_3, newdata = test)),
    pred4_i = exp(predict(gam_add_int_4, newdata = test)),
    pred_avg_i = (pred1_i + pred2_i + pred3_i + pred4_i) / 4,
    pred1_t = exp(predict(gam_tensor_1, newdata = test)),
    pred2_t = exp(predict(gam_tensor_2, newdata = test)),
    pred3_t = exp(predict(gam_tensor_3, newdata = test)),
    pred4_t = exp(predict(gam_tensor_4, newdata = test)),
    pred_avg_t = (pred1_t + pred2_t + pred3_t + pred4_t) / 4
  )


# format annotation text for plot
metrics_text_i <- paste0(
  "MAE = ", round(mae(test$toc_doc, test$pred_avg_i), 2),
  "\nRMSE = ", round(rmse(test$toc_doc, test$pred_avg_i), 2),
  "\nBias = ", round(bias(test$toc_doc, test$pred_avg_i), 2),
  "\nR² = ", round(summary(lm(test$pred_avg_i ~ test$toc_doc))$r.squared, 2)
)

# format annotation text for plot
metrics_text_t <- paste0(
  "MAE = ", round(mae(test$toc_doc, test$pred_avg_t), 2),
  "\nRMSE = ", round(rmse(test$toc_doc, test$pred_avg_t), 2),
  "\nBias = ", round(bias(test$toc_doc, test$pred_avg_t), 2),
  "\nR² = ", round(summary(lm(test$pred_avg_t ~ test$toc_doc))$r.squared, 2)
)


# pred - obs scatter plot!
ggplot(test) +
  geom_point(aes(y = pred_avg_i, x = toc_doc),
             alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (holdout)",
       subtitle = "additive/interaction GAM: TOC ~ DOY + distance upstream") +
  annotate("text", x = 1, y = 12, 
           hjust = 0, vjust = 1, 
           label = metrics_text_i, size = 4) +
  theme_bw(base_size = 14) +
  scale_color_viridis_c() +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12))

ggplot(test) +
  geom_point(aes(y = pred_avg_t, x = toc_doc),
             alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (holdout)",
       subtitle = "simple tensor GAM: TOC ~ DOY + distance upstream") +
  annotate("text", x = 1, y = 12, 
           hjust = 0, vjust = 1, 
           label = metrics_text_t, size = 4) +
  theme_bw(base_size = 14) +
  scale_color_viridis_c() +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12))
```

Okay, these are pretty darn close in performance. For now, let's go
with the additive interaction GAM for the rest of these plots.

Let's dig into that with year
```{r}
ggplot(test, aes(y = pred_avg_i, x = toc_doc, color = factor(year))) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (holdout)",
       subtitle = "additive/interaction GAM: TOC ~ DOY + distance upstream",
       color = "year") +
  annotate("text", x = 1, y = 12, 
           hjust = 0, vjust = 1, 
           label = metrics_text_i, size = 4) +
  theme_bw(base_size = 14) +
  scale_color_viridis_d() +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12))
```

And also look at site / year
```{r, fig.dim=c(12, 8)}
# for kicks, look at performance by site and DOY
ggplot(test, aes(y = pred_avg_i, x = toc_doc)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (holdout)",
       subtitle = "additive/interaction GAM: TOC ~ DOY + distance upstream") +
  theme_bw(base_size = 14) +
  scale_color_viridis_c() +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12)) +
  facet_wrap(year ~ site_name)

```

And let's look at DOY

```{r, fig.dim=c(12, 8)}
ggplot(test, aes(y = pred_avg_i, x = toc_doc, color = DOY)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (holdout)",
       subtitle = "additive/interaction GAM: TOC ~ DOY + distance upstream",
       color = "day of year") +
  theme_bw(base_size = 14) +
  scale_color_viridis_c() +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12)) +
  facet_wrap(year ~ site_name)
```

There's clearly some variation at start/end of season that isn't being captured 
in the model. Based on our other work, this probably has something to do with 
snowpack/melt patterns.

# Visualize the decision space for interpretability

GAMs are pretty interpretable when they have fewer interacting variables. In this
case, we can just create a synthetic space/time space and apply the model across
it to visualize what happens up the canyon over time.

```{r}
# Make a prediction grid across DOY and distance
newdata <- expand.grid(
  DOY = seq(min(mainstem$DOY), max(mainstem$DOY)),
  distance_upstream_km = seq(min(mainstem$distance_upstream_km),
                             max(mainstem$distance_upstream_km))
)

# Get predictions from your GAMs for the grid
newdata$fit1_i <- exp(predict(gam_add_int_1, newdata = newdata))
newdata$fit2_i <- exp(predict(gam_add_int_2, newdata = newdata))
newdata$fit3_i <- exp(predict(gam_add_int_3, newdata = newdata))
newdata$fit4_i <- exp(predict(gam_add_int_4, newdata = newdata))

# create ensemble prediction
newdata <- newdata %>% 
  mutate(ave_fit = (fit1_i + fit2_i + fit3_i + fit4_i) / 4)

# heatmap + contour plot
ggplot(newdata, aes(x = DOY, y = distance_upstream_km, fill = ave_fit)) +
  geom_tile() +
  geom_contour(aes(z = ave_fit), color = "white", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    fill = "Predicted TOC",
    x = "Day of Year",
    y = "Distance upstream from Canyon Mouth (km)"
  ) +
  theme_minimal(base_size = 14) 

```

And let's save these models as the baseline:

```{r}

write_rds(gam_add_int_1, here("data/derived/models/TOC_distributed/TOC_GAM_s_ti_fit1_v2026-02-04.rds"))
write_rds(gam_add_int_2, here("data/derived/models/TOC_distributed/TOC_GAM_s_ti_fit2_v2026-02-04.rds"))
write_rds(gam_add_int_2, here("data/derived/models/TOC_distributed/TOC_GAM_s_ti_fit3_v2026-02-04.rds"))
write_rds(gam_add_int_2, here("data/derived/models/TOC_distributed/TOC_GAM_s_ti_fit4_v2026-02-04.rds"))

```






