---
title: "Baseline TOC model development in CLP"
date: "2026-01-28"
output: html_document
---

# Purpsoe

The purpose of this markdown is to explore the relationship between TOC and 
snowmelt signals within the CLP, with the intention of creating a baseline 
forecast model to measure further model development against.

```{r, echo = F, error = F, message = F, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
library(tidyverse)
library(here)
library(mgcv)
library(Metrics)
```

```{r}
source(here("src/collation/load_grab_sample_data.R"))
# grab chemistry data
chem <- load_grab_sample_data() %>% 
  filter(site_code != "prw")
# and location info
locations <- read_csv(here("data/raw/spatial/metadata/water_sampling_sites.csv"))

chem <- left_join(chem, locations)
```

# Exploring DOC/TOC interoperability

Look at relationship between TOC and DOC, does this get us some "more" data that
we can use DOC "as" TOC. 

```{r}
chem %>% 
  filter(!is.na(TOC) & !is.na(DOC)) %>% 
  ggplot(., aes(x = TOC, y = DOC)) + 
  geom_point() + 
  facet_wrap(site_code ~ .)
```

The answer is, kind of yes. Exceptions: basically everything after Canyon Mouth. 
Let's also look at water source type for this:

```{r}
chem %>% 
  filter(!is.na(TOC) & !is.na(DOC)) %>% 
  ggplot(., aes(x = TOC, y = DOC)) + 
  geom_point() + 
  facet_wrap(location_type ~ .)
```

Let's look at mainstem here:

```{r}
chem %>% 
  filter(!is.na(TOC) & !is.na(DOC) & location_type == "Mainstem") %>% 
  ggplot(., aes(x = TOC, y = DOC)) + 
  geom_point() + 
  facet_wrap(Campaign ~ .)
```

Okay, this aligns with the relationship falling apart after the Canyon Mouth.

```{r}
chem %>% 
  filter(!is.na(TOC) & !is.na(DOC) & grepl("Upper", Campaign)) %>% 
  ggplot(., aes(x = TOC, y = DOC)) + geom_point() + 
  facet_wrap(site_code ~ .)
```

Let's also look at the South Fork

```{r}
chem %>% 
  filter(!is.na(TOC) & !is.na(DOC) & Campaign == "South Fork") %>% 
  ggplot(., aes(x = TOC, y = DOC)) + geom_point() + 
  facet_wrap(site_code ~ .)
```

Great, so let's look at DOC across the upper sites and South Fork:

```{r}
chem %>% 
  filter(grepl("Upper", Campaign) | Campaign == "South Fork") %>% 
  mutate(dist_name = paste(distance_upstream_km, site_code, sep = " - ")) %>% 
  ggplot(., aes(x = yday(Date), y = DOC, color = year(Date))) + 
  geom_point() + 
  facet_wrap(dist_name ~.)
```

Okay, these are all right-on - the correlation is something like 0.97, the intercept
is nearly zero and the slope is nearly 1, so we're just going to say that for these
sites, DOC = TOC. 

And removing post-fire 2021, 2022, since we'll assume that DOC is not as dominant 
of a part of TOC during this time, or at the very least might be a different
proportion of TOC during this recovery time. (We only have DOC-TOC matches 
post-fire.)

```{r}
chem %>% 
  filter(!year(Date) %in% c(2021, 2022) & grepl("Upper|South", Campaign)) %>% 
  mutate(dist_name = paste(distance_upstream_km, site_name, sep = " - "),
         ordered_name = factor(site_name, levels = c("Poudre at Bellvue Diversion",
                                                     "Poudre Above North Fork",
                                                     "Poudre at Manner's Bridge",
                                                     "Poudre Below S. Fork",
                                                     "Poudre Below Rustic",
                                                     "Little Beaver",
                                                     "Pennock (Signal Mountain)",
                                                     "Poudre below Poudre Falls",
                                                     "Poudre Above Joe Wright",
                                                     "Sawmill"))) %>%
  ggplot(., aes(x = yday(Date), y = DOC, color = year(Date))) + 
  geom_point() + 
  facet_wrap(site_name + distance_upstream_km + Campaign ~.)
```

The consistency between 2015 and post-fire leads me to believe that there is 
probably consistency in the relationship between DOC and TOC in 2015 data and 
post-fire data, since it doesn't seem significantly larger or smaller than pre-fire.

For now, let's just limit this to Mainstem - Upper, Upper Tributaries, and South 
Fork Campaigns and Mainstem, Stream, Inflow location types

```{r}
chem %>% 
  filter(grepl("Upper|South", Campaign) & 
           location_type %in% c("Mainstem", "Stream", "Inflow") &
           !year(Date) %in% c(2021, 2022)) %>% 
  mutate(dist_name = paste(distance_upstream_km, site_name, sep = " - "),
         ordered_name = factor(site_name, levels = c("Poudre at Bellvue Diversion",
                                                     "Poudre Above North Fork",
                                                     "Poudre at Manner's Bridge",
                                                     "Poudre Below S. Fork", 
                                                     "Poudre Below Rustic",
                                                     "Little Beaver",
                                                     "Poudre South Fork",
                                                     "Pennock (Signal Mountain)",
                                                     "Hourglass Inflow",
                                                     "Beaver",
                                                     "Comanche Inflow",
                                                     "Poudre at Sleeping Elephant",
                                                     "Poudre below Poudre Falls",
                                                     "Poudre Above Joe Wright",
                                                     "Sawmill"))) %>% 
  ggplot(., aes(x = yday(Date), y = DOC, color = year(Date))) + 
  geom_point() + 
  facet_wrap(ordered_name + distance_upstream_km ~.)
```

# GAM time

Great, now let's play with a GAM fit, for now, we won't do a TVT, because I just 
want to see if it works. I have played with including the Upper Tribs, but they
don't add much value here, so leaving out.

```{r}
mainstem <- chem %>% 
  filter(Campaign %in% c("Mainstem - Upper", "South Fork") & 
           location_type %in% c("Mainstem", "Stream", "Inflow")) %>% 
  mutate(DOY = yday(Date),
         year = year(Date), 
         toc_doc = if_else(!is.na(TOC), TOC, DOC),
         # to play with categorical variables, they failed.
         identity = if_else(Campaign == "South Fork", 1, 0),
         fire_signal = if_else(year %in% c(2021, 2022), 1, 0),
         # recode the flier in south fork
         toc_doc = if_else(toc_doc > 9 & site_name == "Poudre South Fork", 
                           NA, 
                           toc_doc)) %>% 
  # limit time, remove fire signal, remove flier
  filter(!is.na(toc_doc), between(DOY, 100, 320), fire_signal == 0 )

# see what this looks like
TOC_GAM <- gam(toc_doc ~ te(DOY, distance_upstream_km, 
                            bs = c("cc", "tp"),
                            k = c(10, 5)),
               data = mainstem)

gam.check(TOC_GAM)

plot(TOC_GAM, scheme = 2, too.far = 0.1, shade = TRUE)
vis.gam(TOC_GAM, view = c("DOY", "distance_upstream_km"),
        plot.type = "persp", theta = 30, phi = 30,
        ticktype = "detailed")

```

Okay, this seems to work pretty well. We do, however, want to do some thoughtful
TVT here to keep our error estimates honest. First let's look at annual distribution
since we want to test in 2025.

```{r}
mainstem %>% 
summarize(n = n(), 
.by = year)
```

Maybe we actually need to split by S Fork and Mainstem so we don't mess with the
huge contributions of 2024 data or super minimal years like... all others.

```{r}
counts <- mainstem %>% 
  summarise(n = n(),
            .by = c(year, Campaign)) %>% 
  arrange(year)
counts
```

We'll do some kinda random, but not really random splits here:

- Test: 2010, 2017, 2025
- CV 1: 2008, 2013, 2018
- CV 2: 2009, 2014, 2019
- CV 3: 2011, 2015, 2023
- CV 4: 2012, 2016, 2024

```{r}
test_years <- c(2010, 2017, 2025)
cv_1_years <- c(2008, 2013, 2018)
cv_2_years <- c(2009, 2014, 2019)
cv_3_years <- c(2011, 2015, 2023)
cv_4_years <- c(2012, 2016, 2024)
```

GAM diagnostics are interpreted using the gam.check() function. We want to make
sure that:

1) the model converged (second line of the output), the convergnece value is very small,
and ideally model rank is last (26/26 for instance)
2) We are confident we are not finding a local minima in the convergence ("Hessian was positive difinitive")
3) that the k is ~ 1 and the p value > 0.05
4) we want the lowest k (default is 25 or c(5, 5)) while not maxing out the wiggles
which is determined by k' (max number of wiggles) and edf (actual wiggles used)

Some notes on interpretation and tuning:

- If the p-value < 0.05, you need to add more complexity to the model (by increasing k)
- The lower the k means it's less likely to be over fit. 
- As long as the p value > 0.05 and k ~ 1, it's unlikely to be underfit.
- A higher k risks being overfit, and not generalizable. 
- If your edf is very close to k' (like, within 1 unit), you probably need to increase k

My rule of thumb for making sure you don't overfit: 
if the edf value of your model is less than the k' of a lower k, you might be overfit:

edf = 20, k' = 48 (implying the k = c(7,7), or 7x7-1), you'd want to try k = c(6,6),
which would have a k' of 35 (6x6-1), and ideally the edf would still be low with
a smaller parameterization space.


## CV Model 1 

```{r}
valid_1 <- mainstem %>% filter(year %in% cv_1_years)
train_1 <- anti_join(mainstem, valid_1) %>% 
  filter(!(year %in% test_years))

# fit gam
fit_1 <- gam(toc_doc ~ te(DOY, distance_upstream_km, 
                          bs = c("cc", "tp"),
                          k = c(10, 5)), 
             data = train_1, method = "REML") 

# check gam diagnostics
gam.check(fit_1)

# edf is low relative to k', k index is nearly 1, let it slide. the data are 
# noisy, so it makes sense that we're not optimizing the wiggles

# visualize 2d/3d
vis.gam(fit_1,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "contour",
        color = "topo",
        too.far = 0.1)

vis.gam(fit_1,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_1$pred <- predict(fit_1, newdata = valid_1)

# and report performance
mae(valid_1$toc_doc, valid_1$pred)
rmse(valid_1$toc_doc, valid_1$pred)
bias(valid_1$toc_doc, valid_1$pred)
```

## CV Model 2

```{r}
valid_2 <- mainstem %>% filter(year %in% cv_2_years)
train_2 <- anti_join(mainstem, valid_2) %>% 
  filter(!(year %in% test_years))

# fit gam
fit_2 <- gam(toc_doc ~ te(DOY, distance_upstream_km, 
                          bs = c("cc", "tp"),
                          k = c(10, 5)), 
             data = train_2, method = "REML") 

gam.check(fit_2)

# good to go!

# visualize 2d/3d
vis.gam(fit_2,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "contour",
        color = "topo",
        too.far = 0.1)

vis.gam(fit_2,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_2$pred <- predict(fit_2, newdata = valid_2)
ggplot(valid_2, aes(x = toc_doc, y = pred)) +
  geom_point( ) +
  theme_bw()

# and report performance
mae(valid_2$toc_doc, valid_2$pred)
rmse(valid_2$toc_doc, valid_2$pred)
bias(valid_2$toc_doc, valid_2$pred)
```


## CV Model 3

```{r}
valid_3 <- mainstem %>% filter(year %in% cv_3_years)
train_3 <- anti_join(mainstem, valid_3) %>% 
  filter(!(year %in% test_years))

# fit gam
fit_3 <- gam(toc_doc ~ te(DOY, distance_upstream_km, 
                          bs = c("cc", "tp"),
                          k = c(10, 5)), 
             data = train_3, method = "REML") 

# check gam diagnostics
gam.check(fit_3) 

# good to go , but a LOT of noise at low TOC

# visualize 2d/3d
vis.gam(fit_3,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "contour",
        color = "topo",
        too.far = 0.1)

vis.gam(fit_3,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_3$pred <- predict(fit_3, newdata = valid_3)

ggplot(valid_3, aes(x = toc_doc, y = pred)) +
  geom_point( ) +
  theme_bw()

train_3 %>% 
  mutate(pred = predict(fit_3, newdata = train_3)) %>% 
ggplot(., aes(x = toc_doc, y = pred, color = factor(year))) +
  geom_point( ) +
  facet_grid(site_name ~ .) +
  theme_bw() +
  scale_color_viridis_d()
ggplot(train_3, aes(x = DOY, y = toc_doc, color = factor(year))) +
  geom_point() +
  scale_color_viridis_d()

# and report performance
mae(valid_3$toc_doc, valid_3$pred)
rmse(valid_3$toc_doc, valid_3$pred)
bias(valid_3$toc_doc, valid_3$pred)
```


## CV Model 4

```{r}
valid_4 <- mainstem %>% filter(year %in% cv_4_years)
train_4 <- anti_join(mainstem, valid_4) %>% 
  filter(!(year %in% test_years))

# fit gam
fit_4 <- gam(toc_doc ~ te(DOY, distance_upstream_km, 
                          bs = c("cc", "tp"),
                          k = c(10, 5)), 
             data = train_4, method = "REML") 
# check gam diagnostics
gam.check(fit_4)  

# visualize 2d/3d
vis.gam(fit_4,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "contour",
        color = "topo",
        too.far = 0.1)

vis.gam(fit_4,
        view = c("DOY", "distance_upstream_km"),
        plot.type = "persp",
        theta = 30, phi = 30,
        ticktype = "detailed")

# apply to validation
valid_4$pred <- predict(fit_4, newdata = valid_4)

# and report performance
mae(valid_4$toc_doc, valid_4$pred)
rmse(valid_4$toc_doc, valid_4$pred)
bias(valid_4$toc_doc, valid_4$pred)
```


# Apply ensemble to test set

```{r}
test <- mainstem %>% filter(year %in% test_years)

# apply ensemble
test <- test %>%
  mutate(
    pred1 = predict(fit_1, newdata = test),
    pred2 = predict(fit_2, newdata = test),
    pred3 = predict(fit_3, newdata = test),
    pred4 = predict(fit_4, newdata = test),
    pred_avg = (pred1 + pred2 + pred3 + pred4) / 4
  )

# calculate performance metrics
mae_val  <- mae(test$toc_doc, test$pred_avg)
rmse_val <- rmse(test$toc_doc, test$pred_avg)
bias_val <- bias(test$toc_doc, test$pred_avg)

# r2 from linear regression
r2_val <- summary(lm(test$toc_doc ~ test$pred_avg))$r.squared

# format annotation text for plot
metrics_text <- paste0(
  "MAE = ", round(mae_val, 2),
  "\nRMSE = ", round(rmse_val, 2),
  "\nBias = ", round(bias_val, 2),
  "\nR² = ", round(r2_val, 2)
)


# pred - obs scatter plot!
ggplot(test, aes(y = pred_avg, x = toc_doc)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (holdout)",
       subtitle = "TOC ~ DOY + distance upstream") +
  annotate("text", x = 1, y = 10, 
           hjust = 0, vjust = 1, 
           label = metrics_text, size = 4) +
  theme_bw(base_size = 14) +
  scale_color_viridis_c() +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12))

# with DOY
ggplot(test, aes(y = pred_avg, x = toc_doc, color = DOY)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (holdout)",
       subtitle = "TOC ~ DOY + distance upstream") +
  annotate("text", x = 1, y = 10, 
           hjust = 0, vjust = 1, 
           label = metrics_text, size = 4) +
  theme_bw(base_size = 14) +
  scale_color_viridis_c() +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12))

# for kicks, look at performance by site and DOY
ggplot(test, aes(y = pred_avg, x = toc_doc, color = DOY)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Observed TOC (mg/L)",
       y = "Predicted TOC (ensemble, mg/L)",
       title = "GAM Ensemble Predictions vs Observed (holdout)",
       subtitle = "TOC ~ DOY + distance upstream") +
  theme_bw(base_size = 14) +
  scale_color_viridis_c() +
  coord_cartesian(xlim = c(1, 12), ylim = c(1, 12)) +
  facet_wrap(site_name ~ .)

```

There's clearly some variation at start/end of season that isn't being captured 
in the model. Based on our other work, this probably has something to do with 
snowpack/melt patterns.

# Visualize the decision space for interpretability

GAMs are pretty interpretable when they have fewer interacting variables. In this
case, we can just create a synthetic space/time space and apply the model across
it to visualize what happens up the canyon over time.

```{r}
# Make a prediction grid across DOY and distance
newdata <- expand.grid(
  DOY = seq(min(mainstem$DOY), max(mainstem$DOY)),
  distance_upstream_km = seq(min(mainstem$distance_upstream_km),
                             max(mainstem$distance_upstream_km))
)

# Get predictions from your GAMs for the grid
newdata$fit1 <- predict(fit_1, newdata = newdata)
newdata$fit2 <- predict(fit_2, newdata = newdata)
newdata$fit3 <- predict(fit_3, newdata = newdata)
newdata$fit4 <- predict(fit_4, newdata = newdata)

# create ensemble prediction
newdata <- newdata %>% 
  mutate(ave_fit = (fit1 + fit2 + fit3 + fit4) / 4)

# heatmap + contour plot
ggplot(newdata, aes(x = DOY, y = distance_upstream_km, fill = ave_fit)) +
  geom_tile() +
  geom_contour(aes(z = ave_fit), color = "white", alpha = 0.7) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    fill = "Predicted TOC",
    x = "Day of Year",
    y = "Distance upstream from Canyon Mouth (km)"
  ) +
  theme_minimal(base_size = 14) 

```

And let's save these models as the baseline:

```{r}

write_rds(fit_1, here("data/derived/models/TOC_distributed/TOC_GAM_fit1_v2026-01-28.rds"))
write_rds(fit_2, here("data/derived/models/TOC_distributed/TOC_GAM_fit2_v2026-01-28.rds"))
write_rds(fit_3, here("data/derived/models/TOC_distributed/TOC_GAM_fit3_v2026-01-28.rds"))
write_rds(fit_4, here("data/derived/models/TOC_distributed/TOC_GAM_fit4_v2026-01-28.rds"))

```






