---
title: "GAM application for intake forecast"
author: "ROSSyndicate"
date: 2026-02-19
output: html_document
---

# Purpose

This script is meant to apply the intake TOC GAM model to assess its 
performance with forecasted Q from HEFS.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(here)
library(cdssr)
library(mgcv)
library(Metrics)
library(verification)
library(arrow)
library(snotelr)
library(tidyverse)
library(cowplot)

```

# Data

Here, we'll grab the data for 2023-2025. Not all of these data are in the test
set, but it will give us an idea of the added uncertainty for a 30-day forecast.
We're using 2023-2025 here because those are the HEFS forecasts easily available 
to us.

```{r}
# intake TOC
fc_intake <- read_csv(here("data/collated/FC_intake/fc_intake_TOC_record_2010-2025.csv")) %>% 
  filter(between(year(date), 2023, 2025))

```

And grab Snotel data

```{r}
# grab snotel sites in the HUC
snotel_sites <- snotel_info() %>%
  filter(grepl("10190007", description))

# Download daily SNOTEL data
snotel_data <- snotel_download(site_id = snotel_sites$site_id,
                               internal = TRUE,
                               metric = TRUE) %>% 
  mutate(date = ymd(date),
         doy = yday(date),
         year = year(date)) %>% 
  filter(between(year, 2023, 2025))

# prep for PCA
snotel_matrix <- snotel_data %>%
  select(date, site_id, snow_water_equivalent) %>% 
  pivot_wider(names_from = site_id, 
              names_prefix = "swe_",
              values_from = snow_water_equivalent) %>% 
  .[complete.cases(.), ]

# load PCA
swe_pca <- read_rds(here("data/derived/models/TOC_intake/SWE_PCA_v2026-02-18.rds"))

# apply to data, select only PC1
snotel <- snotel_matrix %>% 
  mutate(swe_pca = predict(swe_pca, .)[,"PC1"]) %>% 
  select(date, swe_pca)

# calculate 14-day melt
snotel <- snotel %>% 
  arrange(date) %>% 
  mutate(date_m14 = date - days(14)) %>%
  # join for differences
  left_join(snotel %>%
              select(date, swe_pca) %>%
              rename(date_m14 = date,
                     swe_pca_lag14 = swe_pca),
            by = "date_m14") %>%
  # compute melt rate
  mutate(melt_rate_14d = (swe_pca_lag14 - swe_pca) / 14) %>% 
  select(-c(date_m14, swe_pca_lag14))
```

Grab forecasted data (for our operational statistics) - these data were gathered in 
`modeling/external_forecasts/HEFS_canyon_mouth_23-25.Rmd`

```{r}
forecasted_q <- read_parquet(here("data/raw/external_datasets/flow_forecast_HEFS/HEFS_2023_2025_forecast.parquet")) %>% 
  # each of the forecasts start at noon, and we want the daily average, so we'll 
  # subtract 12 to get an average of the "next day"
  mutate(date_24h = date(date_time - hours(12))) %>% 
  summarize(across(c(ens_1990:ens_2018, ens_2019),
                   ~ mean(.x)),
            .by = c(forecast_date, date_24h))

# pivot to viretical to make the probabilistic forecast
forecasted_q_vert <- forecasted_q %>% 
  pivot_longer(cols = c(ens_1990:ens_2019),
               names_to = "ensemble_member",
               values_to = "canyon_mouth_cfs") %>% 
  rename(date = forecast_date)
```

Create datasets to forecast on

```{r}
# join with location data, and reduce columns
toc_q <- left_join(fc_intake, forecasted_q_vert,
                   relationship = "one-to-many") %>%
  filter(!is.na(canyon_mouth_cfs)) %>% 
  select(date, date_24h, year, canyon_mouth_cfs)  %>% 
  mutate(doy = yday(date_24h))

toc_q_swe <- left_join(toc_q, snotel) %>%
  filter(!is.na(swe_pca)) %>% 
  select(date, date_24h, year, canyon_mouth_cfs, swe_pca, melt_rate_14d)  %>% 
  mutate(doy = yday(date_24h))

```

# Load models

```{r}
q_doy_int <- read_rds(here("data/derived/models/TOC_intake/TOC_intake_GAM_Q_doy_inter_v2026-02-18.rds"))
q_doy_swe <- read_rds(here("data/derived/models/TOC_intake/TOC_intake_GAM_Q_swe-melt14_v2026-02-18.rds"))
```

And apply them to the dataset

```{r}
toc_q_swe <- toc_q_swe %>% 
  mutate(pred_doy_int = predict(q_doy_int, .),
         pred_doy_swe = predict(q_doy_swe, .))
```

# Additive and Interaction GAM Q/DOY

And visualize performance at a few time horizons, remember `date` is the day forecast 
is made and `date_24` is the forecasted day.

```{r, fig.dim=c(9, 6)}
d1 <- toc_q_swe %>% 
  filter(date == date_24h) %>% 
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_int)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Predicted TOC (mg/L)\nfrom HEFS ensemble, 1 day horizon",
       title = "Intake TOC Model Ensemble")
d2 <- toc_q_swe %>% 
  mutate(date = date + days(1)) %>% 
  filter(date == date_24h) %>%
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_int)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Predicted TOC (mg/L)\nfrom HEFS ensemble, 2 day horizon",
       title = "Intake TOC Model Ensemble")
d3 <- toc_q_swe %>% 
  mutate(date = date + days(2)) %>% 
  filter(date == date_24h) %>%
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_int)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Predicted TOC (mg/L)\nfrom HEFS ensemble, 3 day horizon",
       title = "Intake TOC Model Ensemble")
d4 <- toc_q_swe %>% 
  mutate(date = date + days(3)) %>% 
  filter(date == date_24h) %>%
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_int)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Predicted TOC (mg/L)\nfrom HEFS ensemble, 4 day horizon",
       title = "Intake TOC Model Ensemble")
d5 <- toc_q_swe %>% 
  mutate(date = date + days(4)) %>% 
  filter(date == date_24h) %>%
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_int)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Predicted TOC (mg/L)\nfrom HEFS ensemble, 5 day horizon",
       title = "Intake TOC Model Ensemble")
d10 <- toc_q_swe %>% 
  mutate(date = date + days(9)) %>% 
  filter(date == date_24h) %>%
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_int)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Predicted TOC (mg/L)\nfrom HEFS ensemble, 10 day horizon",
       title = "Intake TOC Model Ensemble")

plot_grid(d1, d2, d3, d4, d5, d10, nrow = 2)
```


And get some performance stats, including Continuous Ranked Probability Score 
(CRPS) which is better for probabilistic forecasts, but considered "comparable"
to MAE in determinisitic forecasts.

```{r, fig.dim=c(6, 3), message = FALSE}

for_crps <- map(seq(1:30),
    ~ toc_q_swe %>% 
      mutate(date = date + days(.x-1)) %>% 
      filter(date == date_24h) %>%
      inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
      mutate(horizon = .x) %>% 
      summarize(mean_toc = mean(pred_doy_int), 
                sd_toc = sd(pred_doy_int),
                n = n(),
                .by = c(horizon, date, mean_TOC))) %>% 
  bind_rows() %>% 
  filter(n > 10) 

calc_crps <- for_crps %>% 
  rowwise() %>% 
  mutate(crps_daily = crps(mean_TOC, c(mean_toc, sd_toc))$crps) %>% 
  ungroup() %>% 
  select(horizon, crps_daily, date)

ggplot(calc_crps, aes(x = horizon, y = crps_daily)) +
  geom_boxplot(aes(group = horizon)) +
  theme_bw() +
  labs(y = "continuous ranked probability score (mg/L)\ncomparable to MAE",
       x = "forecast horizon",
       title = "CRPS for intake TOC forecast",
       subtitle = "GAM HEFS") +
  scale_x_continuous(n.breaks = 30, minor_breaks = seq(1:30))

```

Look at performance in June only

```{r, fig.dim=c(6, 3), message = FALSE}
for_crps_jun <- map(seq(1:30),
    ~ toc_q_swe %>% 
      filter(month(date) == 6) %>% 
      mutate(date = date + days(.x-1)) %>% 
      filter(date == date_24h) %>%
      inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
      mutate(horizon = .x) %>% 
      summarize(mean_toc = mean(pred_doy_int), 
                sd_toc = sd(pred_doy_int),
                n = n(),
                .by = c(horizon, date, mean_TOC))) %>% 
  bind_rows() 

calc_crps_jun <- for_crps_jun %>% 
  rowwise() %>% 
  mutate(crps_daily = crps(mean_TOC, c(mean_toc, sd_toc))$crps) %>% 
  ungroup() %>% 
  select(horizon, crps_daily, date)

ggplot(calc_crps_jun, aes(x = horizon, y = crps_daily)) +
  geom_boxplot(aes(group = horizon)) +
  theme_bw() +
  labs(y = "continuous ranked probability score (mg/L)\ncomparable to MAE",
       x = "forecast horizon",
       title = "CRPS for intake TOC forecast, June performance only",
       subtitle = "GAM HEFS") +
  scale_x_continuous(n.breaks = 30, minor_breaks = seq(1:30))
```

# Additive Interaction and SWE melt



```{r, fig.dim=c(9, 6)}
d1_melt <- toc_q_swe %>% 
  filter(date == date_24h) %>% 
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_swe)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Average Predicted TOC (mg/L)\nfrom HEFS ensemble + melt, 1 day horizon",
       title = "Intake TOC Model Ensemble")
d2_melt <- toc_q_swe %>% 
  mutate(date = date + days(1)) %>% 
  filter(date == date_24h) %>%
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_swe)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Average Predicted TOC (mg/L)\nfrom HEFS ensemble + melt, 2 day horizon",
       title = "Intake TOC Model Ensemble")
d3_melt <- toc_q_swe %>% 
  mutate(date = date + days(2)) %>% 
  filter(date == date_24h) %>%
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_swe)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Average Predicted TOC (mg/L)\nfrom HEFS ensemble + melt, 3 day horizon",
       title = "Intake TOC Model Ensemble")
d4_melt <- toc_q_swe %>% 
  mutate(date = date + days(3)) %>% 
  filter(date == date_24h) %>%
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_swe)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Average Predicted TOC (mg/L)\nfrom HEFS ensemble + melt, 4 day horizon",
       title = "Intake TOC Model Ensemble")
d5_melt <- toc_q_swe %>% 
  mutate(date = date + days(4)) %>% 
  filter(date == date_24h) %>%
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_swe)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Average Predicted TOC (mg/L)\nfrom HEFS ensemble + melt, 5 day horizon",
       title = "Intake TOC Model Ensemble")
d10_melt <- toc_q_swe %>% 
  mutate(date = date + days(9)) %>% 
  filter(date == date_24h) %>%
  inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
  ggplot(., aes(x = mean_TOC, y = pred_doy_swe)) + 
  geom_boxplot(aes(group = date), position = position_dodge(), width = 0.2, alpha = 0.5) +
  theme_bw() +
  geom_abline(intercept = 0, slope = 1, lty = 2, color = "red") +
  coord_cartesian(xlim = c(0, 12.5), ylim = c(0, 12.5)) +
  labs(x = "Measured TOC (mg/L)",
       y = "Average Predicted TOC (mg/L)\nfrom HEFS ensemble + melt, 10 day horizon",
       title = "Intake TOC Model Ensemble")
plot_grid(d1_melt, d2_melt, d3_melt, d4_melt, d5_melt, d10_melt, nrow = 2)
```


And get some performance stats, including Continuous Ranked Probability Score 
(CRPS) which is better for probabilistic forecasts, but considered "comparable"
to MAE in determinisitic forecasts.

```{r, fig.dim=c(6, 3), message = FALSE}
for_crps_melt <- map(seq(1:30),
    ~ toc_q_swe %>% 
      mutate(date = date + days(.x-1)) %>% 
      filter(date == date_24h) %>%
      inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
      mutate(horizon = .x) %>% 
      summarize(mean_toc = mean(pred_doy_swe), 
                sd_toc = sd(pred_doy_swe),
                n = n(),
                .by = c(horizon, date, mean_TOC))) %>% 
  bind_rows() %>% 
  filter(n > 10) 

calc_crps_melt <- for_crps_melt %>% 
  rowwise() %>% 
  mutate(crps_daily = crps(mean_TOC, c(mean_toc, sd_toc))$crps) %>% 
  ungroup() %>% 
  select(horizon, crps_daily, date)

ggplot(calc_crps_melt, aes(x = horizon, y = crps_daily)) +
  geom_boxplot(aes(group = horizon)) +
  theme_bw() +
  labs(y = "continuous ranked probability score (mg/L)\ncomparable to MAE",
       x = "forecast horizon",
       title = "CRPS for intake TOC forecast",
       subtitle = "GAM HEFS + SWE melt") +
  scale_x_continuous(n.breaks = 30, minor_breaks = seq(1:30))

```

Just look at CRPS scores for June only

```{r, fig.dim=c(6, 3), message = FALSE}
for_crps_melt_jun <- map(seq(1:30),
    ~ toc_q_swe %>% 
      filter(month(date) == 6) %>% 
      mutate(date = date + days(.x-1)) %>% 
      filter(date == date_24h) %>%
      inner_join(., fc_intake %>% select(date, mean_TOC)) %>% 
      mutate(horizon = .x) %>% 
      summarize(mean_toc = mean(pred_doy_swe), 
                sd_toc = sd(pred_doy_swe),
                n = n(),
                .by = c(horizon, date, mean_TOC))) %>% 
  bind_rows() 

calc_crps_melt_jun <- for_crps_melt_jun %>% 
  rowwise() %>% 
  mutate(crps_daily = crps(mean_TOC, c(mean_toc, sd_toc))$crps) %>% 
  ungroup() %>% 
  select(horizon, crps_daily, date)

ggplot(calc_crps_melt_jun, aes(x = horizon, y = crps_daily)) +
  geom_boxplot(aes(group = horizon)) +
  theme_bw() +
  labs(y = "continuous ranked probability score (mg/L)\ncomparable to MAE",
       x = "forecast horizon",
       title = "CRPS for intake TOC forecast: June only",
       subtitle = "GAM HEFS + SWE melt") +
  scale_x_continuous(n.breaks = 30, minor_breaks = seq(1:30))
```

