---
title: "Intake daily TOC baseline"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(here)
library(cdssr)
library(snotelr)
library(mgcv)
library(Metrics)
library(tidyverse)

```

# Purpose

This document explores some options for modeling daily TOC data from the City of 
Fort Collins' intake.

# Data

```{r}
# intake TOC
fc_intake <- read_csv(here("data/collated/FC_intake/fc_intake_TOC_record_2010-2025.csv"))

# read in flow data at canyon mouth
flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = "2010-01-01",
                       end_date = "2025-12-31",
                       timescale = "day") %>%
  select(datetime, canyon_mouth_cfs = value) %>% 
  mutate(date = date(datetime), 
         year = year(date),
         doy = yday(date)) 

# grab sites in the HUC
snotel_sites <- snotel_info() %>%
  filter(grepl("10190007", description))

# Download daily SNOTEL data
snotel_data <- snotel_download(site_id = snotel_sites$site_id,
                               internal = TRUE,
                               metric = TRUE) %>% 
  mutate(date = ymd(date),
         doy = yday(date),
         year = year(date))

# prep for PCA
snotel_matrix <- snotel_data %>%
  select(date, site_id, snow_water_equivalent) %>% 
  pivot_wider(names_from = site_id, 
              names_prefix = "swe_",
              values_from = snow_water_equivalent) %>% 
  .[complete.cases(.), ]

# run PCA
pca_snotel <- snotel_matrix %>% 
  select(starts_with("swe_")) %>% 
  prcomp(., center = TRUE, scale. = TRUE) 

```

And let's look at the pca

```{r}
summary(pca_snotel)

pca_snotel$rotation

autoplot(pca_snotel, loadings = TRUE, loadings.label = TRUE,variance_percentage = T)

```



```{r}
# make the dataframe
snotel <- tibble(swe_pca = pca_snotel$x[,1], date = snotel_matrix$date) 

# calculate melt from PCA statistic
snotel <- snotel %>% 
  arrange(date) %>% 
  mutate(date_m14 = date - days(14), 
         date_m7 = date - days(7),
         date_m3 = date - days(3)) %>%
  # join for differences
  left_join(snotel %>%
              select(date, swe_pca) %>%
              rename(date_m14 = date,
                     swe_pca_lag14 = swe_pca),
            by = "date_m14") %>%
  left_join(snotel %>%
              select(date, swe_pca) %>%
              rename(date_m7 = date,
                     swe_pca_lag7 = swe_pca),
            by = "date_m7") %>%
  left_join(snotel %>% 
              select(date, swe_pca) %>% 
              rename(date_m3 = date,
                     swe_pca_lag3 = swe_pca)) %>% 
  # compute melt rate
  mutate(melt_rate_14d = (swe_pca_lag14 - swe_pca) / 14,
    melt_rate_7d = (swe_pca_lag7 - swe_pca) / 7,
    melt_rate_3d = (swe_pca_lag3 - swe_pca) / 3) %>% 
  select(-c(date_m14, swe_pca_lag14,
            date_m7, swe_pca_lag7, 
            date_m3, swe_pca_lag3))

intake_flow_swe <- left_join(fc_intake, flow_data) %>% 
  left_join(snotel)
```

# Visualize TOC

```{r}
ggplot(intake_flow_swe, aes(x = date, y = mean_TOC)) +
  geom_point() +
  theme_bw() +
  labs(x = "day of year",
       y = "intake TOC (mg/L)")

intake_flow_swe %>% 
  summarize(max_TOC = max(mean_TOC),
            mean_TOC = mean(mean_TOC),
            n_TOC = n(), 
            .by = year) %>% 
  arrange(year)
```

# Pull testing data

Given the 16 years of data, we're going to be cautious from the top and pull out
a test set of 2 years of data.

```{r}
test_years <- c(2018, 2024)

test <- intake_flow_swe %>% 
  filter(year %in% test_years)

training <- anti_join(intake_flow_swe, test)
```

# Play with training data

Just using day of year.

```{r}
gam_doy <- gam(mean_TOC ~ s(doy, bs = "cc", k = 20),
               data = training, 
               method = "REML")

gam.check(gam_doy)
```

If we increase k substantially (to 20 here), we start to see some meaningful 
model diagnositcs (k' < edf, p value increasing indicating that the model is 
capturing more variation), but there's a lot of variation at the high end.

Using doy + canyon mouth as additive.

```{r}
gam_doy_q <- gam(mean_TOC ~ s(doy, bs = "cc") + s(canyon_mouth_cfs),
                 data = training,
                 method = "REML")

gam.check(gam_doy_q)
```

This looks really good! Model diagnostics are good too. 

Let's add some complexity using some interaction for doy/canyon mouth

```{r}
gam_doy_q_inter <- gam(
  mean_TOC ~ 
    s(doy, bs = "cc") +
    s(canyon_mouth_cfs) +
    ti(doy, canyon_mouth_cfs, bs = c("cc", "tp")),
  data   = training,
  method = "REML"
)

gam.check(gam_doy_q_inter)
```

Even better still!

Now, let's add some snowpack info.

```{r}
gam_doy_q_inter_swe <- gam(
  mean_TOC ~ 
    s(doy, bs = "cc") +
    s(canyon_mouth_cfs) +
    ti(doy, canyon_mouth_cfs, bs = c("cc", "tp")) +
    s(swe_pca, bs = "ts"),
  data   = training,
  method = "REML"
)

gam.check(gam_doy_q_inter_swe)
```

Okay, we might be adding a bit too much complexity here, the interaction term
decreased in significance and the swe values don't look great diagnostically.

For kicks, let's drop the interaction term.

```{r}
gam_doy_q_swe <- gam(
  mean_TOC ~ 
    s(doy, bs = "cc") +
    s(canyon_mouth_cfs) +
    s(swe_pca, bs = "ts"),
  data   = training,
  method = "REML"
)

gam.check(gam_doy_q_swe)
```

A little better in diagnostics, let's do an interaction term for q and swe.

```{r}
gam_doy_q_swe_inter <- gam(
  mean_TOC ~ 
    s(doy, bs = "cc") +
    s(canyon_mouth_cfs) +
    s(swe_pca, bs = "ts") +
    ti(canyon_mouth_cfs, swe_pca, bs = c("ts", "ts")),
  data   = training,
  method = "REML"
)

gam.check(gam_doy_q_swe_inter)
```

Interesting, that interaction seems to really tighten up some of the upper-end
values. The diagnostics aren't great for swe alone, but the edf is not maxed (it's
about a point lower than the k') so we could argue that it's okay to leave in the
swe pca data.

Let's try using melt rate, with interaction term for melt rate and canyon mouth

```{r}
gam_doy_q_swe_melt <- gam(
  mean_TOC ~ 
    s(doy, bs = "cc") +
    s(canyon_mouth_cfs, bs = "tp") +
    s(melt_rate_7d, bs = "tp") +
    ti(melt_rate_7d, canyon_mouth_cfs, bs = "tp"),
  data   = training,
  method = "REML"
)

gam.check(gam_doy_q_swe_melt)
```

Same but 14 day melt rate

```{r}
gam_doy_q_swe_melt_14 <- gam(
  mean_TOC ~ 
    s(doy, bs = "cc") +
    s(canyon_mouth_cfs, bs = "tp") +
    s(melt_rate_14d, bs = "tp") +
    ti(melt_rate_14d, canyon_mouth_cfs, bs = "tp"),
  data   = training,
  method = "REML"
)

gam.check(gam_doy_q_swe_melt_14)
```

Interesting. That 14 day melt rate seems to help tighten up the upper end a little.


# Apply to test set

```{r}

test$pred_TOC_swe_inter <- predict(gam_doy_q_swe_inter, test)
test$pred_TOC_melt <- predict(gam_doy_q_swe_melt_14, test)
test$pred_TOC_simple <- predict(gam_doy_q_inter, test)

ggplot(test, aes(x = mean_TOC, y = pred_TOC_swe_inter)) +
  geom_point() +
  theme_bw() +
  geom_abline(slope = 1, intercept = 0) +
  coord_cartesian(xlim = c(1,12.5), ylim = c(1, 12.5)) +
  labs(x = "measured TOC @ intake (mg/L)",
       y = "predicted TOC @ intake (mg/L)")

rmse(test$mean_TOC, test$pred_TOC_swe_inter)
mae(test$mean_TOC, test$pred_TOC_swe_inter)
bias(test$mean_TOC, test$pred_TOC_swe_inter)

ggplot(test, aes(x = mean_TOC, y = pred_TOC_melt)) +
  geom_point() +
  theme_bw() +
  geom_abline(slope = 1, intercept = 0) +
  coord_cartesian(xlim = c(1,12.5), ylim = c(1, 12.5)) +
  labs(x = "measured TOC @ intake (mg/L)",
       y = "predicted TOC @ intake (mg/L)")

rmse(test$mean_TOC, test$pred_TOC_melt)
mae(test$mean_TOC, test$pred_TOC_melt)
bias(test$mean_TOC, test$pred_TOC_melt)

ggplot(test, aes(x = mean_TOC, y = pred_TOC_simple)) +
  geom_point() +
  theme_bw() +
  geom_abline(slope = 1, intercept = 0) +
  coord_cartesian(xlim = c(1,12.5), ylim = c(1, 12.5)) +
  labs(x = "measured TOC @ intake (mg/L)",
       y = "predicted TOC @ intake (mg/L)")

rmse(test$mean_TOC, test$pred_TOC_simple)
mae(test$mean_TOC, test$pred_TOC_simple)
bias(test$mean_TOC, test$pred_TOC_simple)

```

So definitely some value in including melt statistics, but maybe not swe as 
integrated in the example above. Given this, we'll export both the additive
interaction GAM (simple model) and the 14-day snow melt GAM for operational 
testing.

# Export models

Including the pca model for the snotel data.

```{r}
write_rds(gam_doy_q_swe_melt_14, here("data/derived/models/TOC_intake/TOC_intake_GAM_Q_swe-melt14_v2026-02-18.rds"))
write_rds(gam_doy_q_inter, here("data/derived/models/TOC_intake/TOC_intake_GAM_Q_doy_inter_v2026-02-18.rds"))
write_rds(pca_snotel, here("data/derived/models/TOC_intake/SWE_PCA_v2026-02-18.rds"))
```

