---
title: "Testing NOAA HEFS discharge forecast retrieval"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

# Background

Hydrologic Ensemble Forecast System (HEFS) is a system developed by the National Weather Service (NWS) to
provide ensemble streamflow predictions. HEFS has been implemented at NWS
River Forecast Centers (RFCs) for over 3,000 locations throughout the United
States. HEFS integrates meteorological forecasts with hydrologic modeling to
produce probabilistic streamflow forecasts and other hydrological variables.
The system assesses the potential impacts of high-flow events and assesses
flood risk. Most importantly, HEFS is a valuable tool for decision-makers
and emergency managers, providing them with the necessary information to
make informed decisions about water-related activities and potential flood
hazards.


Read more about HEFS and the associated API
here: https://water.noaa.gov/about/api
and here: https://github.com/NOAA-OWP/data-service-notebooks/blob/master/HEFS/0_executive_overview.ipynb
Short blurb about the HEFS API from that link:
The experimental NWPS HEFS Forecast API can be used to retrieve real-time
and historical HEFS forecast data for NWPS locations. The forecast data
includes complete ensembles and the set of ensemble time-step quantiles. The
HEFS is a service developed by the National Weather Service (NWS) to provide
ensemble streamflow predictions. HEFS has been implemented at NWS River
Forecast Centers (RFCs) for over 3,000 locations throughout the United
States. HEFS integrates meteorological forecasts with hydrologic modeling to
produce probabilistic streamflow forecasts and other hydrological variables.
The service assesses the potential impacts of high-flow events and assesses
flood risk. The service currently serves 10 days of data. Guidance from
individual RFCs may not be available at all times.

```{r}
# Import libraries
library(httr)
library(jsonlite)
library(sf)
library(leafem)
library(mapview)
library(nhdplusTools)
library(tidyverse)

# Import functions
source(here("src/hefs/get_hefs_gauges.R"))
source(here("src/hefs/get_hefs_ensembles.R"))
source(here("src/hefs/get_hefs_observed.R"))
source(here("src/hefs/get_hefs_quantiles.R"))
source(here("src/hefs/get_hefs_ensembles_cbrfc.R"))
```

# Visualize Availabe Sites

Using the CLP HUC8 watershed, let's go through the process to get the Cache La Poudre
site at the canyon mouth. 

```{r}
# Test workflow
huc8 <- get_huc(id="10190007", type = "huc08") # poudre

# Download and plot gauge sites in AOI
gauge_sites <- get_hefs_gauges(huc8)

# plot in space
mapview(huc8, 
        layer.name = "CLP HUC8",
        col.regions = "hotpink",
        alpha.regions = 0.2) +
  mapview(gauge_sites %>% select(lid, geometry),
          layer.name = "HEFS sites",
          legend = NULL)
```

# Current data acqusition and visualization

```{r}
# From that map, the FTDC2 is the gauge we care most about
ID <- "FTDC2"

# Pull in recent observed and current quantile forecast data
observed <- get_hefs_observed(ID)
forecast <- get_hefs_quantiles(ID)

current_forecast <- bind_rows(observed, forecast)

# Plot
ggplot(current_forecast) +
  geom_line(aes(x = date_time, y = discharge_cfs, color = group)) +
  geom_ribbon(aes(x = date_time, ymin = min_value, ymax = max_value, fill = group), alpha = 0.3) +
  geom_line(aes(x = date_time, y = Q0.5, color = group)) +
  theme_bw() +
  labs(y = "Discharge (cfs)", x = NULL) +
  scale_color_manual("", values = c("tomato","black")) +
  scale_fill_manual("", values = c("tomato","black")) +
  scale_x_datetime(minor_breaks = "1 day")

```

## Attempt to get hindcast data

```{r}
# Get hindcast data and compare with observed
hindcast_date = Sys.Date() - days(10)
hindcast_forecast <- get_hefs_ensemble(ID, date = hindcast_date) %>%
  summarize(median_value = median(value),
            max_value = max(value),
            min_value = min(value),
            .by = c(date_time, forecast_date, group))
ten_day_hindcast <- bind_rows(observed, hindcast_forecast) %>%
  filter(between(date_time, ymd(hindcast_date), ymd(hindcast_date) + days(10)))

ggplot(ten_day_hindcast) +
  geom_line(aes(x = date_time, y = discharge_cfs, color = group)) +
  geom_ribbon(aes(x = date_time, ymin = min_value, ymax = max_value, fill = group), alpha = 0.3) +
  geom_line(aes(x = date_time, y = median_value, color = group)) +
  theme_bw() +
  labs(y = "Discharge (cfs)", x = NULL) +
  scale_color_manual("", values = c("tomato","black")) +
  scale_fill_manual("", values = c("tomato","black")) +
  scale_x_datetime(minor_breaks = "1 day")
```


# Get Colorado Basin River Forecast Center hindcast

The real time functions above only provide data for up to the previous 14 days, 
sometimes less. The following method allows us to get the archived data from the 
Colorado Basin River Forecast Center.

```{r}
rand_date <- get_hefs_ensembles_cbrfc(ID = "FTDC2",
                                      date = "2024-05-11")

ggplot(rand_date) +
  geom_ribbon(aes(x = date_time, ymin = min_ens, ymax = max_ens, fill = "Min-Max"), alpha = 0.4) +
  geom_ribbon(aes(x = date_time, ymin = q_25, ymax = q_75, fill = "Q25-Q75"), alpha = 0.4) +
  geom_line(aes(x = date_time, y = q_50, color = "Q50")) +
  theme_bw() +
  labs(y = "Discharge (cfs)", x = NULL, color = "median") +
  guides(fill = guide_legend(reverse = TRUE)) +
  scale_fill_manual("ensemble", 
                    values = c("lightblue", "dodgerblue", "blue", "navy"), 
                    na.translate = FALSE) +
  scale_x_datetime(minor_breaks = "1 day")

```

This method also allows for gathering today's forecast by leaving the date as `NULL`:

```{r}

todays_forecast <- get_hefs_ensembles_cbrfc(ID = "FTDC2",
                                            date = NULL) %>% 
  mutate(group = "Forecast")


current_forecast_cbrfc <- bind_rows(observed, todays_forecast)

# Plot
ggplot(current_forecast_cbrfc) +
  geom_line(aes(x = date_time, y = discharge_cfs)) +
  geom_ribbon(aes(x = date_time, ymin = min_ens, ymax = max_ens, fill = "Min-Max"), alpha = 0.4) +
  geom_ribbon(aes(x = date_time, ymin = q_25, ymax = q_75, fill = "Q25-Q75"), alpha = 0.4) +
  geom_line(aes(x = date_time, y = q_50, color = "Q50")) +
  theme_bw() +
  labs(y = "Discharge (cfs)", x = NULL, color = "median") +
  guides(fill = guide_legend(reverse = TRUE)) +
  scale_fill_manual("ensemble", 
                    values = c("lightblue", "dodgerblue", "blue", "navy"), 
                    na.translate = FALSE) +
  scale_x_datetime(minor_breaks = "1 day")

```
