---
title: "Accessing the NWM forecasts"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

# Purpose 

This Rmd is meant to outline how to access and extract NWM Guidance forecasts. 
NWM Guidance is different from the NWM Official Forecast, which is only available
for some locations during some time periods (that we don't really understand).

The nwmTools package, which seemed like the most straightforward package to use
here is super outdated and no longer works without some finessing (finessing is 
included here). 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nhdplusTools)
library(sf)
# for now, don't worry about grabbing nwmTools, I've grabbed the necessary
# functions and edited them to work.
# install.packages("remotes")
# remotes::install_github('mikejohnson51/nwmTools')
# library(nwmTools)
library(tidyverse)
library(glue)
library(rvest)
library(ncdf4)
library(httr)
```

## nwmTools functions

These functions came from the nwmTools package, but most have been edited

```{r}
validate <- function(complete, field, value){
  # check for field in source file
  if(field %in% names(complete) & !is.null(value)){
    # get all the options from the file for that field
    opts = unique(complete[[field]])
    # check to see that the value is in the list of field values
    if(any(grepl(value, opts))){
      return(filter(complete, grepl(!!value, get(field))))
    } else {
      stop(glue("{value} not a valid {field}. Choose from: {paste(opts, collapse = ', ')}"))
    }
  } else {
    return(complete)
  }
}

nwm_filter <- function(source, version = NULL, config = NULL, ensemble = NULL, 
                      output= NULL, domain = NULL, date = NULL){
  
  startDate <- endDate <- NULL
  
  # Source ------------------------------------------------------------------
  
  meta = validate(nwmTools::nwm_data, "source", source) %>% 
    validate("domain", domain) %>% 
    validate("version", version) %>% 
    validate("output", output) %>% 
    validate("config", config) %>% 
    validate("ensemble", ensemble) 
  
  
  # Date ------------------------------------------------------------------ 
  
  if(!is.null(date)){
    
    meta = meta %>% 
      mutate(startDate = ifelse(startDate == "..", 
                                "1900-01-01",
                                startDate),
             endDate = ifelse(endDate == "..",  as.character(Sys.Date()), endDate))
    
    tmp = filter(meta, ymd(date, tz = "UTC") >= 
                   as.POSIXct(startDate, tz = "UTC"),
                 ymd(date, tz = "UTC") <= 
                   as.POSIXct(endDate, tz = "UTC"))
    
    if(nrow(tmp) == 0){ 
      stop("date not within allowed range: ", meta$startDate, "--", meta$endDate) 
    } else {
      meta = tmp 
    }
  }
  
  if(nrow(meta) > 1){warning('More then one viable source found...', call. = FALSE)}
  
  meta
}

#' Download Remote Files
#' @param fileList fileList object
#' @param outdir directory to write files
#' @return data.frame
#' @export

download_files = function(fileList, outdir = "."){
  
  fileList$outfiles = paste0(outdir,"/", basename(fileList$urls))
  
  for(i in 1:nrow(fileList)){
    if(!file.exists(fileList$outfiles[i])){
      httr::GET(fileList$urls[i], 
                write_disk(fileList$outfiles[i]),
                httr::progress())
    }
  }

  return(fileList)

}

#' Get NOMADs File List
#' @inheritParams get_gcp_urls
#' @param version server version (prod or para)
#' @return data.frame
#' @export

get_nomads_urls = function(config = "medium_range",
                           domain = "conus",
                           date = NULL, 
                           hour = NULL,
                           minute = "00",
                           num, 
                           ensemble = NULL, 
                           output = "channel_rt",
                           version = "prod",
                           outdir = NULL) {
  
  if (config == "medium_range_blend") version <- "v3.0" 
  
  meta = nwm_filter(source = "nomads", 
                    version = version, 
                    config = config, 
                    output = output,
                    ensemble = ensemble, 
                    domain = domain)
  
  meta = meta[1,]
  
  tmp     = glue(strsplit(meta$http_pattern, 'nwm\\.')[[1]][1])
  
  dates   = html_attr(html_elements(read_html(tmp), "a"), "href")
  dates = gsub("/", "", gsub("nwm.", "", dates))
  
  if(is.null(date)){ 
    YYYYMMDD = tail(dates,1)
    YYYYMMDD2 = tail(dates, 2)[1]
  } else if(!gsub("-", "", date) %in% dates){
    stop(date, " not avaliable")
  } else {
    YYYYMMDD = gsub("-", "", date)
    YYYYMMDD2 = gsub("-", "", date)
  }
  
  files   = tryCatch({
    date <<- YYYYMMDD
    html_attr(html_elements(read_html(glue('{tmp}/nwm.{YYYYMMDD}/{meta$bucket}/')), "a"), "href")
  }, error = function(e){
    date <<- YYYYMMDD2
    html_attr(html_elements(read_html(glue('{tmp}/nwm.{YYYYMMDD2}/{meta$bucket}/')), "a"), "href")
  })
  
  files  = grep(meta$output, files, value = TRUE)
  files  = grep(".nc$", files, value = TRUE)
  
  if(is.null(hour)){
    current = strsplit(tail(files, 1), "\\.")[[1]][[2]]
    hour = gsub("t", "", gsub("z", "", current))
  } 
  
  files  = grep(paste0("t", hour, "z"), files, value = TRUE)[1:num]
  dates  = ymd_hm(paste0(date, hour, "00")) + hours(0:(num-1))
  
  df = data.frame(dateTime = dates, 
                  urls     =  glue('{tmp}/nwm.{date}/{meta$bucket}/{files}'),
                  output   =  output)
  
  if(!is.null(outdir)){
    df = download_files(df,  outdir = outdir)
  }
  
  return(df)
  
}

```


# Extracting data

This is my version of the get_timeseries() function in the nwmTools package that
doesn't work as written.

```{r}
read_nwm_streamflow <- function(ncfile, 
                                comid, 
                                varname = "streamflow") {
  
  # Open file
  nc <- nc_open(ncfile)
  
  # Extract feature ids
  feature_ids <- ncvar_get(nc, "feature_id")
  
  # Convert COMID to numeric if needed
  comid <- as.numeric(comid)
  
  # Locate match
  idx <- which(feature_ids == comid)
  if (length(idx) == 0) {
    warning(paste("COMID", comid, "not found in feature_id"))
    return(tibble())
  }
  
  # Pull the time dimension from the global attributes
  forecast_datetime <- ncatt_get(nc, 0) %>% 
    .$model_output_valid_time %>% 
    # convert to datetime object
    ymd_hms(., tz = "UTC") %>% 
    with_tz(., tz = "Etc/GMT+7")

  # Pull the valid time dimension from the global attributes
  valid_datetime <- ncatt_get(nc, 0) %>% 
    .$model_initialization_time %>% 
    # convert to datetime object
    ymd_hms(., tz = "UTC")%>% 
    with_tz(., tz = "Etc/GMT+7")
  
  # Extract value matrix for the given varname
  value_mat <- ncvar_get(nc, varname)
  if (nrow(value_mat) < idx) {
    stop("Index outside streamflow matrix bounds")
  }
  
  value <- value_mat[idx]
  
  nc_close(nc)
  
  # Return tidy tibble
  tibble(
    valid_datetime = valid_datetime,
    forecast_datetime = forecast_datetime,
    value = value,
    variable = varname, 
    comid = comid
  )
  
}

```

# Example run

Here, we'll grab forecasts for the Cache La Poudre canyon mouth.

Short-range forecast is 18h in advance
Mid-range forecast is 10d in advance in 1h increments

Note, this only works for the previous ~ 24 h of forecasts.

```{r}
# 12948 is big thompson with forecast info: https://water.noaa.gov/gauges/ESSC2
# 2900705 is `Reach ID` for canyon mouth https://water.noaa.gov/gauges/FTDC2
comid <- "2900705" 

urls <- get_nomads_urls(config = "medium_range_blend", #short_range, medium_range, medium_range_blend
                        date = "20251207", # date of forecast
                        #ensemble = NA, # if medium_range, must supply ensemble number
                        output = "channel_rt", 
                        outdir = tempdir(),
                        num = 24)

files <- download_files(urls, outdir = tempdir())

```

And then we'll extract the data from these for our comid of choice.
```{r}
data <- map(files$outfiles, 
    ~ read_nwm_streamflow(ncfile = .x, 
                          comid = comid, 
                          varname = "streamflow")) %>% 
  bind_rows()

```

Data from here is reported in cms and the data we are comparing to is in cfs, 
let's do a quick unit conversion.

```{r}
data <- data %>% 
  mutate(q_cfs = value * 35.3147) 

ggplot(data, aes(x = forecast_datetime, y = q_cfs)) +
  geom_line()
```

Yay! That matches the NWM Guidance page.
