---
title: "extract_MODIS_snow_data"
output: html_document
---

The purpose of this markdown is to grab MODIS snow cover data using the rgee
package and summarize it for a catchment (defined by locations of monitored 
discharge). Catchments will be defined using nhdplusTools.

MODIS Snow Cover: https://doi.org/10.5067/MODIS/MOD10A1.061

```{r setup}
source("Scratch/pySetup.R")
# A warning message here is okay!

library(rgee)
library(sf)
library(nhdplusTools)
library(geojsonio)
library(data.table)
library(mapview)
library(dataRetrieval)
library(signal)
library(ggthemes)
library(tidyverse)

# Authenticate your GEE account - only need to do once
# ee_Authenticate(user = "therossyndicate@gmail.com")

# strange workaround here to get rgee to initialize
# see notes here: https://github.com/r-spatial/rgee/issues/355
ee$Initialize(project = "poudre-dss")

```


# define functions

Now, define some helpful functions. Note, these can be added to src/ later
if we end up using this function.

```{r define_functions}

#' Returns watershed sf object for a given USGS NWIS site_no
#' @param site_no USGS site number
#' @return gage_basin sf object of watershed associated with USGS site

nldi_finder <- function(site_no) {
  
  #get those gages' watersheds using get_nldi_basin in the {nhdplusTools}.
  nldi_nwis <- list(featureSource = "nwissite",
                    featureID = paste0("USGS-", site_no))
  
  gage_basin <- tryCatch({invisible(
    suppressMessages(
      get_nldi_basin(nldi_feature = nldi_nwis) %>%
        mutate(site_no = site_no)))
  }, 
  error = function(e) {
    NULL
  })
  
  return(gage_basin)
}



# Get nwis discharge sites within an AOI - only returns sites
# with recent data within the last week

get_recent_nwis <- function(aoi, dist = 0) {
  
  aoi %>%
    st_buffer(dist) %>%
    st_bbox() %>%
    as.vector() %>%
    round(., digits = 7) %>%
    paste(collapse = ",") %>%
    whatNWISdata(bBox = .) %>%
    st_as_sf(coords = c("dec_long_va", "dec_lat_va"),
             crs = "EPSG:4269") %>% 
    filter(data_type_cd %in% c("dv","uv"),
           parm_cd %in% c("00060"),
           end_date > as.Date(Sys.Date() - days(7))) 
  
}


# Get discharge data for NWIS site_nos
get_discharge <- function(site_nos) { 
  
  readNWISdv(
    siteNumbers = site_nos,
    parameterCd = "00060",
    startDate = "1980-01-01",
    endDate = Sys.Date())      
  
}

```

# uCLP Network

Read in uCLP Sonde Locations and NWIS locations 

```{r}

sonde_locs <- read_csv("data/metadata/sonde_location_metadata.csv") %>%
  mutate(lat = str_split(lat_long, pattern = ",", simplify = TRUE)[,1] %>% as.numeric(),
         lng = str_split(lat_long, pattern = ",", simplify = TRUE)[,2] %>% as.numeric()) %>%
  st_as_sf(coords = c("lng","lat"), crs = 4326)

search_aoi <- sonde_locs %>%
  st_bbox() %>%
  st_as_sfc()

nwis_gages <- get_recent_nwis(search_aoi, dist = 0) %>% 
  filter(data_type_cd == "dv")

watersheds <- nwis_gages$site_no %>%
  unique() %>%
  map(~ nldi_finder(site_no = .)) %>%
  bind_rows() %>% 
  left_join(., nwis_gages %>% st_drop_geometry(), by = "site_no")

# Pull discharge because why not
discharge <- get_discharge(nwis_gages$site_no %>% unique())

# preview data
mapview(watersheds,
        zcol = "station_nm",
        alpha.regions = 0.2) +
  mapview(nwis_gages,
          zcol = "station_nm",
          alpha.regions = 1,
          cex = 4) +
  mapview(sonde_locs,
          col.regions = "tomato",
          cex = 4)

```



# Example: define AOI as watershed and pull in modis

```{r define_aoi}
# for now, just start with first watershed
aoi_sf <- watersheds[1,] %>% 
  select(geometry)

# create an eeFeatureCollection for use in GEE
aoi <- aoi_sf %>%
  sf_as_ee() 
```


```{r modis_3}

# MODIS MOD10A1 (Terra) and MYD10A1 (Aqua)
mod10 <- ee$ImageCollection("MODIS/061/MOD10A1")
myd10 <- ee$ImageCollection("MODIS/061/MYD10A1")

# Combine collections and select both NDSI and the QA band
modis_snow <- mod10$merge(myd10)$select(
  c("NDSI_Snow_Cover", "NDSI_Snow_Cover_Basic_QA") 
)

# Define time range (using your provided range)
start_date <- "2022-01-01"
end_date <- "2024-01-01"

# COMBINED IMAGE PROCESSING FUNCTION WITH QA MASKING AND MEAN NDSI
get_prop_snow <- function(img) {
  
  # 1. DEFINE MASK
  qa_band <- img$select("NDSI_Snow_Cover_Basic_QA")
  # Keep observations where QA is 0 (Best quality) or 1 (Good quality).
  is_good_qa <- qa_band$lte(1) 
  
  ndsi_band <- img$select("NDSI_Snow_Cover")
  
  # 2. APPLY MASKING, CLIPPING, AND DATE SETTING
  img_processed <- ndsi_band %>%
    ee$Image$clip(aoi) %>%
    ee$Image$updateMask(is_good_qa) %>% # Apply the good quality mask
    ee$Image$set("date", img$date()$format("YYYY-MM-dd"))
  
  # 3. PERFORM REDUCEREGION CALCULATIONS
  
  # A. Total clear pixels
  total_clear_pixels <- img_processed$reduceRegion(
    reducer = ee$Reducer$count(),
    geometry = aoi,
    scale = 500,
    bestEffort = TRUE
  )$getNumber("NDSI_Snow_Cover")
  
  # B. Snow pixels (Use stricter threshold of NDSI > 40 to fix 100% issue)
  snow_pixels <- img_processed$gt(40)$reduceRegion( 
    reducer = ee$Reducer$count(),
    geometry = aoi,
    scale = 500,
    bestEffort = TRUE
  )$getNumber("NDSI_Snow_Cover")
  
  # C. MEAN NDSI (New variable for inspection)
  mean_ndsi <- img_processed$reduceRegion(
    reducer = ee$Reducer$mean(),
    geometry = aoi,
    scale = 500,
    bestEffort = TRUE
  )$getNumber("NDSI_Snow_Cover")
  
  # Calculate proportion
  prop <- snow_pixels$divide(total_clear_pixels)
  
  # 4. SET FINAL PROPERTIES
  return(img_processed$set(
    "n_pixels_clear", total_clear_pixels,
    "n_snow_pixels", snow_pixels,
    "snow_cover_fraction", prop,
    "mean_ndsi", mean_ndsi
  ))
}

# Apply date filter, then apply the single comprehensive map function
snow_summary <- modis_snow$
  filterDate(start_date, end_date)$
  map(get_prop_snow)


# Define the column names (5 columns now)
columns_to_extract <- c("date", "n_pixels_clear", "n_snow_pixels", "snow_cover_fraction", "mean_ndsi")
num_inputs <- length(columns_to_extract) # Should be 5

# Extract data
snow_list <- snow_summary$reduceColumns(
  reducer = ee$Reducer$toList(num_inputs),
  selectors = columns_to_extract
)$getInfo()

# 1. Convert the raw Python list structure to a standard R list
snow_r_list <- ee_utils_py_to_r(snow_list)

# 2. Extract the data (which is in the element named 'list') and use 
#    matrix/data.frame methods to restructure the concatenated list into columns.
#    The number of columns is 4 (the argument passed to Reducer.toList).
raw_data <- unlist(snow_r_list$list)

snow_df <- data.frame(
  matrix(
    raw_data,
    ncol = 5,
    byrow = TRUE 
  )
)

# Assign names
colnames(snow_df) <- c("date", "n_pixels_clear", "n_snow_pixels", "snow_cover_fraction", "mean_ndsi")

# Create final DF
snow_df <- snow_df %>%
  mutate(date = as.Date(date),
         across(.cols = c(n_pixels_clear, n_snow_pixels, snow_cover_fraction, mean_ndsi),
                .fns = ~ as.numeric(.)),
         # PERCENT SNOW CALCULATION: Snow Pixels divided by TOTAL CLEAR PIXELS
         # 1. Rolling Mean (frollmean) 
         roll_mean_ndsi = frollmean(mean_ndsi, 7, align = "right"), 
         # 2. Savitzky-Golay Filter 
         # p = polynomial order (2 or 3 is typical)
         # n = filter length (must be odd, e.g., 7 for a week of data)
         smooth_ndsi = sgolayfilt(x = mean_ndsi,
                                  # Polynomial order (3 = cubic smoothing)
                                  p = 3,  
                                  # Filter length (7-day window, must be odd)
                                  n = 31),
         # 3. Melt calculation (using the newly created smooth_ndsi)
         melt = frollmean(c(NA, diff(smooth_ndsi)), 7, align = "right"),
         melt = if_else(melt < 0, abs(melt), 0)) %>% 
  left_join(discharge %>% filter(site_no == watersheds[1,]$site_no), 
            by = c("date" = "Date"))

# Plotting
ggplot(snow_df, aes(x = date)) +
  geom_line(aes(y = mean_ndsi, color = "NDSI"), alpha = 0.2) +
  geom_line(aes(y = roll_mean_ndsi, color = "NDSI_7day"), alpha = 0.5) +
  geom_line(aes(y = smooth_ndsi, color = "NDSI_SGfilt")) +
  geom_line(aes(y = 10*melt, color = "Est. Melt")) +
  geom_line(aes(y = X_00060_00003, color = "Discharge"), lwd = 1) +
  theme_bw() +
  labs(x = "", y = "NDSI (-) and Discharge (cfs)") +
  scale_color_colorblind()

```
