---
title: "extract_greeley_data"
author: "Danny Duncan"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = '100%')

library(readxl)
library(tidyxl)
library(purrr)
library(here)
library(kableExtra)
library(gghighlight)
library(cdssr)
library(tidyverse)
library(reticulate)
library(parallel)

source(here("src/setup_libraries.R"))
```

# Overview

This document performs some initial data structure analysis on the raw Greeley plant chemical data. The goal is to understand the data structure, transform the data into a format suitable for analysis, and perform some basic checks to ensure the transformations are correct.

# Data Load Preparation

```{r data load prep}
# Load path to raw Greeley data folder
raw_path <- here("data/raw/chem/greeley_plant_chem")

# List all Excel files in the raw data folder
file_list <- list.files(raw_path, 
                        pattern = "\\.xlsx$", 
                        full.names = TRUE)
```

# Data Extraction

To extract data from the raw Greeley files, we will use the tidyxl package.

## Retrieve Raw Data
```{r retrieve & transform data}
# Create a list of our desired variables
vars_actual <- c("DATE","TIME", "TURBIDITY","CHLORINE","pH","ALKALINITY","COND.","T.D.S.","HACHTOC", "TEMPCEL", "POUDRE", "CBT")

# Create a regex pattern to filter for our desired variables in the character column of the tidyxl dataframe
var_list <- paste(c("DATE", "TIME", "TURBIDITY", "CHLORINE", "PH", "ALKALINITY", "COND.", "T.D.S.", "HACHTOC", "TEMPCEL", "POUDRE", "CBT"),
                  collapse = "|")

# Create a regex pattern to filter for monthly sheets in the sheet column of the tidyxl dataframe
months = paste(c("JAN", "FEB", "MAR", "APR", "MAY", "JUN", "JUL", "AUG", "SEP", "OCT", "NOV", "DEC"), 
                  collapse = "|")

# Map through all excel files and extract the tidyxl dataframes for each file, then combine into a single dataframe
for (x in file_list) {
  # Uses xlsx_cells() to extract all cells that contain data
  tidyxl_df <- xlsx_cells(paste(x), include_blank_cells = FALSE)
  # Combine all tidyxl dataframes into a single dataframe
  if (!exists("rawData_combined")) {
    rawData_combined <- tidyxl_df}
  # Creates combined dataframe if it doesn't exist
  else {
    rawData_combined <- bind_rows(rawData_combined, tidyxl_df)
  }
}

# Create a list of all monthly sheets we would like to keep for analysis
desirable_sheet_list <- unique(rawData_combined$sheet[grepl(months, rawData_combined$sheet, ignore.case = TRUE)])

# Filter our newly created rawData_combined dataframe:
filtered_tidyxl <-
  # The character column contains unnecessary leading empty space before the value; gsub() removes this empty space for uniformity
  as.data.frame(apply(rawData_combined,2, function(x) gsub("\\s+", "", x))) %>%
  # Filter for sheets that contain month abbreviations in their name
  filter(sheet %in% desirable_sheet_list) %>%
  # Filter for values in the character column that match our list of variables of interest (vars_actual)
  filter(character %in% vars_actual) %>%
  # Filter for necessary information
  select(sheet, address, character)

print(head(filtered_tidyxl))
```

The filtered_tidyxl dataframe contains all cell values and information for our variables of interest across all monthly sheets. This is a good starting point for where we need to look for these data in each sheet.

## Filter & Refine
```{r refine filtered data}
# Continue to further refine our filtered data and transform it into wide format
filtered_tidyxl_wide <- filtered_tidyxl %>%
  # Group by monthly sheet and variables
  group_by(sheet, character) %>%
  # Slice the first occurrence for each variable.
  slice(1) %>% 
  ungroup() %>%
  # Make the data wide
  pivot_wider(names_from = character, values_from = address) %>%
  # Create a year and month column by extracting from the sheet name using regular expressions.
  mutate(Month = str_extract(sheet, "(?<=^BEL)[A-Z]+(?=\\d{2}$)")) %>%
  mutate(Year  = paste0("20", str_extract(sheet, "\\d{2}$")))

print(head(filtered_tidyxl_wide))
```

The filtered_tidyxl_wide dataframe contains the first cell location for each variable within each sheet. This will be used as an index for extracting the values of each variable of interest in the next section.

## Extract WQ Data
```{r extract data with tidyxl}
# Helper: convert excel column letter(s) to number (e.g., "A" -> 1, "AD" -> 30)
col_to_num <- function(col_letter) {
  col_letter <- toupper(col_letter)
  chars <- rev(strsplit(col_letter, "")[[1]])
  sum(sapply(seq_along(chars), function(i) {
    (match(chars[i], LETTERS)) * 26^(i - 1)
  }))
}

# Helper: parse a single cell reference like "AD5" into col number and row number
parse_cell <- function(cell_str) {
  col_letter <- gsub("\\d+", "", cell_str)
  row_num    <- as.integer(gsub("[A-Za-z]+", "", cell_str))
  list(
    col = col_to_num(col_letter),
    row = row_num
  )
}

# Define variable columns (everything except sheet, Month, Year)
var_cols <- setdiff(names(filtered_tidyxl_wide), c("sheet", "Month", "Year"))

# Number of rows to extract from the start cell. 35 ensures accounts for row gaps between the variable name and the data, as well as the <= 31 rows of data we want to extract.
extract_rows <- 35

# Read the full workbook once
cells <- rawData_combined

# # Extract only the cells that fall within the specified ranges
# extractedData_base <- bind_rows(lapply(seq_len(nrow(filtered_tidyxl_wide)), function(i) {
#   row_data <- filtered_tidyxl_wide[i, ]
#   sheet_name <- as.character(row_data$sheet)
# 
#   bind_rows(lapply(var_cols, function(var_name) {
#     cell_str <- as.character(row_data[[var_name]])
# 
#     if (is.na(cell_str) || cell_str == "") return(NULL)
# 
#     ref <- parse_cell(cell_str)
#     start_row <- ref$row
#     end_row   <- start_row + extract_rows
# 
#     cells %>%
#       filter(
#         sheet == sheet_name,
#         col == ref$col,
#         row >= start_row,
#         row <= end_row
#       ) %>%
#       mutate(variable = var_name) %>%
#       select(sheet, variable, address, row, col, content, numeric, character, data_type)
#   }))
# }))

extractedData <- filtered_tidyxl_wide %>%
  pmap_dfr(function(...) {
    row_data <- tibble(...)
    sheet_name <- as.character(row_data$sheet)

    map_dfr(var_cols, function(var_name) {
      cell_str <- as.character(row_data[[var_name]])

      if (is.na(cell_str) || cell_str == "") return(tibble())

      ref <- parse_cell(cell_str)
      start_row <- ref$row
      end_row   <- start_row + extract_rows

      cells %>%
        filter(
          sheet == sheet_name,
          col == ref$col,
          row >= start_row,
          row <= end_row
        ) %>%
        mutate(variable = var_name) %>%
        select(sheet, variable, address, row, content, numeric, character, data_type)
    })
  })

# finalData_cleaned <- extractedData %>%
#   filter(data_type == "numeric") %>% 
#   select(sheet, variable, row, content) %>%
#   pivot_wider(names_from = variable, values_from = content) %>% 
#   mutate(Month = str_extract(sheet, "(?i)(JAN|FEB|MAR|APR|MAY|JUN|JUL|AUG|SEP|OCT|NOV|DEC)")) %>%
#   mutate(Year  = paste0("20", str_extract(sheet, "\\d{2}$"))) %>% 
#   mutate(Day = DATE) %>%
#   mutate(Date = parse_date_time(paste(Year, Month, Day), orders = "Ybd")) %>%
#   select(Date, Year, Month, Day, TIME, ALKALINITY, CHLORINE, COND., `T.D.S.`, TEMPCEL, TURBIDITY, pH, sheet, row)
```

One problem with the commented-out code above is that some of the DATE (day) values are in character format instead of numeric. To fix this, we need to create a function which uses an if_else statement to create the day column.

## Clean Data
```{r clean data}
cleanedData <- extractedData %>%
  # Edit the numeric column so that if the value is NA but the character column contains a numeric value, it will place that value in the numeric column instead.
  mutate(numeric = if_else(is.na(numeric) & !is.na(as.numeric(character)),
                         as.numeric(character),
                         numeric)) %>%
  # Change the data_type to "numeric" for rows that were just transformed.
  mutate(data_type = if_else(!is.na(numeric) & !is.na(character),
                         "numeric",
                         data_type)) %>%
  # Filter for only numeric rows
  filter(data_type == "numeric") %>% 
  select(sheet, variable, row, numeric) %>%
  # Remove duplicate rows before pivoting
  distinct(sheet, variable, row, .keep_all = TRUE) %>%
  pivot_wider(names_from = variable, values_from = numeric) %>%
  mutate(Month = toupper(str_extract(sheet, "(?i)(JAN|FEB|MAR|APR|MAY|JUN|JUL|AUG|SEP|OCT|NOV|DEC)"))) %>%
  mutate(Year  = paste0("20", str_extract(sheet, "\\d{2}$"))) %>%
  mutate(Day = DATE) %>%
  mutate(DTG = parse_date_time(paste(Year, Month, Day), orders = "Ybd")) %>%
  select(DTG, Year, Month, Day, TIME, ALKALINITY, CHLORINE, COND., `T.D.S.`, TEMPCEL, TURBIDITY, POUDRE, CBT, pH, sheet, row, col) %>% 
  drop_na(DTG)

print(head(cleanedData))
```

Unfortunately, some of the DTG are still failing to parse. This might be because some DATE (day) values aren't present in a few of the original datasheets. We'll have to explore these to see if there isn't a day column within the actual sheet, or if there's a bug in the code that's causing this issue.

# Data Exploration
```{r}
# For now, lets see how the data looks
cleanedData_long <- cleanedData %>% 
  select(!c(sheet, row, Year, Month, Day)) %>%
  pivot_longer(cols = c(ALKALINITY, CHLORINE, COND., `T.D.S.`, TEMPCEL, TURBIDITY, POUDRE, CBT, pH), 
               names_to = "variable", values_to = "value")

# Graph all variables over time to see if there are any obvious issues with the data
cleanedData_long %>% 
  ggplot(aes(x = DTG, y = value, colour = variable)) +
  geom_line() +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Greeley Plant Chemical Data Over Time",
       x = "Date",
       y = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal()

# Graphing Alkalinity alongside Conductivity
cleanedData_long %>% 
  filter(variable == c("ALKALINITY", "COND.")) %>%
  ggplot(aes(x = DTG, y = value, colour = variable)) +
  geom_line() +
  scale_x_datetime(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Greeley Plant Chemical Data Over Time",
       x = "Date",
       y = "Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal()
```

