---
title: "extract_MODIS_snow_data"
output: html_document
---

The purpose of this markdown is to grab MODIS snow cover data using the rgee
package and summarize it for a catchment (defined by locations of monitored 
discharge). Catchments will be defined using nhdplusTools.

MODIS Snow Cover: https://doi.org/10.5067/MODIS/MOD10A1.061

```{r setup}
library(rgee)
library(sf)
library(nhdplusTools)
library(tidyverse)
library(geojsonio)
library(data.table)
library(mapview)
library(dataRetrieval)
library(purrr)
library(signal)
library(ggthemes)
library(ggpubr)
library(zoo)

source("Scratch/pySetup.R")


# Example pySetup that this was copied from (Aquasat):
# https://github.com/AquaSat/AquaMatch_lakeSR/blob/main/python/pySetup.R
# Note, on the first run, you can just manually go in and run the lines of 
# code in pySetup that 

# see also rgee setup instructions:
# https://cran.r-project.org/web/packages/rgee/vignettes/rgee01.html#6_Installation

# Initialize Earth Engine
ee_Initialize()
```


# define functions

Now, define some helpful functions. Note, these can be added to src/ later
if we end up using this function.

```{r define_functions}

#' Returns watershed sf object for a given USGS NWIS site_no
#' @param site_no USGS site number
#' @return gage_basin sf object of watershed associated with USGS site

nldi_finder <- function(site_no) {
  #get those gages' watersheds using get_nldi_basin in the {nhdplusTools}.
  nldi_nwis <- list(featureSource = "nwissite",
                    featureID = paste0("USGS-", site_no))

  gage_basin <-
    tryCatch({
      invisible(
        suppressMessages(
          nhdplusTools::get_nldi_basin(nldi_feature = nldi_nwis) %>%
            sf::st_transform(., 4269) %>%
            dplyr::mutate(site_no = site_no)))
    }, error = function(e) {
      NA
    })

  return(gage_basin)
}



# Get nwis discharge sites within an AOI - only returns sites
# with recent data within the last week

get_recent_nwis <- function(aoi, dist = 0) {
  aoi %>%
  st_buffer(dist) %>%
  st_bbox() %>%
  as.vector() %>%
  round(., digits = 7) %>%
  paste(collapse = ",") %>%
      dataRetrieval::whatNWISdata(bBox = .) %>%
  st_as_sf(coords = c("dec_long_va", "dec_lat_va"), crs = 4269) %>%
  dplyr::filter(data_type_cd %in% c("dv","uv"),
                parm_cd %in% c("00060"),
                end_date > as.Date(Sys.Date() - days(7))) 
}


# Get discharge data for NWIS site_nos
get_discharge <- function(site_nos) { 
  dataRetrieval::readNWISdv(
    siteNumbers = site_nos,
    parameterCd = "00060",
    startDate = "1980-01-01",
    endDate = Sys.Date())      
}



#' @title Low-Pass Filter Smoothing using Fast Fourier Transform (FFT)
#'
#' @description
#' Smooths a noisy time series by applying a low-pass filter in the frequency domain.
#' High-frequency components (noise) are attenuated by setting their FFT coefficients to zero,
#' and the signal is then reconstructed using the Inverse FFT.
#'
#' @param signal A numeric vector representing the noisy time series data.
#' @param cutoff An integer specifying the **Low-Pass Cutoff frequency** component to retain.
#'   This value determines the degree of smoothing. Components from the beginning (low
#'   frequencies) up to `cutoff` are kept. A smaller number results in a tighter,
#'   smoother filter (more noise removed).
#'
#' @return A numeric vector, the same length as the input `signal`, containing the
#'   low-pass filtered (smoothed) time series.

fft_smooth <- function(signal, cutoff = 5) {
  
  # Perform Fast Fourier Transform (FFT)
  fft_result <- fft(signal)
  
  N <- length(signal)
  
  # Check for valid cutoff value
  if (cutoff >= N / 2) {
    warning("Cutoff is too large. Setting cutoff to floor(N/2) - 1.")
    cutoff <- floor(N / 2) - 1
  }
  
  # Zero out the high-frequency components: Set coefficients from the cutoff 
  # index up to the corresponding symmetrical point (N - cutoff + 1) to zero. 
  # This retains the low frequencies and their necessary complex conjugates.
  fft_result[cutoff:(N - cutoff + 1)] <- 0
  
  # Perform the Inverse FFT (IFFT) and normalize
  # We take the real part (Re) and divide by N for proper normalization.
  smoothed_signal <- Re(fft(fft_result, inverse = TRUE)) / N
  
  return(smoothed_signal)
}



```

# uCLP Network

Read in uCLP Sonde Locations and NWIS locations 

```{r}

sonde_locs <- read.csv("data/metadata/sonde_location_metadata.csv") %>%
  dplyr::mutate(
    lat = str_split(lat_long, pattern = ",", simplify = TRUE)[,1] %>% as.numeric(),
    lng = str_split(lat_long, pattern = ",", simplify = TRUE)[,2] %>% as.numeric()) %>%
  st_as_sf(coords = c("lng","lat"), crs = 4326)

search_aoi <- sonde_locs %>%
  st_bbox() %>%
  st_as_sfc() %>%
  st_as_sf()

nwis_gages <- get_recent_nwis(search_aoi, dist = 0) %>% 
  dplyr::filter(data_type_cd == "dv")

watersheds <- nwis_gages$site_no %>%
  unique() %>%
  purrr::map_dfr(~nldi_finder(site_no = .)) %>%
  left_join(nwis_gages  %>% st_drop_geometry(), by = "site_no")

# Pull discharge because why not
discharge <- get_discharge(nwis_gages$site_no %>% unique())

# preview data
mapview(watersheds,
        zcol = "station_nm",
        alpha.regions = 0.2) +
  mapview(nwis_gages,
          zcol = "station_nm",
          alpha.regions = 1,
          cex = 4) +
  mapview(sonde_locs,
          col.regions = "tomato",
          cex = 4)

```



# Example: define AOI as watershed and pull in modis

```{r define_aoi}
# for now, just start with first watershed
aoi_sf <- watersheds[1,] %>% dplyr::summarize()

aoi <- aoi_sf %>%
  sf_as_ee()# lower poudre gage


#aoi <- nldi_finder("06746110") %>% # Joe Wright Reservoir Gage (small area for testing)
#  sf_as_ee()
```


```{r modis_3}

# MODIS MOD10A1 (Terra) and MYD10A1 (Aqua)
mod10 <- ee$ImageCollection("MODIS/061/MOD10A1")
myd10 <- ee$ImageCollection("MODIS/061/MYD10A1")

# Combine collections and select both NDSI and the QA band
modis_snow <- mod10$merge(myd10)$select(
    c("NDSI_Snow_Cover", "NDSI_Snow_Cover_Basic_QA") 
)

# Define time range (using your provided range)
start_date <- "2022-01-01"
end_date <- "2024-01-01"

# COMBINED IMAGE PROCESSING FUNCTION WITH QA MASKING AND MEAN NDSI
get_prop_snow <- function(img) {
  
    # 1. DEFINE MASK
    qa_band <- img$select("NDSI_Snow_Cover_Basic_QA")
    # Keep observations where QA is 0 (Best quality) or 1 (Good quality).
    is_good_qa <- qa_band$lte(0) 
    
    ndsi_band <- img$select("NDSI_Snow_Cover")
    
    # 2. APPLY MASKING, CLIPPING, AND DATE SETTING
    img_processed <- ndsi_band %>%
        ee$Image$clip(aoi) %>%
        ee$Image$updateMask(is_good_qa) %>% # Apply the good quality mask
        ee$Image$set("date", img$date()$format("YYYY-MM-dd"))
    
    # 3. PERFORM REDUCEREGION CALCULATIONS
    
    # A. Total clear pixels
    total_clear_pixels <- img_processed$reduceRegion(
        reducer = ee$Reducer$count(),
        geometry = aoi,
        scale = 500,
        bestEffort = TRUE
    )$getNumber("NDSI_Snow_Cover")
    
    # B. Snow pixels (Use stricter threshold of NDSI > 40 to fix 100% issue)
    snow_pixels <- img_processed$gt(40)$reduceRegion( 
        reducer = ee$Reducer$count(),
        geometry = aoi,
        scale = 500,
        bestEffort = TRUE
    )$getNumber("NDSI_Snow_Cover")
    
    # C. MEAN NDSI (New variable for inspection)
    mean_ndsi <- img_processed$reduceRegion(
        reducer = ee$Reducer$mean(),
        geometry = aoi,
        scale = 500,
        bestEffort = TRUE
    )$getNumber("NDSI_Snow_Cover")
    
    # Calculate proportion
    prop <- snow_pixels$divide(total_clear_pixels)
    
    # 4. SET FINAL PROPERTIES
    return(img_processed$set(
        "n_pixels_clear", total_clear_pixels,
        "n_snow_pixels", snow_pixels,
        "snow_cover_fraction", prop,
        "mean_ndsi", mean_ndsi
    ))
}

# Apply date filter, then apply the single comprehensive map function
snow_summary <- modis_snow$
    filterDate(start_date, end_date)$
    map(get_prop_snow)


# Define the column names (5 columns now)
columns_to_extract <- c("date", "n_pixels_clear", "n_snow_pixels", "snow_cover_fraction", "mean_ndsi")
num_inputs <- length(columns_to_extract) # Should be 5

# Extract data
snow_list <- snow_summary$reduceColumns(
    reducer = ee$Reducer$toList(num_inputs),
    selectors = columns_to_extract
)$getInfo()

# 1. Convert the raw Python list structure to a standard R list
snow_r_list <- rgee::ee_utils_py_to_r(snow_list)

# 2. Extract the data (which is in the element named 'list') and use 
#    matrix/data.frame methods to restructure the concatenated list into columns.
#    The number of columns is 4 (the argument passed to Reducer.toList).
raw_data <- unlist(snow_r_list$list)

snow_df <- data.frame(
    matrix(
        raw_data,
        ncol = 5,
        byrow = TRUE
    )
) %>%
  setNames(c("date", "n_pixels_clear", "n_snow_pixels", "snow_cover_fraction", "mean_ndsi")) %>%
  mutate(
    date = as.Date(date),
    n_pixels_clear = as.numeric(n_pixels_clear), 
    n_snow_pixels = as.numeric(n_snow_pixels),
    snow_cover_fraction = as.numeric(snow_cover_fraction),
    mean_ndsi = as.numeric(mean_ndsi)
  ) %>%
  dplyr::filter(n_pixels_clear > quantile(n_pixels_clear, .4, na.rm = TRUE)) %>%
  # First, fill gaps
  tidyr::complete(date = seq.Date(min(date, na.rm = TRUE), max(date, na.rm = TRUE), by = "day")) %>%
  dplyr::mutate(mean_ndsi = na.approx(mean_ndsi)) %>%
  # CORRECT PERCENT SNOW CALCULATION: Snow Pixels divided by TOTAL CLEAR PIXELS
  dplyr::mutate(
    # 1. Rolling Mean (frollmean) - Unchanged
    roll_mean_ndsi = frollmean(mean_ndsi, 14, align = "right"), 
    # 2. Savitzky-Golay Filter 
    # p = polynomial order (2 or 3 is typical)
    # n = filter length (must be odd, e.g., 7 for a week of data)
    smooth_ndsi = signal::sgolayfilt(
      x = mean_ndsi,
      p = 2,  # Polynomial order (3 = cubic smoothing)
      n = 31   # Filter length (7-day window, must be odd)
    ),
     smooth_ndsi_der = -18 * signal::sgolayfilt(
      x = mean_ndsi,
      p = 2,  # Polynomial order (3 = cubic smoothing)
      n = 31,   # Filter length (7-day window, must be odd)
      m = 1 # first derivative
    ),
    fft_ndsi = fft_smooth(mean_ndsi, cutoff = 20),
    # 3. Melt calculation (using the newly created roll_mean_ndsi) - Unchanged
    #melt = frollmean(c(NA, diff(smooth_ndsi)), 7, align = "right")
     melt = c(NA, diff(roll_mean_ndsi))
  ) %>%
  dplyr::mutate(melt = ifelse(melt < 0, -melt, 0),
                smooth_ndsi_der = ifelse(smooth_ndsi_der < 0, 0, smooth_ndsi_der)) %>%
  dplyr::left_join(discharge %>% dplyr::filter(site_no == watersheds[1,]$site_no), 
                   by = c("date" = "Date"))





# Plot
ggarrange(
ggplot(snow_df, aes(x = date)) +
  geom_line(aes(y = mean_ndsi, color = "NDSI"), alpha = 0.6, lwd = 1) +
  geom_line(aes(y = roll_mean_ndsi, color = "NDSI_14dayMA"), alpha = 0.5, lwd = 1) +
  geom_line(aes(y = smooth_ndsi, color = "NDSI_SGolay"), lwd = 1) +
  geom_line(aes(y = fft_ndsi, color = "NDSI_FFT"), lwd = 1) +
  #geom_line(aes(y = 10*melt, color = "Est. Melt")) +
  #geom_smooth(method = "loess", se = FALSE, aes(y = 10*melt, color = "Est. Melt (Loess)"), span = .01) +
    geom_line(aes(y = X_00060_00003, color = "Discharge"), lwd = 1, alpha = 0.5) +
  theme_bw() +
  labs(x = "", y = "NDSI (-) and Discharge (cfs)") +
  scale_color_colorblind(name = ""),
ggplot(snow_df, aes(x = date)) +
  #geom_line(aes(y = mean_ndsi, color = "NDSI"), alpha = 0.2) +
  #geom_line(aes(y = roll_mean_ndsi, color = "NDSI_7day"), alpha = 0.5) +
  #geom_line(aes(y = smooth_ndsi, color = "NDSI_SGfilt")) +
  geom_line(aes(y = X_00060_00003, color = "Discharge"), lwd = 1, alpha = 0.5) +
  #geom_line(aes(y = 15*melt, color = "Mv. Avg. Melt"), lwd = 1, alpha = 0.8) +
  geom_smooth(method = "loess", se = FALSE, aes(y = 15*melt, color = "Melt_14dayMA_plus_Loess"), span = .05, alpha = 0.8, lwd = 1) +
  geom_line(aes(y = smooth_ndsi_der, color = "Melt_SGolay"), lwd = 1) +
  theme_bw() +
  labs(x = "", y = "NDSI Melt (-) and Discharge (cfs)") +
  scale_color_colorblind(name = ""),
ncol = 1, align = "hv"
)

```
