---
title: "extract_MODIS_snow_data"
output: html_document
---

This markdown explores a workflow to grab MODIS snow cover data using the rgee package and summarizes it for a catchment to return daily values. UCLP sonde locations are used to create a search AOI for which to find NWIS gauges. Then, watershed polygons associated with each gauge are read in using nhdplusTools. Watershed polygon are then set as the AOI for which which to summarize daily MODIS NDSI snow cover statistics for a user-defined period of time - e.g., the mean daily snow cover for the AOI. The processing function filters to only include good and high quality pixels as defined by a generic QA band associated with the dataset. A melt proxy variable is then calculated from the daily snow cover fraction by differencing the NDSI snow cover through time. Because the NDSI snow cover is very noisy, smoothing must occur prior to calculating melt. At the end of the script, cloud cover is explored using a cloud band to consider how much clouds contribute to noise. 

MODIS Snow Cover: https://doi.org/10.5067/MODIS/MOD10A1.061

# Requirements

This workflow requires that you have access to the ROSSyndicate Google Account (ask for PW if you need it) and that you have set up a symlink to the Poudre_ROSS Sharepoint Document Library. See the Sharepoint site for directions to set this up.

# Setup
```{r setup}
source("Scratch/pySetup.R")
# A warning message here is okay!

library(rgee)
library(sf)
library(nhdplusTools)
library(geojsonio)
library(data.table)
library(mapview)
library(dataRetrieval)
library(signal)
library(ggthemes)
library(zoo)
library(plotly)
library(ggpubr)
library(tidyverse)

# Authenticate your GEE account - only need to do once
#ee_Authenticate(user = "therossyndicate@gmail.com")

# strange workaround here to get rgee to initialize
# see notes here: https://github.com/r-spatial/rgee/issues/355
ee$Initialize(project = "poudre-dss")

```


# Define functions

Now, define some helpful functions. Note, these can be added to src/ later

```{r define_functions}

#' Returns watershed sf object for a given USGS NWIS site_no
#' @param site_no USGS site number
#' 
#' @return gage_basin sf object of watershed associated with USGS site
#' 
nldi_finder <- function(site_no) {
  
  #get those gages' watersheds using get_nldi_basin in the {nhdplusTools}.
  nldi_nwis <- list(featureSource = "nwissite",
                    featureID = paste0("USGS-", site_no))
  
  gage_basin <- tryCatch({invisible(
    suppressMessages(
      get_nldi_basin(nldi_feature = nldi_nwis) %>%
        mutate(site_no = site_no)))
  }, 
  error = function(e) {
    NULL
  })
  
  return(gage_basin)
}

#' Function that pulls in NWIS gauge sites with discharge data recorded within
#' the last week
#' @param aoi sf_geometry of area of interest to search for gauges in
#' @param dist buffer distance around AOI to search for gauges
#' 
#' @return sf_dataframe of gauge sites with some metadata about gauge
#' 
get_recent_nwis <- function(aoi, dist = 0) {
  
  aoi %>%
    st_buffer(dist) %>%
    st_bbox() %>%
    as.vector() %>%
    round(., digits = 7) %>%
    paste(collapse = ",") %>%
    whatNWISdata(bBox = .) %>%
    st_as_sf(coords = c("dec_long_va", "dec_lat_va"),
             crs = "EPSG:4269") %>% 
    filter(data_type_cd %in% c("dv","uv"),
           parm_cd %in% c("00060"),
           end_date > as.Date(Sys.Date() - days(7))) 
  
}


# Function that pulls discharge data for NWIS gauge sites given site number
#' @param site_nos a character array of USGS site numbers
#' 
#' @return A dataframe containing discharge data for site. 
#' 
get_discharge <- function(site_nos) { 
  
  readNWISdv(
    siteNumbers = site_nos,
    parameterCd = "00060",
    startDate = "1980-01-01",
    endDate = Sys.Date()) %>%
    rename(c("discharge_cfs" = "X_00060_00003",
             "status" = "X_00060_00003_cd"))
  
}

#' A filter function that smooths noisy time series by applying a low-pass 
#' filter in the frequency domain. High-frequency components (noise) are 
#' attenuated by setting their FFT coefficients to zero, and signal is then 
#' reconstructed using the Inverse FFT.
#'
#' @param signal A numeric vector representing the noisy time series data.
#' @param cutoff An integer specifying the **Low-Pass Cutoff frequency** component to retain.
#'   This value determines the degree of smoothing. Components from the beginning (low
#'   frequencies) up to `cutoff` are kept. A smaller number results in a tighter,
#'   smoother filter (more noise removed).
#'
#' @return A numeric vector, the same length as the input `signal`, containing the
#'   low-pass filtered (smoothed) time series.
#'   
fft_smooth <- function(signal, cutoff = 5) {
  
  # Perform Fast Fourier Transform (FFT)
  fft_result <- fft(signal)
  
  N <- length(signal)
  
  # Check for valid cutoff value
  if (cutoff >= N / 2) {
    warning("Cutoff is too large. Setting cutoff to floor(N/2) - 1.")
    cutoff <- floor(N / 2) - 1
  }
  
  # Zero out the high-frequency components: Set coefficients from the cutoff 
  # index up to the corresponding symmetrical point (N - cutoff + 1) to zero. 
  # This retains the low frequencies and their necessary complex conjugates.
  fft_result[cutoff:(N - cutoff + 1)] <- 0
  
  # Perform the Inverse FFT (IFFT) and normalize
  # We take the real part (Re) and divide by N for proper normalization.
  smoothed_signal <- Re(fft(fft_result, inverse = TRUE)) / N
  
  return(smoothed_signal)
}


```

# uCLP Network

Read in uCLP Sonde Locations and NWIS locations 

```{r}

sonde_locs <- read_csv("data/metadata/sonde_location_metadata.csv") %>%
  mutate(lat = str_split(lat_long, pattern = ",", simplify = TRUE)[,1] %>% as.numeric(),
         lng = str_split(lat_long, pattern = ",", simplify = TRUE)[,2] %>% as.numeric()) %>%
  st_as_sf(coords = c("lng","lat"), crs = 4326)

# Define search aoi based on bounding box of sonde locations
search_aoi <- sonde_locs %>%
  st_bbox() %>%
  st_as_sfc()

# Read in NWIS gauge locations that have data available from the last week
# apply a buffer distance of 0 - if relevant gauges are upstream of the
# bounding box of sondes, which is likely, we can increase distance later
nwis_gages <- get_recent_nwis(search_aoi, dist = 0) %>% 
  filter(data_type_cd == "dv")

# Now, read in watersheds for gauges
watersheds <- nwis_gages$site_no %>%
  unique() %>%
  map(~ nldi_finder(site_no = .)) %>%
  bind_rows() %>% 
  left_join(., nwis_gages %>% st_drop_geometry(), by = "site_no")

# preview data
mapview(watersheds,
        zcol = "station_nm",
        alpha.regions = 0.2) +
  mapview(nwis_gages,
          zcol = "station_nm",
          alpha.regions = 1,
          cex = 4) +
  mapview(sonde_locs,
          col.regions = "tomato",
          cex = 4)

```

We'll also pull in discharge data for the gauges for comparison with snow later on
```{r discharge}

# Pull in discharge data for gauges
discharge <- get_discharge(nwis_gages$site_no %>% unique())

ggplot(discharge, aes(x = Date, y = discharge_cfs, color = site_no)) +
  geom_line() +
  geom_point() +
  scale_color_colorblind(name = "USGS Site No.") +
  theme_bw() +
  labs(y = "Discharge (cfs)", x = "") +
  facet_wrap(~site_no, scales = "free")

```


# MODIS

Now, begin workflow to pull in MODIS Snow Cover data.


## AOI
First, define an area of interest (AOI). Here, I select the first watershed in the NWIS watersheds as it's small and therefore quick for testing. 

```{r define_aoi}
# for now, just start with first watershed
n <- 1
aoi_sf <- watersheds[n,] %>% 
  select(geometry)

# Create an eeFeatureCollection for use in GEE
aoi <- aoi_sf %>%
  sf_as_ee()

coords <- st_centroid(aoi_sf) %>% st_coordinates()

mapview(aoi_sf)

```


### - MODIS crop + filter + retrieve NDSI

Now, use the rgee package to pull in MODIS Normalized Difference Snow Index timeseries for the AOI. Snow has strong visible reflectance and strong short-wave IR absorbing characteristics. The Normalized Difference Snow Index (NDSI) is an effective way to distinguish snow from many other surface features. This is analagous to the similar index for vegetation, the Normalized Difference Vegetation Index (NDVI). Additionally, some snow/cloud discrimination is accomplished using the NDSI.The associated dataset includes snow cover, quality assurance (QA) SDSs, latitude and longitude SDSs, and local and global attributes. The NDSI snow cover algorithm identifies snow-covered land and snow-covered ice on inland water. Read more about it here: https://modis.gsfc.nasa.gov/data/dataprod/mod10.php and here: https://modis.gsfc.nasa.gov/data/atbd/atbd_mod10.pdf.


Read in and plot the first 60 images as a composite

```{r modis}

#NDSI is produced from both the Terra and Aqua satellites, so start by bringing 
# them both in and merging them. Specifically, the two image collections are:
#     - MODIS MOD10A1 (Terra), and 
#     - MYD10A1 (Aqua)
mod10 <- ee$ImageCollection("MODIS/061/MOD10A1")
myd10 <- ee$ImageCollection("MODIS/061/MYD10A1")

# Now, merge the collections and select the Normalized Difference Snow Index (NDSI)
# and a predefined quality (QA) band.
modis_snow <- mod10$merge(myd10)$select(
  c("NDSI_Snow_Cover", "NDSI_Snow_Cover_Basic_QA") 
)

# Define a time range for which to pull modis data - for now, keeping this
# small for testing.
start_date <- "2020-01-01"
end_date <- "2024-01-01"

# Plot composite of NDSI snow cover for first 60 images just to see what it looks like

Map$setCenter(coords[1], coords[2], 13)
Map$addLayer(
  eeObject = modis_snow$limit(60)$mosaic() %>% ee$Image$clip(aoi),
  visParams = list(
    bands = c("NDSI_Snow_Cover"),
    min = 0.0,
    max = 100.0,
    palette = c("black", '#0dffff', "#0524ff", "#ffffff")
  ),
  name = "Snow Cover"
) +
  Map$addLegend(
     visParams = list(
    bands = c("NDSI_Snow_Cover"),
    min = 0.0,
    max = 100.0,
    palette = c("black", '#0dffff', "#0524ff", "#ffffff")
  ),
  name = "Snow Cover"
  )

```

Now, define a processing function to extract statistics for daily rasters throughout the time period for the AOI. 
Perform filtering using the NDSI_Snow_Cover_Basic_QA band - keep all pixels that are considered "good quality".

Note: A NDSI threshold filter was tested in this script but was determined to not be useful (indicated by "gt" for greater than threshold in the following code). It was kept in because this script is exploratory and others may want to test this. Applying this threshold effectively removes no and low snow pixels which made it impossible to see changes in snow cover for the AOI below that threshold. Because we're interested in melt, we need to retain now and low snow pixels.

```{r MODIS snow_properties}

# Next, define a combined image processing function that performs QA masking
# and calculates the mean NDSI for an area of interest. 

get_prop_snow <- function(img) {
  
    # 1. Define Bands
    qa_band <- img$select("NDSI_Snow_Cover_Basic_QA")
    ndsi_band <- img$select("NDSI_Snow_Cover")
    
    # 2. Define masks
    is_good_qa <- qa_band$lte(1) # QA > 0 (Best quality) or 1 (Good quality).
    ndsi_gt <- ndsi_band$gte(50) # NDSI > 50 for snow pixels
    
    # 3. Apply masking, clipping, and date setting
    img_processed <- ndsi_band %>%
        ee$Image$clip(aoi) %>%
        ee$Image$updateMask(is_good_qa) %>% # Apply the good quality mask
        ee$Image$set("date", img$date()$format("YYYY-MM-dd"))
    
    img_processed_gt <- ndsi_band %>%
        ee$Image$clip(aoi) %>%
        ee$Image$updateMask(ndsi_gt) %>% # Apply the NDSI threshold mask
        ee$Image$updateMask(is_good_qa) %>% # Apply the good quality mask
        ee$Image$set("date", img$date()$format("YYYY-MM-dd"))
    
    # 4. Perform reduceregion calculations 
    
    # A. Total clear pixels
    total_clear_pixels <- img_processed$reduceRegion(
        reducer = ee$Reducer$count(),
        geometry = aoi,
        scale = 500,
        bestEffort = TRUE
    )$getNumber("NDSI_Snow_Cover")
    
    # B. Snow pixels (Use stricter threshold of NDSI > 40 to fix 100% issue)
    snow_pixels <- img_processed_gt$reduceRegion( 
        reducer = ee$Reducer$count(),
        geometry = aoi,
        scale = 500,
        bestEffort = TRUE
    )$getNumber("NDSI_Snow_Cover")
    
    # C. MEAN NDSI
    mean_ndsi <- img_processed$reduceRegion(
        reducer = ee$Reducer$mean(),
        geometry = aoi,
        scale = 500,
        bestEffort = TRUE
    )$getNumber("NDSI_Snow_Cover")
    
    # D. MEAN NDSI greater than threshold (40 in this case) 
    mean_ndsi_gt <- img_processed_gt$reduceRegion(
        reducer = ee$Reducer$mean(),
        geometry = aoi,
        scale = 500,
        bestEffort = TRUE
    )$getNumber("NDSI_Snow_Cover")
    
    # E. MEDIAN NDSI
    med_ndsi <- img_processed$reduceRegion(
        reducer = ee$Reducer$median(),
        geometry = aoi,
        scale = 500,
        bestEffort = TRUE
    )$getNumber("NDSI_Snow_Cover")
    
    
    # Calculate proportion
    prop <- snow_pixels$divide(total_clear_pixels)
    
    # 4. SET FINAL PROPERTIES
    return(img_processed$set(
        "n_pixels_clear", total_clear_pixels,
        "n_snow_pixels", snow_pixels,
        "snow_cover_fraction", prop,
        "mean_ndsi", mean_ndsi,
        "mean_ndsi_gt", mean_ndsi_gt,
        "med_ndsi", med_ndsi
    ))
}

# Apply date filter, then apply the single comprehensive map function
snow_summary <- modis_snow$
  filterDate(start_date, end_date)$
  map(get_prop_snow)


# Define the column names (5 columns now)
columns_to_extract <- c("date", "n_pixels_clear", "n_snow_pixels", "snow_cover_fraction", "mean_ndsi", "mean_ndsi_gt", "med_ndsi")
num_inputs <- length(columns_to_extract) 

# Extract data
snow_list <- snow_summary$reduceColumns(
  reducer = ee$Reducer$toList(num_inputs),
  selectors = columns_to_extract
)$getInfo()

# 1. Convert the raw Python list structure to a standard R list
snow_r_list <- ee_utils_py_to_r(snow_list)

# 2. Extract the data (which is in the element named 'list') and use 
#    matrix/data.frame methods to restructure the concatenated list into columns.
#    The number of columns is 4 (the argument passed to Reducer.toList).
raw_data <- unlist(snow_r_list$list)

snow_df <- data.frame(
    matrix(
        raw_data,
        ncol = num_inputs,
        byrow = TRUE
    )
) %>%
# Assign names
  setNames(columns_to_extract) %>%
 
 # Clean final DF
  mutate(
    date = as.Date(date),
    across(.cols = columns_to_extract[2:num_inputs],
           .fns = ~ as.numeric(.))
    ) %>%
    # Only keep clear pixels - can comment out if not
    filter(n_pixels_clear > quantile(n_pixels_clear, .4, na.rm = TRUE)) %>%
    # Now fill gaps for any missing dates with interpolated value.
    tidyr::complete(date = seq.Date(min(date, na.rm = TRUE), max(date, na.rm = TRUE), by = "day")) %>%
    mutate(mean_ndsi = na.approx(mean_ndsi)) %>%
    # NDSI Smoothing
    mutate(
    # 1. Moving Average (frollmean)
    ndsi_ma = frollmean(mean_ndsi, 14, align = "right"), 
    # 2. Savitzky-Golay Filter 
    ndsi_sg = signal::sgolayfilt(
      x = mean_ndsi,
      p = 1,  # Polynomial order (3 = cubic smoothing)
      n = 41),   # Filter length (7-day window, must be odd)
    # 3. Fourier transform low frequency filter
    ndsi_fft = fft_smooth(mean_ndsi, cutoff = 30),
    # Calculate melt as negative changes in snow cover
    melt_ma = c(NA, diff(ndsi_ma)),
    melt_sg = -18 * signal::sgolayfilt(
      x = mean_ndsi,
      p = 1,  # Polynomial order (3 = cubic smoothing)
      n = 41,   # Filter length (7-day window, must be odd)
      m = 1 # first derivative
    ),
    melt_fft = c(NA, diff(ndsi_fft))
  ) %>%
  dplyr::mutate(#melt_ma_ma = frollmean(melt_ma, 14, align = right),
                melt_ma = ifelse(melt_ma < 0, -melt_ma, 0),
                #melt_ma_ma = ifelse(melt_ma_ma < 0, -melt_ma_ma, 0),
                melt_fft = ifelse(melt_fft < 0, -melt_fft, 0),
                melt_sg = ifelse(melt_sg < 0, 0, melt_sg)) %>%
  dplyr::left_join(discharge %>% dplyr::filter(site_no == watersheds[n,]$site_no), 
                   by = c("date" = "Date"))

```

### - Plot NDSI 

Note, this plot shows the issue with using an NDSI threshold (only retains high snowcover cells)

```{r plot_NDSI}

# Just view NDSI Data through time
  ggplotly(
  ggplot(snow_df, aes(x = date)) +
    geom_line(aes(y = mean_ndsi, color = "Mean NDSI")) +
    geom_line(aes(y = mean_ndsi_gt, color = "Mean NDSI > 50")) +
    geom_line(aes(y = med_ndsi, color = "Median NDSI")) +
    geom_point(aes(y = mean_ndsi, color = "Mean NDSI")) +
    geom_point(aes(y = mean_ndsi_gt, color = "Mean NDSI > 50")) +
    geom_point(aes(y = med_ndsi, color = "Median NDSI")) +
    theme_bw() +
    scale_color_colorblind(name = ""))
```


# Now plot NDSI and melt with discharge
```{r plot_NDSI_MELT_DISCHARGE}


# Create multiplier for scaling melt to plots - note melt is a change in fraction
# so it is unitless. NDSI is a percentage snow cover so it is unitless also.
  
mult <- max(snow_df$discharge_cfs, na.rm = TRUE)

ggarrange(
  ggplot(snow_df, aes(x = date)) +
  geom_line(aes(y = mult * mean_ndsi, color = "NDSI"), alpha = 0.6, lwd = 1) +
  geom_line(aes(y = mult * ndsi_ma, color = "NDSI_14dayMA"), alpha = 0.5, lwd = 1) +
  geom_line(aes(y = mult * ndsi_sg, color = "NDSI_SGolay"), lwd = 1) +
  geom_line(aes(y = mult * ndsi_fft, color = "NDSI_FFT"), lwd = 1) +
  #geom_line(aes(y = 10*melt, color = "Est. Melt")) +
  #geom_smooth(method = "loess", se = FALSE, aes(y = 10*melt, color = "Est. Melt (Loess)"), span = .01) +
    geom_line(aes(y = discharge_cfs, color = "Discharge"), lwd = 1, alpha = 0.5) +
  theme_bw() +
  labs(x = "", y = "NDSI (-) and Discharge (cfs)") +
  scale_color_colorblind(name = ""),
ggplot(snow_df, aes(x = date)) +
  geom_line(aes(y = discharge_cfs, color = "Discharge"), lwd = 1, alpha = 0.9) +
  geom_line(aes(y = melt_ma * mult / max(melt_ma, na.rm = TRUE), color = "Melt_MA"), lwd = 0.75, alpha = 0.2) +
  #geom_smooth(method = "loess", se = FALSE, aes(y = 30 * mult * melt, color = "Melt_14dayMA_plus_Loess"), span = .05, alpha = 0.8, lwd = 1) +
  geom_line(aes(y = melt_sg * mult / max(melt_sg, na.rm = TRUE), color = "Melt_SGolay"), lwd = 0.75) +
  geom_line(aes(y = melt_fft * mult / max(melt_fft, na.rm = TRUE), color = "Melt_FFT"), lwd = 0.75) +
  theme_bw() +
  labs(x = "", y = "NDSI Melt (-) and Discharge (cfs)") +
  scale_color_colorblind(name = ""),
ncol = 1, align = "hv"
)



```

# Explore Clouds
Some of the noise in NDSI might simply be added by QA filtering that removes cloud cover bands - and other low quality data. Here, we'll take a closer look at the number of image pixels that are removed for cloud cover during the time period.

```{r}

# This time, merge the collections and select the NDSI_Snow_Cover_Class. Then, 
# clip to the AOI,and filter to the selected date range.

mod_tab <- "Snow_Albedo_Daily_Tile_Class" # 150 = cloud
cloud_val <- 150

# mod_tab <- "NDSI_Snow_Cover_Class" 
# cloud_val <- 250

cloud_col <- mod10$merge(myd10)$select(
    c(mod_tab))$map(
      function(img) {
        return(img$clip(aoi))
      }
    )$filterDate(start_date, end_date)

# Apply the mapping function to the collection
cloud_binary_col <- cloud_col$map(
  function(img) {
    # Select the classification band
    class_band <- img$select(mod_tab)
    # Use the eq() method to create the binary mask (1 where class == 250, 0 otherwise)
    binary_mask <- class_band$eq(cloud_val)
    # Rename the band for clarity and return the new image
    return(binary_mask$toInt8()$rename("Cloud_Binary"))
  }
)

# Sum all images to get the total cloud count for each grid cell
cloud_cover_days <- cloud_binary_col$sum()$toInt32()

# Also count the total number of data values for each grid cell
total_observations <- cloud_binary_col$count()$toDouble()

# 3. Calculate the Cloud Fraction (Cloud Sum / Total Observations)
cloud_fraction_image <- cloud_cover_days$divide(total_observations)$rename("Cloud_Fraction")

# Find the actual max and min pixel values within the AOI
max_value_ee <- cloud_fraction_image$reduceRegion(
  reducer = ee$Reducer$max(),
  geometry = aoi,
  scale = 500, # Use the native resolution of MODIS data
  bestEffort = TRUE
)

min_value_ee <- cloud_fraction_image$reduceRegion(
  reducer = ee$Reducer$min(),
  geometry = aoi,
  scale = 500, # Use the native resolution of MODIS data
  bestEffort = TRUE
)

max_cloud_vars <- max_value_ee$get('Cloud_Fraction')$getInfo()
min_cloud_vars <- min_value_ee$get('Cloud_Fraction')$getInfo()

# Now, plot the resulting fraction of time that each pixel was flagged as
# bad data due to cloud cover 
Map$setCenter(coords[1], coords[2], 13)
Map$addLayer(
  # 1. Limit the collection to the first 8 images
  # 2. Use sum() to combine them into a single image where pixel values are added
  eeObject = cloud_fraction_image,
  visParams = list( 
    # Note: The band name will change after summation, often defaulting to the original name
    # or the only band present. If an error occurs, try omitting 'bands' or using the default 'sum'.
    bands = c("Cloud_Fraction"), 
    min = min_cloud_vars,
    max = max_cloud_vars, 
    palette = c("white","darkblue")
  )
) +
Map$addLegend(
  visParams = list(
    min = min_cloud_vars,
    max = max_cloud_vars, 
    palette = c("white","darkblue")),
             name = "Fraction of time cloud cover"
  ) #+
#Map$addLayer(eeObject = sel_watershed,
#             visParams = list(
#               palette = c("seagreen")
#             ),
#             opacity = 0.3) 

```


Code is provided to plot elevation for the AOI. Not surprisingly, cloud cover correlates to elevation.The question is then, does this impact the resulting average melt we are calculating? For an AOI that spans a large elevation range, higher elevation cells will more often be screened out due to cloud cover. If they are tend to be screened out during melt periods, then melt estimates will be biased low,  or towards low elevation areas. For example, it might record no melt during spring if high elevation areas are cloudy and low elevation areas, which have already melted, record minimal change in snow cover. This analysis could be expanded to look at seasonal cloud cover fractions. Alternatively, elevation could be incorporated into the analysis when calculating the mean to achieve a representative sample across elevations. For now, this is likely unnecessary.
```{r elevation}

#library(elevatr)

#aoi_elev <- elevatr::get_elev_raster(aoi_sf, z = 14)
#mapview(aoi_elev) + mapview(watersheds[1,])


```

