---
title: "Accessing the NWM forecasts"
author: "ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

# Purpose 

This Rmd is meant to outline how to access and extract NWM forecasts. 

As written, this is a bit of a dead end. Unsure where historical data are stored
and there isn't a great how-to for acquiring NWM data, and the nwmTools package
is super outdated and no longer works without some finessing (which is included 
here). For now, just committing as progress, but not as functional code. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nhdplusTools)
library(sf)
# for now, don't worry about grabbing nwmTools, I've grabbed the necessary
# functions and edited them to work... kind of.
# install.packages("remotes")
# remotes::install_github('mikejohnson51/nwmTools')
# library(nwmTools)
library(tidyverse)
library(glue)
library(rvest)
library(ncdf4)
library(httr)
```

## nwmTools functions

These functions came from the nwmTools package, but most have been edited

```{r}
validate <-  function(complete, field, value){
  # check for field in source file
  if(field %in% names(complete) & !is.null(value)){
    # get all the options from the file for that field
    opts = unique(complete[[field]])
    # check to see that the value is in the list of field values
    if(any(grepl(value, opts))){
      return(filter(complete, grepl(!!value, get(field))))
    } else {
      stop(glue("{value} not a valid {field}. Choose from: {paste(opts, collapse = ', ')}"))
    }
  } else {
    return(complete)
  }
}

nwm_filter = function(source, version = NULL, config = NULL, ensemble = NULL, 
                      output= NULL, domain = NULL, date = NULL){
  
  startDate <- endDate <- NULL
  
  # Source ------------------------------------------------------------------
  
  meta = validate(nwmTools::nwm_data, "source", source) %>% 
    validate("domain", domain) %>% 
    validate("version", version) %>% 
    validate("output", output) %>% 
    validate("config", config) %>% 
    validate("ensemble", ensemble) 
  
  
  # Date ------------------------------------------------------------------ 
  if(!is.null(date)){
    
    meta = meta %>% 
      mutate(startDate = ifelse(startDate == "..", 
                                "1900-01-01",
                                startDate),
             endDate = ifelse(endDate == "..",  as.character(Sys.Date()), endDate))
    
    tmp = filter(meta, ymd(date, tz = "UTC") >= 
                   as.POSIXct(startDate, tz = "UTC"),
                 ymd(date, tz = "UTC") <= 
                   as.POSIXct(endDate, tz = "UTC"))
    
    if(nrow(tmp) == 0){ 
      stop("date not within allowed range: ", meta$startDate, "--", meta$endDate) 
    } else {
      meta = tmp 
    }
  }
  
  
  if(nrow(meta) > 1){warning('More then one viable source found...', call. = FALSE)}
  
  meta
}

#' Download Remote Files
#' @param fileList fileList object
#' @param outdir directory to write files
#' @return data.frame
#' @export

download_files = function(fileList, outdir = "."){
  
  fileList$outfiles = paste0(outdir,"/", basename(fileList$urls))
  
  for(i in 1:nrow(fileList)){
    if(!file.exists(fileList$outfiles[i])){
      httr::GET(fileList$urls[i], 
                write_disk(fileList$outfiles[i]),
                httr::progress())
    }
  }

  return(fileList)

}

#' Get NOMADs File List
#' @inheritParams get_gcp_urls
#' @param version server version (prod or para)
#' @return data.frame
#' @export

get_nomads_urls = function(config = "medium_range",
                           domain = "conus",
                           date = NULL, 
                           hour = NULL,
                           minute = "00",
                           num, 
                           ensemble = NULL, 
                           output = "channel_rt",
                           version = "prod",
                           outdir = NULL) {
  
  if (config == "medium_range_blend") version <- "v3.0" 
  
  meta = nwm_filter(source = "nomads", 
                    version = version, 
                    config = config, 
                    output = output,
                    ensemble = ensemble, 
                    domain = domain)
  
  meta = meta[1,]
  
  tmp     = glue(strsplit(meta$http_pattern, 'nwm\\.')[[1]][1])
  
  dates   = html_attr(html_elements(read_html(tmp), "a"), "href")
  dates = gsub("/", "", gsub("nwm.", "", dates))
  
  if(is.null(date)){ 
    YYYYMMDD = tail(dates,1)
    YYYYMMDD2 = tail(dates, 2)[1]
  } else if(!gsub("-", "", date) %in% dates){
    stop(date, " not avaliable")
  } else {
    YYYYMMDD = gsub("-", "", date)
    YYYYMMDD2 = gsub("-", "", date)
  }
  
  files   = tryCatch({
    date <<- YYYYMMDD
    html_attr(html_elements(read_html(glue('{tmp}/nwm.{YYYYMMDD}/{meta$bucket}/')), "a"), "href")
  }, error = function(e){
    date <<- YYYYMMDD2
    html_attr(html_elements(read_html(glue('{tmp}/nwm.{YYYYMMDD2}/{meta$bucket}/')), "a"), "href")
  })
  
  files  = grep(meta$output, files, value = TRUE)
  files  = grep(".nc$", files, value = TRUE)
  
  if(is.null(hour)){
    current = strsplit(tail(files, 1), "\\.")[[1]][[2]]
    hour = gsub("t", "", gsub("z", "", current))
  } 
  
  files  = grep(paste0("t", hour, "z"), files, value = TRUE)[1:num]
  dates  = ymd_hm(paste0(date, hour, "00")) + hours(0:(num-1))
  
  df = data.frame(dateTime = dates, 
                  urls     =  glue('{tmp}/nwm.{date}/{meta$bucket}/{files}'),
                  output   =  output)
  
  if(!is.null(outdir)){
    df = download_files(df,  outdir = outdir)
  }
  
  return(df)
  
}

```


# Extracting data

This is my version of the get_timeseries() function in the nwmTools package that
doesn't work as written. I don't think all the assumptions are right (particularly)
about valid time and the scaling factor/offset, but there's also very little 
guidance about how to use the data. 

```{r}
read_nwm_streamflow <- function(ncfile, 
                                comid, 
                                varname = "streamflow") {
  
  # Open file
  nc <- nc_open(ncfile)
  
  # Extract feature ids
  feature_ids <- ncvar_get(nc, "feature_id")
  
  # Convert COMID to numeric if needed
  comid <- as.numeric(comid)
  
  # Locate match
  idx <- which(feature_ids == comid)
  if (length(idx) == 0) {
    warning(paste("COMID", comid, "not found in feature_id"))
    return(tibble())
  }
  
  # Pull the time dimension from the global attributes
  forecast_datetime <- ncatt_get(nc, 0) %>% 
    .$model_output_valid_time %>% 
    # convert to datetime object
    ymd_hms(.)
  
  # Extract value matrix for the given varname
  value_mat <- ncvar_get(nc, varname)
  if (nrow(value_mat) < idx) {
    stop("Index outside streamflow matrix bounds")
  }
  # Get the scaling factor and offset for the var name
  var_attributes <- ncatt_get(nc, varname)
  scale_factor <- var_attributes$scale_factor
  add_offset <- var_attributes$add_offset
  # calculate the value using scaling factor and offset
  value <- (value_mat[idx]*scale_factor) + add_offset
  
  # If the variable is streamflow, there is also a "nudge" variable
  # that needs to be accounted for
  if (varname == "streamflow") {
    nudge_mat <- ncvar_get(nc, "nudge")
    nudge_attributes <- ncatt_get(nc, "nudge")
    nudge_scale <- nudge_attributes$scale_factor
    nudge_offset <- nudge_attributes$add_offset
    nudge_value <- (nudge_mat[idx]*nudge_scale) + nudge_offset
    value <- value + nudge_value
  }
  
  nc_close(nc)
  
  # Return tidy tibble
  tibble(
    datetime = forecast_datetime,
    value = value,
    variable = varname, 
    comid = comid
  )
  
}


```

# Example run

Here, we'll grab forecasts for the Cache La Poudre canyon mouth.

Short-range forecast is 18h in advance
Mid-range forecast is 10d in advance in 6h (?) increments


```{r}
# 12948 is big thompson with forecast info: https://water.noaa.gov/gauges/ESSC2
# 2900705 is `Reach ID` for canyon mouth https://water.noaa.gov/gauges/FTDC2
comid <- "2900705" 

urls <- get_nomads_urls(config = "medium_range_blend", #short_range, medium_range, medium_range_blend
                        #date = "20251119", # date of forecast
                        #ensemble = NA, # if medium_range, must supply ensemble number
                        output = "channel_rt", 
                        outdir = tempdir(),
                        num = 18)

files <- download_files(urls, outdir = tempdir())

```

And then we'll extract the data from these for our comid of choice.
```{r}
map(files$outfiles, 
    ~ read_nwm_streamflow(ncfile = .x, 
                          comid = comid, 
                          varname = "streamflow")) %>% 
  bind_rows()

```

These values are wrong (compare to those at the web address above)... but leaving for now.
