---
title: "Fort Collins Intake Data Explore"
output: html_document
date: "2026-04-29"
editor_options: 
  markdown: 
    wrap: 80
---

# Purpose

The purpose of this document is to explore the data provided to us from the City
of Fort Collins at their intake/Munroe Canal for turbidity, TOC, alkalinity.

```{r, echo = F, error = F, message = F, warning = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
library(readxl)
library(here)
library(kableExtra)
library(gghighlight)

# remotes::install_github("https://github.com/anguswg-ucsb/cdssr")
library(cdssr)
library(tidyverse)
```

# Total Organic Carbon

## TOC from the FC intake

We have intake TOC data since 2019.

```{r}

toc <- read_excel(here("data/raw/fc_intake_data/Intake TOC.xlsx")) %>% 
  mutate(date = date(`Sampled Date`)) %>% 
  select(date, parameter = `Parameter Name`, result = `Corrected Result`) %>% 
  mutate(doy = yday(date),
         year = year(date))

ggplot(toc, aes(x = doy, y = result, color = factor(year))) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

# limited to days 100-200
ggplot(toc, aes(x = doy, y = result, color = factor(year))) +
  coord_cartesian(xlim = c(100, 200)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

```

## Raw Water TOC

This is data from the raw water station INSIDE the FC treatment facility, but
the data record starts in 2010.

```{r}
rw_toc <- read_csv(here("data/raw/fc_intake_data/Poudre_Raw_Water_Sample_Station_TOC_2010-2025.csv")) %>% 
  mutate(date = mdy(`Sampled Date`)) %>% 
  select(date, TOC = `Corrected Result`) %>% 
  filter(!is.na(TOC) & TOC > 1) %>%  # one oddball here.
  mutate(doy = yday(date),
         year = year(date))

ggplot(rw_toc, aes(x = doy, y = TOC, color = factor(year))) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

```

Interesting to note that the 2013 flood is very evident in these data - see the
spike in late season TOC.

Let's do a quick check of the data at intake and the data at the RW station to
see how interchangable the values are.

```{r}
full_join(toc, rw_toc) %>% 
  ggplot(aes(x = TOC, y = result)) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey60") +
  geom_point() +
  labs(x = "TOC at raw water sink", y = "TOC at intake") +
  theme_bw()

```

These are pretty darn close to interchangable, so if there are data available at
the intake and not at the RW sink, let's use it, and generally take an average
if there are multiple values available.

## Collated TOC

```{r}
fc_TOC_record <- full_join(toc, rw_toc) %>% 
  pivot_longer(cols = c(TOC, result),
               names_to = "source",
               values_to = "TOC") %>% 
  filter(!is.na(TOC)) %>% 
  summarise(mean_TOC = mean(TOC), 
            .by = c(year, doy, date))

ggplot(fc_TOC_record, aes(x = doy, y = mean_TOC, color = factor(year))) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(fc_TOC_record$date, na.rm = T),
                       end_date = max(fc_TOC_record$date, na.rm = T),
                       timescale = "day")%>%
  mutate(year = factor(year(datetime)), 
         doy = yday(datetime))

# faceted
ggplot() +
  geom_area(data = flow_data,
            aes(x = doy, y = value/200), # divide flow by a silly amount for visualization
            fill = "gray70",
            alpha = 0.5) +
  geom_point(data = fc_TOC_record, 
             aes(x = doy, y = mean_TOC, color = factor(year))) +
  facet_wrap(year ~ ., scale = "free_y") +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

# write_csv(fc_TOC_record, here("data/collated/FC_intake/fc_intake_TOC_record_2010-2025.csv"))
```

# Alkalinity

## Alkalinity from intake

Unlinke to TOC, alkalinity at the intake is available since 2012.

```{r}
alk <- read_csv(here("data/raw/fc_intake_data/Poudre_Intake_Alkalinity.csv")) %>% 
  mutate(date = mdy(`Sampled Date`)) %>% 
  select(date, parameter = `Parameter Name`, result = `Corrected Result`) %>% 
  mutate(doy = yday(date),
         year = year(date))

ggplot(alk, aes(x = doy, y = result, color = factor(year))) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_d() +
  geom_abline(slope = 0, intercept = 60, lty = 2, color = "orange") +
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

# limit to doy 100-200
ggplot(alk, aes(x = doy, y = result, color = factor(year))) +
  coord_cartesian(xlim = c(100, 200)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_d()+
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(alk$date, na.rm = T),
                       end_date = max(alk$date, na.rm = T),
                       timescale = "day")%>%
  mutate(year = factor(year(datetime)), 
         doy = yday(datetime))

# facet
ggplot() +
  geom_area(data = flow_data,
            aes(x = doy, y = value/50),
            fill = "gray70",
            alpha = 0.5) +
  geom_point(data = alk, aes(x = doy, y = result, color = factor(year))) +
  facet_wrap(year ~ ., scales = "free_y") +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

```

Alkalinity data at the intake is mostly Apr-Jun, and during this time there has
never been a sample that is higher than the treatment matrix cutoff (60 mg/L),
the only thing that we could do here is try to relate this to some other
variable to give a warning that it might be a high alkalinity year in the
spring.

In the data that Sam has explored and anecdotally, alkalinity is more of an 
issue for Greeley whose intake is further downstream.

## Raw Water Alkalinity

We also have the record of Alkalinity at the RW sink.

```{r}
rw_alk <- read_csv(here("data/raw/fc_intake_data/Poudre_Raw_Water_Sample_Station_Alkalinity.csv")) %>% 
  mutate(date = mdy(`Sampled Date`)) %>% 
  select(date, alkalinity = `Corrected Result`) %>% 
  #filter(!is.na(TOC) & TOC > 1) %>%  # one oddball here.
  mutate(doy = yday(date),
         year = year(date))

ggplot(rw_alk, aes(x = doy, y = alkalinity, color = factor(year))) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

```

We still don't see anything higher than the 60 mg/L, but helpful to see the
pattern.

Let's do a quick check of the data at intake and the data at the RW station to
see how interchangable the values are.

```{r}
full_join(alk, rw_alk) %>% 
  ggplot(aes(x = alkalinity, y = result)) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey60") +
  geom_point() +
  labs(x = "Alkalinity at raw water sink", y = "Alkalinity at intake") +
  theme_bw() +
  coord_cartesian(xlim = c(5, 50), ylim = c(5,50))

```

Also right on like TOC. We'll again take an average if there are multiple values
available for a given date.

## Collated Alkalinity

```{r}
fc_alk_record <- full_join(alk, rw_alk) %>% 
  pivot_longer(cols = c(alkalinity, result),
               names_to = "source",
               values_to = "alkalinity") %>% 
  filter(!is.na(alkalinity)) %>% 
  summarise(mean_alk = mean(alkalinity), 
            .by = c(year, doy, date))

ggplot(fc_alk_record, aes(x = doy, y = mean_alk, color = factor(year))) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "Alkalinity (mg/L CaCO3)", color = "year")

# write_csv(fc_alk_record, here("data/collated/FC_intake/fc_intake_alk_record_2010-2025.csv"))
```

Those peaks in 2012 data are very likely real (after High Park Fire).

# Turbidity

We have 15-minute data from the City at both the Intake and Munroe Canal. These
data need a little QAQC love, so we'll take care of that. For Munroe Canal - not
sure if this is in the canal or in the river at the canal. This is City of FC
alternate Poudre source if the intake proper had an issue.

## Collation and QAQC

```{r}
intake_turb <- read_csv(here("data/raw/fc_intake_data/Intake_Turbidity_2010-2025.csv"),
                        col_names = c("datetime", "intake_turb_NTU"),
                        skip = 1,
                        col_select = c("datetime", "intake_turb_NTU"))

munroe_turb <- read_csv(here("data/raw/fc_intake_data/Munroe_Turbidity_2013-2025.csv"),,
                        col_names = c("datetime", "munroe_turb_NTU"),
                        skip = 1,
                        col_select = c("datetime", "munroe_turb_NTU")) 

fc_turb <- full_join(intake_turb, munroe_turb) %>% 
  pivot_longer(cols = c(munroe_turb_NTU, intake_turb_NTU),
               names_to = "location",
               values_to = "NTU") %>% 
  mutate(NTU = case_when(NTU <= 0 ~ NA, 
                         NTU >= 1000 ~ NA,
                         round(NTU, 0) == 655 & location == "munroe_turb_NTU" ~ NA,
                         round(NTU, 0) == 630 & location == "munroe_turb_NTU" ~ NA,
                         round(NTU, 0) == 555 & location == "munroe_turb_NTU" ~ NA,
                         .default = NTU))

```

This still needs a little massaging, so let's look year-by-year.

### First Pass QAQC

-   2010 some low-NTU noise during spring runnoff, noise october to end of year

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2010-01-01"), 
                                  ymd("2011-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2010-04-01"), 
                                  ymd("2010-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2010-05-05"),
                               ymd("2010-07-01")) &
                         NTU < 3, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2010-04-01"), 
                                  ymd("2010-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2010-10-01"), 
                                  ymd("2011-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2010-10-01"), 
                                  ymd("2010-11-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2010-10-01"),
                                 ymd("2010-10-06")) &
                           NTU > 3.5 ~ NA,
                         between(datetime, 
                                 ymd("2010-10-06"),
                                 ymd("2010-10-13")) &
                           NTU > 3 ~ NA,
                         between(datetime, 
                                 ymd("2010-10-13"),
                                 ymd("2010-10-14")) &
                           NTU > 5 ~ NA,
                         between(datetime, 
                                 ymd("2010-10-14"),
                                 ymd("2010-11-01")) &
                           NTU > 3 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2010-10-01"), 
                                  ymd("2010-11-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2010-11-01"), 
                                  ymd("2010-12-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2010-11-02"),
                                 ymd("2010-12-01")) &
                           NTU > 2 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2010-11-01"), 
                                  ymd("2010-12-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2010-12-01"), 
                                  ymd("2011-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2010-12-01"),
                                 ymd("2011-01-01")) &
                           NTU > 10 ~ NA,
                         between(datetime, 
                                 ymd("2010-12-12"),
                                 ymd("2010-12-19")) &
                           NTU > 2 ~ NA,
                         between(datetime, 
                                 ymd("2010-12-20"),
                                 ymd("2011-01-01")) &
                           NTU > 2 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2010-10-01"), 
                                  ymd("2011-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


```

</details>

-   2011 early season noise, noise during runoff, a spike in Sept that may not be real

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-01-01"), 
                                  ymd("2012-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-01-01"), 
                                  ymd("2011-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2011-01-01"),
                                 ymd("2011-04-01")) &
                           NTU > 3 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-01-01"), 
                                  ymd("2011-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-04-01"), 
                                  ymd("2011-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2011-04-01"),
                                 ymd("2011-05-01")) &
                           NTU > 3 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-05-01"), 
                                  ymd("2011-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2011-05-01"),
                                 ymd("2011-05-09")) &
                           NTU > 4 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-06-01"), 
                                  ymd("2011-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2011-05-30"),
                                 ymd("2011-08-01")) &
                           NTU < 3 ~ NA,
                         between(datetime, 
                                 ymd("2011-06-05"),
                                 ymd("2011-06-25")) &
                           NTU > 50 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-04-01"), 
                                  ymd("2011-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-07-01"), 
                                  ymd("2011-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-08-15"), 
                                  ymd("2011-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2011-09-01"),
                               ymd("2011-09-06")),
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-07-01"), 
                                  ymd("2011-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2011-10-01"), 
                                  ymd("2012-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

```

There is still some noise late sept-october, but that requires a bit more precision
removal than we are going to apply right now.

</details>

-   2012 has a pretty significant data gap after high turbidity, might want to
leave out.

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-01-01"), 
                                  ymd("2013-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-01-01"), 
                                  ymd("2012-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-01-01"), 
                                  ymd("2012-02-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2012-01-05"),
                               ymd("2012-01-14")),
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-01-01"), 
                                  ymd("2012-02-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-02-01"), 
                                  ymd("2012-03-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2012-02-06"),
                                 ymd("2012-02-11")) ~ NA, 
                         between(datetime, 
                                 ymd("2012-02-20"),
                                 ymd("2012-04-01")) & 
                           NTU > 3 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-02-01"), 
                                  ymd("2012-03-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-03-01"), 
                                  ymd("2012-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-04-01"), 
                                  ymd("2012-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-04-01"), 
                                  ymd("2012-05-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2012-04-01"),
                                 ymd("2012-04-22"))  &
                           NTU > 10 ~ NA, 
                         between(datetime, 
                                 ymd("2012-04-24"),
                                 ymd("2012-05-01")) & 
                           NTU > 10 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-04-01"), 
                                  ymd("2012-05-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-05-01"), 
                                  ymd("2012-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2012-05-21"),
                                 ymd("2012-05-24")) ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-05-01"), 
                                  ymd("2012-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-06-01"), 
                                  ymd("2012-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2012-06-01"),
                                 ymd("2012-07-01")) &
                           NTU > 12.5 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-04-01"), 
                                  ymd("2012-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-07-01"), 
                                  ymd("2012-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2012-07-01"),
                                 ymd("2012-09-18")) ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-07-01"), 
                                  ymd("2012-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-09-24"), 
                                  ymd("2012-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2012-09-24"), 
                                 ymd("2012-10-01"))  &
                           NTU >= 100 ~ NA,
                         .default = NTU)) 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-07-01"), 
                                  ymd("2012-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2012-10-01"), 
                                  ymd("2013-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

```

</details>

-   Prior to 2013, data was generally no higher than 100 NTU, starting in 2013
the range increases to 1000. Maybe leave out data prior to 2013? That's also
post-High Park Fire, so maybe it's real? It could also just be a sensor
change?

-   intermittent data looks suspect starting mid june

<details>
<summary>Click to expand plots</summary>
```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-01-01"), 
                                  ymd("2014-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-01-01"), 
                                  ymd("2013-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-01-01"), 
                                  ymd("2013-02-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2013-01-11"), 
                               ymd("2013-01-17")), 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-01-01"), 
                                  ymd("2013-02-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-02-01"), 
                                  ymd("2013-03-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2013-02-11"), 
                                 ymd("2013-02-12")) ~ NA,
                         between(datetime, 
                                 ymd("2013-02-01"), 
                                 ymd("2013-03-01")) &
                           NTU > 50 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-02-01"), 
                                  ymd("2013-03-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-03-01"), 
                                  ymd("2013-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2013-03-15"), 
                               ymd("2013-04-01")) &
                         NTU < 1, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-03-01"), 
                                  ymd("2013-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-04-01"), 
                                  ymd("2013-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-04-01"), 
                                  ymd("2013-05-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2013-04-26"), 
                               ymd("2013-05-01")) &
                         (NTU < 3 | NTU > 25), 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-04-01"), 
                                  ymd("2013-05-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-06-01"), 
                                  ymd("2013-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-08-01"), 
                                  ymd("2013-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-07-01"), 
                                  ymd("2013-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-09-01"), 
                                  ymd("2013-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-10-01"), 
                                  ymd("2014-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2013-10-01"), 
                                  ymd("2014-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

# honestly, all this data looks terrible and noisy
fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2013-06-01"), 
                               ymd("2014-01-01")),
                       NA, 
                       NTU))

```

</details>

-   2014 munroe Apr data should be recoded, some near-zero values during runoff to be recoded, look closer at some of the late-season peaks in intake, especially compared to munroe

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-04-01"), 
                                  ymd("2014-05-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ .) +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2014-04-07"), 
                               ymd("2014-04-15")) &
                         location == "munroe_turb_NTU", 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-04-21"), 
                                  ymd("2014-05-01")) &
                            location == "intake_turb_NTU"), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ .) +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2014-04-21"), 
                               ymd("2014-05-01")) &
                         location == "intake_turb_NTU" &
                         NTU > 100, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-04-01"), 
                                  ymd("2014-05-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ .) +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-05-01"), 
                                  ymd("2014-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ .) +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2014-05-01"), 
                                 ymd("2014-06-01")) &
                           (NTU < 5 | NTU > 500) ~ NA,
                         between(datetime,
                                 ymd("2014-05-01"), 
                                 ymd("2014-05-05")) &
                           NTU > 50 ~ NA,
                         location == "munroe_turb_NTU" &
                           between(datetime,
                                   ymd("2014-05-18"),
                                   ymd("2014-05-20")) ~ NA,
                         location == "munroe_turb_NTU" &
                           between(datetime,
                                   ymd("2014-05-25"),
                                   ymd("2014-05-27")) ~ NA,
                         .default = NTU)
  )

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-05-01"), 
                                  ymd("2014-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-04-01"), 
                                  ymd("2014-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-06-01"), 
                                  ymd("2014-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(location == "intake_turb_NTU" &
                           between(datetime,
                                   ymd("2014-06-14"),
                                   ymd("2014-06-16")) ~ NA,
                         location == "intake_turb_NTU" &
                           between(datetime,
                                   ymd("2014-06-17"),
                                   ymd("2014-06-20")) ~ NA,
                         .default = NTU)
  )

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-06-01"), 
                                  ymd("2014-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-07-01"), 
                                  ymd("2014-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-07-01"), 
                                  ymd("2014-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(location == "intake_turb_NTU" &
                           between(datetime,
                                   ymd("2014-07-11"),
                                   ymd("2014-07-18")) ~ NA,
                         .default = NTU)
  )

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-07-01"), 
                                  ymd("2014-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-07-15"), 
                                  ymd("2014-08-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-07-15"), 
                                  ymd("2014-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-08-15"), 
                                  ymd("2014-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-09-15"), 
                                  ymd("2014-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2014-08-26"), 
                                 ymd("2014-08-27")) &
                           NTU > 10 ~ NA,
                         between(datetime,
                                 ymd("2014-09-30"), 
                                 ymd("2014-10-03")) &
                           NTU > 10 ~ NA,
                         
                         .default = NTU),
  )

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-07-01"), 
                                  ymd("2014-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2014-10-01"), 
                                  ymd("2015-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ ., scales = "free_y") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

```


-   2015: near-zero noise during snowmelt, Sept 2015-Jun 2016 munroe; is munroe an order of magnitude off?

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-01-01"), 
                                  ymd("2016-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  facet_grid(location ~ .) +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

# munroe is wrong all year
fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime,
                               ymd("2015-01-01"), 
                               ymd("2016-01-01")) &
                         location == "munroe_turb_NTU", 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-01-01"), 
                                  ymd("2016-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-02-01"), 
                                  ymd("2015-03-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime,
                               ymd("2015-02-21"), 
                               ymd("2015-02-24")) &
                         location == "intake_turb_NTU", 
                       NA, 
                       NTU))
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-02-01"), 
                                  ymd("2015-03-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-03-15"), 
                                  ymd("2015-06-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime,
                               ymd("2015-03-15"),
                               ymd("2015-06-15")) &
                         NTU < 2, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-01-01"), 
                                  ymd("2015-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-07-01"), 
                                  ymd("2015-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-07-01"), 
                                  ymd("2015-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime,
                               ymd("2015-07-21"), 
                               ymd("2015-07-22")) &
                         location == "intake_turb_NTU", 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-07-01"), 
                                  ymd("2015-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-08-15"), 
                                  ymd("2015-10-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2015-08-15"),
                                 ymd("2015-10-15")) &
                           location == "intake_turb_NTU" & 
                           NTU > 40 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-07-01"), 
                                  ymd("2015-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-10-01"), 
                                  ymd("2016-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

```

</details>

-   2016: errant data ~ 325 at munroe, near zero throughout snowmelt, Munroe Oct 2016-end of year

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-01-01"), 
                                  ymd("2017-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-01-01"), 
                                  ymd("2016-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-01-01"), 
                                  ymd("2016-02-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2016-01-01"),
                                 ymd("2016-02-04")) ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-01-01"), 
                                  ymd("2016-02-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-02-01"), 
                                  ymd("2016-03-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2016-02-15"),
                                 ymd("2016-03-01")) &
                           NTU > 10 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-02-01"), 
                                  ymd("2016-03-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-03-01"), 
                                  ymd("2016-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-04-01"), 
                                  ymd("2016-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-04-01"), 
                                  ymd("2016-05-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2016-01-01"),
                                 ymd("2016-04-02")) &
                           location == "munroe_turb_NTU" ~ NA,
                         between(datetime, 
                                 ymd("2016-04-25"),
                                 ymd("2016-04-28")) &
                           location == "munroe_turb_NTU" ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-04-01"), 
                                  ymd("2016-05-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-05-01"), 
                                  ymd("2016-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2016-05-01"), 
                               ymd("2016-06-01")) &
                         location == "munroe_turb_NTU" &
                         NTU > 300, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-05-01"), 
                                  ymd("2016-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2016-05-08"), 
                                 ymd("2016-05-27")) &
                           location == "munroe_turb_NTU" ~ NA,
                         between(datetime, 
                                 ymd("2016-05-01"), 
                                 ymd("2016-07-01")) &
                           location == "intake_turb_NTU" &
                           (NTU < 2 | NTU > 150) ~ NA,
                         between(datetime, 
                                 ymd("2016-05-10"), 
                                 ymd("2016-05-11")) &
                           location == "intake_turb_NTU" ~ NA,
                         between(datetime, 
                                 ymd("2016-05-17"), 
                                 ymd("2016-05-18")) &
                           location == "intake_turb_NTU" ~ NA,
                         between(datetime, 
                                 ymd("2016-05-24"), 
                                 ymd("2016-05-26")) &
                           location == "intake_turb_NTU" ~ NA,
                         between(datetime, 
                                 ymd("2016-06-12"), 
                                 ymd("2016-06-24")) &
                           location == "intake_turb_NTU" ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-04-01"), 
                                  ymd("2016-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-07-01"), 
                                  ymd("2016-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-07-01"), 
                                  ymd("2016-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-07-01"), 
                                  ymd("2016-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-08-01"), 
                                  ymd("2016-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-09-01"), 
                                  ymd("2016-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-10-01"), 
                                  ymd("2017-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


# these all look pretty sus - they are low NTU, so unhelpful for model creation,
# therefore, removal is okay from a modeling standpoint.

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2016-09-01"), 
                                 ymd("2017-01-01")) ~ NA,
                         .default = NTU))




```

</details>

-   2017: end of year at intake

<details>
<summary>Click to expand plots</summary>

```{r}

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2017-01-01"), 
                                  ymd("2018-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2017-12-01"),
                               ymd("2018-01-01")) &
                         location == "intake_turb_NTU" , 
                       NA, 
                       NTU))


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2017-01-01"), 
                                  ymd("2018-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2017-03-01"),
                                 ymd("2017-04-01")) &
                           location == "intake_turb_NTU" &
                           NTU > 15 ~ NA,
                         between(datetime, 
                                 ymd("2017-01-01"),
                                 ymd("2017-04-03")) &
                           location == "munroe_turb_NTU" ~ NA,
                         .default = NTU))

```

</details>

-   2018: Nov 2017-Apr 2018 intake, snowmelt near-zero

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2018-01-01"), 
                                  ymd("2019-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2018-01-01"),
                               ymd("2018-04-01")) &
                         location == "intake_turb_NTU" , 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2018-01-01"), 
                                  ymd("2019-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2018-04-01"), 
                                  ymd("2018-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2018-04-01"),
                                 ymd("2018-04-08")) &
                           location == "munroe_turb_NTU" ~ NA, 
                         between(datetime, 
                                 ymd("2018-05-04"),
                                 ymd("2018-05-08")) &
                           location == "munroe_turb_NTU" ~ NA, 
                         between(datetime, 
                                 ymd("2018-05-01"),
                                 ymd("2018-05-15")) &
                           location == "munroe_turb_NTU" & 
                           NTU > 50 ~ NA, 
                         between(datetime, 
                                 ymd("2018-05-25"),
                                 ymd("2018-05-30")) &
                           location == "munroe_turb_NTU" & 
                           NTU > 50 ~ NA, 
                         between(datetime, 
                                 ymd("2018-06-01"),
                                 ymd("2018-06-05")) &
                           NTU < 1 ~ NA, 
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2018-04-01"), 
                                  ymd("2018-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2018-07-01"), 
                                  ymd("2018-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2018-07-10"),
                                 ymd("2018-07-16")) &
                           location == "munroe_turb_NTU" ~ NA, 
                         between(datetime, 
                                 ymd("2018-07-01"),
                                 ymd("2018-07-05")) &
                           location == "munroe_turb_NTU" & 
                           NTU > 150 ~ NA, 
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2018-07-01"), 
                                  ymd("2018-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2018-10-01"), 
                                  ymd("2019-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


```

</details>

-   2019: runoff noise

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2019-01-01"), 
                                  ymd("2020-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2019-05-13"), 
                                  ymd("2019-05-20"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2019-05-15"), 
                               ymd("2019-05-18")) &
                         location == "intake_turb_NTU" , 
                       NA, 
                       NTU))


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2019-06-01"), 
                                  ymd("2019-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2019-06-08"), 
                               ymd("2019-06-11")) &
                         location == "munroe_turb_NTU" , 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2019-06-20"), 
                                  ymd("2019-06-24"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2019-06-21"), 
                               ymd("2019-06-23")) &
                         location == "intake_turb_NTU" , 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2019-04-01"), 
                                  ymd("2019-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2019-04-01"), 
                               ymd("2019-07-01")) &
                         location == "munroe_turb_NTU" &
                         NTU > 300, 
                       NA, 
                       NTU))

```

</details>


-   2020: munroe noise at start of record, late season high values at intake

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2020-01-01"), 
                                  ymd("2021-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2020-04-01"), 
                                  ymd("2020-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2020-05-01"), 
                                 ymd("2020-06-01")) &
                           NTU < 3 ~ NA, 
                         between(datetime, 
                                 ymd("2020-06-01"), 
                                 ymd("2020-07-01")) &
                           NTU < 1.5 ~ NA, 
                         
                         .default = NTU))


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2020-06-15"), 
                                  ymd("2020-06-20"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2020-06-16"), 
                               ymd("2020-06-18")) &
                         location == "munroe_turb_NTU", 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2020-10-01"), 
                                  ymd("2021-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2020-10-10"), 
                                  ymd("2020-10-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2020-10-12"), 
                                 ymd("2021-01-01")) &
                           location == "munroe_turb_NTU" ~ NA,
                         between(datetime, 
                                 ymd("2020-10-13"), 
                                 ymd("2020-10-14")) &
                           location == "intake_turb_NTU" ~ NA,
                         .default = NTU))

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2020-01-01"), 
                               ymd("2021-01-10")) &
                         location == "intake_turb_NTU" &
                         NTU > 500, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2020-10-01"), 
                                  ymd("2021-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


```

</details>

-   2021: 1000 values, munroe sensor failure in late summer, near-zero during runoff, back online after sensor offline intake

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2021-01-01"), 
                                  ymd("2022-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2021-01-01"), 
                               ymd("2022-01-01")) &
                         NTU > 990, NA, NTU))


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2021-08-19"), 
                                  ymd("2021-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2021-07-19"), 
                               ymd("2021-08-01")) &
                         location == "munroe_turb_NTU", 
                       NA, 
                       NTU))


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2021-04-01"), 
                                  ymd("2021-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2021-05-01"), 
                               ymd("2021-07-01")) &
                         location == "intake_turb_NTU" &
                         NTU < 3, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2021-08-09"), 
                                  ymd("2021-08-16"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2021-08-10"), 
                               ymd("2021-08-14")) &
                         location == "intake_turb_NTU",        
                       NA, 
                       NTU))


```

</details>


-   2022: 

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2022-01-01"), 
                                  ymd("2023-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2022-01-01"), 
                                 ymd("2023-01-01")) &
                           NTU > 990 ~ NA,
                         between(datetime, 
                                 ymd("2022-01-01"), 
                                 ymd("2022-05-01")) &
                           NTU > 200 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2022-01-01"), 
                                  ymd("2023-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2022-05-01"), 
                                  ymd("2022-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2022-05-01"), 
                               ymd("2022-06-01")) &
                         NTU > 200,
                       NA,
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2022-05-01"), 
                                  ymd("2022-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2022-06-01"), 
                                  ymd("2022-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2022-06-01"), 
                                 ymd("2022-07-01")) &
                           NTU > 300 ~ NA,
                         between(datetime, 
                                 ymd("2022-06-01"), 
                                 ymd("2022-08-01")) &
                           NTU < 2  ~ NA,
                         between(datetime, 
                                 ymd("2022-07-15"), 
                                 ymd("2022-08-01")) &
                           NTU > 200 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2022-06-01"), 
                                  ymd("2022-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2022-10-01"), 
                                  ymd("2022-11-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2022-10-27"), 
                               ymd("2023-01-01")) &
                         location == "munroe_turb_NTU",
                       NA,
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2022-10-01"), 
                                  ymd("2022-11-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")
```

</details>

-   2023

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-01-01"), 
                                  ymd("2024-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-01-01"), 
                                  ymd("2023-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2023-01-01"), 
                               ymd("2023-04-01")) &
                         NTU > 100,
                       NA,
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-01-01"), 
                                  ymd("2023-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-04-01"), 
                                  ymd("2023-05-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2023-04-01"), 
                                 ymd("2023-04-03")) &
                           NTU > 40 ~ NA,
                         between(datetime, 
                                 ymd("2023-04-01"), 
                                 ymd("2023-04-08")) &
                           location == "munroe_turb_NTU" ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-04-01"), 
                                  ymd("2023-05-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-05-01"), 
                                  ymd("2023-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2023-05-01"), 
                                 ymd("2023-06-01")) &
                           NTU > 990 ~ NA,
                         between(datetime, 
                                 ymd("2023-05-01"), 
                                 ymd("2023-06-01")) &
                           NTU < 2 ~ NA,
                         between(datetime, 
                                 ymd("2023-05-14"), 
                                 ymd("2023-05-18")) &
                           location == "munroe_turb_NTU" ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-05-01"), 
                                  ymd("2023-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-06-01"), 
                                  ymd("2023-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2023-06-01"), 
                                 ymd("2023-07-01")) &
                           NTU < 2 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-06-01"), 
                                  ymd("2023-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,200))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-07-01"), 
                                  ymd("2023-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2023-07-01"), 
                                 ymd("2023-08-01")) &
                           NTU > 990 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-07-01"), 
                                  ymd("2023-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2023-07-01"), 
                                 ymd("2023-07-19")) &
                           location == "intake_turb_NTU" ~ NA,
                         .default = NTU))


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-07-01"), 
                                  ymd("2023-08-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") +
  coord_cartesian(ylim = c(0,100))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-08-01"), 
                                  ymd("2023-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2023-08-01"), 
                                 ymd("2023-09-01")) &
                           NTU < 2 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-08-01"), 
                                  ymd("2023-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-12-01"), 
                                  ymd("2024-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

```

</details>

-   2024

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-01-01"), 
                                  ymd("2025-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-01-01"), 
                                  ymd("2024-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2024-01-01"), 
                               ymd("2024-04-01")) &
                         NTU > 100,
                       NA,
                       NTU))


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-01-01"), 
                                  ymd("2024-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-04-01"), 
                                  ymd("2024-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2024-04-01"), 
                                 ymd("2024-07-01")) &
                           NTU > 500 ~ NA,
                         between(datetime, 
                                 ymd("2024-04-01"), 
                                 ymd("2024-07-01")) &
                           NTU < 2 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-04-01"), 
                                  ymd("2024-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-07-01"), 
                                  ymd("2024-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2024-07-01"), 
                                 ymd("2024-10-01")) &
                           NTU > 500 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-07-01"), 
                                  ymd("2024-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-08-01"), 
                                  ymd("2024-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2024-08-21"), 
                                 ymd("2024-08-26")) &
                           location == "munroe_turb_NTU" ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-08-01"), 
                                  ymd("2024-09-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 


ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-09-01"), 
                                  ymd("2024-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2024-09-24"), 
                                 ymd("2024-09-30")) &
                           location == "munroe_turb_NTU" ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-09-01"), 
                                  ymd("2024-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

```

</details>

- 2025

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-01-01"), 
                                  ymd("2026-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-01-01"), 
                                  ymd("2025-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, 
                               ymd("2025-01-01"), 
                               ymd("2025-04-01")) &
                         NTU > 60,
                       NA,
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-01-01"), 
                                  ymd("2025-04-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-04-01"), 
                                  ymd("2025-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2025-04-01"), 
                                 ymd("2025-04-04")) &
                           location == "munroe_turb_NTU" ~ NA,
                         between(datetime, 
                                 ymd("2025-04-15"), 
                                 ymd("2025-07-01")) &
                           NTU < 2 ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-04-01"), 
                                  ymd("2025-07-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-07-01"), 
                                  ymd("2025-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-09-01"), 
                                  ymd("2025-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2025-09-25"), 
                                 ymd("2025-09-29")) &
                           location == "munroe_turb_NTU" ~ NA,
                         .default = NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-07-01"), 
                                  ymd("2025-10-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-10-01"), 
                                  ymd("2026-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-10-01"), 
                                  ymd("2025-11-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

fc_turb <- fc_turb %>% 
  mutate(NTU = case_when(between(datetime, 
                                 ymd("2025-10-06"), 
                                 ymd("2026-01-01")) &
                           location == "munroe_turb_NTU" ~ NA,
                         .default = NTU))
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2025-10-01"), 
                                  ymd("2026-01-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom") 

```

</details>

### Final check

Plot to re-check cleaning year-by-year

<details open>
<summary>Click to collapse year-by-year </summary> 

```{r}
flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(fc_turb$datetime, na.rm = T),
                       end_date = max(fc_turb$datetime, na.rm = T),
                       timescale = "day")%>%
  mutate(year = factor(year(datetime)), 
         doy = yday(datetime))

# plot year-by-year for outliers and errors
map(2010:2025, \(y) {
  start_date <- paste0(y, "-01-01")
  end_date <- paste0(y + 1, "-01-01")
  ggplot() +
    geom_area(data = flow_data %>% 
                filter(between(datetime, 
                               ymd(start_date), 
                               ymd(end_date))),
              aes(x = datetime, y = value/10),
              fill = "blue",
              alpha = 0.3) +
    geom_point(data = fc_turb %>% filter(between(datetime, 
                                                 ymd(start_date), 
                                                 ymd(end_date))), 
               aes(x = datetime, y = NTU, color = location)) +
    labs(title = y,
         y = "turbidity (NTU)")  +
    scale_color_viridis_d() +
    theme_bw() +
    theme(legend.position = "bottom")
})

```

</details>

Okay, that's a lot better, still some noise, but generally okay. 

### Daily Quantiles

For the purpose of visualization and a dataset for training and validation, we'll 
get some quantiles (95%, 75%, 50%, 25% and 5%) and plot in ribbons, with 
Q from the canyon mouth. These quantiles assume that min/max might be a bit misleading
due to the noise in the data.


```{r}
turb_q <- fc_turb %>% 
  mutate(date = date(datetime),
         doy = yday(date),
         year = year(date)) %>%
  filter(!is.na(NTU)) %>% 
  summarize(p05 = quantile(NTU, 0.05),
            p25 = quantile(NTU, 0.25),
            p50 = quantile(NTU, 0.50),
            p75 = quantile(NTU, 0.75),
            p95 = quantile(NTU, 0.95),
            n = n(),
            .by = c(year, doy, date, location))

# data that have the same quantile values on a given day or that have low n 
# should be excluded here
turb_q_filt <- turb_q %>% 
  # we want to make sure we have at least 75% of the days values
  filter(n > 96*0.75) %>% 
  # and that all the values are not the same (indicating something is wrong with the sensor)
  filter(p05 != p95) %>% 
  filter(p25 != p75)

date_sequence <- seq.Date(min(turb_q$date), max(turb_q$date)) %>% 
  tibble(date = .) %>% 
  mutate(doy = yday(date),
         year = year(date)) %>% 
  cross_join(., tibble(location = c("munroe_turb_NTU", "intake_turb_NTU")))

turb_q_filt <- left_join(date_sequence, turb_q_filt)

```


```{r, fig.height=12, fig.width=8}
# with Q
ggplot(turb_q_filt, aes(x = doy)) +
  geom_area(data = flow_data,
            aes(x = doy, y = value/30),
            fill = "lightblue",
            alpha = 0.5) +
  # 5–95% band
  geom_ribbon(aes(ymin = p05, ymax = p95, fill = "5 - 95%")) +
  # 25–75% band
  geom_ribbon(aes(ymin = p25, ymax = p75, fill = "25 - 75%")) +
  # Median line
  geom_line(aes(y = p50, color = "median"), color = "black") +
  # 100 NTU threshold
  geom_abline(aes(color = "100 NTU threshold"), color = "orange", 
              slope = 0, intercept = 100, lty = 2) +
  facet_grid(year ~ location) +
  coord_cartesian(xlim = c(75, 275),
                  ylim = c(0, 150)) +
  labs(y = "turbidity (NTU)",
       x = "Day of Year",
       title = "Daily Turbidity Quantile Bands",
       fill = "Quantile Bands") +
  scale_fill_manual(values = c("5 - 95%" = "grey85",
                               "25 - 75%" = "grey60"),
                    breaks = c("5 - 95%", "25 - 75%")) +
  theme_bw() +
  theme(legend.position = "bottom")

```

For kicks, let's look at the relationship

```{r}
turb_q_filt %>% 
  select(date, year, p50, location) %>% 
  filter(year != 2015) %>% 
  pivot_wider(names_from = location,
              values_from = p50) %>% 
  ggplot(., aes(x = intake_turb_NTU, y = munroe_turb_NTU, color = factor(year))) +
  geom_point() +
  theme_bw() +
  labs(x = "median daily turbidity at intake (NTU)",
       y = "median daily turbidity at Munroe (NTU)",
       color = "year") +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey60") +
  scale_color_viridis_d() +
  coord_cartesian(xlim = c(0, 120),
                  ylim = c(0, 120))

```

Still some outliers, but overall, this looks like we've done sufficient QAQC to
move on. For kicks, let's just flag any data where there is greater than a 100%
difference between Munroe and the Intake. These data might be real, but they
also might not and we may wish to keep them out of the training dataset.

We'll do this using symmetric differencing at the median value.

```{r}
flagged <- turb_q_filt %>%
  pivot_wider(id_cols = c(date, doy, year),
              names_from = location,
              values_from = p50) %>%
  mutate(pct_diff = abs(munroe_turb_NTU - intake_turb_NTU) /
           ((munroe_turb_NTU + intake_turb_NTU) / 2),
         flag_diff = if_else(pct_diff > 1,
                             "suspect",
                             NA)) %>% 
  select(date, pct_diff, flag_diff) 


turb_q_flagged <- full_join(turb_q_filt, flagged)

turb_q_flagged %>% 
  select(date, year, p50, location, flag_diff) %>% 
  pivot_wider(names_from = location,
              values_from = p50) %>% 
  ggplot(., aes(x = intake_turb_NTU, y = munroe_turb_NTU)) +
  geom_point() +
  theme_bw() +
  labs(x = "median daily turbidity at intake (NTU)",
       y = "median daily turbidity at Munroe (NTU)",
       color = "year") +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey60") +
  scale_color_viridis_d() +
  coord_cartesian(xlim = c(0, 120),
                  ylim = c(0, 120)) +
  gghighlight(is.na(flag_diff))

# write_csv(turb_q_flagged, here("data/collated/FC_intake/fc_intake_munroe_turbidity_quantile_record_2010-2025_v2026-04-29.csv"))
```

Let's add those flags to the 15-minute data, too.

```{r}
fc_turb_flagged <- fc_turb %>% 
  mutate(date = date(datetime)) %>% 
  left_join(., flagged %>% select(date, flag_diff)) %>% 
  select(-date)

# write_csv(fc_turb_flagged, here("data/collated/FC_intake/fc_intake_munroe_turbidity_15min_record_2010-2025_v2026-04-29.csv"))

```

And let's tally how many times each quantile was above the 50 NTU, which is when
City of FC starts to monitor for high turbidity:

```{r}
turb_q_flagged %>% 
  filter(year != 2015, location == "intake_turb_NTU") %>% 
  pivot_longer(cols = c(p05, p25, p50, p75, p95), 
               names_to = "quantile", 
               values_to = "turb_ntu") %>% 
  filter(turb_ntu > 50) %>% 
  count(quantile, year) %>% 
  pivot_wider(names_from = quantile, values_from = n) %>% 
  # no instances of p05 above 50
  select(year, p95, p75, p50, p25) %>% 
  arrange(year) %>% 
  kable(caption = "Number of Days Turbidity Quantiles Exceeded 50 NTU",
        align = "c") %>% 
  kable_styling(bootstrap_options = c("striped"),
                full_width = TRUE,
                position = "center")  
```

Generally speaking, we will only be using the data aggregated to a daily timestep
for the intake data and we will not be using the Munroe data at all due to how
noisy it is.

