---
title: "Fort Collins Intake Data Explore"
output: html_document
date: "2025-11-20"
---

# Purpose

The purpose of this document is to explore the data provided to us from the City
of Fort Collins at their intake/Munroe Canal for turbidity, TOC, alkalinity.

```{r, echo = F, error = F, message = F}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
library(readxl)
library(here)
library(kableExtra)

# remotes::install_github("https://github.com/anguswg-ucsb/cdssr")
library(cdssr)
library(tidyverse)
```

## Total Organic Carbon

```{r}

toc <- read_excel(here("data/raw/fc_intake_data/Intake TOC.xlsx")) %>% 
  mutate(date = date(`Sampled Date`)) %>% 
  select(date, parameter = `Parameter Name`, result = `Corrected Result`) %>% 
  mutate(doy = yday(date),
         year = year(date))

ggplot(toc, aes(x = doy, y = result, color = factor(year))) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

# limited to days 100-200
ggplot(toc, aes(x = doy, y = result, color = factor(year))) +
  coord_cartesian(xlim = c(100, 200)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(toc$date, na.rm = T),
                       end_date = max(toc$date, na.rm = T),
                       timescale = "day")%>%
  mutate(year = factor(year(datetime)), 
         doy = yday(datetime))

# faceted
ggplot() +
  geom_area(data = flow_data,
            aes(x = doy, y = value/200),
            fill = "gray70",
            alpha = 0.5) +
  geom_point(data = toc, aes(x = doy, y = result, color = factor(year))) +
  coord_cartesian(xlim = c(100, 200)) +
  facet_wrap(year ~ ., scale = "free_y") +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

```


## Alkalinity

```{r}
alk <- read_csv(here("data/raw/fc_intake_data/Poudre_Intake_Alkalinity.csv")) %>% 
  mutate(date = mdy(`Sampled Date`)) %>% 
  select(date, parameter = `Parameter Name`, result = `Corrected Result`) %>% 
  mutate(doy = yday(date),
         year = year(date))

ggplot(alk, aes(x = doy, y = result, color = factor(year))) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_d() +
  geom_abline(slope = 0, intercept = 60, lty = 2, color = "orange") +
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

# limit to doy 100-200
ggplot(alk, aes(x = doy, y = result, color = factor(year))) +
  coord_cartesian(xlim = c(100, 200)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_d()+
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(alk$date, na.rm = T),
                       end_date = max(alk$date, na.rm = T),
                       timescale = "day")%>%
  mutate(year = factor(year(datetime)), 
         doy = yday(datetime))

# facet
ggplot() +
  geom_area(data = flow_data,
            aes(x = doy, y = value/50),
            fill = "gray70",
            alpha = 0.5) +
  geom_point(data = alk, aes(x = doy, y = result, color = factor(year))) +
  facet_wrap(year ~ ., scales = "free_y") +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

```

Alkalinity data is mostly Apr-Jun, and during this time there has never been a
sample that is higher than the treatment matrix cutoff (60 mg/L), the only thing
that we could do here is try to relate this to some other variable to give a warning
that it might be a high alkalinity year in the spring.

## Turbidity

```{r}
turb <- read_csv(here("data/raw/fc_intake_data/Intake and Munroe NTU.csv")) %>% 
  select(SampleDatetime, 
         Intake_NTU = N311_Intake_NTU, 
         Munroe_NTU = N4118_Munroe_NTU) %>% 
  mutate(datetime = ymd_hms(SampleDatetime),
         date = date(datetime),
         doy = yday(date),
         year = year(date))

# looks like there is both high frequency and daily data here based on the
# parsing issues

# high-frequency data
turb_rt <- turb %>% 
  filter(!is.na(datetime)) %>% 
  mutate(Munroe_NTU = case_when(Munroe_NTU <= 0 ~ NA_real_,
                                Munroe_NTU >= 327.67 ~ NA_real_, 
                                .default = Munroe_NTU),
         Intake_NTU = case_when(Intake_NTU <= 0 ~ NA_real_,
                                Intake_NTU >= 990 ~ NA_real_,
                                .default = Intake_NTU))

ggplot(turb_rt, aes(x = datetime)) +
  geom_point(aes(y = Intake_NTU, color = "Intake")) +
  geom_point(aes(y = Munroe_NTU, color = "Munroe")) +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = NULL, y = "turbidity (NTU)", color = "location")


turb_daily <- turb %>% 
  filter(is.na(datetime)) %>% 
  mutate(date = ymd(SampleDatetime),
         doy = yday(date),
         year = year(date)) %>% 
  select(-SampleDatetime, -datetime)

ggplot(turb_daily, aes(x = date)) +
  geom_point(aes(y = Intake_NTU, color = "Intake")) +
  geom_point(aes(y = Munroe_NTU, color = "Munroe")) +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = NULL, y = "turbidity (NTU)", color = "location")

ggplot(turb_daily, aes(x = date)) +
  geom_point(aes(y = Intake_NTU, color = "Intake")) +
  geom_point(aes(y = Munroe_NTU, color = "Munroe")) +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = NULL, y = "turbidity (NTU)", color = "location")


```

Based on the artifacts in the data, it looks like the daily data is just a 
summary of the real time, we can drop it and summarize from the RT ourselves.

Boxplots look pretty rough for these data, so we'll just plot the 95%, 75% 50%, 
25% and 5% in ribbons, with and without Q from the canyon mouth

```{r}
turb_q <- turb_rt %>%
  summarize(p05 = quantile(Intake_NTU, 0.05, na.rm = TRUE),
            p25 = quantile(Intake_NTU, 0.25, na.rm = TRUE),
            p50 = quantile(Intake_NTU, 0.50, na.rm = TRUE),
            p75 = quantile(Intake_NTU, 0.75, na.rm = TRUE),
            p95 = quantile(Intake_NTU, 0.95, na.rm = TRUE),
            .by = c(year, doy, date))

ggplot(turb_q, aes(x = doy)) +
  # 5–95% band
  geom_ribbon(aes(ymin = p05, ymax = p95, fill = "5 - 95%")) +
  # 25–75% band
  geom_ribbon(aes(ymin = p25, ymax = p75, fill = "25 - 75%")) +
  # Median line
  geom_line(aes(y = p50, color = "median"), color = "black") +
  # 100 NTU threshold
  geom_abline(aes(color = "100 NTU threshold"), color = "orange", 
              slope = 0, intercept = 100, lty = 2) +
  facet_grid(year ~ ., scales = "free_y") +
  coord_cartesian(xlim = c(75, 275)) +
  labs(y = "Intake Turbidity (NTU)",
       x = "Day of Year",
       title = "Daily Turbidity Quantile Bands",
       fill = "Quantile Bands") +
  scale_fill_manual(values = c("5 - 95%" = "grey85",
                               "25 - 75%" = "grey60"),
                    breaks = c("5 - 95%", "25 - 75%")) +
  theme_bw() +
  theme(legend.position = "bottom")

flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(turb_q$date, na.rm = T),
                       end_date = max(turb_q$date, na.rm = T),
                       timescale = "day")%>%
  mutate(year = factor(year(datetime)), 
         doy = yday(datetime))

# with Q
ggplot(turb_q, aes(x = doy)) +
  geom_area(data = flow_data,
            aes(x = doy, y = value/30),
            fill = "lightblue",
            alpha = 0.5) +
  # 5–95% band
  geom_ribbon(aes(ymin = p05, ymax = p95, fill = "5 - 95%")) +
  # 25–75% band
  geom_ribbon(aes(ymin = p25, ymax = p75, fill = "25 - 75%")) +
  # Median line
  geom_line(aes(y = p50, color = "median"), color = "black") +
  # 100 NTU threshold
  geom_abline(aes(color = "100 NTU threshold"), color = "orange", 
              slope = 0, intercept = 100, lty = 2) +
  facet_grid(year ~ ., scales = "free_y") +
  coord_cartesian(xlim = c(75, 275)) +
  labs(y = "Intake Turbidity (NTU)",
       x = "Day of Year",
       title = "Daily Turbidity Quantile Bands",
       fill = "Quantile Bands") +
  scale_fill_manual(values = c("5 - 95%" = "grey85",
                               "25 - 75%" = "grey60"),
                    breaks = c("5 - 95%", "25 - 75%")) +
  theme_bw() +
  theme(legend.position = "bottom")

```

And look at Munroe Canal - not sure if this is in the canal or in the river
at the canal. This is their alternate Poudre source if the intake proper had an
issue.

```{r}
turb_q_munroe <- turb_rt %>%
  summarize(p05 = quantile(Munroe_NTU, 0.05, na.rm = TRUE),
            p25 = quantile(Munroe_NTU, 0.25, na.rm = TRUE),
            p50 = quantile(Munroe_NTU, 0.50, na.rm = TRUE),
            p75 = quantile(Munroe_NTU, 0.75, na.rm = TRUE),
            p95 = quantile(Munroe_NTU, 0.95, na.rm = TRUE),
            .by = c(year, doy))

ggplot(turb_q_munroe, aes(x = doy)) +
  # 5–95% band
  geom_ribbon(aes(ymin = p05, ymax = p95, fill = "5 - 95%")) +
  # 25–75% band
  geom_ribbon(aes(ymin = p25, ymax = p75, fill = "25 - 75%")) +
  # Median line
  geom_line(aes(y = p50, color = "median"), color = "black") +
  # 100 NTU threshold
  geom_abline(aes(color = "100 NTU threshold"), color = "orange", 
              slope = 0, intercept = 100, lty = 2) +
  facet_grid(year ~ ., scales = "free_y") +
  coord_cartesian(xlim = c(75, 275)) +
  labs(y = "Munroe Canal Turbidity (NTU)",
       x = "Day of Year",
       title = "Daily Turbidity Quantile Bands",
       fill = "Quantile Bands") +
  scale_fill_manual(values = c("5 - 95%" = "grey85",
                               "25 - 75%" = "grey60"),
                    breaks = c("5 - 95%", "25 - 75%")) +
  theme_bw() +
  theme(legend.position = "bottom")

```

And let's tally how many times each quantile was above the 50 NTU, which is when
City of FC starts to monitor for high turbidity:

```{r}
turb_q %>% 
  pivot_longer(cols = c(p05, p25, p50, p75, p95), 
               names_to = "quantile", 
               values_to = "turb_ntu") %>% 
  filter(turb_ntu > 50) %>% 
  count(quantile, year) %>% 
  pivot_wider(names_from = quantile, values_from = n) %>% 
  select(year, p95, p75, p50, p25) %>% 
  arrange(year) %>% 
  kable(caption = "Number of Days Turbidity Quantiles Exceeded 50 NTU",
        align = "c") %>% 
  kable_styling(bootstrap_options = c("striped"),
                full_width = TRUE,
                position = "center")  
```

