---
title: "Fort Collins Intake Data Explore"
output: html_document
date: "2026-01-08"
editor_options: 
  markdown: 
    wrap: 80
---

# Purpose

The purpose of this document is to explore the data provided to us from the City
of Fort Collins at their intake/Munroe Canal for turbidity, TOC, alkalinity.

```{r, echo = F, error = F, message = F, warning = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
library(readxl)
library(here)
library(kableExtra)
library(gghighlight)

# remotes::install_github("https://github.com/anguswg-ucsb/cdssr")
library(cdssr)
library(tidyverse)
```

# Total Organic Carbon

## TOC from the FC intake

We have intake TOC data since 2019.

```{r}

toc <- read_excel(here("data/raw/fc_intake_data/Intake TOC.xlsx")) %>% 
  mutate(date = date(`Sampled Date`)) %>% 
  select(date, parameter = `Parameter Name`, result = `Corrected Result`) %>% 
  mutate(doy = yday(date),
         year = year(date))

ggplot(toc, aes(x = doy, y = result, color = factor(year))) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

# limited to days 100-200
ggplot(toc, aes(x = doy, y = result, color = factor(year))) +
  coord_cartesian(xlim = c(100, 200)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

```

## Raw Water TOC

This is data from the raw water station INSIDE the FC treatment facility, but
the data record starts in 2010.

```{r}
rw_toc <- read_csv(here("data/raw/fc_intake_data/Poudre_Raw_Water_Sample_Station_TOC_2010-2025.csv")) %>% 
  mutate(date = mdy(`Sampled Date`)) %>% 
  select(date, TOC = `Corrected Result`) %>% 
  filter(!is.na(TOC) & TOC > 1) %>%  # one oddball here.
  mutate(doy = yday(date),
         year = year(date))

ggplot(rw_toc, aes(x = doy, y = TOC, color = factor(year))) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

```

Interesting to note that the 2013 flood is very evident in these data - see the
spike in late season TOC.

Let's do a quick check of the data at intake and the data at the RW station to
see how interchangable the values are.

```{r}
full_join(toc, rw_toc) %>% 
  ggplot(aes(x = TOC, y = result)) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey60") +
  geom_point() +
  labs(x = "TOC at raw water sink", y = "TOC at intake") +
  theme_bw()

```

These are pretty darn close to interchangable, so if there are data available at
the intake and not at the RW sink, let's use it, and generally take an average
if there are multiple values available.

## Collated TOC

```{r}
fc_TOC_record <- full_join(toc, rw_toc) %>% 
  pivot_longer(cols = c(TOC, result),
               names_to = "source",
               values_to = "TOC") %>% 
  filter(!is.na(TOC)) %>% 
  summarise(mean_TOC = mean(TOC), 
            .by = c(year, doy, date))

ggplot(fc_TOC_record, aes(x = doy, y = mean_TOC, color = factor(year))) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(fc_TOC_record$date, na.rm = T),
                       end_date = max(fc_TOC_record$date, na.rm = T),
                       timescale = "day")%>%
  mutate(year = factor(year(datetime)), 
         doy = yday(datetime))

# faceted
ggplot() +
  geom_area(data = flow_data,
            aes(x = doy, y = value/200), # divide flow by a silly amount for visualization
            fill = "gray70",
            alpha = 0.5) +
  geom_point(data = fc_TOC_record, 
             aes(x = doy, y = mean_TOC, color = factor(year))) +
  facet_wrap(year ~ ., scale = "free_y") +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "TOC (mg/L)", color = "year")

# write_csv(fc_TOC_record, here("data/collated/FC_intake/fc_intake_TOC_record_2010-2025.csv"))
```

# Alkalinity

## Alkalinity from intake

Unlinke to TOC, alkalinity at the intake is available since 2012.

```{r}
alk <- read_csv(here("data/raw/fc_intake_data/Poudre_Intake_Alkalinity.csv")) %>% 
  mutate(date = mdy(`Sampled Date`)) %>% 
  select(date, parameter = `Parameter Name`, result = `Corrected Result`) %>% 
  mutate(doy = yday(date),
         year = year(date))

ggplot(alk, aes(x = doy, y = result, color = factor(year))) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_d() +
  geom_abline(slope = 0, intercept = 60, lty = 2, color = "orange") +
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

# limit to doy 100-200
ggplot(alk, aes(x = doy, y = result, color = factor(year))) +
  coord_cartesian(xlim = c(100, 200)) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_d()+
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(alk$date, na.rm = T),
                       end_date = max(alk$date, na.rm = T),
                       timescale = "day")%>%
  mutate(year = factor(year(datetime)), 
         doy = yday(datetime))

# facet
ggplot() +
  geom_area(data = flow_data,
            aes(x = doy, y = value/50),
            fill = "gray70",
            alpha = 0.5) +
  geom_point(data = alk, aes(x = doy, y = result, color = factor(year))) +
  facet_wrap(year ~ ., scales = "free_y") +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

```

Alkalinity data at the intake is mostly Apr-Jun, and during this time there has
never been a sample that is higher than the treatment matrix cutoff (60 mg/L),
the only thing that we could do here is try to relate this to some other
variable to give a warning that it might be a high alkalinity year in the
spring.

In the data that Sam has explored and anecdotally, alkalinity is more of an 
issue for Greeley whose intake is further downstream.

## Raw Water Alkalinity

We also have the record of Alkalinity at the RW sink.

```{r}
rw_alk <- read_csv(here("data/raw/fc_intake_data/Poudre_Raw_Water_Sample_Station_Alkalinity.csv")) %>% 
  mutate(date = mdy(`Sampled Date`)) %>% 
  select(date, alkalinity = `Corrected Result`) %>% 
  #filter(!is.na(TOC) & TOC > 1) %>%  # one oddball here.
  mutate(doy = yday(date),
         year = year(date))

ggplot(rw_alk, aes(x = doy, y = alkalinity, color = factor(year))) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "alkalinity (mg/L CaCO3)", color = "year")

```

We still don't see anything higher than the 60 mg/L, but helpful to see the
pattern.

Let's do a quick check of the data at intake and the data at the RW station to
see how interchangable the values are.

```{r}
full_join(alk, rw_alk) %>% 
  ggplot(aes(x = alkalinity, y = result)) +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey60") +
  geom_point() +
  labs(x = "Alkalinity at raw water sink", y = "Alkalinity at intake") +
  theme_bw() +
  coord_cartesian(xlim = c(5, 50), ylim = c(5,50))

```

Also right on like TOC. We'll again take an average if there are multiple values
available for a given date.

## Collated Alkalinity

```{r}
fc_alk_record <- full_join(alk, rw_alk) %>% 
  pivot_longer(cols = c(alkalinity, result),
               names_to = "source",
               values_to = "alkalinity") %>% 
  filter(!is.na(alkalinity)) %>% 
  summarise(mean_alk = mean(alkalinity), 
            .by = c(year, doy, date))

ggplot(fc_alk_record, aes(x = doy, y = mean_alk, color = factor(year))) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_color_viridis_d() +
  labs(x = "day of year", y = "Alkalinity (mg/L CaCO3)", color = "year")

# write_csv(fc_alk_record, here("data/collated/FC_intake/fc_intake_alk_record_2010-2025.csv"))
```

Those peaks in 2012 data are very likely real (after High Park Fire).

# Turbidity

We have 15-minute data from the City at both the Intake and Munroe Canal. These
data need a little QAQC love, so we'll take care of that. For Munroe Canal - not
sure if this is in the canal or in the river at the canal. This is City of FC
alternate Poudre source if the intake proper had an issue.

## Collation and QAQC

```{r}
intake_turb <- read_csv(here("data/raw/fc_intake_data/Intake_Turbidity_2010-2025.csv"),
                        col_names = c("datetime", "intake_turb_NTU"),
                        skip = 1,
                        col_select = c("datetime", "intake_turb_NTU"))

munroe_turb <- read_csv(here("data/raw/fc_intake_data/Munroe_Turbidity_2013-2025.csv"),,
                        col_names = c("datetime", "munroe_turb_NTU"),
                        skip = 1,
                        col_select = c("datetime", "munroe_turb_NTU")) 

fc_turb <- full_join(intake_turb, munroe_turb) %>% 
  pivot_longer(cols = c(munroe_turb_NTU, intake_turb_NTU),
               names_to = "location",
               values_to = "NTU") %>% 
  mutate(NTU = case_when(NTU <= 0 ~ NA, 
                         NTU >= 1000 ~ NA,
                         round(NTU, 0) == 655 & location == "munroe_turb_NTU" ~ NA,
                         round(NTU, 0) == 630 & location == "munroe_turb_NTU" ~ NA,
                         round(NTU, 0) == 555 & location == "munroe_turb_NTU" ~ NA,
                         .default = NTU))

```

This still needs a little massaging, so let's look year-by-year.

### First Pass QAQC

<details open>
<summary>Click to collapse year-by-year plots</summary>

```{r}
# plot year-by-year for outliers and errors
map(2010:2025, \(y) {
  start_date <- paste0(y, "-01-01")
  end_date <- paste0(y + 1, "-01-01")
  ggplot(fc_turb %>% filter(between(datetime, 
                                    ymd(start_date), 
                                    ymd(end_date))), 
         aes(x = datetime, y = NTU, color = location)) +
    geom_point() +
    labs(title = y) +
    scale_color_viridis_d() +
    theme_bw() +
    theme(legend.position = "bottom")
  
})
```

</details>

-   2012 has a pretty significant data gap after high turbidity, might want to
leave out.

-   Prior to 2013, data was generally no higher than 100 NTU, starting in 2013
the range increases to 1000. Maybe leave out data prior to 2013? That's also
post-High Park Fire, so maybe it's real? It could also just be a sensor
change?

A few places to recode errant data:

-   Sept 2015-Jun 2016

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-09-01"), 
                                  ymd("2015-10-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-04-15"), 
                                  ymd("2016-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2015-09-15"), ymd("2016-06-01")) &
                         location == "munroe_turb_NTU" &
                         NTU > 300, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2015-09-01"), 
                                  ymd("2015-10-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-04-15"), 
                                  ymd("2016-06-01"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

```

</details>

-   Oct 2016-Apr 2017

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-10-15"), 
                                  ymd("2016-11-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2017-03-20"), 
                                  ymd("2017-04-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2016-10-24"), ymd("2017-04-03")) &
                         location == "munroe_turb_NTU" &
                         NTU > 500, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2016-10-15"), 
                                  ymd("2016-11-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2017-03-20"), 
                                  ymd("2017-04-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

```

</details>

-   Nov 2017-Apr 2018

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2017-10-15"), 
                                  ymd("2017-11-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2018-03-20"), 
                                  ymd("2018-04-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2017-11-06"), ymd("2018-04-09")) &
                         location == "munroe_turb_NTU" &
                         NTU > 500, 
                       NA, 
                       NTU),
         NTU = if_else(between(datetime, ymd("2017-12-01"), ymd("2018-04-01")) &
                         location == "intake_turb_NTU" &
                         NTU > 200, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2017-10-15"), 
                                  ymd("2017-11-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2018-03-20"), 
                                  ymd("2018-04-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")


```

</details>

-   Oct 2023 - Apr 2024

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-10-15"), 
                                  ymd("2023-11-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-03-20"), 
                                  ymd("2024-04-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2023-10-23"), ymd("2024-04-08")) &
                         location == "munroe_turb_NTU" &
                         NTU > 600, 
                       NA, 
                       NTU))

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2023-10-15"), 
                                  ymd("2023-11-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fc_turb %>% filter(between(datetime, 
                                  ymd("2024-03-20"), 
                                  ymd("2024-04-15"))), 
       aes(x = datetime, y = NTU, color = location)) +
  geom_point() +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "bottom")

```

</details>

### Second Pass

Plot year-by-year to re-check cleaning

<details open>
<summary>Click to collapse year-by-year plots</summary>

```{r}
# plot year-by-year for outliers and errors
map(2010:2025, \(y) {
  start_date <- paste0(y, "-01-01")
  end_date <- paste0(y + 1, "-01-01")
  ggplot() +
    geom_area(data = flow_data %>% 
                filter(between(datetime, 
                               ymd(start_date), 
                               ymd(end_date))),
              aes(x = datetime, y = value/10),
              fill = "blue",
              alpha = 0.3) +
    geom_point(data = fc_turb %>% filter(between(datetime, 
                                                 ymd(start_date), 
                                                 ymd(end_date))), 
               aes(x = datetime, y = NTU, color = location)) +
    labs(title = y,
         y = "turbidity (NTU)") +
    scale_color_viridis_d() +
    theme_bw() +
    theme(legend.position = "bottom")
})

```

</details>

Still a few artifacts here:

-   Sept 2011 before data gap

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2011-09-01"), 
                             ymd("2011-10-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2011-09-01"), 
                                               ymd("2011-10-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2011-09-04"), ymd("2011-09-06")) &
                         NTU > 10, 
                       NA, 
                       NTU))

ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2011-09-01"), 
                             ymd("2011-10-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2011-09-01"), 
                                               ymd("2011-10-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

```

</details>

-   Apr 2013 before data gap

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2013-04-15"), 
                             ymd("2013-05-15"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2013-04-15"), 
                                               ymd("2013-05-15"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2013-04-29"), ymd("2013-05-01")) &
                         NTU > 10, 
                       NA, 
                       NTU))

ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2013-04-15"), 
                             ymd("2013-05-15"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2013-04-15"), 
                                               ymd("2013-05-15"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")
```

</details>

-   Munroe 2015 Apr - Jul miscalibrated

If you look at the quantiles, the relationship between the two locations for
this period is super different from every other year:

```{r, echo=FALSE}
turb_q <- fc_turb %>% 
  mutate(date = date(datetime),
         doy = yday(date),
         year = year(date)) %>%
  filter(!is.na(NTU)) %>% 
  summarize(p05 = quantile(NTU, 0.05),
            p25 = quantile(NTU, 0.25),
            p50 = quantile(NTU, 0.50),
            p75 = quantile(NTU, 0.75),
            p95 = quantile(NTU, 0.95),
            n = n(),
            .by = c(year, doy, date, location))

# data that have the same quantile values on a given day or that have low n 
# should be excluded here
turb_q_filt <- turb_q %>% 
  # we want to make sure we have at least 75% of the days values
  filter(n > 96*0.75) %>% 
  # and that all the values are not the same (indicating something is wrong with the sensor)
  filter(p05 != p95) %>% 
  filter(p25 != p75)

date_sequence <- seq.Date(min(turb_q$date), max(turb_q$date)) %>% 
  tibble(date = .) %>% 
  mutate(doy = yday(date),
         year = year(date)) %>% 
  cross_join(., tibble(location = c("munroe_turb_NTU", "intake_turb_NTU")))

turb_q_filt <- left_join(date_sequence, turb_q_filt)

turb_q_filt %>% 
  select(date, year, p50, location) %>% 
  pivot_wider(names_from = location,
              values_from = p50) %>% 
  ggplot(., aes(x = intake_turb_NTU, y = munroe_turb_NTU, color = factor(year))) +
  geom_point() +
  theme_bw() +
  labs(x = "median daily turbidity at intake (NTU)",
       y = "median daily turbidity at Munroe (NTU)",
       color = "year") +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey60") +
  scale_color_viridis_d() +
  coord_cartesian(xlim = c(0, 120),
                  ylim = c(0, 120)) +
  gghighlight(year == 2015)
```

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2015-04-01"), 
                             ymd("2015-07-31"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2015-04-01"), 
                                               ymd("2015-07-31"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 week") +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2015-04-06"), ymd("2015-07-16")) &
                         location == "munroe_turb_NTU",
                       NA, 
                       NTU))

ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2015-04-01"), 
                             ymd("2015-07-31"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2015-04-01"), 
                                               ymd("2015-07-31"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 week") +
  theme(legend.position = "bottom")

```

</details>

-   Apr 2016 Munroe (after data gap)

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2016-04-01"), 
                             ymd("2016-05-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2016-04-01"), 
                                               ymd("2016-05-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2016-04-25"), ymd("2016-04-28")) &
                         location == "munroe_turb_NTU" &
                         NTU < 10,
                       NA, 
                       NTU))

ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2016-04-01"), 
                             ymd("2016-05-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2016-04-01"), 
                                               ymd("2016-05-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")
```

</details>

-   May 2019 intake (flatlined near 0)

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2019-05-01"), 
                             ymd("2019-06-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2019-05-01"), 
                                               ymd("2019-06-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  labs(y = "turbidity (NTU)") +
  scale_color_viridis_d() +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2019-05-15"), ymd("2019-05-18")) &
                         location == "intake_turb_NTU" &
                         NTU < 8,
                       NA, 
                       NTU))

ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2019-05-01"), 
                             ymd("2019-06-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2019-05-01"), 
                                               ymd("2019-06-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  labs(y = "turbidity (NTU)") +
  scale_color_viridis_d() +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

```

</details>

-   Jun 2020 Munroe (after data gap)

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2020-06-01"), 
                             ymd("2020-07-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2020-06-01"), 
                                               ymd("2020-07-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2020-06-16"), ymd("2020-06-18")) &
                         location == "munroe_turb_NTU" &
                         NTU >10,
                       NA, 
                       NTU))

ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2020-06-01"), 
                             ymd("2020-07-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2020-06-01"), 
                                               ymd("2020-07-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

```

</details>

-   Aug 2021 intake

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2021-08-01"), 
                             ymd("2021-09-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2021-08-01"), 
                                               ymd("2021-09-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2021-08-10"), ymd("2021-08-13")) &
                         location == "intake_turb_NTU" &
                         NTU < 5,
                       NA, 
                       NTU))

ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2021-08-01"), 
                             ymd("2021-09-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2021-08-01"), 
                                               ymd("2021-09-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")
```

</details>

-   Jul 2021 Munroe

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2021-07-01"), 
                             ymd("2021-08-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2021-07-01"), 
                                               ymd("2021-08-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2021-07-21"), ymd("2021-07-28")) &
                         location == "munroe_turb_NTU" &
                         NTU > 10,
                       NA, 
                       NTU))

ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2021-07-01"), 
                             ymd("2021-08-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2021-07-01"), 
                                               ymd("2021-08-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")
```

</details>

-   Apr 2025 Munroe

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2025-04-01"), 
                             ymd("2025-05-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2025-04-01"), 
                                               ymd("2025-05-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2025-04-02"), ymd_hm("2025-04-02 12:00")) &
                         location == "munroe_turb_NTU",
                       NA, 
                       NTU))

ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2025-04-01"), 
                             ymd("2025-05-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2025-04-01"), 
                                               ymd("2025-05-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

```

</details>

-   Sept 2025 Munroe

<details>
<summary>Click to expand plots</summary>

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2025-09-01"), 
                             ymd("2025-10-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2025-09-01"), 
                                               ymd("2025-10-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

fc_turb <- fc_turb %>% 
  mutate(NTU = if_else(between(datetime, ymd("2025-09-25"), ymd_hm("2025-09-28 12:00")) &
                         location == "munroe_turb_NTU",
                       NA, 
                       NTU))

ggplot() +
  geom_area(data = flow_data %>% 
              filter(between(datetime, 
                             ymd("2025-09-01"), 
                             ymd("2025-10-01"))), 
            aes(x = datetime, y = value/10),
            fill = "blue",
            alpha = 0.3) +
  geom_point(data = fc_turb %>% filter(between(datetime, 
                                               ymd("2025-09-01"), 
                                               ymd("2025-10-01"))), 
             aes(x = datetime, y = NTU, color = location)) +
  scale_color_viridis_d() +
  labs(y = "turbidity (NTU)") +
  theme_bw() +
  scale_x_datetime(minor_breaks = "1 day") +
  theme(legend.position = "bottom")

```

</details>

### Final check

Plot to re-check cleaning year-by-year

<details open>
<summary>Click to collapse year-by-year </summary> 

```{r}
# plot year-by-year for outliers and errors
map(2010:2025, \(y) {
  start_date <- paste0(y, "-01-01")
  end_date <- paste0(y + 1, "-01-01")
  ggplot() +
    geom_area(data = flow_data %>% 
                filter(between(datetime, 
                               ymd(start_date), 
                               ymd(end_date))),
              aes(x = datetime, y = value/10),
              fill = "blue",
              alpha = 0.3) +
    geom_point(data = fc_turb %>% filter(between(datetime, 
                                                 ymd(start_date), 
                                                 ymd(end_date))), 
               aes(x = datetime, y = NTU, color = location)) +
    labs(title = y,
         y = "turbidity (NTU)")  +
    scale_color_viridis_d() +
    theme_bw() +
    theme(legend.position = "bottom")
})

```

</details>

Okay, that's a lot better, still some noise, but generally okay. There's some
persistent near-zero noise sometimes during runoff, but I think that with
quantiles we should be okay including those in any training/validation data.

### Daily Quantiles

For the purpose of visualization and a dataset for training and validation, we'll 
get some quantiles (95%, 75%, 50%, 25% and 5%) and plot in ribbons, with 
Q from the canyon mouth. These quantiles assume that min/max might be a bit misleading
due to the noise in the data.


```{r}
turb_q <- fc_turb %>% 
  mutate(date = date(datetime),
         doy = yday(date),
         year = year(date)) %>%
  filter(!is.na(NTU)) %>% 
  summarize(p05 = quantile(NTU, 0.05),
            p25 = quantile(NTU, 0.25),
            p50 = quantile(NTU, 0.50),
            p75 = quantile(NTU, 0.75),
            p95 = quantile(NTU, 0.95),
            n = n(),
            .by = c(year, doy, date, location))

# data that have the same quantile values on a given day or that have low n 
# should be excluded here
turb_q_filt <- turb_q %>% 
  # we want to make sure we have at least 75% of the days values
  filter(n > 96*0.75) %>% 
  # and that all the values are not the same (indicating something is wrong with the sensor)
  filter(p05 != p95) %>% 
  filter(p25 != p75)

date_sequence <- seq.Date(min(turb_q$date), max(turb_q$date)) %>% 
  tibble(date = .) %>% 
  mutate(doy = yday(date),
         year = year(date)) %>% 
  cross_join(., tibble(location = c("munroe_turb_NTU", "intake_turb_NTU")))

turb_q_filt <- left_join(date_sequence, turb_q_filt)

```


```{r, fig.height=12, fig.width=8}
flow_data <- get_sw_ts(abbrev = "CLAFTCCO",
                       start_date = min(turb_q$date, na.rm = T),
                       end_date = max(turb_q$date, na.rm = T),
                       timescale = "day")%>%
  mutate(year = factor(year(datetime)), 
         doy = yday(datetime))

# with Q
ggplot(turb_q_filt, aes(x = doy)) +
  geom_area(data = flow_data,
            aes(x = doy, y = value/30),
            fill = "lightblue",
            alpha = 0.5) +
  # 5–95% band
  geom_ribbon(aes(ymin = p05, ymax = p95, fill = "5 - 95%")) +
  # 25–75% band
  geom_ribbon(aes(ymin = p25, ymax = p75, fill = "25 - 75%")) +
  # Median line
  geom_line(aes(y = p50, color = "median"), color = "black") +
  # 100 NTU threshold
  geom_abline(aes(color = "100 NTU threshold"), color = "orange", 
              slope = 0, intercept = 100, lty = 2) +
  facet_grid(year ~ location) +
  coord_cartesian(xlim = c(75, 275),
                  ylim = c(0, 150)) +
  labs(y = "turbidity (NTU)",
       x = "Day of Year",
       title = "Daily Turbidity Quantile Bands",
       fill = "Quantile Bands") +
  scale_fill_manual(values = c("5 - 95%" = "grey85",
                               "25 - 75%" = "grey60"),
                    breaks = c("5 - 95%", "25 - 75%")) +
  theme_bw() +
  theme(legend.position = "bottom")

```

For kicks, let's look at the relationship

```{r}
turb_q_filt %>% 
  select(date, year, p50, location) %>% 
  filter(year != 2015) %>% 
  pivot_wider(names_from = location,
              values_from = p50) %>% 
  ggplot(., aes(x = intake_turb_NTU, y = munroe_turb_NTU, color = factor(year))) +
  geom_point() +
  theme_bw() +
  labs(x = "median daily turbidity at intake (NTU)",
       y = "median daily turbidity at Munroe (NTU)",
       color = "year") +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey60") +
  scale_color_viridis_d() +
  coord_cartesian(xlim = c(0, 120),
                  ylim = c(0, 120))

```

Still some outliers, but overall, this looks like we've done sufficient QAQC to
move on. For kicks, let's just flag any data where there is greater than a 100%
difference between Munroe and the Intake. These data might be real, but they
also might not and we may wish to keep them out of the training dataset.

We'll do this using symmetric differencing at the median value.

```{r}
flagged <- turb_q_filt %>%
  pivot_wider(id_cols = c(date, doy, year),
              names_from = location,
              values_from = p50) %>%
  mutate(pct_diff = abs(munroe_turb_NTU - intake_turb_NTU) /
           ((munroe_turb_NTU + intake_turb_NTU) / 2),
         flag_diff = if_else(pct_diff > 1,
                             "suspect",
                             NA)) %>% 
  select(date, pct_diff, flag_diff) 


turb_q_flagged <- full_join(turb_q_filt, flagged)

turb_q_flagged %>% 
  select(date, year, p50, location, flag_diff) %>% 
  pivot_wider(names_from = location,
              values_from = p50) %>% 
  ggplot(., aes(x = intake_turb_NTU, y = munroe_turb_NTU)) +
  geom_point() +
  theme_bw() +
  labs(x = "median daily turbidity at intake (NTU)",
       y = "median daily turbidity at Munroe (NTU)",
       color = "year") +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey60") +
  scale_color_viridis_d() +
  coord_cartesian(xlim = c(0, 120),
                  ylim = c(0, 120)) +
  gghighlight(is.na(flag_diff))

# write_csv(turb_q_flagged, here("data/collated/FC_intake/fc_intake_munroe_turbidity_quantile_record_2010-2025.csv"))
```

And let's tally how many times each quantile was above the 50 NTU, which is when
City of FC starts to monitor for high turbidity:

```{r}
turb_q_flagged %>% 
  filter(year != 2015, location == "intake_turb_NTU") %>% 
  pivot_longer(cols = c(p05, p25, p50, p75, p95), 
               names_to = "quantile", 
               values_to = "turb_ntu") %>% 
  filter(turb_ntu > 50) %>% 
  count(quantile, year) %>% 
  pivot_wider(names_from = quantile, values_from = n) %>% 
  # no instances of p05 above 50
  select(year, p95, p75, p50, p25) %>% 
  arrange(year) %>% 
  kable(caption = "Number of Days Turbidity Quantiles Exceeded 50 NTU",
        align = "c") %>% 
  kable_styling(bootstrap_options = c("striped"),
                full_width = TRUE,
                position = "center")  
```

