---
title: "original_grab_sample_collation"
author: "Sam Struthers- CSU ROSSyndicate"
date: "`r Sys.Date()`"
output: html_document
---

# Setup libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(dplyr.summarise.inform = FALSE)
# load libraries
source("src/setup_libraries.R")
```
# Pull in & clean up grab sample data

This has been migrated to `src/generate_grab_sample` scripts and should use these scripts to generate updated chemistry datasets. 

## ROSS

```{r}

#discrete sample data and location metadata from most recent pub

source("src/pull_ROSS_zenodo_data.R")
zenodo_data <- pull_ROSS_zenodo_data(data_version = "v2025.07.01", DOI = "15883685", save_folder_dir = here("data", "raw", "chem", "ross_clp_chem"))

# Grab the water chemistry dataset

ross_water_chem <- zenodo_data[["most_recent_chem"]]%>%
  mutate(site_code = tolower(site_code), 
         #fixing site names based on sonde deployments with ROSS/Virridy sondes
         site_code = case_when(site_code == "archery" & DT_mst <= ymd("2024-11-30") ~ "archery_virridy",
                          site_code == "timberline" & DT_mst <= ymd("2024-11-30") ~ "riverbend_virridy",
                          site_code == "prospect" & DT_mst <= ymd("2024-11-30") ~ "cottonwood_virridy",
                          #updating to new names
                          site_code == "timberline" & DT_mst >= ymd("2024-11-30") ~ "riverbend",
                          site_code == "prospect" & DT_mst >= ymd("2024-11-30") ~ "cottonwood",
                          site_code == "lincoln" ~ "udall", 
                          site_code == "legacy" ~ "salyer",
                          site_code == "boxelder" ~ "elc",
                          site_code == "tamasag" ~ "bellvue",
                          T ~ site_code))%>%
  filter(DT_mst >= as_date("2023-09-01"))%>%
  distinct()%>%
  dplyr::select(site_code, 
         Date, 
         DT_sample = DT_mst,
         TOC, TN, DOC, ChlA, NO3, NH4, PO4, SC,Cl, lab_turb = Turbidity, TSS,DT_mst_char)%>%
  mutate(collector = "ROSS")
```

## Fort Collins UCLP Data

This is given to us by the City of Fort Collins Watershed Team (Diana Schmidt and Jared Heath). 
2025 Dataset is not yet finalized but can likely still be used for our first models


```{r}

FC_chem_2022_2024 <- read_csv(here("data", "raw","chem", "fc_clp_chem", "UCLP_CWQMP_database_2022-24.csv"), show_col_types = F)%>%
  mutate(Date = as.POSIXct(Date, format = "%m/%d/%y"), 
         NO3 = NO3_N/ 0.2259, 
         SC = TDS/ 0.65, #converting TDS to SC
         Cl = as.numeric(Cl), 
         lab_turb = as.numeric(Turbidity),
         TN = as.numeric(TN_calc),
         ChlA = NA_integer_,
         DOC = NA_integer_, 
         NH4 = NA_integer_, 
         TSS = NA_integer_,
         site_code = tolower(ShortDesc))%>%
  #filter to the period of our sensor deployment and sites where we have sensors
  filter(site_code %in% c("chd", 'pbd', "sfm") & Date >= as_date("2023-09-01"))%>%
  #grab columns of interest
  dplyr::select(site_code, Date,
         TOC, TN, DOC, ChlA, NO3, NH4, PO4 = oPhos, SC,Cl, lab_turb, TSS)%>%
  mutate(collector = "FC")

FC_field_data_2022_2024 <- readxl::read_xlsx(path = here("data", "raw","chem", "fc_clp_chem", "UCLP_field_data_2022-24.xlsx"))%>%
  mutate(
    Date = as.Date(`Date (MM/DD/YYYY)`),
    # Combine date and time into one datetime column from MT to MST for consistency 
    dt_mst = with_tz(ymd_hms(paste(
      format(`Date (MM/DD/YYYY)`, "%Y-%m-%d"), 
      format(`Time (HH:MM:SS)`, "%H:%M:%S")
    ), tz = "America/Denver"), tzone = "MST"),
    # Remove numbers from site name
    site_code = tolower(gsub("\\d+", "", `Site Name`)), 
    DT_mst_char = as.character(dt_mst)
  )%>%
  dplyr::select(site_code, Date, DT_sample = dt_mst, DT_mst_char)
  
# Combine the chemistry data with the field data
FC_chem_2022_2024 <- FC_chem_2022_2024 %>%
  left_join(FC_field_data_2022_2024, by = c("site_code", "Date")) 


# Repeat for 2025 data (TOC only!)

FC_chem_2025 <- readxl::read_xlsx(path = here("data", "raw","chem", "fc_clp_chem", "UCLP_CWQMP_TOC_202504-202506.xlsx"))%>%
  filter(grepl(x = `Parameter Name`, pattern = "Total Organic Carbon", ignore.case = TRUE), 
          `Entered Unit` == "mg/L",
         grepl(x = `Location Name`, pattern = "CHD|PBD|SFM|PBR", ignore.case = TRUE))%>%
  dplyr::select(
    Date = `Sampled Date`, 
    site_name = `Location Name`,
    value = `Corrected Result`
  )%>%
  mutate(site_code = case_when(
    grepl("PBR", site_name, ignore.case = TRUE) ~ "pbr",
    grepl("CHD", site_name, ignore.case = TRUE) ~ "chd",
    grepl("PBD", site_name, ignore.case = TRUE) ~ "pbd",
    grepl("SFM", site_name, ignore.case = TRUE) ~ "sfm",
    TRUE ~ NA_character_),
         TOC = as.numeric(value))%>%
  mutate(collector = "FC")%>%
  dplyr::select(site_code, Date, TOC, collector)

FC_field_data_2025 <- readxl::read_xlsx(path = here("data", "raw", "chem", "fc_clp_chem", "UCLP_field_data_04-062025.xlsx"))%>%
  mutate(
    Date = as.Date(`Date (MM/DD/YYYY)`),
    # Combine date and time into one datetime column from MT to MST for consistency 
    dt_mst = with_tz(ymd_hms(paste(
      format(`Date (MM/DD/YYYY)`, "%Y-%m-%d"), 
      format(`Time (HH:MM:SS)`, "%H:%M:%S")
    ), tz = "America/Denver"), tzone = "MST"),
    # Remove numbers from site name
    site_code = tolower(gsub("\\d+", "", `Site Name`)), 
    DT_mst_char = as.character(dt_mst)
  )%>%
  dplyr::select(site_code, Date, DT_sample = dt_mst, DT_mst_char)


FC_chem_2025 <- FC_chem_2025 %>%
  left_join(FC_field_data_2025, by = c("site_code", "Date"))%>%
  na.omit()


FC_chem <- bind_rows(FC_chem_2022_2024, FC_chem_2025)


#write_csv(FC_chem, "data/FC_clp_chem/FC_testing_data.csv")
```

## Combine data
```{r}

water_chem <- bind_rows(ross_water_chem, FC_chem)

#save for later use
write_parquet(water_chem, here("data", "collated", "chem", "ROSS_FC_water_chemistry.parquet"))


# ggplot(water_chem%>%
#          filter(site_code %in% c("sfm", "chd", "pbr")), aes(x = Date, y = TOC, color = collector)) +
#   geom_point() +
#   geom_line()+
#   facet_wrap(~site_code, scales = "free_y") +
#   labs(title = "Water Chemistry Data from Ross and FC",
#        x = "Date",
#        y = "TOC (mg/L)") +
#   theme_minimal()

```
