---
title: "City of Fort Collins Turbidity Data QAQC"
output: html_document
---

# Objective

The purpose of this notebook is to QAQC the turbidity data provided to us from 
the City of Fort Collins at their intake/Munroe Canal. Some exploration on this
data set has been done already so we will be applying whatever tools are applicable 
for autoQAQC from `{ross.wq.tools}`.

We will be building on the turbidity QAQC that was started in `scratch/FC_Intake_Explore.Rmd`

# Environment Set Up
```{r}
library(tidyverse)
library(ross.wq.tools)
library(here)
library(plotly)
library(timetk)
```

# Turbidity data collation
```{r}
intake_turb <- read_csv(here("data/raw/fc_intake_data/Intake_Turbidity_2010-2025.csv"),
                        col_names = c("datetime", "intake_turb_NTU"),
                        skip = 1,
                        col_select = c("datetime", "intake_turb_NTU"),
                        show_col_types = F)

munroe_turb <- read_csv(here("data/raw/fc_intake_data/Munroe_Turbidity_2013-2025.csv"),,
                        col_names = c("datetime", "munroe_turb_NTU"),
                        skip = 1,
                        col_select = c("datetime", "munroe_turb_NTU"),
                        show_col_types = F) 

fc_turb <- full_join(intake_turb, munroe_turb, by = "datetime") %>% 
  pivot_longer(cols = c(munroe_turb_NTU, intake_turb_NTU),
               names_to = "location",
               values_to = "NTU") 

rm(intake_turb, munroe_turb)
```

# Manual QAQC

Manual QAQC decisions have already been made in `scratch/FC_Intake_Explore.Rmd`.
Here we want to reproduce those decisions. 

```{r}
fc_turb <- fc_turb %>%
  mutate(NTU = case_when(NTU <= 0 ~ NA,
                         NTU >= 1000 ~ NA,
                         round(NTU, 0) == 655 & location == "munroe_turb_NTU" ~ NA,
                         round(NTU, 0) == 630 & location == "munroe_turb_NTU" ~ NA,
                         round(NTU, 0) == 555 & location == "munroe_turb_NTU" ~ NA,
                         .default = NTU)) %>%
  mutate(flag = case_when(
    # 2011-09-04 to 2011-09-06: both, spike with no flow signal (FC_Intake_Explore.Rmd L613)
    between(datetime, ymd("2011-09-04"), ymd("2011-09-06")) &
      NTU > 10 ~ "fail",
    # 2013-04-29 to 2013-05-01: both, spike before data gap (FC_Intake_Explore.Rmd L665)
    between(datetime, ymd("2013-04-29"), ymd("2013-05-01")) &
      NTU > 10 ~ "fail",
    # 2015-04-06 to 2015-07-16: munroe miscalibrated sensor (entire window) (FC_Intake_Explore.Rmd L767)
    between(datetime, ymd("2015-04-06"), ymd("2015-07-16")) &
      location == "munroe_turb_NTU" ~ "fail",
    # 2015-09-15 to 2016-06-01: munroe winter off-season spikes (FC_Intake_Explore.Rmd L358)
    between(datetime, ymd("2015-09-15"), ymd("2016-06-01")) &
      location == "munroe_turb_NTU" & NTU > 300 ~ "fail",
    # 2016-04-25 to 2016-04-28: munroe spurious near-zero values after data gap (FC_Intake_Explore.Rmd L819)
    between(datetime, ymd("2016-04-25"), ymd("2016-04-28")) &
      location == "munroe_turb_NTU" & NTU < 10 ~ "fail",
    # 2016-10-24 to 2017-04-03: munroe winter off-season spikes (FC_Intake_Explore.Rmd L411)
    between(datetime, ymd("2016-10-24"), ymd("2017-04-03")) &
      location == "munroe_turb_NTU" & NTU > 500 ~ "fail",
    # 2017-11-06 to 2018-04-09: munroe winter off-season spikes (FC_Intake_Explore.Rmd L464)
    between(datetime, ymd("2017-11-06"), ymd("2018-04-09")) &
      location == "munroe_turb_NTU" & NTU > 500 ~ "fail",
    # 2017-12-01 to 2018-04-01: intake winter off-season spikes (FC_Intake_Explore.Rmd L469)
    between(datetime, ymd("2017-12-01"), ymd("2018-04-01")) &
      location == "intake_turb_NTU" & NTU > 200 ~ "fail",
    # 2019-05-15 to 2019-05-18: intake flatlined near zero during runoff (FC_Intake_Explore.Rmd L871)
    between(datetime, ymd("2019-05-15"), ymd("2019-05-18")) &
      location == "intake_turb_NTU" & NTU < 8 ~ "fail",
    # 2020-06-16 to 2020-06-18: munroe spike after data gap (FC_Intake_Explore.Rmd L924)
    between(datetime, ymd("2020-06-16"), ymd("2020-06-18")) &
      location == "munroe_turb_NTU" & NTU > 10 ~ "fail",
    # 2021-07-21 to 2021-07-28: munroe spike (FC_Intake_Explore.Rmd L1029)
    between(datetime, ymd("2021-07-21"), ymd("2021-07-28")) &
      location == "munroe_turb_NTU" & NTU > 10 ~ "fail",
    # 2021-08-10 to 2021-08-13: intake flatlined near zero (FC_Intake_Explore.Rmd L977)
    between(datetime, ymd("2021-08-10"), ymd("2021-08-13")) &
      location == "intake_turb_NTU" & NTU < 5 ~ "fail",
    # 2023-10-23 to 2024-04-08: munroe winter off-season spikes (FC_Intake_Explore.Rmd L523)
    between(datetime, ymd("2023-10-23"), ymd("2024-04-08")) &
      location == "munroe_turb_NTU" & NTU > 600 ~ "fail",
    # 2025-04-02 to 2025-04-02 12:00: munroe artifact at start of record (FC_Intake_Explore.Rmd L1081)
    between(datetime, ymd("2025-04-02"), ymd_hm("2025-04-02 12:00")) &
      location == "munroe_turb_NTU" ~ "fail",
    # 2025-09-25 to 2025-09-28 12:00: munroe artifact (FC_Intake_Explore.Rmd L1133)
    between(datetime, ymd("2025-09-25"), ymd_hm("2025-09-28 12:00")) &
      location == "munroe_turb_NTU" ~ "fail",
    .default = NA
  )) 
```

Now let's visualize these manual decisions

```{r, fig.height=12, fig.width=12}
ggplot(fc_turb, aes(x = datetime, y = NTU)) +
  geom_point(size = 0.5, color = "grey40") +
  geom_point(data = fc_turb %>% filter(flag == "fail"),
             aes(x = datetime, y = NTU),
             shape = 1, size = 2, color = "red") +
  facet_wrap(~ location, nrow = 2, scales = "free_y") +
  labs(x = "", y = "turbidity (NTU)") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Automatic QAQC

Format the data to match what we expect in ross.wq.tools

```{r}
turb_list <- fc_turb %>%
  mutate(parameter = "Turbidity", units = "NTU") %>%
  select(DT_round = datetime, site = location, parameter, mean = NTU, units,
         manual_flag = flag) %>%
  split(f = list(.$site, .$parameter), sep = "-") %>%
  discard(~nrow(.) == 0)
```

Generate summary statistics once for use in both threshold table and flagging.

```{r}
turb_stats <- turb_list %>%
  map(~arrange(., DT_round)) %>%
  map(~generate_summary_statistics(.))
```

Make the seasonal threshold table for each of these sites

```{r}
turb_seasonal_thresholds <- turb_stats %>%
  map(function(df) {
    df %>%
      filter(is.na(manual_flag), !is.na(mean)) %>%
      make_threshold_table()
  }) %>%
  bind_rows()
```

Build the network structure that is expected for network check

```{r}
fc_intake_site_order <- list(
  fc_intake = c("munroe_turb_NTU", "intake_turb_NTU")
)
```

Single parameter QAQC flags

Note: `add_repeat_flag` is not used here because there are many instances of
consecutive repeated values that contain actual data.

Note: `add_drift_flag` is not used here because it splits the record on site
visit flags, which are not available for this dataset.

Note: `add_na_flag` is not used here because missing data is already
represented as NA in the mean column.

```{r}
turb_auto_flagged <- turb_stats %>%
  map(function(df) {
    df %>%
      data.table::data.table() %>%
      add_seasonal_flag(threshold_table = turb_seasonal_thresholds)
  })
```

Network check
```{r}
turb_network_flagged <- turb_auto_flagged %>%
  map(~network_check(df = .,
                     intrasensor_flags_arg = turb_auto_flagged,
                     site_order_arg = fc_intake_site_order)) %>%
  bind_rows()
```

```{r, fig.height=12, fig.width=12}
threshold_lines <- turb_network_flagged %>%
  select(DT_round, site, season) %>%
  left_join(turb_seasonal_thresholds %>%
              select(site, season, t_mean01, t_mean99),
            by = c("site", "season"))

ggplot(turb_network_flagged, aes(x = DT_round, y = mean)) +
  geom_point(size = 0.5, color = "grey40") +
  geom_point(data = turb_network_flagged %>% filter(!is.na(auto_flag)),
             aes(x = DT_round, y = mean),
             shape = 1, size = 2, color = "red") +
  geom_line(data = threshold_lines, aes(x = DT_round, y = t_mean99),
            color = "blue", linewidth = 0.5) +
  geom_line(data = threshold_lines, aes(x = DT_round, y = t_mean01),
            color = "blue", linewidth = 0.5, linetype = "dashed") +
  facet_wrap(~ site, nrow = 2, scales = "free_y") +
  labs(x = "", y = "turbidity (NTU)") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Why `ross.wq.tools` does not work well for this dataset

`ross.wq.tools` was designed for in-stream continuous monitoring sondes in a
river network with mWater field visit records. This dataset violates several of
those assumptions:

**Slope-based flagging (`add_seasonal_flag` slope violation)**

`make_threshold_table` computes p99 of positive slopes and values from the historical record, 
pooled across all years within each season. For in-stream turbidity, the dominant 
source of large slopes and values are real runoff events. Because the
threshold is derived from the same data that contains real events, the p99 cutoff
is set by the most extreme real events and still flags the next tier of real
events below them. There is no way to separate sensor artifacts from real
hydrological signal using slope alone without site visit records to split the
record on cleaning events.

**Drift flagging (`add_drift_flag`)**

`add_drift_flag` splits the time series on site visit flags before computing
rolling R² values. Without site visit records, the entire 15-year record is
treated as one continuous window, making the drift detection unreliable.

**Repeat flagging (`add_repeat_flag`)**

At low turbidity during baseflow, the sensor resolution causes many consecutive
identical readings that are real measurements, not stuck-sensor artifacts.
`add_repeat_flag` cannot distinguish these from true sensor malfunctions.

**Network check (`network_check`)**

`network_check` was designed for a linear stream network where upstream and
downstream sites share the same water parcel with a predictable travel time and 
benefits from having an upstream and downstream site for every site. 

# Trying {timetk} anomalize

This also did not work, but it was a suggestion from Sam.

```{r}
turb_anomalize <- fc_turb %>%
  mutate(NTU = if_else(!is.na(flag) & flag == "fail", NA_real_, NTU)) %>%
  filter(!is.na(NTU), NTU > 0) %>%
  group_by(location) %>%
  timetk::anomalize(
    .date_var    = datetime,
    .value       = NTU,
    .frequency   = "auto",
    .trend       = "auto",
    .method      = "stl",
    .iqr_alpha   = 0.05,
    .max_anomalies = 0.05
  )
```

```{r, fig.height=12, fig.width=12}
# Plotting this work
turb_anomalize %>%
  timetk::plot_anomalies(
    .date_var    = datetime,
    .facet_ncol  = 1,
    .interactive = FALSE
  ) +
  labs(x = "", y = "turbidity (NTU)") +
  theme_bw()
```

