---
title: "City of Fort Collins Turbidity Data QAQC"
output: html_document
---

# Objective

The purpose of this notebook is to QAQC the turbidity data provided to us from 
the City of Fort Collins at their intake/Munroe Canal. Some exploration on this
data set has been done already so we will be applying whatever tools are applicable 
for autoQAQC from `{ross.wq.tools}`.

We will be building on the turbidity QAQC that was started in `scratch/FC_Intake_Explore.Rmd`

# Environment Set Up
```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(ross.wq.tools)
library(here)
library(plotly)
library(timetk)
library(changepoint)
library(ggchangepoint)
```

# Turbidity data collation
```{r}
intake_turb <- read_csv(here("data/raw/fc_intake_data/Intake_Turbidity_2010-2025.csv"),
                        col_names = c("datetime", "intake_turb_NTU"),
                        skip = 1,
                        col_select = c("datetime", "intake_turb_NTU"),
                        show_col_types = F)

munroe_turb <- read_csv(here("data/raw/fc_intake_data/Munroe_Turbidity_2013-2025.csv"),,
                        col_names = c("datetime", "munroe_turb_NTU"),
                        skip = 1,
                        col_select = c("datetime", "munroe_turb_NTU"),
                        show_col_types = F) 

fc_turb <- full_join(intake_turb, munroe_turb, by = "datetime") %>% 
  pivot_longer(cols = c(munroe_turb_NTU, intake_turb_NTU),
               names_to = "location",
               values_to = "NTU") 

rm(intake_turb, munroe_turb)
```

# Manual QAQC

Manual QAQC decisions have already been made in `scratch/FC_Intake_Explore.Rmd`.
Here we want to reproduce those decisions. 

```{r}
fc_turb <- fc_turb %>%
  mutate(NTU = case_when(NTU <= 0 ~ NA,
                         NTU >= 1000 ~ NA,
                         round(NTU, 0) == 655 & location == "munroe_turb_NTU" ~ NA,
                         round(NTU, 0) == 630 & location == "munroe_turb_NTU" ~ NA,
                         round(NTU, 0) == 555 & location == "munroe_turb_NTU" ~ NA,
                         .default = NTU)) %>%
  mutate(flag = case_when(
    # 2011-09-04 to 2011-09-06: both, spike with no flow signal (FC_Intake_Explore.Rmd L613)
    between(datetime, ymd("2011-09-04"), ymd("2011-09-06")) &
      NTU > 10 ~ "fail",
    # 2013-04-29 to 2013-05-01: both, spike before data gap (FC_Intake_Explore.Rmd L665)
    between(datetime, ymd("2013-04-29"), ymd("2013-05-01")) &
      NTU > 10 ~ "fail",
    # 2015-04-06 to 2015-07-16: munroe miscalibrated sensor (entire window) (FC_Intake_Explore.Rmd L767)
    between(datetime, ymd("2015-04-06"), ymd("2015-07-16")) &
      location == "munroe_turb_NTU" ~ "fail",
    # 2015-09-15 to 2016-06-01: munroe winter off-season spikes (FC_Intake_Explore.Rmd L358)
    between(datetime, ymd("2015-09-15"), ymd("2016-06-01")) &
      location == "munroe_turb_NTU" & NTU > 300 ~ "fail",
    # 2016-04-25 to 2016-04-28: munroe spurious near-zero values after data gap (FC_Intake_Explore.Rmd L819)
    between(datetime, ymd("2016-04-25"), ymd("2016-04-28")) &
      location == "munroe_turb_NTU" & NTU < 10 ~ "fail",
    # 2016-10-24 to 2017-04-03: munroe winter off-season spikes (FC_Intake_Explore.Rmd L411)
    between(datetime, ymd("2016-10-24"), ymd("2017-04-03")) &
      location == "munroe_turb_NTU" & NTU > 500 ~ "fail",
    # 2017-11-06 to 2018-04-09: munroe winter off-season spikes (FC_Intake_Explore.Rmd L464)
    between(datetime, ymd("2017-11-06"), ymd("2018-04-09")) &
      location == "munroe_turb_NTU" & NTU > 500 ~ "fail",
    # 2017-12-01 to 2018-04-01: intake winter off-season spikes (FC_Intake_Explore.Rmd L469)
    between(datetime, ymd("2017-12-01"), ymd("2018-04-01")) &
      location == "intake_turb_NTU" & NTU > 200 ~ "fail",
    # 2019-05-15 to 2019-05-18: intake flatlined near zero during runoff (FC_Intake_Explore.Rmd L871)
    between(datetime, ymd("2019-05-15"), ymd("2019-05-18")) &
      location == "intake_turb_NTU" & NTU < 8 ~ "fail",
    # 2020-06-16 to 2020-06-18: munroe spike after data gap (FC_Intake_Explore.Rmd L924)
    between(datetime, ymd("2020-06-16"), ymd("2020-06-18")) &
      location == "munroe_turb_NTU" & NTU > 10 ~ "fail",
    # 2021-07-21 to 2021-07-28: munroe spike (FC_Intake_Explore.Rmd L1029)
    between(datetime, ymd("2021-07-21"), ymd("2021-07-28")) &
      location == "munroe_turb_NTU" & NTU > 10 ~ "fail",
    # 2021-08-10 to 2021-08-13: intake flatlined near zero (FC_Intake_Explore.Rmd L977)
    between(datetime, ymd("2021-08-10"), ymd("2021-08-13")) &
      location == "intake_turb_NTU" & NTU < 5 ~ "fail",
    # 2023-10-23 to 2024-04-08: munroe winter off-season spikes (FC_Intake_Explore.Rmd L523)
    between(datetime, ymd("2023-10-23"), ymd("2024-04-08")) &
      location == "munroe_turb_NTU" & NTU > 600 ~ "fail",
    # 2025-04-02 to 2025-04-02 12:00: munroe artifact at start of record (FC_Intake_Explore.Rmd L1081)
    between(datetime, ymd("2025-04-02"), ymd_hm("2025-04-02 12:00")) &
      location == "munroe_turb_NTU" ~ "fail",
    # 2025-09-25 to 2025-09-28 12:00: munroe artifact (FC_Intake_Explore.Rmd L1133)
    between(datetime, ymd("2025-09-25"), ymd_hm("2025-09-28 12:00")) &
      location == "munroe_turb_NTU" ~ "fail",
    .default = NA
  )) 
```

Now let's visualize these manual decisions

```{r, fig.height=12, fig.width=12}
ggplot(fc_turb, aes(x = datetime, y = NTU)) +
  geom_point(size = 0.5, color = "grey40") +
  geom_point(data = fc_turb %>% filter(flag == "fail"),
             aes(x = datetime, y = NTU),
             shape = 1, size = 2, color = "red") +
  facet_wrap(~ location, nrow = 2, scales = "free_y") +
  labs(x = "", y = "turbidity (NTU)") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Automatic QAQC

Format the data to match what we expect in ross.wq.tools

```{r}
turb_list <- fc_turb %>%
  mutate(parameter = "Turbidity", units = "NTU") %>%
  select(DT_round = datetime, site = location, parameter, mean = NTU, units,
         manual_flag = flag) %>%
  split(f = list(.$site, .$parameter), sep = "-") %>%
  discard(~nrow(.) == 0)
```

Generate summary statistics once for use in both threshold table and flagging.

```{r}
turb_stats <- turb_list %>%
  map(~arrange(., DT_round)) %>%
  map(~filter(., is.na(manual_flag))) %>%
  map(~generate_summary_statistics(.))
```

Make the seasonal threshold table for each of these sites

```{r}
turb_seasonal_thresholds <- turb_stats %>%
  map(function(df) {
    df %>%
      filter(is.na(manual_flag), !is.na(mean)) %>%
      make_threshold_table()
  }) %>%
  bind_rows()
```

Build the network structure that is expected for network check

```{r}
fc_intake_site_order <- list(
  fc_intake = c("munroe_turb_NTU", "intake_turb_NTU")
)
```

Single parameter QAQC flags

Note: `add_repeat_flag` is not used here because there are many instances of
consecutive repeated values that contain actual data.

Note: `add_drift_flag` is not used here because it splits the record on site
visit flags, which are not available for this dataset.

Note: `add_na_flag` is not used here because missing data is already
represented as NA in the mean column.

```{r}
turb_auto_flagged <- turb_stats %>%
  map(function(df) {
    df %>%
      data.table::data.table() %>%
      add_seasonal_flag(threshold_table = turb_seasonal_thresholds)
  })
```

Do the network check
```{r}
turb_network_flagged <- turb_auto_flagged %>%
  map(~network_check(df = .,
                     intrasensor_flags_arg = turb_auto_flagged,
                     site_order_arg = fc_intake_site_order)) %>%
  bind_rows()
```

```{r, fig.height=12, fig.width=12}
threshold_lines <- turb_network_flagged %>%
  select(DT_round, site, season) %>%
  left_join(turb_seasonal_thresholds %>%
              select(site, season, t_mean01, t_mean99),
            by = c("site", "season"))

ggplot(turb_network_flagged, aes(x = DT_round, y = mean)) +
  geom_point(size = 0.5, color = "grey40") +
  geom_point(data = turb_network_flagged %>% filter(!is.na(auto_flag)),
             aes(x = DT_round, y = mean),
             shape = 1, size = 2, color = "red") +
  geom_line(data = threshold_lines, aes(x = DT_round, y = t_mean99),
            color = "blue", linewidth = 0.5) +
  geom_line(data = threshold_lines, aes(x = DT_round, y = t_mean01),
            color = "blue", linewidth = 0.5, linetype = "dashed") +
  facet_wrap(~ site, nrow = 2, scales = "free_y") +
  labs(x = "", y = "turbidity (NTU)") +
  theme_bw() +
  theme(legend.position = "bottom")
```


# Trying {timetk} anomalize

This also did not work, but it was a suggestion from Sam.

```{r}
turb_anomalize <- fc_turb %>%
  mutate(NTU = if_else(!is.na(flag) & flag == "fail", NA_real_, NTU)) %>%
  filter(!is.na(NTU), NTU > 0) %>%
  group_by(location) %>%
  timetk::anomalize(
    .date_var    = datetime,
    .value       = NTU,
    .frequency   = "auto",
    .trend       = "auto",
    .method      = "stl",
    .iqr_alpha   = 0.05,
    .max_anomalies = 0.05
  )
```

```{r, fig.height=12, fig.width=12}
# Plotting this work
turb_anomalize %>%
  timetk::plot_anomalies(
    .date_var    = datetime,
    .facet_ncol  = 1,
    .interactive = FALSE
  ) +
  labs(x = "", y = "turbidity (NTU)") +
  theme_bw()
```

#### Parameter sweep for anomalize

Three levers: `.iqr_alpha` and `.max_anomalies` control flagging sensitivity; `.trend` controls how wide a window is used to extract the trend before anomaly detection runs on the remainder — too wide and drift gets absorbed, too narrow and storm events bleed into the remainder. (`.method = "twitter"` is documented in the timetk docs but not yet implemented — only `"stl"` works.)

```{r}
anomalize_params <- list(
  # baseline
  list(method = "stl",     iqr_alpha = 0.05, max_anoms = 0.05, trend = "auto"),
  # loosen alpha/max_anoms with stl
  list(method = "stl",     iqr_alpha = 0.20, max_anoms = 0.20, trend = "auto"),
  list(method = "stl",     iqr_alpha = 0.50, max_anoms = 0.50, trend = "auto"),
  list(method = "stl",     iqr_alpha = 0.75, max_anoms = 0.75, trend = "auto"),
  # vary trend window with mid alpha
  list(method = "stl",     iqr_alpha = 0.20, max_anoms = 0.20, trend = "3 months"),
  list(method = "stl",     iqr_alpha = 0.20, max_anoms = 0.20, trend = "1 year"),
  # longer trend windows with high alpha
  list(method = "stl",     iqr_alpha = 0.50, max_anoms = 0.50, trend = "3 months"),
  list(method = "stl",     iqr_alpha = 0.50, max_anoms = 0.50, trend = "1 year"),
  list(method = "stl",     iqr_alpha = 0.75, max_anoms = 0.75, trend = "3 months"),
  list(method = "stl",     iqr_alpha = 0.75, max_anoms = 0.75, trend = "1 year")
)

turb_anomalize_sweep <- anomalize_params %>%
  map(function(p) {
    fc_turb %>%
      mutate(NTU = if_else(!is.na(flag) & flag == "fail", NA_real_, NTU)) %>%
      filter(!is.na(NTU), NTU > 0) %>%
      group_by(location) %>%
      timetk::anomalize(
        .date_var      = datetime,
        .value         = NTU,
        .frequency     = "auto",
        .trend         = p$trend,
        .method        = p$method,
        .iqr_alpha     = p$iqr_alpha,
        .max_anomalies = p$max_anoms
      ) %>%
      mutate(method = p$method, iqr_alpha = p$iqr_alpha,
             max_anoms = p$max_anoms, trend = p$trend)
  }) %>%
  bind_rows() %>%
  mutate(params = paste0(method, " | alpha=", iqr_alpha,
                         " / max=", max_anoms, " / trend=", trend))
```

```{r}
turb_anomalize_sweep %>%
  group_by(location, params) %>%
  summarise(
    n_total     = n(),
    n_anomaly   = sum(anomaly == "Yes", na.rm = TRUE),
    pct_flagged = round(n_anomaly / n_total * 100, 2),
    .groups = "drop"
  )
```

```{r, fig.height=30, fig.width=12}
turb_anomalize_sweep %>%
  ggplot(aes(x = datetime, y = observed)) +
  geom_point(size = 0.3, color = "grey50") +
  geom_point(
    data = ~filter(., anomaly == "Yes"),
    aes(x = datetime, y = observed),
    shape = 1, size = 2, color = "red"
  ) +
  facet_grid(params ~ location, scales = "free_y") +
  labs(x = "", y = "turbidity (NTU)",
       title = "anomalize sensitivity sweep") +
  theme_bw(base_size = 9)
```

# Changepoint detection for sensor cleaning events

Sensor cleanings produce abrupt mean-shifts in turbidity. With no field visit records to split on, changepoint detection on mean NTU is a candidate for inferring when cleanings occurred. Using `{changepoint}` with PELT (Pruned Exact Linear Time) to find multiple changepoints per location.

Aggregate to daily median before running changepoint detection. At sub-hourly resolution PELT reacts to every storm spike; daily medians suppress that noise so only persistent level shifts (cleanings) survive.

```{r}
turb_daily_cp <- fc_turb %>%
  mutate(NTU = if_else(!is.na(flag) & flag == "fail", NA_real_, NTU)) %>%
  filter(!is.na(NTU), NTU > 0) %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(location, date) %>%
  summarise(NTU_median = median(NTU, na.rm = TRUE), .groups = "drop") %>%
  arrange(location, date)
```

Fit PELT on daily medians with `minseglen = 14` (two weeks minimum between changepoints) and an MBIC penalty, which is stricter than BIC and favors fewer, more confident changepoints.

```{r}
cp_results <- turb_daily_cp %>%
  split(.$location) %>%
  map(function(df) {
    cpt_fit <- cpt.mean(df$NTU_median, method = "PELT", penalty = "MBIC", minseglen = 14)
    cp_indices <- cpts(cpt_fit)
    if (length(cp_indices) == 0) return(tibble())
    tibble(
      location    = df$location[[1]],
      cp_date     = df$date[cp_indices],
      mean_before = map_dbl(cp_indices, ~mean(df$NTU_median[max(1, . - 14):.])),
      mean_after  = map_dbl(cp_indices, ~mean(df$NTU_median[.:(min(nrow(df), . + 14))]))
    )
  }) %>%
  bind_rows() %>%
  mutate(mean_shift = mean_after - mean_before)

cp_results
```

Build background rectangles — one per segment between changepoints, colored by the direction of the shift that started that segment.

```{r}
date_range <- turb_daily_cp %>%
  group_by(location) %>%
  summarise(start = min(date), end = max(date), .groups = "drop")

cp_rects <- cp_results %>%
  arrange(location, cp_date) %>%
  group_by(location) %>%
  mutate(
    seg_start = cp_date,
    seg_end   = lead(cp_date, default = date_range$end[date_range$location == location[[1]]]),
    shift_dir = if_else(mean_shift < 0, "down", "up")
  ) %>%
  ungroup()
```

Plot changepoints as colored background regions over the daily median series.

```{r, fig.height=12, fig.width=12}
ggplot(turb_daily_cp, aes(x = date, y = NTU_median)) +
  geom_rect(
    data = cp_rects,
    aes(xmin = seg_start, xmax = seg_end, ymin = -Inf, ymax = Inf, fill = shift_dir),
    inherit.aes = FALSE, alpha = 0.25
  ) +
  geom_line(linewidth = 0.3, color = "grey30") +
  scale_fill_manual(values = c("down" = "steelblue", "up" = "tomato")) +
  facet_wrap(~ location, nrow = 2, scales = "free_y") +
  labs(x = "", y = "daily median turbidity (NTU)", fill = "shift direction",
       title = "PELT changepoints on daily medians (MBIC penalty, minseglen = 14 days)") +
  theme_bw() +
  theme(legend.position = "bottom")
```

Seasonality of downward shifts — if cleaning events are real they should cluster in spring/fall maintenance windows.

```{r, fig.height=6, fig.width=10}
cp_results %>%
  filter(mean_shift < 0) %>%
  mutate(month = lubridate::month(cp_date, label = TRUE)) %>%
  ggplot(aes(x = month)) +
  geom_bar() +
  facet_wrap(~ location) +
  labs(x = "month of changepoint", y = "count",
       title = "Seasonality of downward mean-shifts (candidate cleanings)") +
  theme_bw()
```

# Why `ross.wq.tools` does not work well for this dataset

`ross.wq.tools` was designed for in-stream continuous monitoring sondes in a
river network with mWater field visit records. This dataset violates several of
those assumptions:

**Slope-based flagging (`add_seasonal_flag` slope violation)**

`make_threshold_table` computes p99 of positive slopes and values from the historical record,
pooled across all years within each season. For in-stream turbidity, the dominant
source of relatively large slopes and values are real runoff events, since the majority of 
turbidity data we received was low-slope and low-value. Because the
threshold is derived from data that contains real events and errors that look
very similar statistically, the p99 cutoff is set by both the most extreme real events 
and errors in readings. Since we are unable to separate out these two signals for 
flagging the threshold table approach we have used for our own network does not 
work well here. I will note here that the thresholds in the PWQN probably work 
better because the data that we use to generate those thresholds are very well 
vetted and manually verified. 

TLDR: There is no way to separate sensor artifacts from real
hydrological signal using slope alone without site visit records or additional sensors
(e.g. Depth or SC), AND most of this data is dominated by low slopes and low values,
so p99 captures small deviations from those patterns.

**Drift flagging (`add_drift_flag`)**

`add_drift_flag` splits the time series on site visit flags before computing
rolling R² values. Without site visit records, the entire 15-year record is
treated as one continuous window, making the drift detection impossible with 
our current methods. 

An idea going forward: Some changepoint approach could probably be added to the 
autoQAQC to make decisions about that. Comparison to up/downstream data would also 
be useful here, but some idea about how correlated these data sets are temporally
would be a more "correct" thing to think about when building this out. 

**Repeat flagging (`add_repeat_flag`)**

At low turbidity during baseflow, the sensor resolution causes many consecutive
identical readings that are real measurements, not stuck-sensor artifacts.
`add_repeat_flag` cannot distinguish these from true sensor malfunctions.

**Network check (`network_check`)**

`network_check` was designed for a linear stream network where upstream and
downstream sites share the same water parcel with a predictable travel time and
benefits from having an upstream and downstream site for every site. 

We could in theory plug this into the PWQN, I personally just didn't spend the time
to do that for this exercise since I am not super familiar with these data streams.

**Other issues that were resolved in this script, but should be cleaned up in the package to make our life easier in the future for AutoQAQC of other data streams**

## Explicit assumptions (stated in code or docs)

**15-minute sampling interval is hardcoded.**

`generate_summary_statistics` divides slopes by `15` (`(front1 - mean)/15`).
`network_check` uses `width_fun = 17` ("17 observations at 15-minute intervals = 4-hour window").
`add_drift_flag` uses `win_short <- 96L` and `win_long <- 288L` (~1 day and ~3 days at 15-min).
All data is not guaranteed to be regular 15-minute intervals.

**Site visit / field visit records are required.**

`add_drift_flag` explicitly splits the time series on site visit flags before computing rolling R²
— without them, the entire 15-year record is one unbroken window and drift detection is unreliable.

This is actually a huge thing that should be looked into when working on drift detection
and correction.

**Linear stream network with upstream/downstream relationships.**

`network_check` is designed for a network where a water parcel flows from an upstream site to a
downstream site with predictable travel time. I am unsure of the relationship with
Munroe Canal and the Intake. 

I will also note here that some temporal correlation relationship can be used to 
make the network check more reliable.

**One site and one parameter per dataframe.**

`add_seasonal_flag` and `generate_summary_statistics` both call `unique(na.omit(df$site))` and
`unique(na.omit(df$parameter))` — they assume a pre-split list structure, which required a manual
`split()` + `discard()` step. Major headache that should be removed from the package. 
The package should have some input structure expectation to aid in generalizability.

**Specific column naming convention required.**

The package expects `DT_round`, `site`, `parameter`, `mean`, `flag`, `manual_flag`. The FC data
came in as `datetime`, `location`, `NTU`, requiring explicit renaming. Another 
point that should be updated to make it easier to interact with different data structures 
in the future.

**mWater is the source of field records.**

`grab_mWater_cleaning_notes`, `load_mWater`, etc. are all hardcoded to mWater. There is no path
for non-mWater field records and no way to incorporate field note data that does 
not jibe with what we've made. 

## Implicit assumptions (baked into logic, not documented)

**Thresholds can be separated from real signal.**

`make_threshold_table` in this case uses non-manually-flagged data to compute p99 value and slope thresholds.
For in-stream turbidity, real storm events *are* the extreme values, so p99 is set by the largest
real events — the next tier of real events will always be flagged.

**Repeated values are sensor artifacts, not real data.**

`add_repeat_flag` flags any value equal to its immediate neighbor. At low turbidity during
baseflow, sensor resolution produces many legitimate consecutive identical readings.

**Temperature and Depth are co-measured at every site.**

`intersensor_check` validates slope violations by cross-referencing concurrent Temperature and
Depth flags. Without those parameters, the cross-validation step silently does nothing useful.

**Simultaneous flags at neighboring sites = real event.**

`network_check` removes flags when they overlap within a 4-hour rolling window at an upstream or
downstream site. For FC, relationship between Munroe and Intake is unknown (to me), though
it is safe to assume that they behave the same way. 

**NAs represent gaps in an otherwise continuous record.**

`add_na_flag` flags NAs as "missing data." For FC, NAs were already used to remove out-of-range
values (≤0, ≥1000, stuck values) — flagging them as missing data conflates two different data
quality issues.

**Season definitions match the ROSS network's hydrological regime.**

`generate_summary_statistics` hardcodes `winter_baseflow` (Dec–Apr), `snowmelt` (May–Jun),
`monsoon` (Jul–Sep), `fall_baseflow` (Oct–Nov). These reflect a mountain stream watershed and
may not match other intake turbidity dynamics.

**`data.table` format is required before `add_seasonal_flag`.**

The notebook had to explicitly call `data.table::data.table()` before passing data to
`add_seasonal_flag` — this dependency is not documented in the function's `@param` or examples.

# Next Steps for `{ross.wq.tools}`

The issues encountered in this workflow point to several concrete improvements needed in the
package to support data streams outside the PWQN. These are grouped by theme.

## Generalization

**Decouple the hardcoded 15-minute interval.**

The slope calculation in `generate_summary_statistics` (`(front1 - mean)/15`), the rolling window
sizes in `add_drift_flag` (`win_short = 96L`, `win_long = 288L`), and the network check window
(`width_fun = 17`) all assume 15-minute data. These should be derived from the actual timestamp
spacing in the data, or exposed as user-configurable arguments, so the package works with any
sampling frequency.

**Make season definitions configurable.**

`generate_summary_statistics` hardcodes four seasons tied to the Cache la Poudre hydrological
regime. This should be a user-supplied lookup table or argument so the package can be applied to
watersheds with different seasonal dynamics.

**Decouple field records from mWater.**

`add_drift_flag`, `add_field_flag`, and `add_malfunction_flag` all depend on mWater-formatted
field records. There is currently no path for datasets without mWater records. The package should
accept a generic field notes dataframe with a documented schema, with mWater being one way to
populate it — not the only way.

**Standardize but do not enforce column names.**

The package silently requires `DT_round`, `site`, `parameter`, `mean`, `flag`. Working with
external data requires manual renaming before any package function can be called. The package
should either accept a column mapping argument or provide a documented helper that validates and
renames input data to the expected schema.

## Input validation and documentation

**Document the `data.table` input requirement for `add_seasonal_flag`.**

The function silently requires a `data.table` object but this is not stated in the `@param`
documentation. Either document it explicitly or coerce internally so the caller does not need to
know.

**Replace disabled examples with `@examplesIf` or `\donttest{}`.**

All flagging functions currently have `# Examples are temporarily disabled`. These should use
`@examplesIf interactive()` so they run in `R CMD check` where possible, catching breaking changes
earlier.

**Add early input validation.**

Functions currently rely on implicit downstream errors when inputs are malformed. Adding
`stopifnot()` checks (or `cli::cli_abort()` for messages with runtime values) for required columns
and correct data types would make errors easier to diagnose.

## Flagging logic

**Revisit `add_repeat_flag` for low-resolution sensors.**

The current logic flags any value equal to its immediate neighbor as a stuck sensor. For sensors
with coarse resolution measuring stable parameters (e.g., low turbidity at baseflow), this
produces many false positives. Consider requiring a minimum run length (e.g., N consecutive
identical values) before flagging.

**Separate value removal from missingness in `add_na_flag`.**

NAs in external datasets may represent values that were already removed during manual QAQC, not
true data gaps. `add_na_flag` conflates these two cases. The package should either distinguish
them via a separate removal flag or document this limitation clearly so users know to handle it
upstream.

**Make `intersensor_check` gracefully handle missing parameters.**

The function assumes Temperature and Depth are always present. When they are absent it silently
does nothing. It should either warn the user or skip gracefully with a message.

## Network check

**Incorporate temporal lag into `network_check`.**

The current implementation checks for flag overlap within a fixed ±2-hour window, assuming
simultaneous response across sites. A cross-correlation or user-supplied travel-time offset would
make the check more reliable for networks where upstream/downstream sites do not respond
synchronously.
