---
title: "Creating a Baseline Chlorophyll-a Model"
output: html_document
---

# Purpose

The purpose of this markdown is meant to create a baseline chl-a model across the 
mainstem of the upper CLP.

```{r, echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
library(readxl)
library(here)
library(mgcv)
library(Metrics)
library(sf)
library(nhdplusTools)
library(mapview)
library(arrow)
library(ross.wq.tools)
library(cdssr)
library(scales)
library(snotelr)
# remotes::install_github("rossyndicate/ross.wq.tools")
library(tidyverse)

version_date = Sys.Date()
source(here("src/collation/load_grab_sample_data.r"))
```

# Chlorophyll-a Grab Samples

Get all chl-a grab samples from our collated data set (`load_grab_sample_data()`)

```{r}
# grab some specific columns
grab_chla <- load_grab_sample_data() %>% 
  select(site_code, site_name, date = Date,
         ChlA, Field_DO_mgL:distance_upstream_km, Lat:flag) %>% 
  filter(!is.na(ChlA), !Campaign %in% c("Mainstem - Lower", "Lower Tributary")) %>% 
  mutate(year = year(date))

# check for flags
unique(grab_chla$flag)
```

Plot the data to see what it looks like

```{r}
ggplot(grab_chla, aes(x = yday(date), y = ChlA, color = factor(year))) +
  geom_point()+
  facet_grid(location_type ~ ., scales = "free_y") +
  theme_bw()

ggplot(grab_chla, aes(x = yday(date), y = ChlA, color = factor(year))) +
  geom_point()+
  facet_grid(Campaign ~ ., scales = "free_y") +
  theme_bw()

```

And also by site/location type

```{r}
walk(unique(grab_chla$location_type), function(loc) {
  
  gg <- grab_chla %>% 
    filter(location_type == loc) %>% 
    ggplot(. , aes(x = yday(date), y = ChlA, color = factor(year(date)))) +
    geom_point() +
    labs(title = loc) +
    facet_grid(site_name ~ ., scales = "free_y")
  print(gg)
  
})

```

And look at just the mainstem upper poudre, which is what we care about

```{r}
grab_chla %>% 
  filter(Campaign == "Mainstem - Upper") %>% 
  ggplot(. , aes(x = yday(date), 
                 y = ChlA, 
                 color = site_code)) +
  facet_grid(year(date) ~ .) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_d()

```

And we'll add q here, since we know there is a temp/q connection:

```{r}
#Read in flow data
flow_data <- get_sw_ts(
  abbrev = "CLAFTCCO",
  start_date = min(grab_chla$date, na.rm = T),
  end_date = max(grab_chla$date, na.rm = T),
  timescale = "day"
) %>%
  rename(canyon_mouth_cfs = value) %>%
  mutate(canyon_mouth_cfs_rescale = rescale(canyon_mouth_cfs, to = c(0,10)))%>%
  select(datetime, canyon_mouth_cfs, canyon_mouth_cfs_rescale) %>% 
  mutate(date = date(datetime)) %>% 
  summarise(mean_cfs = mean(canyon_mouth_cfs),
            mean_rescale = mean(canyon_mouth_cfs_rescale),
            .by = date)

ggplot() +
  geom_area(data = flow_data %>% 
              arrange(date) %>%
              mutate(year = year(date)) %>% 
              filter(year > 2021),
            aes(x = yday(date), y = mean_rescale, group = 1),
            fill = "gray70",
            alpha = 0.5) +
  geom_point(data = grab_chla %>%
               filter(Campaign == "Mainstem - Upper") %>% mutate(year = year(date)),
             aes(x = yday(date),
                 y = ChlA,
                 color = site_code)) +
  facet_grid(year(date) ~ .) +
  theme_bw() +
  scale_color_viridis_d()

```

And let's look at snotel

```{r}
site_meta <- snotel_info()
huc8_code <- "10190007"
sites_in_huc8 <- site_meta %>%
  filter(grepl(huc8_code, description))
# Extract site IDs to a vector
site_ids <- sites_in_huc8$site_id

# Download daily SNOTEL data
snotel_data <- snotel_download(site_id = site_ids,
                               internal = TRUE,
                               metric = TRUE) %>% 
  mutate(date = ymd(date))

snotel_data <- snotel_data %>% 
  mutate(doy = yday(date),
         year = year(date))

snotel_data %>% 
  filter(between(year, 2022, 2025)) %>% 
  ggplot(., aes(x = doy, y = snow_water_equivalent, color = factor(site_id))) +
  geom_line() +
  facet_grid(year ~ .)
labs(title = paste("SWE for HUC 8", huc8_code),
     x = "Date",
     y = "Snow Water Equivalent (mm)") +
  theme_minimal()
```

And add it in without chl-a data:

```{r}
ggplot() +
  geom_area(data = flow_data %>% 
              arrange(date) %>%
              mutate(year = year(date)) %>% 
              filter(year > 2021),
            aes(x = yday(date), y = mean_rescale, group = 1),
            fill = "gray70",
            alpha = 0.5) +
  geom_line(data = snotel_data %>%
              filter(between(year, 2022, 2025)) %>%
              mutate(swe_scaled = rescale(snow_water_equivalent, c(0, 10))),
            aes(x = doy, y = swe_scaled, lty = factor(site_name))) +
  geom_point(data = grab_chla %>%
               filter(Campaign == "Mainstem - Upper") %>% mutate(year = year(date)),
             aes(x = yday(date),
                 y = ChlA,
                 color = site_name)) +
  labs(x = "day of year", 
       y = "chlorophyll a (ug/L)\nrescaled Q at canyon mouth (grey)\nrescaled SWE (lines)",
       color = "chl-a sample site",
       lty = "snotel site") +
  facet_grid(year(date) ~ .) +
  theme_bw() +
  scale_color_viridis_d() 

```

Given that there is also a presumed relationship between temp and discharge and
chlorophyl-a, let's look at that:

```{r}
temp_q <- snotel_data %>%
  filter(between(year, 2022, 2025)) %>%
  summarize(mean_temp = mean(temperature_mean, na.rm = T), 
            .by = c(date, doy, year)) %>% 
  left_join(., flow_data) %>% 
  mutate(temp_q = mean_temp/mean_cfs) %>% 
  filter(!is.na(temp_q))

ggplot() +
  geom_area(data = temp_q,
            aes(x = doy, y = rescale(temp_q, c(0,10)), group = 1),
            fill = "gray70",
            alpha = 0.5) +
  geom_point(data = grab_chla %>%
               filter(Campaign == "Mainstem - Upper") %>% mutate(year = year(date)),
             aes(x = yday(date),
                 y = ChlA,
                 color = site_name)) +
  labs(x = "day of year", 
       y = "chlorophyll a (ug/L)\nrescaled temp/Q (grey)",
       color = "monitoring site") +
  facet_grid(year(date) ~ .) +
  theme_bw() +
  scale_color_viridis_d() 
```

This makes me think the driving force is not temp AND q related, but is more 
dominantly controlled by Q and the connection of how water moves through the system.
The hypothesis here, is that when Q is high, it's moving through high-elevation
reservoirs and it "scrapes" algal growth from the outflow and it then travels
down the river. 

Given that, let's look at at some of the data from sensor networks that have 
data just below reservoirs or that are inputs that may help us understand where 
some of the higher values in chlorophyll-a might be coming from.

```{r}
stations <- tribble(
  ~abbrev, ~name,
  "JOEBELCO", "JW below JW",
  "JWCCHACO", "JW below Chambers",
  "LAPLODCO", "Poudre below Long Draw",
  "LAPTUNCO", "Laramie Tunnel",
  "CLANSECO", "North Fork"
)

streamflow_data <- map(stations$abbrev, 
                       function(station){
                         get_sw_ts(abbrev = station, 
                                   start_date = "2022-01-01", 
                                   end_date = "2025-12-31")
                       }) %>% 
  bind_rows()  %>% 
  mutate(date = ymd(datetime)) %>% 
  select(date, abbrev, value, meas_unit) %>% 
  mutate(value = if_else(value == -999,
                         NA,
                         value),
         doy = yday(date),
         year = year(date)) %>% 
  full_join(., stations) 
  

ggplot(streamflow_data, aes(x = doy, y = value, lty = name)) +
  geom_line() +
  facet_grid(year ~ .) +
  theme_bw()
```

```{r}
ggplot() +
  geom_line(data = streamflow_data %>% 
              mutate(cfs_rescaled = rescale(value, c(0,10))),
            aes(x = doy, y = cfs_rescaled, lty = name)) +
  geom_point(data = grab_chla %>%
               filter(Campaign == "Mainstem - Upper") %>% 
               mutate(year = year(date)),
             aes(x = yday(date),
                 y = ChlA,
                 color = site_name)) +
  labs(x = "day of year", 
       y = "chlorophyll a (ug/L)\nrescaled temp/Q (grey)",
       color = "monitoring site") +
  facet_grid(year ~ .) +
  theme_bw() +
  scale_color_viridis_d() 

```

Well, maybe there is something here, but it's all anecdotal. We could probably 
also dig into this a bit more, but not right now. 

And let's compare the reservoir chl-a and outflow chl-a, just to understand what
relationship is there, if any.

```{r}
grab_chla %>% 
  filter(location_type %in% c("Reservoir", "Outflow")) %>% 
  mutate(res_name = str_remove(site_name, " Reservoir| Outflow")) %>% 
  select(res_name, date, ChlA, location_type) %>% 
  pivot_wider(names_from = "location_type", values_from = "ChlA") %>% 
  filter(!is.na(Reservoir) & !is.na(Outflow)) %>% 
  ggplot(., aes(x = Reservoir, y = Outflow, color = yday(date))) +
  geom_point() +
  labs(x = "Reservoir Chlorophyll-a (ug/L)",
       y = "Outflow Chlorophyll-a (ug/L)") +
  theme_bw() +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") 
```

And let's break that out by reservoir:

```{r}
grab_chla %>% 
  filter(location_type %in% c("Reservoir", "Outflow")) %>% 
  mutate(res_name = str_remove(site_name, " Reservoir| Outflow")) %>% 
  select(res_name, date, ChlA, location_type) %>% 
  pivot_wider(names_from = "location_type", values_from = "ChlA") %>% 
  filter(!is.na(Reservoir) & !is.na(Outflow)) %>% 
  ggplot(., aes(x = Reservoir, y = Outflow, color = yday(date))) +
  geom_point() +
  labs(x = "Reservoir Chlorophyll-a (ug/L)",
       y = "Outflow Chlorophyll-a (ug/L)") +
  theme_bw() +
  facet_grid(res_name ~ .) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, lty = 2, color = "grey80") 
```



# Calculate some additional metrics per site

Grab some info from NHD that we need for some crawling!

```{r}
# Poudre HUC
huc8 <- get_huc(id="10190007", type = "huc08")
mapview(huc8)

# Pull all NHDPlusV2 flowines within Poudre Watershed
flowlines <- get_nhdplus(AOI = huc8, realization = "flowline", t_srs = 4326)
mapview(flowlines)

# Grab all waterbodies
waterbodies <- get_waterbodies(AOI = huc8, t_srs = 4326)
mapview(waterbodies)
# filter to those intersecting with flowlines
waterbodies <- waterbodies[flowlines, ]
mapview(waterbodies)

# remove the one waterbody for the fish hatchery
fish <- "17496521"
waterbodies <- waterbodies %>% 
  filter(comid != fish)

# Pull in site coordinates
xy <- grab_chla %>%
  select(site_code, site_name, Lat, Long, Campaign, location_type) %>% 
  distinct() %>% 
  st_as_sf(coords = c('Long','Lat'), crs = 4326) %>% 
  rowwise() %>% 
  mutate(comid = get_nhdplus(AOI = geometry, t_srs = 4326) %>%
           st_drop_geometry() %>%
           pull(comid)) %>% 
  ungroup()

mapview(flowlines) + mapview(waterbodies) + mapview(xy)

```

Calculate the distance to the nearest upstream reservoir. Basically looking for
instances where there will be changes based on proximity to reservoir.

```{r}
# NOTE: resolution not great for some sites... Some samples are along the same 
# flowline section, and are therefore represented with the same information. 

distance_to_up_waterbody <- function(site) {
  
  #one site
  one_site <- xy %>% filter(site_code == site) %>% first()
  
  # if the site type is a reservoir, return a value of 0
  if (one_site$location_type == "Reservoir") {
    return(tibble("site_code" = site, "dist_to_up_wb_km" = 0))
  }
  
  # get flowlines upstream of point
  upstream_of_point <- get_UT(flowlines, one_site$comid)
  # and grab those as a sf
  upstream_fl <- get_nhdplus(comid = upstream_of_point) %>% 
    get_tocomid(., add = TRUE) %>%
    mutate(ID=comid, toID=tocomid, length=lengthkm)
  # filter waterbodies for those that intersect the fl
  upstream_wb <- waterbodies[upstream_fl, ] %>% 
    filter(onoffnet == 1)
  
  # if there is at least one wb upstream, calculate the distance
  if(nrow(upstream_wb) >= 1) {
    art_path_outlet <- map_dfr(.x = upstream_wb$comid,
                               ~ tryCatch(get_wb_outlet(.x, upstream_fl),
                                          error = function(e) {} ))
    end_point <- get_node(art_path_outlet) %>% 
      mutate(comid = art_path_outlet$comid)
    mapview(upstream_wb) + mapview(end_point) + mapview(one_site$geometry)
    distance <- get_pathlength(upstream_fl) %>% 
      rename(comid = ID) %>% 
      right_join(end_point, by = "comid") %>% 
      arrange(pathlength) %>% 
      first() %>% 
      mutate(site_code = site) %>% 
      select(site_code, dist_to_up_wb_km = pathlength)
    
  } else {
    # otherwise, store distance as NA for the site code
    distance <- tibble("site_code" = site, "dist_to_up_wb_km" = NA)
  }
  
  return(distance)
  
}

distance_upstream_waterbody <- map(xy$site_code,
                                   ~ distance_to_up_waterbody(.)) %>% 
  bind_rows()

```

Calculate the distance to the nearest upstream tributary - basically looking for
possible step changes in chemistry and nutrients.

```{r}
distance_to_major_trib <- function(site) {
  
  # grab site info by site_code
  one_site <- xy %>% filter(site_code == site) %>% first()
  
  # get flowline for site
  flowline <- get_nhdplus(AOI = one_site)
  
  # make sure there are upstream tribs
  if (flowline$streamorde > 1) {
    
    # get comids for upstream mainstem
    mainstem <- get_UM(flowlines, flowline$comid)
    
    # get flowlines for upstream mainstem
    mainstem_fl <- get_nhdplus(comid = mainstem) %>% 
      get_tocomid(., add = TRUE) %>%
      arrange(comid) %>% 
      mutate(ID = comid, 
             toID = tocomid, 
             length = lengthkm) %>%
      arrange(totdasqkm) %>% 
      # calculate the difference in drainage area between segments
      mutate(diff_da = diff(c(0, totdasqkm))) %>% 
      select(comid, ID, toID, length, diff_da)
    
    # determine where the largest change is - this might not work in every basin
    # but it worked for all that I tried in CLP that were on the mainstem
    mainstem_big_trib <- mainstem_fl %>%
      arrange(-diff_da) %>% 
      slice(1)
    
    # get the flowline of the comid with largest change
    trib_fl <- get_nhdplus(comid = mainstem_big_trib$comid) %>%
      get_tocomid(., add = TRUE) %>%
      mutate(ID=comid, toID=tocomid, length=lengthkm) %>% 
      select(comid, ID, toID, length)
    
    # calculate distance from point to that segment
    distance <- get_pathlength(mainstem_fl) %>% 
      filter(ID == trib_fl$comid) %>%
      pull(pathlength)
    
    one_site <- one_site %>%
      mutate(dist_to_maj_trib = distance)
    
  } else {
    
    one_site <- one_site %>%
      mutate(dist_to_maj_trib = NA)
    
  }
  
  return(one_site)
  
}

distance_meta <- map(xy$site_code, ~ distance_to_major_trib(.)) %>% 
  bind_rows()

```

And add that info to the grab data.

```{r}
grab_site_info <- grab_chla %>% 
  left_join(., 
            distance_meta %>% select(site_code, dist_to_maj_trib), 
            relationship = "many-to-many") %>% 
  left_join(., 
            distance_upstream_waterbody,
            relationship = "many-to-many")
```

# Simplistic models for Chlorophyll-a

Try some simplistic methods for predicting Chlorophyll-a. The relationship looks
like a shallow polynomial equation, so we'll just give that a shot.

```{r}
chla_ms_site_info <- grab_site_info %>% 
  filter(Campaign == "Mainstem - Upper") %>% 
  mutate(doy = yday(date))

chl_model <- lm(log(ChlA) ~ poly(doy, degree=3), 
                data = chla_ms_site_info)
summary(chl_model)

plot(chl_model$residuals)

chla_ms_site_info <- chla_ms_site_info %>% 
  mutate(baseline_chla = predict(chl_model, newdata = .),
         residuals = exp(baseline_chla) - ChlA)

ggplot(chla_ms_site_info, aes(x = doy)) +
  geom_point(aes(y = ChlA)) +
  geom_point(aes(y = exp(baseline_chla)), color = "blue") +
  theme_bw()

ggplot(chla_ms_site_info, aes(x = dist_to_maj_trib, y = residuals)) +
  geom_point() +
  theme_bw()

ggplot(chla_ms_site_info, aes(x = dist_to_up_wb_km, y = residuals)) +
  geom_point() +
  theme_bw()

ggplot(chla_ms_site_info, aes(x = distance_upstream_km, y = residuals)) +
  geom_point() +
  theme_bw()

```

It's usually pretty bad practice to apply polynomial equation to environmental
data (because there generally is not an explainable reason for it), this relationship
is probably related to solar radiation and water temperature, which have a similar
pattern that is better characterized by a sinusoidal curve. Let's try an NLS model
with those assumptions. We're going to do this more rigorously by holding out the
2025 data to test this development on.

```{r}
test <- chla_ms_site_info %>% 
  filter(year(date) == 2025 & site_code %in% c("pbr","pbd","sfm","pman","chd") )

train <- anti_join(chla_ms_site_info, test)

fit_nls <- nls(
  log(ChlA) ~ A * sin(2 * pi * (doy - C) / 365) + D,
  data = train,
  start = list(
    A = (max(log(train$ChlA)) - min(log(train$ChlA))) / 2, # amplitude
    D = mean(log(train$ChlA)), # mean level
    C = 100 # phase shift guess (day)
  )
)

summary(fit_nls)
plot(fit_nls)
AIC(fit_nls)
BIC(fit_nls)

train1 <- train %>% filter(year(date) %in% c(2022, 2025))
train2 <- train %>% filter(year(date) %in% c(2023, 2025))
train3 <- train %>% filter(year(date) %in% c(2024, 2025))

fit_nls1 <- nls(
  log(ChlA) ~ A * sin(2 * pi * (doy - C) / 365) + D,
  data = train1,
  start = list(
    A = (max(log(train1$ChlA)) - min(log(train1$ChlA))) / 2, # amplitude
    D = mean(log(train1$ChlA)), # mean level
    C = 100 # phase shift guess (day)
  )
)

summary(fit_nls1)
plot(fit_nls1)
AIC(fit_nls1)
BIC(fit_nls1)

fit_nls2 <- nls(
  log(ChlA) ~ A * sin(2 * pi * (doy - C) / 365) + D,
  data = train2,
  start = list(
    A = (max(log(train2$ChlA)) - min(log(train2$ChlA))) / 2, # amplitude
    D = mean(log(train2$ChlA)), # mean level
    C = 100 # phase shift guess (day)
  )
)

summary(fit_nls2)
plot(fit_nls2)
AIC(fit_nls2)
BIC(fit_nls2)

fit_nls3 <- nls(
  log(ChlA) ~ A * sin(2 * pi * (doy - C) / 365) + D,
  data = train3,
  start = list(
    A = (max(log(train3$ChlA)) - min(log(train3$ChlA))) / 2, # amplitude
    D = mean(log(train3$ChlA)), # mean level
    C = 100 # phase shift guess (day)
  )
)

summary(fit_nls3)
plot(fit_nls3)
AIC(fit_nls3)
BIC(fit_nls3)

```

And let's see what this looks like when compared with the observed data.

```{r}
train <- train %>% 
  mutate(baseline_chla = exp(predict(fit_nls, newdata = .)),
         residuals = baseline_chla - ChlA)

ggplot(train, aes(x = doy)) +
  geom_point(aes(y = ChlA)) +
  geom_point(aes(y = baseline_chla), color = "blue") +
  theme_bw()

ggplot(train, aes(x = dist_to_maj_trib, y = residuals)) +
  geom_point() +
  theme_bw()

ggplot(train, aes(x = dist_to_up_wb_km, y = residuals)) +
  geom_point() +
  theme_bw()

ggplot(train, aes(x = distance_upstream_km, y = residuals)) +
  geom_point() +
  theme_bw()
```


And also as the ensemble

```{r}
train <- train %>% 
  mutate(baseline_chla_1 = exp(predict(fit_nls1, newdata = .)),
         baseline_chla_2 = exp(predict(fit_nls2, newdata = .)),
         baseline_chla_3 = exp(predict(fit_nls3, newdata = .)),
         baseline_chla_ens = (baseline_chla_1 + baseline_chla_2 + baseline_chla_3)/3,
         residuals_ens = baseline_chla_ens - ChlA)

ggplot(train, aes(x = doy)) +
  geom_point(aes(y = ChlA)) +
  geom_point(aes(y = baseline_chla_ens), color = "blue") +
  theme_bw()

ggplot(train, aes(x = dist_to_maj_trib, y = residuals_ens)) +
  geom_point() +
  theme_bw()

ggplot(train, aes(x = dist_to_up_wb_km, y = residuals_ens)) +
  geom_point() +
  theme_bw()

ggplot(train, aes(x = distance_upstream_km, y = residuals_ens)) +
  geom_point() +
  theme_bw()
```


These residuals look fine enough - the ensemble looks a bit better, let's apply
that ensemble to the test set.

```{r}
test <- test %>% 
  mutate(baseline_chla_1 = exp(predict(fit_nls1, newdata = .)),
         baseline_chla_2 = exp(predict(fit_nls2, newdata = .)),
         baseline_chla_3 = exp(predict(fit_nls3, newdata = .)),
         baseline_chla_ens = (baseline_chla_1 + baseline_chla_2 + baseline_chla_3)/3,
         residuals_ens = baseline_chla_ens - ChlA,
         baseline_chla = exp(predict(fit_nls, newdata = .)),
         residuals = baseline_chla - ChlA)

ggplot(test, aes(x = doy)) +
  geom_point(aes(y = ChlA)) +
  geom_point(aes(y = baseline_chla), color = "blue") +
  geom_point(aes(y = baseline_chla_ens), color = "orange") +
  theme_bw()

ggplot(test, aes(x = dist_to_maj_trib, y = residuals_ens)) +
  geom_point() +
  theme_bw()

ggplot(test, aes(x = dist_to_up_wb_km, y = residuals_ens)) +
  geom_point() +
  theme_bw()

ggplot(test, aes(x = distance_upstream_km, y = residuals_ens)) +
  geom_point() +
  theme_bw()
```

Okay, we're generally under predicting chl-a, but that's okay, we know this 
won't be perfect.

Calculate performance metrics.

```{r}
rmse_nls = rmse(test$ChlA, test$baseline_chla)
mae_nls = mae(test$ChlA, test$baseline_chla)
bias_nls = bias(test$ChlA, test$baseline_chla)

rmse_nls_ens = rmse(test$ChlA, test$baseline_chla_ens)
mae_nls_ens = mae(test$ChlA, test$baseline_chla_ens)
bias_nls_ens = bias(test$ChlA, test$baseline_chla_ens)
```

NLS performance:

RMSE: `r round(rmse_nls, 2)`
MAE: `r round(mae_nls, 2)`
Bias: `r round(bias_nls, 2)`

NLS ensemble performance:

RMSE: `r round(rmse_nls_ens, 2)`
MAE: `r round(mae_nls_ens, 2)`
Bias: `r round(bias_nls_ens, 2)`

Performance is much better in the ensemble. This makes sense because the AIC 
and BIC values were all smaller than the non-ensemble version (because there was
a lot less variability to try to explain). We'll export the ensemble baselines.

```{r}
# write_rds(fit_nls1, 
#           here(paste0("data/derived/models/chla_baseline/working/sinusoidal_ens1_doy_v",
#                       version_date,
#                       ".rds")))
# 
# write_rds(fit_nls2, 
#           here(paste0("data/derived/models/chla_baseline/working/sinusoidal_ens2_doy_v",
#                       version_date,
#                       ".rds")))
# 
# write_rds(fit_nls3, 
#           here(paste0("data/derived/models/chla_baseline/working/sinusoidal_ens3_doy_v",
#                       version_date,
#                       ".rds")))
```

# Plot fit metrics

```{r}
ggplot(test, aes(x = doy)) +
  geom_point(aes(y = ChlA, color = "Observed")) +
  geom_point(aes(y = baseline_chla_ens, color = "Baseline")) +
  scale_color_manual(values = c("Observed" = "black",
                                "Baseline" = "orange")) +  theme_bw() +
  labs(y = "Chlorophyll a (ug/L)",
       x = "Day of Year",
       color = "data source") +
  theme(
    legend.position = c(0.05, 0.95),   # x, y in NPC coordinates
    legend.justification = c(0, 1),   # anchor the legend box top-left
    legend.background = element_rect(fill = alpha("white", 0.7))
  )

# format annotation text for plot
metrics_text <- paste0(
  "MAE = ", round(mae_nls_ens, 2),
  "\nRMSE = ", round(rmse_nls_ens, 2),
  "\nBias = ", round(bias_nls_ens, 2)
)

ggplot(test, aes(x = baseline_chla_ens, y = ChlA)) +
  geom_point() +
  theme_bw() + 
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  coord_cartesian(xlim = c(0, 10), ylim = c(0, 10)) +
  labs(x = "Baseline Chlorophyll a Ensemble (ug/L)", 
       y = "Measured Chlorophyll a (ug/L)") +
  annotate("text", x = 8, y = 2, 
           hjust = 0, vjust = 1, 
           label = metrics_text, size = 4)

```


