---
title: "Creating a Baseline Chlorophyll-a Model"
output: html_document
---

# Purpose

The purpose of this markdown is meant to create a baseline chl-a model across the 
mainstem of the upper CLP.

```{r, echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE)
library(readxl)
library(here)
library(mgcv)
library(Metrics)
library(sf)
library(nhdplusTools)
library(mapview)
library(arrow)
library(ross.wq.tools)
# remotes::install_github("rossyndicate/ross.wq.tools")
library(tidyverse)

version_date = Sys.Date()
source(here("src/collation/load_grab_sample_data.r"))
```

# Chlorophyll-a Grab Samples

Get all chl-a grab samples from our collated data set (`load_grab_sample_data()`)

```{r}
# grab some specific columns
grab_chla <- load_grab_sample_data() %>% 
  select(site_code, site_name, date = Date,
         ChlA, Field_DO_mgL:distance_upstream_km, Lat:flag) %>% 
  filter(!is.na(ChlA), !Campaign %in% c("Mainstem - Lower", "Lower Tributary")

# check for flags
unique(grab_chla$flag)
```

Plot the data to see what it looks like

```{r}
ggplot(grab_chla, aes(x = yday(date), y = ChlA)) +
  geom_point()+
  facet_grid(location_type ~ ., scales = "free_y") +
  theme_bw()

ggplot(grab_chla, aes(x = yday(date), y = ChlA)) +
  geom_point()+
  facet_grid(Campaign ~ ., scales = "free_y") +
  theme_bw()

```

And also by site/location type

```{r}
walk(unique(grab_chla$location_type), function(loc) {
  
  gg <- grab_chla %>% 
    filter(location_type == loc) %>% 
    ggplot(. , aes(x = yday(date), y = ChlA, color = factor(year(date)))) +
    geom_point() +
    facet_wrap(site_name ~ ., scales = "free_y")
  print(gg)
  
})

```

And look at just the mainstem upper poudre, which is what we care about

```{r}
grab_chla %>% 
  filter(Campaign == "Mainstem - Upper") %>% 
  ggplot(. , aes(x = yday(date), 
                 y = ChlA, 
                 color = site_code, 
                 shape = factor(year(date)))) +
  geom_point() +
  theme_bw() +
  scale_color_viridis_d()

```

# Calculate some additional metrics per site

Grab some info from NHD that we need for some crawling!

```{r}
# Poudre HUC
huc8 <- get_huc(id="10190007", type = "huc08")
mapview(huc8)

# Pull all NHDPlusV2 flowines within Poudre Watershed
flowlines <- get_nhdplus(AOI = huc8, realization = "flowline", t_srs = 4326)
mapview(flowlines)

# Grab all waterbodies
waterbodies <- get_waterbodies(AOI = huc8, t_srs = 4326)
mapview(waterbodies)
# filter to those intersecting with flowlines
waterbodies <- waterbodies[flowlines, ]
mapview(waterbodies)

# Pull in site coordinates
xy <- grab_chla %>%
  select(site_code, site_name, Lat, Long, Campaign, location_type) %>% 
  distinct() %>% 
  st_as_sf(coords = c('Long','Lat'), crs = 4326) %>% 
  rowwise() %>% 
  mutate(comid = get_nhdplus(AOI = geometry, t_srs = 4326) %>%
           st_drop_geometry() %>%
           pull(comid)) %>% 
  ungroup()

mapview(flowlines) + mapview(waterbodies) + mapview(xy)

```

Calculate the distance to the nearest upstream reservoir. Basically looking for
instances where there will be changes based on proximity to reservoir.

```{r}
# NOTE: resolution not great for some sites 
# (see Comanche/Hourglass complex)... Some samples are along the same flowline 
# section, and are therefore represented with the same information. 

distance_to_up_waterbody <- function(site) {
  #one site
  one_site <- xy %>% filter(site_code == site) %>% first()
  # get flowlines upstream of point
  upstream_of_point <- get_UT(flowlines, one_site$comid)
  # and grab those as a sf
  upstream_fl <- get_nhdplus(comid = upstream_of_point) %>% 
    get_tocomid(., add = TRUE) %>%
    mutate(ID=comid, toID=tocomid, length=lengthkm)
  # filter waterbodies for those that intersect the fl
  upstream_wb <- waterbodies[upstream_fl, ] %>% 
    filter(onoffnet == 1)
  
  # if there is at least one wb upstream, calculate the distance
  if(nrow(upstream_wb) >= 1) {
    art_path_outlet <- map_dfr(.x = upstream_wb$comid,
                               ~ tryCatch(get_wb_outlet(.x, upstream_fl),
                                          error = function(e) {} ))
    end_point <- get_node(art_path_outlet) %>% 
      mutate(comid = art_path_outlet$comid)
    mapview(upstream_wb) + mapview(end_point) + mapview(one_site$geometry)
    distance <- get_pathlength(upstream_fl) %>% 
      rename(comid = ID) %>% 
      right_join(end_point, by = "comid") %>% 
      arrange(pathlength) %>% 
      first() %>% 
      mutate(site_code = site) %>% 
      select(site_code, dist_to_up_wb_km = pathlength)
    
  } else {
    # otherwise, store distance as NA for the site code
    distance <- tibble("site_code" = site, "dist_to_up_wb_km" = NA)
  }
  
  return(distance)
  
}

distance_upstream_waterbody <- map(xy$site_code,
                                   ~ distance_to_up_waterbody(.)) %>% 
  bind_rows()

```

Calculate the distance to the nearest upstream tributary - basically looking for
possible step changes in chem/nuts.

```{r}
distance_to_major_trib <- function(site) {
  
  # grab site info by site_code
  one_site <- xy %>% filter(site_code == site) %>% first()
  
  # get flowline for site
  flowline <- get_nhdplus(AOI = one_site)
  
  # make sure there are upstream tribs
  if (flowline$streamorde > 1) {
    
    # get comids for upstream mainstem
    mainstem <- get_UM(flowlines, flowline$comid)
    
    # get flowlines for upstream mainstem
    mainstem_fl <- get_nhdplus(comid = mainstem) %>% 
      get_tocomid(., add = TRUE) %>%
      arrange(comid) %>% 
      mutate(ID = comid, 
             toID = tocomid, 
             length = lengthkm) %>%
      arrange(totdasqkm) %>% 
      # calculate the difference in drainage area between segments
      mutate(diff_da = diff(c(0, totdasqkm))) %>% 
      select(comid, ID, toID, length, diff_da)
    
    # determine where the largest change is - this might not work in every basin
    # but it worked for all that I tried in CLP that were on the mainstem
    mainstem_big_trib <- mainstem_fl %>%
      arrange(-diff_da) %>% 
      slice(1)
    
    # get the flowline of the comid with largest change
    trib_fl <- get_nhdplus(comid = mainstem_big_trib$comid) %>%
      get_tocomid(., add = TRUE) %>%
      mutate(ID=comid, toID=tocomid, length=lengthkm) %>% 
      select(comid, ID, toID, length)
    
    # calculate distance from point to that segment
    distance <- get_pathlength(mainstem_fl) %>% 
      filter(ID == trib_fl$comid) %>%
      pull(pathlength)
    
    one_site <- one_site %>%
      mutate(dist_to_maj_trib = distance)
    
  } else {
    
    one_site <- one_site %>%
      mutate(dist_to_maj_trib = NA)
    
  }
  
  return(one_site)
  
}

distance_meta <- map(xy$site_code, ~ distance_to_major_trib(.)) %>% 
  bind_rows()

```

And add that info to the grab data.

```{r}
grab_site_info <- grab_chla %>% 
  left_join(., 
            distance_meta %>% select(site_code, dist_to_maj_trib), 
            relationship = "many-to-many") %>% 
  left_join(., 
            distance_upstream_waterbody,
            relationship = "many-to-many")
```

# Simplistic models for Chlorophyll-a

Try some simplistic methods for predicting Chlorophyll-a. The relationship looks
like a shallow polynomial equation, so we'll just give that a shot.

```{r}
chla_ms_site_info <- grab_site_info %>% 
  filter(Campaign == "Mainstem - Upper") %>% 
  mutate(doy = yday(date))

chl_model <- lm(log(ChlA) ~ poly(doy, degree=3), 
                data = chla_ms_site_info)
summary(chl_model)

plot(chl_model$residuals)

chla_ms_site_info <- chla_ms_site_info %>% 
  mutate(baseline_chla = predict(chl_model, newdata = .),
         residuals = exp(baseline_chla) - ChlA)

ggplot(chla_ms_site_info, aes(x = doy)) +
  geom_point(aes(y = ChlA)) +
  geom_point(aes(y = exp(baseline_chla)), color = "blue") +
  theme_bw()

ggplot(chla_ms_site_info, aes(x = dist_to_maj_trib, y = residuals)) +
  geom_point() +
  theme_bw()

ggplot(chla_ms_site_info, aes(x = dist_to_up_wb_km, y = residuals)) +
  geom_point() +
  theme_bw()

ggplot(chla_ms_site_info, aes(x = distance_upstream_km, y = residuals)) +
  geom_point() +
  theme_bw()

```

It's usually pretty bad practice to apply polynomial equation to environmental
data (because there generally is not an explainable reason for it), this relationship
is probably related to solar radiation and water temperature, which have a similar
pattern that is better characterized by a sinusoidal curve. Let's try an NLS model
with those assumptions. We're going to do this more rigorously by holding out the
2025 data to test this development on.

```{r}
test <- chla_ms_site_info %>% 
  filter(year(date) == 2025 & site_code %in% c("pbr","pbd","sfm","pman","chd") )
  

train <- anti_join(chla_ms_site_info, test)

fit_nls <- nls(
  log(ChlA) ~ A * sin(2 * pi * (doy - C) / 365) + D,
  data = train,
  start = list(
    A = (max(log(train$ChlA)) - min(log(train$ChlA))) / 2, # amplitude
    D = mean(log(train$ChlA)), # mean level
    C = 100 # phase shift guess (day)
  )
)

summary(fit_nls)
plot(fit_nls)
AIC(fit_nls)
BIC(fit_nls)
```

And let's see what this looks like when compared with the observed data.

```{r}
train <- train %>% 
  mutate(baseline_chla = exp(predict(fit_nls, newdata = .)),
         residuals = baseline_chla - ChlA)

ggplot(train, aes(x = doy)) +
  geom_point(aes(y = ChlA)) +
  geom_point(aes(y = baseline_chla), color = "blue") +
  theme_bw()

ggplot(train, aes(x = dist_to_maj_trib, y = residuals)) +
  geom_point() +
  theme_bw()

ggplot(train, aes(x = dist_to_up_wb_km, y = residuals)) +
  geom_point() +
  theme_bw()

ggplot(train, aes(x = distance_upstream_km, y = residuals)) +
  geom_point() +
  theme_bw()
```

These residuals look fine enough, let's apply to the test set.

```{r}
test <- test %>% 
  mutate(baseline_chla = exp(predict(fit_nls, newdata = .)),
         residuals = baseline_chla - ChlA)

ggplot(test, aes(x = doy)) +
  geom_point(aes(y = ChlA)) +
  geom_point(aes(y = baseline_chla), color = "blue") +
  theme_bw()

ggplot(test, aes(x = dist_to_maj_trib, y = residuals)) +
  geom_point() +
  theme_bw()

ggplot(test, aes(x = dist_to_up_wb_km, y = residuals)) +
  geom_point() +
  theme_bw()

ggplot(test, aes(x = distance_upstream_km, y = residuals)) +
  geom_point() +
  theme_bw()
```

Okay, we're generally under predicting chl-a, but that's okay, we know this 
won't be perfect.

Calculate performance metrics.

```{r}
rmse_nls = rmse(test$ChlA, test$baseline_chla)
mae_nls = mae(test$ChlA, test$baseline_chla)
bias_nls = bias(test$ChlA, test$baseline_chla)
```

Let's compare this with the same polynomial equation as above and see if the 
AIC/BIC could be lower.

```{r}
fit_poly <- lm(log(ChlA) ~ poly(doy, degree = 3), data = train)

summary(fit_poly)
plot(fit_poly$residuals)

AIC(fit_poly)
BIC(fit_poly)

train <- train %>% 
  mutate(baseline_chla_p = exp(predict(fit_nls, newdata = .)),
         residuals_p = baseline_chla_p - ChlA)

ggplot(train, aes(x = doy)) +
  geom_point(aes(y = ChlA)) +
  geom_point(aes(y = baseline_chla), color = "blue") +
  theme_bw()

```

And apply to the test/calculate metrics

```{r}
test <- test %>% 
  mutate(baseline_chla_p = exp(predict(fit_poly, newdata = .)),
         residuals_p = baseline_chla_p - ChlA)

ggplot(test, aes(x = doy)) +
  geom_point(aes(y = ChlA)) +
  geom_point(aes(y = baseline_chla_p), color = "blue") +
  theme_bw()

ggplot(test, aes(x = dist_to_maj_trib, y = residuals_p)) +
  geom_point() +
  theme_bw()

ggplot(test, aes(x = dist_to_up_wb_km, y = residuals_p)) +
  geom_point() +
  theme_bw()

ggplot(test, aes(x = distance_upstream_km, y = residuals_p)) +
  geom_point() +
  theme_bw()

rmse_poly = rmse(test$ChlA, test$baseline_chla_p)
mae_poly = mae(test$ChlA, test$baseline_chla_p)
bias_poly = bias(test$ChlA, test$baseline_chla_p)

```
Poly performance: 
RMSE: `r round(rmse_poly,2)`
MAE: `r round(mae_poly,2)`
Bias: `r round(bias_poly,2)`
NLS performance: 
RMSE: `r round(rmse_nls,2)`
MAE: `r round(mae_nls,2)`
Bias: `r round(bias_nls,2)`
Performance is basically the same across both models. Because the nls model 
incoporates a sinusoidal relationship which is more justifiable than the poly
model, let's stick with that one. 

```{r}
write_rds(fit_nls, 
          here(paste0("data/derived/models/chla_baseline/working/sinusoidal_doy_v", 
                      version_date,
                      ".rds")))
```

